### 2018-08-06 hitting_times.Rmd
hitting time theory

# Finding $G$ from $H$

Suppose that $G$ is the generator matrix of an irreducible continuous-time Markov chain $X$,
so that $G_{ij} \ge 0$ for $i \neq j$ and
$$ G 1 = 0 $$.

Let $\tau_{j} = \inf\{t \ge 0 : X_t = j\}$ and $H_{ij} = \E[\tau_j \,|\, X_0 = i]$.
Then we know that
$$
    (G H)_{ij} = -1 \qquad \text{for} \; i \neq j ,
$$
i.e.,
$$
    GH = - 1 1^T + \diag(x).
$$

What is $x$?  Well, note that
$$
    G = (\diag(x) - 1 1^T) H^{-1}
$$
and so
$$ \begin{aligned}
    0 &= G1 \\
    &= (\diag(x) - 1 1^T) H^{-1} x 
\end{aligned} $$
and so
$$
    x_i = \frac{ 1^T H^{-1} 1 }{ (H^{-1} 1)_i } .
$$
Furthermore,
the [Random Target Lemma](https://www.stat.berkeley.edu/users/aldous/RWG/Book_Ralph/Ch2.S2.html)
tells us that $H \pi = 1$, so thus $H^{-1} 1 = \pi$, the stationary distribution of the chain.
Therefore, $x_i = 1/\pi_i$,
and in fact $1^T H^{-1} 1 = 1$.

In summary,
$$
    G = (\diag(1/H^{-1} 1) - 1 1^T) H^{-1} .
$$

# Finding $G$ from $R$

The resistance distance is given by
$R = H + H^T$, so if we can modify $H$ by a skew-symmetric matrix,
we won't change $R$.
Can we do this and retain a valid $G$?
I.e., does $R$ uniquely specify $G$?
Let $Z + Z^T = 0$. 
Since $G$ is a continuous function of $H$, if all entries of $G$ are nonzero
we can find a small enough $\epsilon$ such that the $G'$ corresponding to $H' = H + \epsilon Z$
is also a valid generator matrix.
If $G$ has zero entries this may not be true.
To find a direction that we can change $H$ in that doesn't mess this up,
we could differentiate $dG/dH$.

### 2018-08-07 isolation_by_coalescence.tex
first draft

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath,amssymb,amsthm}
\usepackage[hidelinks]{hyperref}
\usepackage{graphicx}
\usepackage[utf8]{inputenc}
\usepackage{xcolor}
\usepackage{natbib}

\bibliographystyle{plainnat}

\newcommand{\Dcom}{\widetilde D}

\newcommand{\R}{\mathbb{R}}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\DeclareMathOperator{\var}{\mathop{\mbox{Var}}}
\DeclareMathOperator{\cov}{\mathop{\mbox{Cov}}}
\DeclareMathOperator{\diag}{\mathop{\mbox{diag}}}

\renewcommand{\P}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\given}{\,\vert\,}
\newcommand{\supp}{\mathop{\mbox{supp}}}
\newcommand{\sgn}{\mathop{\mbox{sgn}}}
\newcommand{\EE}[1]{\E\!\left[#1\right]}
\newcommand{\PP}[1]{\P\!\left\{#1\right\}}
\newcommand{\PPP}[2]{\P_{#1}\!\left\{#2\right\}}
\newcommand{\EEE}[2]{\E_{#1}\!\left[#2\right]}
\newcommand{\EEsup}[2]{\E^{#1}\!\left[#2\right]}

\newcommand{\bone}{\mathbf{1}}

\newcommand{\plr}[1]{{\em \color{blue} #1}}

\begin{document}

\title{Are populations like a circuit? The relationship between isolation by distance and isolation by resistance}
\author{Erik Lundgren\footnote{Computational Biology, University of Southern California}, and
        Peter L. Ralph\footnote{Institute for Ecology and Evolution, University of Oregon}}
\maketitle

\begin{abstract}
\end{abstract}

% NOTE: I tend to use semantic linebreaks.
% Please excuse the ragged right margins in the source.

%%%%%%%%%%%%%%%%%%%%%%
\section*{Introduction}

Genetic relatedness is determined by past gene flow,
a product of individual or gamete movements across geographic space.
Genomes therefore retain the traces of this movement,
and can contribute to the inference of how a species moves across a landscape,
which is important for understanding how diseases spread, how species adapt,
and how to retain genetic diversity in threatened species.
This is a potentially important source of information,
as direct observation can be difficult \citep[especially of long-distance movement]{dispersal_estimation},
or even impossible, if some of the populations in question no longer exist.

The idea of ``resistance'' is an important tool in the lanscape genomics toolbox.
Introduced by \citet{mcrae},
it makes use of a mathematical equivalence between random walks and electrical resistance
\citep{resistance}:
the expected time for a continuous-time Markov chain that starts at node $x$
to first hit node $y$ and then return to $x$ (the ``commute time'')
is equal to the effective resistance between $x$ and $y$ in an electrical network
whose conductances are given by the movement rates of the Markov chain.
Both electrical resistance and commute time between two nodes
average over all possible paths through the network between the two.
For this reason, it is thought that if genetic distances between samples
from different parts of the landscape
are significantly correlated with the effective resistances between those locations,
with local conductances assigned based on the values of some landscape variable,
then that landscape variable is likely a good indicator of the regions of the map 
where gene flow occurs.
For instance,
if the conductance across a grid cell in a discretization of the landscape
is set equal to its elevation,
and this produces effective resistance values 
that are significantly correlated with genetic distances,
then one might conclude that the species in question tends to disperse more at higher elevations,
perhaps along ridgelines.
\plr{reference some examples}
Other methods \citep{eems} seek to build \emph{de novo} 
a map of conductance values that produces resistances 
most strongly correlated with observed genetic distance,
and then interpret regions of low conductance in the resulting map as barriers to gene flow.

Genetic distances between two genomes sampled from the landscape
do derive from an average across a large number of paths between those points --
the lineages along which each segment of genetic material has been inherited
from their most recent common ancestor.
The more recent this ancestor is,
the smaller the genetic distance will be.
Thanks to recombination, there are a large number of such paths, 
and genetic distance averages over these.
It is reasonable in some situations to model these lineages as random walks across the landscape;
however, the model of genetic distance that one is thus led to
is \emph{coalescence time} of the random walks, 
rather than commute time, the quantity that corresponds to resistance.
This naturally raises the question: 
Are methods that depend on effective resistances being misled?
If so, would using coalescence time do better?
To better contrast the two,
we refer to ``commute time'' rather than ``resistance'' throughout,
although the two concepts are equivalent.

In this paper, we contrast 
coalescence time and commute time,
and develop an inference method for movement rates on a landscape
using both as estimated from genetic distances.
As we will see,
although the two methods are conceptually similar,
commute time inference may get the wrong answer if data derive from a coalescent process,
as does real genetic data.
We also explore the question of identifiability.
Resistance-based methods are often used to predict distances based on a few layers of geographical data
(land cover, slope, etcetera).
Some methods find the combination of landscape layers to best explain genetic distance,
which allows very fine geographic resolution \citep{infer_resistance}.
Others methods such as EEMS \citep{eems} infer conductances without such prior information;
however, the geographical resolution is much coarser.
One reason for this difference in resolution
is that EEMS uses more computationally intensive Bayesian methods.
However, we show that this is a more general problem --
inference of movement rates, in either a coalescence or commute time framework,
is an ill-conditioned problem,
implying the need for regularization to obtain reliable inference.
Using maps for prior information, or restricting the geographical resolution of the results,
are two ways to do this.

\plr{I didn't find myself wanting to explain isolation by distance here:
    I don't think we need to? Do you agree?}

%%%%%%%%%%%%%%%%%%
\section*{Methods}

\subsection*{Genetic distance and coalescence time}

\plr{pasted in: divergence is coalescence of lineages}
The genetic distance between
individuals, which provides a good estimate of the average time to a common ancestor along the genome, or
the average coalescence time. Coalescence is the process in which genes or regions of the genome from two
or more individuals, if traced back in time, eventually share a common origin [Wakeley, 2005]. While we
cannot know exactly when this happens, if we assume the rate of mutation per nucleotide per generation is
relatively constant, we can use the number of differences between two copies of a gene to estimate how much
time has passed since their common ancestor. With enough loci, we can calculate an estimate for
the average coalescence time across the genome [Hudson, 2007].

\plr{pasted in: lineages as random walks}
Even though the population dynamics in forwards time are Markov, the dynamics of lineages traced back
in time (the coalescent process) may not be. For example, if fecundity is variable and population density
is low, two lineages near each other are more likely to share a recent common origin, so knowing how one
lineage moves gives you information about how the other lineage is likely to move. This effect becomes
negligible as population density increases [Barton et al., 2002]), so it should not be an issue if the offspring
of any one parent are typically interspersed with the offspring of many others. The methods I develop here
model the coalescent process as Markov, which should be a good approximation for many systems.

\plr{Pasted in, needs transition; make understandable without the below}
The mean commute time of a location to itself is zero,
so to fit to genetic data, we clearly must also account for local diversity.
\plr{say why}
Coalescence time does not have this property,
because coalescence time accounts for both the first meeting time and the eventual coalescence,
while commute time is more analagous to the first meeting time.
For local diversity, we will also have another set of model parameters, $q$, 
where $q_{i}$ approximates the observed coalescence time in location $i$.
As is done by \citet{petkova2016visualizing},
we use commute time to approximate genetic distance as
\begin{align} \label{eq:hcom}
	D_{ij} \approx \Dcom_{ij} = R_{ij}/4 + q_{i}/2 + q_{j}/2 .
\end{align}
To gain an intuitive feel for why resistance distance is divided by 4, 
consider how, if the movement rates are symmetric, 
a single particle moving from one state to another is similar to (but not the same as)
the first meeting time of two particles, 
one moving forward in time from the first location 
and the second one moving backward in time from the other, 
so it is similar to twice the first meeting time.
Since the commute time is the sum of the hitting times in each direction, 
it is similar to 4 times the first meeting time.
For a conceptual picture of the differences between the coalescence time and commute time,
see Figure \ref{fig:concept_coalcom}. 
The average of the local diversity estimates ($q$) are then added 
to the first meeting time estimate to account for the eventual coalescence after meeting.
%\emph{Briefly motivate by describing "first meeting" plus "subsequent coalescence" -> 
%	commute + diversity approximation.}
In this case, these values
are parameters in the model to be inferred instead of within deme coalescence rates ($\gamma$).
Since the values a $q$ are typically fairly easy to infer 
and requiring them all to be the same can significantly decrease the quality of the fit, 
we will not be using an analogue of Problem Modification \ref{probmod:const-coal} 
(where we would only have one parameter for $q$).
Commute time inference has the advantage of only needing to use the base Markov chain
instead of the product chain,
but it also leads to incorrect inferences in certain circumstances.
Using the base Markov chain instead of the product chain is advantageous 
because, as the grid size increases, it is much more computationally efficient 
to work with a Markov chain with $n$ states than $n^2$ states 
since the matrices would be much smaller.

\begin{figure}
\centering
% \includegraphics[scale=.5]{conceptn}
\caption{Illustration of the conceptual differences
betweeen coalescence and commute time between states 1 and 3 
for a continuous time Markov chain with four linearly connected states. 
Coalescence time (left) is the time for two independently moving particles to meet and coalesce 
(it is possible for them to be in the same location without coalescing).
Commute time is time for a single particle starting in one state to travel to another state 
and then back to the original state.
Note how the time axis is scaled to be faster in the figure showing commute time
than the one for coalescence time in order for them to fit in the same space.}
\label{fig:concept_coalcom}
\end{figure}

For a simple example where the coalescence times and commute time approximations are very different, 
consider a system of three islands in a line,
where lineages currently in the outer islands are much more likely to come from the central island
than the other way around.
In this case, lineages are likely to quickly move to the center and coalesce regardless of starting position,
so coalescence time will be relatively short between all individuals.
Commute time between the two outer locations, on the other hand,
will be much longer since it requires the lineage to leave the center.
If we plug real numbers into this example, say, 
reverse time gene flow rate of $1$ into the center and $0.1$ out of it
with $\gamma=1$ everywhere,
then $H_{13}=H_{31}=2.90$ but $\Dcom_{13}=\Dcom_{31}=7.94$.
However, the fact that $\Dcom$ as predicted from the correct parameters may be far from $H$
does not mean that the parameters inferred using resistance distance methods
will necessarily be very wrong.
In principle, it still might be that the parameters for which $\Dcom$ is closest to $H$ 
are close to the real parameters, 
at least in some situations.

\begin{figure}
\centering
% \includegraphics[scale=.5]{three_state_example}
\caption{A simple example where the coalescence time and commute time based approximation
are very different.
In this case, commute time between the two outer locations will be much longer 
than coalescence time.}
\label{fig:3_state}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Hitting times of Markov chains}

First, we explain how to compute the relevant quantities --
mean commute and coalescence times -- from a Markov chain model of lineage movement.
The Markov chain is defined by its \emph{generator matrix}, denoted $G$,
for which $G_{xy}$ gives the jump rate from $x$ to $y$.
This means that if the Markov chain is at location $X_t$ at time $t$,
then for $x \neq y$, the probability the chain jumps from $x$ to $y$ is proportional to $G_{xy}$,
i.e., $\PP{X_{t+\epsilon} = y \given X_t = y} = \epsilon G_{xy} + O(\epsilon^2)$,
and $G_{xx} = - \sum_{y \neq x} G_{xy}$.
We will need to find the \emph{hitting times} of the chain,
i.e., for each pair of states $x$ and $y$, 
the mean time until the chain first hits $y$ after being started at $x$.
We denote this quantity $H_{xy}$.
Throughout, we assume that all hitting times are finite,
which implies the chain is connected and irreducible.
By conditioning on $X_\epsilon$,
one can use the Markov property to show that
\begin{align} \label{eqn:hitting_sum}
    \sum_y G_{xy} H_{yz} = -1 \qquad \text{for} \; x \neq z,
\end{align}
i.e., if $G_{-z}$ is the matrix with the $z^\text{th}$ row and column removed,
and $\bone$ is the vector composed of all ones,
then the $z^\text{th}$ column of $H$, except for $H_{zz}$,
can be computed as $- (G_{-z})^{-1} \bone$.
This, along with $H_{zz} = 0$, allows computation of $H$
from the movement rates $G$.

Suppose instead we are given the hitting times, $H$, and wish to find the movement rates, $G$.
This is not the situation we are in -- we have either commute or coalescence times --
but it is related.
The Random Target Lemma \citep{aldous}
tells us that the stationary distribution of the Markov chain, denoted $\pi$,
can be recovered from the hitting times by solving $\pi = H^{-1} \bone$.
As shown in Appendix \ref{apx::hitting_calcs},
equation \eqref{eqn:hitting_sum} can be rewritten in matrix form as
\begin{align}
    G H = \diag(1/\pi) - \bone \bone^T ,
\end{align}
which implies that $G$ can be computed directly from $H$ as
\begin{align} \label{eqn:G_from_H}
    G = \left( \diag(1/\pi) - \bone \bone^T \right) H^{-1} .
\end{align}
Therefore, $H$ uniquely determines $G$.

% \paragraph{Commute times}
The \emph{commute times} are a symmetrized version of the hitting times
(which need not be symmetric):
\begin{align} \label{eqn:R_from_H}
\text{\bf (commute time)} \qquad
    R_{xy} = H_{xy} + H_{yx},
\end{align}
or $R = H + H^T$.
We use $R$ to refer to the commute time as a reminder that commute times are also effetive resistances,
and to reserve $C$ for the coalescence time.

Although commute times are uniquely determined by movement rates, the reverse is not true.
Commute times only depend on the symmetric part of $H$ --
given an antisymmetric matrix $Z$ (so that $Z + Z^T = 0$),
any Markov chain that has hitting times given by $H_\epsilon = H + \epsilon Z$ for some $\epsilon$
will have the same commute times.
The resulting hitting times may not be valid --
if we let $G_\epsilon$ be the matrix constructed by applying equation \eqref{eqn:G_from_H} to $H_\epsilon$
(assuming this is invertible),
then it is not guaranteed that the offdiagonal elements of $G_\epsilon$ are all nonnegative, as required.
(It is always true that rows of $G_\epsilon$ sum to zero.)
However, since $G$ and $H$ are continuous functions of each other,
if all offdiagonal elements of $G$ are strictly positive,
then there exists a positive $\epsilon_0$ such that all $G_\epsilon$ for $\epsilon < \epsilon_0$
do define valid Markov chains, all with the same commute times.

% \paragraph{Coalescence times}
The \emph{coalescence times} are defined using \emph{two} copies of the same chain,
as the mean time until coalescence,
if the chains coalesce at rate $\gamma$ when they are in the same place.
Concretely, 
suppose that $X$ and $Y$ are independent Markov chains both moving with movement rates given by $G$,
that coalesce at rate $\gamma_x$ when $X$ and $Y$ are both at $x$.
Then we define $\tau$ to be this coalescence time,
so that 
$\P\{\tau \le t + \epsilon \given \tau > t, \; X_t = Y_t = x\} = \epsilon \gamma_x + O(\epsilon^2)$,
and 
$\P\{\tau \le t + \epsilon \given \tau > t, \; X_t \neq Y_t\} = O(\epsilon^2)$.
Then, the (mean) coalescence time is $C_{xy} = \E[\tau \given X_0 = x, \, Y_0 = y]$.
A similar argument using the Markov property
shows that $C$ satisfies the following equation:
\begin{align}
\text{\bf (coalescence time)} \qquad
    \sum_y \left(G_{xy} C_{yz} + C_{xy} G_{zy}\right)
    &=
    \begin{cases}
        -1                   \quad & \text{if} \; x \neq z, \\
        -1 + \gamma_z C_{zz} \quad & \text{if} \; x = z.
    \end{cases}
\end{align}
In matrix notation, this is
\begin{align} \label{eqn:C_matrix}
    G C + C G^T - \diag(\gamma) \circ C = -\bone \bone^T,
\end{align}
where $\diag(\gamma)$ is the matrix with the vector $\gamma$ on the diagonal and zeros elsewhere,
and $\circ$ is the componentwise product.
This can also be seen by working with the product Markov chain $(X_t, Y_t)$,
whose generator matrix is $G \prod I + I \prod G$,
where $I$ is the identity matrix and $\prod$ is the Kronecker product.

Equation \eqref{eqn:C_matrix} is linear in $G$ and $\gamma$,
and so can be solved with straightforward linear algebra.
If $\gamma$ is given, then solving for $G$ is an example of a XXX equation \citep{matrixeqn}.
However, the solution is again not necessarily unique:
suppose that we have a matrix $Z$ such that $ZH$ is antisymmetric.
Then $(G + Z) C + C (G + Z)^T = GC + CG^T$,
and so a Markov chain with generator matrix $G + Z$ has the same coalescence times
as the original chain.
As before, if all entries of $G$ are nonzero, 
it is always possible to find sufficiently small $Z$
so that $G + Z$ remains the generator of a valid Markov chain,
but this is not guaranteed in general.

\paragraph{Constraints and uniqueness}
We have seen that to find the movement rates of a continuous-time Markov chain
given the coalescence or commute times
we must solve equations \eqref{eqn:C_matrix} or \eqref{eqn:R_from_H} and \eqref{G_from_H} for $G$,
and that these do not hae unique solutions.
The situation is worse when one needs to infer coalescence rates ($\gamma$) as well.
However, usually in applications the locations lie across geographical space,
so that many of the entries of $G$ can be assumed to be zero.
We can get an idea of this by simply counting equations and unknowns.
For concreteness, suppose that the spatial locations
are arranged in a square grid of $n$ locations,
so that each is connected to four others.
(Boundary locations will have fewer, but we omit this detail.)
Since movement rates in each direction can be different,
there are $4n$ free parameters, each corresponding to an offdiagonal entry of $G$.
Coalescence rates provide another $n$ parameters.

Since commute times are symmetric, and $R_{zz} = 0$, 
equation \eqref{eqn:R_from_H} only provides $n (n-1)/2$ informative equations.
This is larger than $4n$, the number of parameters,
for a grid with at least nine nodes (i.e., $3 \times 3$).
The same calculation holds for commute times, 
except that (a) we get an additional $n$ equations from the diagonal entries,
but (b) this is counteracted by the additional $n$ coalescence parameters.

This suggests that although nonuniqueness of solutions may not be a problem in practice,
as long as geography is discretized into sufficiently many locations.
However, another problem appears at finer geographic resolutions:
although solutions may be in principle unique,
the problems become \emph{ill-conditioned}, 
in the sense that arbitrarily small variations in the input times
(even numerical instability)
produce large differences in the inferred rates.
Very roughly speaking, this occurs because as the spatial resolution increases,
the matrix $G$ converges to a second-order elliptic differential operator \citep{pde};
such operators are deformations of the Laplacian \citep{laplacian},
which is well-known to have eigenvalues that decay exponentially \citep{laplacian_eigenvals}.
Information about the landscape contained in higher-order modes
(e.g., finer resolution changes in hitting times)
is then almost obliterated by the small eigenvalues of $G$,
making the inverse problem possible in theory but not in practice.
For more discussion of related problems, see \citet{badtruth,shape_of_a_pop,inverse_prob}.

\subsection*{The simplest example}

Consider a Markov chain with two states, 
where the rate of movement from state 1 to state 2 is $G_{12}$
and the rate of movement from state 2 to state 1 is $G_{21}$,
so the generator matrix is
\begin{align*}
G = 
    \begin{bmatrix}
        -G_{12}  & G_{12} \\
         G_{21}  & -G_{21}
    \end{bmatrix}.
\end{align*}
Writing out equation \ref{eqn:C_matrix}, we get three unique equations
for the coalescence times:
\begin{align*}
   -1 &= -C_{11} G_{12} + C_{21} G_{12} - C_{11} G_{12} + C_{12} G_{12} - C_{11} \gamma_1 \\
   -1 &=  C_{11} G_{21} - C_{21} G_{21} - C_{21} G_{12} + C_{22} G_{12} \\
   -1 &=  C_{12} G_{21} - C_{22} G_{21} + C_{21} G_{21} - C_{22} G_{21} - C_{22} \gamma_2 .
\end{align*}
Given $C$, we have three equations for the four unknowns, 
$G_{12}$, $G_{21}$, $\gamma_{1}$, and $\gamma_{2}$.
By grouping terms together, factoring back into matrix form, 
and writing $C_{ji}$ instead of $C_{ij}$ when $i > j$,
we have
\begin{align*}
\begin{bmatrix}
-2C_{11} + 2C_{12} & 0                 & -C_{11} & 0       \\
-C_{12} + C_{22}   & C_{11} - C_{12}   & 0       & 0       \\
0                  & 2C_{12} - 2C_{22} & 0       & -C_{22}
\end{bmatrix}
\begin{bmatrix}
G_{12} \\ G_{21} \\ \gamma(1) \\ \gamma(2)
\end{bmatrix}
=
\begin{bmatrix}
-1 \\ -1 \\ -1
\end{bmatrix}.
\end{align*}
This is small enough to be solved symbolically. 
The valid solutions to these equations are those with $G_{12}$, $G_{21}$, $\gamma_1$,
and $\gamma_2$ nonnegative.
If $C_{11} \neq C_{12}$, 
we can write the solution with $\gamma_1$ as the free variable:
\begin{align*}
G_{12} &= \frac{1}{2(C_{11} - C_{12})} - \frac{C_{11}}{2(C_{11} - C_{12})}\gamma_1 \\
G_{21} &= \frac{-2(C_{11} - C_{12}) + (C_{12} - C_{22})}{2(C_{11} - C_{12})^2}
	- \frac{C_{11}(C_{12} - C_{22})}{2(C_{11} - C_{12})^2}\gamma_1 \\
\gamma_2 &= \frac{(C_{11} - 2C_{12} + C_{22})^2}{C_{22}(C_{11} - C_{12})^2}
	- \frac{C_{11}(C_{12} - C_{22})^2}{C_{22}(C_{11} - C_{12})^2}\gamma_1.
\end{align*}
This implies that $\gamma_2$ decreases as $\gamma_1$ increases as long as $C_{12} \neq C_{22}$.
This makes sense because
increasing a rate of coalescence cannot make expected coalescence times longer,
so in order to keep coalescence times the same,
if one coalescence rate is increased, another must be decreased.
This also implies that, in order for $\gamma_2$ to be a value other than $0$, 
we must have $C_{11} - 2C_{12} + C_{22} \neq 0$.
In general, if there is a value or range of values of $\gamma_1$
for which the other movement and coalescence parameters are non-negative,
then a solution exists.

\begin{figure}
\centering
% \includegraphics[scale=.7]{valid_range}
\caption{The green shaded region shows 
	the valid range of the parameter values
	for the two state Markov chain given that
	$C_{11}=1$, $C_{12}=2$, and $C_{22}=1.5$}
\label{fig:valid_range}
\end{figure}

To produce a comcrete example, suppose that
$C_{11} = 1$, $C_{12} = 2$, and $C_{22} = 1.5$ in arbitrary time units.
The other parameters will be non-negative 
when $1 \leq \gamma_1 \leq 5$,
as shown in in Figure \ref{fig:valid_range}.
Assuming $\gamma_1 = \gamma_2$ produces a unique solution, 
with $\gamma_1 = \gamma_2 = 1.29$ and $G_{12} = 0.14$ and $G_{21} = 0.93$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Bayesian inference of movement rates}

Mean coalescence times estimated from real data are subject to a number of sources of noise
(which we discuss in more detail later).
Here, we use
the exact solutions based on linear algebra developed above
to develop a Bayesian inference method that accounts for noise in the data
and estimates uncertainty in the resulting estimates.
Suppose that $\hat{C}_{ij} = C + \epsilon_{ij}$,
where $\epsilon$ is Normally distributed
with mean $\mathbf{0}$ and covariance $\Sigma$.
For now, we will make a few assumptions about $\epsilon$.
First, that all $\epsilon_{ij}$ are independent, identically distributed $\mathcal{N}(0,\sigma_{\epsilon}^2)$
for $i \leq j$, and $\epsilon_{ij} = \epsilon_{ji}$.
Additionally, we will assume that $C_{ij}$ is large compared to $\sigma_{\epsilon}$
such that the probability that a value of $\hat{C}$ is ever negative is negligible.
Writing $f(\mathbf{x})$ as the probability density function of 
the multivariate normal distribution $\epsilon$,
the likelihood function for the pair $(G,\gamma)$ is
\begin{align*}
\mathcal{L}(G,\gamma;\hat{C}) &= f(\mathcal{C}(G,\gamma) - \hat{C}) \\
			      &= f(-(G^*-\text{diag$(\Gamma)$})^{-1} \mathbf{1} - \hat{C})
\end{align*}
recalling that $\mathcal{C}$ is the hitting time matrix for a given pair $(G,\gamma)$
defined by equation \ref{eq:solve_H}. 
If $\hat{C}$ is a valid coalescence time matrix,
we can find the maximum likelihood estimate (MLE) of $G$;
however, this is in practice unlikely to happen, 
and the likelihood function is still useful 
for providing information about the uncertainty of $G$.

To complete the Bayesian model,
we place priors on the parameters.
Let's say that $i$ is said to be adjacent to $j$ if $i \neq j$ and $G_{ij}$ is one of 
the values known to be non-zero.
Then we could let the prior probability density for the movement parameters be that of an 
exponential random variable with rate parameter $\lambda_{G}$. 
That is, the prior likelihood of $G$ would be
\begin{align*}
\mathcal{L}_G(G) = 
\begin{cases}
	\lambda_{G}^{m_G} \prod_{i \sim j} e^{-\lambda_{G} G_{ij}} 
		\quad & \text{for $G_{ij} \geq 0 \quad \forall \ i \neq j$} \\
	0 & \text{otherwise}
\end{cases}
\end{align*}
where $i \sim j$ means that $i$ and $j$ are adjacent and $m_G$ is the number of positive values of $G$. 
Similarly, we can put a prior density on $\gamma$:
\begin{align*}
\mathcal{L}_{\gamma}(\gamma) = 
\begin{cases}
	\prod_{i}^{n} \lambda_{\gamma} e^{-\lambda_{\gamma}\gamma_i} 
		\quad & \text{for $\gamma_i \geq 0 \quad \forall \ i$} \\
	0 & \text{otherwise.}
\end{cases}
\end{align*}
Recall that $n$ is the number of states in the single particle Markov chain.
If we assume that all $\epsilon_{ij}$ are IID with $\epsilon_{ij} \sim \mathcal{N}(0,\sigma_{\epsilon}^2)$ 
for $i \leq j$
we can write the likelihood for $\hat{C}$:
\begin{align*}
\mathcal{L}(\hat{C};G,\gamma) = \frac{1}{(2 \pi \sigma_{\epsilon}^2)^{m/2}} \prod_{i \leq j}
	\exp\left\{-\frac{(\mathcal{C}_{ij}(G,\gamma) - \hat{C}_{ij})^2}{2 \sigma_{\epsilon}^2}\right\},
\end{align*}
where $m$ is the number of nonmissing values on or above the diagonal in $\hat{C}$.
Then the posterior distribution is
\begin{align} \label{eq:post}
\mathcal{L}(\hat{C};G,\gamma) \mathcal{L}_{G}(G) \mathcal{L}_{\gamma}(\gamma) = 
	\frac{\lambda_{G}^{m_G} \lambda_{\gamma}^n}{(2 \pi \sigma_{\epsilon}^2)^{m/2}}
	\exp\{-\sum_{i \sim j} \lambda_G G_{ij} -\sum_{i}^n \lambda_{\gamma}\gamma_i
		-\frac{1}{2 \sigma_{\epsilon}^2} \sum_{i \leq j} (\mathcal{C}_{ij}
		(G,\gamma) - \hat{C}_{ij})^2 \}
\end{align}
with the same support over $G$ and $\gamma$ as in their individual priors.
To sample from this distribution, 
we will use the Metropolis-Hastings algorithm \citep{brooks2011handbook}.

\paragraph{Incomplete observations}
With real world data, 
unless the number of states is small or we have a large amount of data,
our pairwise average coalescent time matrix will have a large number of unobserved values.
For example, if our landscape is a $20 \times 20$ grid and we take $100$ samples, 
each from a different grid square,
we would have only $100(101)/2$ out of $400(401)/2$ of the unique values for $H$. 
(Note that we are including the self-comparisons in this example;
if the organisms were haploid or inbreeding were substantial,
our $H$ matrix would have only $99(100)/2$ usable values.)
To carry out inference in this situation,
we simply use Equation \ref{eq:post},
but the rightmost sum will be only over observed entries of H.

%%%%%%%%%%%%%%%%
\subsection*{Test cases and simulations}

\plr{describe graphs and slim sims}

\plr{MCMC methods: pre-burnin, burnin, number of iterations, trace plots}

%%%%%%%%%%%%%%%%%%
\section*{Results}
%%%%%%%%%%%%%%%%%%

\plr{transition paragraph}
% biased migration is bad for commute times: 
%              figure 19 showing the two against each other, and the simple example
% continuous space: can find big barriers but lots of between-replicate variability (life is still hard)
% tortoises: results on small graph, showing strong asymmetry (?)

\plr{Summarizing observations about inference being harder with more noise,
more parameters,
and lower coalescence rate.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Identifying a barrier to gene flow}
\label{sec:5x3b}

\plr{TODO: replace bottom two panels in figure 18 with same thing for commute time}

\plr{pasted in}
We now test the method's ability to locate barriers to migration
on a $5 \times 3$ grid with asymmetric gene flow. 
There are impassable boundaries on the bottom left and top right of the landscape,
but the overall graph remains fairly well connected (see Figure \ref{fig:5x3b_grid}).
Each value of $g$ in the true model is determined randomly as before.
The coalescence rates are set to 1 everywhere, 
and for the first case, we assume that we know it is the same everywhere, 
although the specific value is not known.
Gaussian noise with standard deviation equal to 1/1000 of the mean of $H$ is added to form $\hat{H}$, 
and inference is performed using all values of $\hat{H}$.
All inference in this section is done using
$5 \times 10^6$ iterations of MCMC 
after $4 \times 10^6$ iterations for burn-in and $1 \times 10^6$ iterations for pre-burn-in.
In this case, the mean absolute error of $\hat{g}$ was 0.019, 
while the error of $\hat{\gamma}$ was 0.0028. %\emph{talk more about results here?}
This analysis is repeated in three more situations, 
all with noise set to have standard deviation 1/100 of the mean of $H$.
In the first case, there are no missing values of $\hat{H}$, 
in the second case, values from rows and columns
corresponding to populations 2, 5, 7, and 15 are missing from $\hat{H}$. 
The third case is the same as the second except that
there is a coalescence parameter for each location on the grid 
instead of one parameter for all locations.
With no missing values and noise of 1/100, 
the mean absolute error of $\hat{g}$ was 0.17 and the error of $\hat{\gamma}$ was 0.0066.
With the four missing values,
the mean absolute error of $\hat{g}$ was 0.37 and the error of $\hat{\gamma}$ was 0.033.
With missing values and multiple coalescence parameters, 
the mean absolute error of $\hat{g}$ was 0.52 
and the mean absolute error of $\hat{\gamma}$ was 0.42.
Boxplots of the posterior distributions of $g$ can be seen in Figure \ref{fig:5x3boxplots}.
We can see that the inference method does very well 
in the case with noise of 1/1000.
With each modification, the problem becomes less well defined, 
so it is not surprising that the error increases as well.
Notably, we are still able to identify the barrier 
except when there is data missing from both locations on either side of the barrier, 
as is the case for movement parameters $g_{5}$ and $g_{17}$ with locations 2 and 7.
Additionally, inference of the coalescence rate is very accurate 
when there is only one coalescence rate parameter,
but is much harder with multiple parameters, 
which in turn makes the movement parameters harder to estimate.

%This is repeated with noise set to have standard deviation 1/100 of the mean of $H$ 
%and with data missing from 4 of the 15 locations (\emph{also non-constant coalescence?}).
%In this case, the mean absolute error of $\hat{g}$ was (insert) and (appropriate statement for $\gamma$)
%after (insert) iterations of MCMC.
%\emph{talk more about results here}

\begin{figure}
\centering
% \includegraphics[scale=.6]{5x3b_grid}
\caption{Plot of the grid structure 
for the 5 $\times$ 3 grid with barriers.
Values for the non-zero movement rates 
were chosen by rounding independent draws of an exponential random variable with mean 1 
up to the nearest tenth.}
\label{fig:5x3b_grid}
\end{figure}

\begin{figure}
\centering
% \includegraphics[scale=0.8]{5x3boxplots}
\caption{Posterior distributions for values of $g$ 
for the 5 $\times$ 3 graph with barriers 
for each of the analysis cases.
See Figure \ref{fig:5x3b_grid} to see movement parameter locations.}
\label{fig:5x3boxplots}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Biased migration}
\label{sec:comvcoal}

\plr{transition about how we expect biased migration to produce large differences}

In order to further illustrate the differences between commute time and coalescence time inference,
we test both methods on four different $4 \times 4$ graphs (shown in Figure \ref{fig:4x4_grids}:
one where movement parameters are all equal to 1 (which we will call the uniform case),
one where movement parameters are symmetric but randomly generated, 
one asymmetric graph where all movement parameters are randomly generated,
and one asymmetric graph where there is a bias in moving towards one direction in the graph.
For the symmetric and unbiased asymmetric graphs, movement parameters are chosen randomly as before.
In the biased asymmetric graph, all parameters for moving left or down are 2 while 
all parameters moving right or up are 0.5.
Differences between true mean coalescence times ($H$) and mean commute times ($\Dcom$) 
are shown for each graph in Figure \ref{fig:4x4coalvcomH}. 
The differences are fairly small in the uniform and symmetric cases, moderate in the asymmetric case, 
and extreme in the biased asymmetric case.
%The structure of the graphs themselves can be seen in Figure \ref{fig:4x4_grids}.
Inference is done using $\hat{H}$ with a noise level of 1/200 the mean value of $H$ for each grid.
All trials with each inference method were done using $4 \times 10^6$ iterations of MCMC 
after $1 \times 10^5$  iterations of pre-burn-in and $1 \times 10^6$ iterations of burn-in.
Posterior distributions for each graph and and inference method are shown in Figure \ref{fig:4x4box}.
The coalescence time inference was substantially more accurate in all cases.
The commute time based inference and coalescence time based inference methods had 
mean absolute errors in $\hat{g}$ of 0.30 and 0.12 respectively for the uniform graph,
0.49 and 0.053 for the symmetric graph, 
0.87 and 0.035 for the asymmetric graph,
and 3.27 and 0.053 for the biased asymmetric graph, respectively.
We can see that, as expected based on the differences between $H$ and $\Dcom$,
the coalescence time inference performs very well in all four test cases
while the commute time inference performs fairly well in uniform and symmetric cases,
poorly in the asymmetric case, and very poorly in the biased asymmetric case.
The behavior of the biased asymmetric graph supports our previous thoughts about
situations similar to the three state system seen in Figure \ref{fig:3_state},
where commute times are much longer than coalescence times.
In this case, the commute time inference predicts that many movement parameters
are an order of magnitude larger than they are in truth. 
%and there does not appear to be a correlation between 
%the magnitude of the true and inferred values, 
%indicating that we cannot infer much beyond that the population mixes quickly,
%which was already clear from the data itself.

%\emph{talk more about results here}

\begin{figure}
\centering
% \includegraphics[scale=.8]{4x4_grids}
\caption{Plots of the grid structure 
for the uniform graph, symmetric graph, the asymmetric graph,
and the biased asymmetric graph 
using the igraph R package.}
\label{fig:4x4_grids}
\end{figure}

\begin{figure}
\centering
% \includegraphics[scale=.8]{4x4coalvcomH}
\caption{Plots of the true values of $\Dcom$ vs $H$ 
for the uniform graph, symmetric graph, the asymmetric graph,
and the biased asymmetric graph, 
shown with the $y=x$ line.
See Figure \ref{fig:4x4_grids} to see overall graph structure.}
\label{fig:4x4coalvcomH}
\end{figure}

\begin{figure}
\centering
% \includegraphics[scale=.8]{4x4box}
\caption{Boxplots of the posterior distributions for $g$ for commute time and coalescence time
based inference for the uniform graph, symmetric graph, 
the asymmetric graph, and the biased symmetric graph.
See Figure \ref{fig:4x4_grids} to see movement parameter locations.}
\label{fig:4x4box}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Continuous geographical space}

\plr{transition about discretization and continuous space}

In order to validate the methods in situations 
that are more similar to those that we would see with actual populations,
we performed simulations using SLiM (Selection on Linked Mutations) \citep{haller2016slim}.

\plr{move to methods}
We used SLiM to simulate discrete generations based on the Wright-Fisher model.
Simulations were performed in three different situations:
uniform migration on a continuous square landscape, 
uniform migration on a rectangular landscape with barriers to migration, 
and biased migration on a square landscape.
The simulations performed in this section 
were of populations of 10000 diploid hermaphroditic individuals.
Mates were selected from nearby individuals using a Gaussian kernel 
with standard deviation of $\sigma_{d}$,
truncated at $3 \sigma_{d}$.
In each simulation,
$\sigma_{d}$ was chosen such that there would be an average of 
0.25 individuals per $\sigma_{d}^2$ (2.25 per $(3 \sigma_{d})^2$),
which is 0.005 for a square landscape with one unit of area.
Offspring are dispersed according to a Gaussian distribution away from the mother
using the same truncated Gaussian kernel as before.
If the location of the offspring falls outside of the habitable area, 
it is redrawn until it is in a valid location.
In order to prevent excessive clumpiness of the population density distribution \citep{felsensten1975pain}, 
we also model local competition. 
To do this, we use the same truncated Gaussian kernel,
with the strength of the interaction chosen 
by visual inspection of the density distribution.
Specifically, the new fitness of each individual after accounting for local competition
is equal to $\exp{(-\text{totalStrength}/40)}$,
where totalStrength is the sum of the strength of interations with all individuals
within the cutoff distance according to the Gaussian kernel,
resulting in reduced fitness for individuals in dense regions.
Individuals are chosen to reproduce proportional to their fitness until the next generation.
Clumpiness is similar to what is seen in 
Figure \ref{fig:ind_locs_5x3b_1}

In our simulations, individuals have 10 pairs of chromosomes with 200,000 base pairs each 
with a recombination rate of $10^{-8}$ per generation per base pair.
In the landscapes with barriers and biased migration,
mutation rate was set to $10^{-7}$.
In order to speed up simulation in the case of the landscape with uniform migration,
the mutation rate was lowered to $10^{-8}$.
This should not have a dramatic effect on the pattern of genetic relatedness between individuals
aside from reducing the genetic distance for those simulations by a factor of ten.

In each location, individuals are sampled uniformly from those 
within the middle three quarters in each dimension of the grid space,
as shown for one situation iin Figure \ref{fig:ind_locs_5x3b_1}.
This protocol was chosen as a compromise between sampling all individuals 
from near the center of each location,
which could result in a large number of close relatives being sampled, 
leading to an underestimate of the mean genetic distance for individuals in that location
and uniform sampling from the whole area of each grid, 
which may result in many sampled individuals being very close to other grid squares.
An equal number of individuals are drawn from each location,
but this need not be the case.
We could even have some locations from which no individuals were drawn,
but this lowers performance as discussed in Section \ref{sec:5x3b}.
Mean genetic distances for each pair of locations 
are calculated from the means of all of the pairwise genetic divergences 
between genomes in the two locations, excluding self comparisons.
Standard error of the mean is calculated conservatively using 
the minumum number of independent samples, 
in this case the smallest number of individuals from each location.

Before being passed to the MCMC inference method,
genetic distance values are scaled so that the resulting parameters 
are approximately of order one.
This is accomplished by multiplying the mean genetic distances by a constant
such that the new mean of the mean genetic distances 
is equal to the number of states in the discretization. 
For example, if we discretize a landscape into 16 squares, 
the mean of the mean genetic distances after scaling will be 16.

For the $4 \times 4$ discretizations of the uniform and biased migration cases, 
as well the barrier case, 
inference was performed using $4 \times 10^6$ iterations of MCMC
after $1 \times 10^6$ iterations of pre-burn-in and $3 \times 10^6$ iterations of burn-in. 
For the $2 \times 2$ discretizations, $1 \times 10^6$ iterations were used 
with $1 \times 10^5$ for pre-burn-in and $1 \times 10^6$ for burn-in.

% {Continuous Landscape with a Barrier}
We now revisit the earlier situation where a landscape has locations that are barriers to gene flow,
as in Section \ref{sec:5x3b}.
The landscape is designed to be a continuous version of this situation.
Since the distribution of offspring locations is truncated at three standard deviations
away from the mother, a sufficiently thick uninhabitable area can completely stop migration
directly across it.
As in \ref{sec:5x3b}, though, the landscape is still relatively well connected.
The layout of the landscape can be seen in Figure \ref{fig:ind_locs_5x3b_1},
with the red bars being the uninhabitable regions that serve as barriers to migration.
Unlike the case of a square landscape with no barriers, 
there is not a natual way to discretize it more coarsely 
with grid squares of equal size and shape,
so a $5 \times 3$ grid will be the only discretization used in this section.
We calculate mean genetic distances using 50 individuals 
from each grid location,
here giving us a total of 750 individuals.

\begin{figure}
\centering
% \includegraphics[scale=.8]{ind_locs_5x3b_1}
\caption{Locations of all and selected individuals on the $5 \times 3$ landscape with barriers
in the first replicate.
Locations of individuals used to compute the mean genetic distance matrix are shown in bold points
whereas the locations of other individuals are shown in smaller points.
The red bars show the locations of uninhabitable regions of the landscape.
They are sufficiently thick so that no migration can occur directly across them.}
\label{fig:ind_locs_5x3b_1}
\end{figure}

\begin{figure}
\centering
% \includegraphics[scale=.8]{grid_5x3b_1}
\caption{Medians of the posterior distributions of the inferred values of $g$ 
for the first replicate of the $5 \times 3$ landscape with barriers.
The gene flow rate across the barriers are small, 
but there is no clear pattern amoung the rest of the values.}
\label{fig:grid_5x3b_1}
\end{figure}

\begin{figure}
\centering
% \includegraphics[scale=.8]{post_dists_5x3b_1_new}
\caption{Boxplots of the posterior distributions for $g$ 
for the first replicate of the $5 \times 3$ landscape with barriers.
Boxes for gene flow rates for edges that cross the barrier are colored in red.}
\label{fig:post_dists_5x3b_1}
\end{figure}

The gene flow rates across the barriers are reliably estimated to be very small
%(\emph{maybe reference all 6 supplemental figures here?}),
but there is considerable variation in the other values.
Note, however, that this analysis requires that we guess a reasonable place
to put the divisions between the locations.
For instance, we may wonder whether a river or other geographical feature
restricts gene flow.
We would then discretize the landscape such that individuals on opposite sides of the 
potential obstruction are in separate grid spaces.

\subsection*{Continuous Landscape with Biased Migration}

The third situation that we will discuss is where migration is biased.
We will be using a square landscape as in the first case, 
but the offspring locations of offspring are not centered around the parent.
Instead, the mean of the distribution of the locations
is one tenth of a standand deviation (or 0.05\% of the edge length of the landscape) 
up and to the right of their parent's location.
This results in reverse time gene flow down and to the left 
as in the biased asymmetric graph in Section \ref{sec:comvcoal}.
In terms of population genetics, 
this is very similar to the case where a population is expanding into new territory 
since in both cases, 
offspring tend to be further from the edge of the expansion than their parents.
The difference is that, in our case, 
the population is constrained to be a bounded distance 
away from the edge of the ``expansion''.
In both situations, individuals at the edge may have disproportionately more offspring,
leading to low genetic diversity and strong drift.
This results in substantially more noise in the isolation by distance plots 
as seen in Figure \ref{fig:ibd_comp}. 
The horizontal lines in that figure are likely from comparing genomes of individuals from
two families that have recently increased in size and geographical spread, 
resulting in many pairs of genomes with similar genetic distance 
but varying geographic distance.

As in Section \ref{sec:slim_unif}, we discretize the landscape into
both a $2 \times 2$ grid and a $4 \times 4$ grid for inference.
Looking at the medians of the inferred posteriors of the gene flow rates in the $2 \times 2$ case
(Figure \ref{fig:2x2grid_asb}),
it is immediately clear how strong the bias 
against reverse time gene flow out of the lower left corner is.
That is, hardly any individuals in the lower left area have ancestors
from other regions of the landscape.
This is to be expected as even though the single generation migration bias is small,
having a descendant in a location against the bias after a large number of generations
becomes less and less likely as the number of generations increases.
In the $4 \times 4$ discretization, 
the bottom left also shows very strongly biased gene flow (Figure \ref{fig:grid_bias_4x4_1}).
Most (but not all) of the other pairs also show bias in the expected direction.
Boxplots of the posterior distributions are shown in Figure \ref{fig:post_dists_bias_4x4_1}.
%\emph{maybe talk about/cite some biased random walk stuff here?}

%\emph{talk about some kind of log ratio test for the 4x4 case?}

\begin{figure}
\centering
% \includegraphics[scale=.35]{ibd_comp}
\caption{Comparison of the isolation by distance plots for the square landscape with uniform migration
and the square landscape with biased migration.
Faster coalescence in the biased migration case results in more noticeable horizontal lines,
indicative of families that recently increased in size and geographical spread.}
\label{fig:ibd_comp}
\end{figure}

\begin{figure}
\centering
% \includegraphics[scale=.5]{2x2grid_asb}
\caption{Medians of the posterior distributions of the inferred values of $g$ 
for the first replicate of the $2 \times 2$ discretization of the square landscape with biased migration.
The reverse time gene flow rates out of the bottom left section are very small,
indicating that very few individuals living in the bottom left of the landscape
have ancestors from other regions of the habitat,
consistent with migration biased up and to the right.}
\label{fig:2x2grid_asb}
\end{figure}

\begin{figure}
\centering
% \includegraphics[scale=.8]{grid_bias_4x4_1}
\caption{Medians of the posterior distributions of the inferred values of $g$ 
for the first replicate of the $4 \times 4$ discretization of the square landscape with biased migration.
As in the $2 \times 2$ case, 
the reverse time gene flow rates out of the bottom left section are very small,
indicating that very few individuals living in that area of the landscape
have ancestors from other regions of the graph.
Most, but not all of the other gene flow rates are biased 
in the direction we would expect.}
\label{fig:grid_bias_4x4_1}
\end{figure}

\begin{figure}
\centering
% \includegraphics[scale=.6]{post_dists_bias_4x4_1}
\caption{Boxplots of the posterior distributions for $g$ 
for the first replicate of the $4 \times 4$ discretization of the square landscape with biased migration.
Black boxes are gene flow rates down and to the left 
with each green box being the corresponding gene flow rate up and to the right.}
\label{fig:post_dists_bias_4x4_1}
\end{figure}

\section*{Application of Inference to Mojave Desert Tortoises}

We also apply the method to a population of Mojave desert tortoises (\emph{Gopherus agassizii})
living in California. 
Location and low coverage whole genome sequencing data was collected from 271 individuals 
\citep{shaffer2017desert}.
Pairwise genetic distances between all individuals was obtained from the same study.
The landscape was discretized into 13 regions based on watersheds and the locations of individuals.
The mean coalescence time matrix and standard deviations were calculated as before. 
Since the regions vary substantially in size, 
and may vary substantially in population density, 
we do not assume that coalescence rates are the same everywhere, 
which gives 13 values of $\gamma$. 
Since allowing only nearest neighbor migration gives 50 values of $g$, 
we have a total of 63 unknown parameters.
Because there are $13(14)/2 = 91$ unique values of the mean coalescence time matrix, 
we have more equations than unknowns.

Posterior distributions of gene flow and coalescence rate parameters were inferred
using $4 \times 10^7$ iterations of MCMC 
after $1 \times 10^6$ iterations of pre-burn-in and $3 \times 10^6$ iterations of burn-in.
The landscape and associated graph are shown in Figure \ref{fig:tort_land}. 
Boxplots of the posterior distributions for the gene flow rates can be seen in Figure \ref{fig:tort_post_g} 
and for the coalescence rates can be seen in Figure \ref{fig:tort_post_gam}.
There is substantial variation in the gene flow rates 
but not the coalescence rates, 
with the exception of region 4, which seems to have a relatively slow coalescence rate.
This is not too surprising given its large area and central location.
In general, the results are reasonable. 
For example,  we observe relatively low gene flow between areas with mountains on the border, 
%such as the Paiute and Castle Mountains between regions 1 and 4, 
%the Providence and Granite Mountains between regions 4 and 12, 
%the New York and McCollough Mountains between regions 1 and 9
such as between regions 1 and 4, 1 and 9, 4 and 12, 4 and 5, and 9 and 10.
On the other hand, there is an open connection between 9 and 13, 
and we infer high rates of migration between the two regions.
%say something about regions that don't touch much?
Overall, the model is a better fit than the resistance distance method of \citet{shaffer2017desert}
(Figure \ref{fig:tort_h_comp}).

The strong asymmetries, such as those present between regions 9 and 11 or 9 and 12, 
are probably the result of either source-sink dynamics or historical processes. 
The presence of source-sink dynamics would be of great interest for tortoise conservation efforts; 
however, it is too early in this analysis to use these numbers for that purpose.
For example, we would want to check for the effects of the sample configuration 
(e.g., are gene flow rates between regions 2 and 3 low 
simply because individuals sampled in region 3 are geographically far from region 2) 
as well as perform a sensitivity analysis, 
potentially including simulations on the landscape.
% sensitivity analysis would need to be performed, 
%as well as checking for the effects of the sample configuration.

\begin{figure}
\centering
% \includegraphics[scale=.34]{tort_land5}
\caption{Left: Locations of the sampled tortoises.
Right: Sample maximum likelihood estimates from the posterior distributions of the inferred values of $g$.}
\label{fig:tort_land}
\end{figure}

\begin{figure}
\centering
% \includegraphics[scale=.6]{tort_post_g}
\caption{Boxplots of the posterior distributions for $g$ 
for the Mojave desert tortoise data set 
on the thirteen region discretization of the landscape.}
\label{fig:tort_post_g}
\end{figure}

\begin{figure}
\centering
% \includegraphics[scale=.6]{tort_post_gam}
\caption{Boxplots of the posterior distributions for $\gamma$ 
for the Mojave desert tortoise data set 
on the thirteen region discretization of the landscape.}
\label{fig:tort_post_gam}
\end{figure}

\begin{figure}
\centering
% \includegraphics[scale=.8]{tort_h_comp}
\caption{Mean coalescence times calculated from the sample maximum likelihood model parameters 
plotted against the observed mean coalescence times calculated from the data, 
measured in the mean of pairwise $\pi$.
The mean absolute value relative differences between the two is 0.4\%.}
\label{fig:tort_h_comp}
\end{figure}

%%%%%%%%%%%%%%%%%%%%
\section*{Discussion}

resistance distance can mislead, especially in the presence of biased migration

coalescence time analogue is feasible but more computationally intensive

practical example (tortoises) shows strong asymmetry

but methods dealing with ill-posedness are necessary

With these methods, we provide a framework for
inferring demographic parameters of 
populations living on a habitat spread over a two dimensional landscape 
by using present day location and genetic sequencing information.
We are able to do inference of reverse time gene flow and coalescence rates 
with the matrix of mean coalescence times
in cases where the matrix is noise-free, noisy, or incomplete, 
or when it made after discretization of a simulated or real landscape.
To do this, we assume that the reverse time movements of lineages are Markov
and the population is near stationarity.
Typically, there is more error and uncertainty in the posterior distributions of the parameters
when there is more noise, slower coalescence rate, more missing values,
and multiple parameters for coalescence rate.

While the primary focus is coalescence time based inference,
we also compare the results with that of resistance distance (i.e., commute time) based inference.
Coalescence based inference typically has lower error than commute time based inference,
along with the benefit that you can also infer coalescence rates. 
The commute time plus diversity estimate, $\Dcom$, 
provides a good estimate of the coalescence times, $H$,
when gene flow rates are symmetric ($G_{ij} = G_{ji}$), 
but can be very wrong when they are asymmetric, 
including extreme discrepencies 
when there is an overall bias in the direction of the asymmetry across the landscape 
(Figure \ref{fig:4x4coalvcomH}).
When inference is performed on discrete landscapes 
with varying degrees of migrational asymmetry, 
coalescence time based inference is more accurate than commute time inference in all tested cases
(Figure \ref{fig:4x4box}).
Although performance of the commute time based method is reasonable 
when gene flow rates in the underlying model are symmetric, 
it becomes worse as the gene flow rates become more asymmetric 
to the point where there is little correlation at all with the true values 
when there is biased asymmetric migration.
On the other hand, commute time inference is much less computationally intensive, 
so much larger spatial systems can be used.
If the gene flow rates are symmetric, the higher resolution allowed by commute time inference
may outweigh the loss in accuracy.
The implication of these results for the body of literature using resistance distance methods
is that one may need to use caution when interpreting results.
For example, inferring high gene flow rates between two areas by using resistance distance
could potentially mean that those were areas were actually being fed by a common source.
Therefore, it may be worthwhile to apply a coarse resolution coalescence time based method 
alongside a finer resolution resistance distance method 
in order to check for evidence of asymmetric gene flow patterns.

We also apply the method to data from continuous space simulations. 
While there is substantial variation between the inferred values 
for different simulations with the same underlying demographic parameters, 
we are able to reliably find evidence of biased asymmetric migration (if it is present)
and verify the exitence of a barrier to migration, 
even when the landscape is otherwise relatively well connected.
Further work could involve more precise study of the relationship
between reverse time gene flow rates and forwards time migration rates.

\bibliography{references}

\appendix

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Finding $G$ from $H$}
\label{apx::hitting_calcs}

\plr{edit}

Suppose that $G$ is the generator matrix of an irreducible continuous-time Markov chain $X$,
so that $G_{ij} \ge 0$ for $i \neq j$ and
$$ G 1 = 0 $$.

Let $\tau_{j} = \inf\{t \ge 0 : X_t = j\}$ and $H_{ij} = \E[\tau_j \,|\, X_0 = i]$.
Then we know that
$$
    (G H)_{ij} = -1 \qquad \text{for} \; i \neq j ,
$$
i.e.,
$$
    GH = - 1 1^T + \diag(x).
$$

What is $x$?  Well, note that
$$
    G = (\diag(x) - 1 1^T) H^{-1}
$$
and so
$$ \begin{aligned}
    0 &= G1 \\
    &= (\diag(x) - 1 1^T) H^{-1} x 
\end{aligned} $$
and so
$$
    x_i = \frac{ 1^T H^{-1} 1 }{ (H^{-1} 1)_i } .
$$
Furthermore,
the Random Target Lemma \citep{aldous} % https://www.stat.berkeley.edu/users/aldous/RWG/Book_Ralph/Ch2.S2.html
tells us that $H \pi = 1$, so thus $H^{-1} 1 = \pi$, the stationary distribution of the chain.
Therefore, $x_i = 1/\pi_i$,
and in fact $1^T H^{-1} 1 = 1$.

In summary,
$$
    G = (\diag(1/H^{-1} 1) - 1 1^T) H^{-1} .
$$

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Finding $G$ from $R$}

The resistance distance is given by
$R = H + H^T$, so if we can modify $H$ by a skew-symmetric matrix,
we won't change $R$.
Can we do this and retain a valid $G$?
I.e., does $R$ uniquely specify $G$?
Let $Z + Z^T = 0$. 
Since $G$ is a continuous function of $H$, if all entries of $G$ are nonzero
we can find a small enough $\epsilon$ such that the $G'$ corresponding to $H' = H + \epsilon Z$
is also a valid generator matrix.
If $G$ has zero entries this may not be true.
To find a direction that we can change $H$ in that doesn't mess this up,
we could differentiate $dG/dH$.

\end{document}

### 2018-08-08 isolation_by_coalescence.tex
edited until results

@@
 
-\newcommand{\Dcom}{\widetilde D}
+\newcommand{\comdist}{\widetilde R}
 
@@
 \end{abstract}
-
-% NOTE: I tend to use semantic linebreaks.
-% Please excuse the ragged right margins in the source.
 
@@
 
+The \emph{commute time} of a Markov chain between locations $i$ and $j$
+is the mean time it takes a particle moving according to the rules of the chain, 
+started at $i$, to first reach $j$ and then return.
+We will denote this by $R_{ij}$, to emphasize that it is also the effective resistance
+between $i$ and $j$.
+The \emph{coalescence time} is the mean time it takes two particles started at $i$ and $j$
+to \emph{coalesce},
+which happens at a given rate whenever they are in the same location.
+We will denote this by $C_{ij}$.
+Why should these be related to genetic distance? 
+First, we explain this connection,
+and then discuss their computation and application.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection*{Genetic distance and coalescence time}
 
-\plr{pasted in: divergence is coalescence of lineages}
-The genetic distance between
-individuals, which provides a good estimate of the average time to a common ancestor along the genome, or
-the average coalescence time. Coalescence is the process in which genes or regions of the genome from two
-or more individuals, if traced back in time, eventually share a common origin [Wakeley, 2005]. While we
-cannot know exactly when this happens, if we assume the rate of mutation per nucleotide per generation is
-relatively constant, we can use the number of differences between two copies of a gene to estimate how much
-time has passed since their common ancestor. With enough loci, we can calculate an estimate for
-the average coalescence time across the genome [Hudson, 2007].
-
-\plr{pasted in: lineages as random walks}
-Even though the population dynamics in forwards time are Markov, the dynamics of lineages traced back
-in time (the coalescent process) may not be. For example, if fecundity is variable and population density
-is low, two lineages near each other are more likely to share a recent common origin, so knowing how one
-lineage moves gives you information about how the other lineage is likely to move. This effect becomes
-negligible as population density increases [Barton et al., 2002]), so it should not be an issue if the offspring
-of any one parent are typically interspersed with the offspring of many others. The methods I develop here
-model the coalescent process as Markov, which should be a good approximation for many systems.
-
-\plr{Pasted in, needs transition; make understandable without the below}
-The mean commute time of a location to itself is zero,
-so to fit to genetic data, we clearly must also account for local diversity.
-\plr{say why}
-Coalescence time does not have this property,
-because coalescence time accounts for both the first meeting time and the eventual coalescence,
-while commute time is more analagous to the first meeting time.
-For local diversity, we will also have another set of model parameters, $q$, 
-where $q_{i}$ approximates the observed coalescence time in location $i$.
-As is done by \citet{petkova2016visualizing},
-we use commute time to approximate genetic distance as
-\begin{align} \label{eq:hcom}
-	D_{ij} \approx \Dcom_{ij} = R_{ij}/4 + q_{i}/2 + q_{j}/2 .
+There are a number of ways to calculate measures of dissimilarity using genetic data \citep{distances}.
+In this paper we use mean pairwise divergence:
+for two genomes, it is calculated as the density of sites at which the two differ.
+The genomes differ at a site only if there has been a mutation in some ancestor on the lineages leading
+from the two genomes back to their most recent common ancestor at that site;
+under an infinite-sites model with average mutation rate $\mu$ per site and per meiosis,
+pairwise divergence is therefore an unbiased estimate of the mean time to most recent common ancestor,
+averaged across the genome \citep{hudson2007variance,ralphXX}.
+It is natural to model this process by following the two lineages back through time
+(and across space)
+until they find their most recent common ancestor,
+i.e., until they coalesce.
+It is common to model the geographic distribution of a species
+as a collection of randomly mating populations
+that exchange occasional migrants,
+thus discretizing geographic space.
+If the genome is selectively netural,
+then the motion of these lineages back through time
+forms a Markov process, the \emph{structured coalescent} \citep{wakeley2005coalescent}.
+In this framework, each lineage performs a random walk across the populations,
+with movement probabilities depending on the flux of individuals between populations.
+This provides the link between genetic distance and random walks.
+
+For data, we are given genotypes of individuals and their geographic coordinates;
+then, we divide these individuals into groups,
+and compute the genetic distance between group $i$ and group $j$
+as the mean divergence between a pair of genomes, chosen one from each group.
+We denote the resulting genetic distance $D_{ij}$;
+since divergence is symmetric, $D_{ij} = D_{ji}$.
+We compute $D_{ii}$ by sampling without replacement from group $i$,
+so that generally $D_{ii} > 0$: the diagonal elements measure local diversity.
+As discussed below, we will find commute times so that the mean commute time between
+$i$ and $j$ matches $D_{ij}$ -- but the commute time of a location to itself is zero,
+so we must account for local diversity.
+We do this in the same way as \citet{petkova2016visualizing},
+by introducing another set of model parameters, $q$, 
+that fill the same role as the coalescence rates, $\gamma$.
+Then, we use commute time to approximate genetic distance as
+\begin{align} \label{eq:commute_approx}
+	D_{ij} \approx \comdist{ij} = R_{ij}/4 + q_{i}/2 + q_{j}/2 .
 \end{align}
-To gain an intuitive feel for why resistance distance is divided by 4, 
-consider how, if the movement rates are symmetric, 
-a single particle moving from one state to another is similar to (but not the same as)
-the first meeting time of two particles, 
-one moving forward in time from the first location 
-and the second one moving backward in time from the other, 
-so it is similar to twice the first meeting time.
-Since the commute time is the sum of the hitting times in each direction, 
-it is similar to 4 times the first meeting time.
+The factor of four appears to make commute time more similar to coalescence time:
+the commute time covers the path from $i$ to $j$ twice
+(once in both directions),
+while the coalescence time covers the distance only once,
+and in half the time, because two particles are moving.
+Furthermore, coalescence time includes within-location diversity,
+because particles do not coalesce instantly --
+commute time is most analogous to the first meeting time.
 For a conceptual picture of the differences between the coalescence time and commute time,
 see Figure \ref{fig:concept_coalcom}.
-The average of the local diversity estimates ($q$) are then added 
-to the first meeting time estimate to account for the eventual coalescence after meeting.
-%\emph{Briefly motivate by describing "first meeting" plus "subsequent coalescence" -> 
-%	commute + diversity approximation.}
-In this case, these values
-are parameters in the model to be inferred instead of within deme coalescence rates ($\gamma$).
-Since the values a $q$ are typically fairly easy to infer 
-and requiring them all to be the same can significantly decrease the quality of the fit, 
-we will not be using an analogue of Problem Modification \ref{probmod:const-coal} 
-(where we would only have one parameter for $q$).
-Commute time inference has the advantage of only needing to use the base Markov chain
-instead of the product chain,
-but it also leads to incorrect inferences in certain circumstances.
-Using the base Markov chain instead of the product chain is advantageous 
-because, as the grid size increases, it is much more computationally efficient 
-to work with a Markov chain with $n$ states than $n^2$ states 
-since the matrices would be much smaller.
 
@@
     % \includegraphics[scale=.5]{conceptn}
-\caption{Illustration of the conceptual differences
+    \caption{
+    Illustration of the conceptual differences
     betweeen coalescence and commute time between states 1 and 3 
@@
     Note how the time axis is scaled to be faster in the figure showing commute time
-than the one for coalescence time in order for them to fit in the same space.}
-\label{fig:concept_coalcom}
-\end{figure}
-
-For a simple example where the coalescence times and commute time approximations are very different, 
-consider a system of three islands in a line,
-where lineages currently in the outer islands are much more likely to come from the central island
-than the other way around.
-In this case, lineages are likely to quickly move to the center and coalesce regardless of starting position,
-so coalescence time will be relatively short between all individuals.
-Commute time between the two outer locations, on the other hand,
-will be much longer since it requires the lineage to leave the center.
-If we plug real numbers into this example, say, 
-reverse time gene flow rate of $1$ into the center and $0.1$ out of it
-with $\gamma=1$ everywhere,
-then $H_{13}=H_{31}=2.90$ but $\Dcom_{13}=\Dcom_{31}=7.94$.
-However, the fact that $\Dcom$ as predicted from the correct parameters may be far from $H$
-does not mean that the parameters inferred using resistance distance methods
-will necessarily be very wrong.
-In principle, it still might be that the parameters for which $\Dcom$ is closest to $H$ 
-are close to the real parameters, 
-at least in some situations.
-
-\begin{figure}
-\centering
-% \includegraphics[scale=.5]{three_state_example}
-\caption{A simple example where the coalescence time and commute time based approximation
-are very different.
-In this case, commute time between the two outer locations will be much longer 
-than coalescence time.}
-\label{fig:3_state}
+    than the one for coalescence time in order for them to fit in the same space.
+    } \label{fig:concept_coalcom}
 \end{figure}
@@
 
-First, we explain how to compute the relevant quantities --
+Now, we explain how to compute the relevant quantities --
 mean commute and coalescence times -- from a Markov chain model of lineage movement.
@@
 This can also be seen by working with the product Markov chain $(X_t, Y_t)$,
-whose generator matrix is $G \prod I + I \prod G$,
-where $I$ is the identity matrix and $\prod$ is the Kronecker product.
+whose generator matrix is $G \otimes I + I \otimes G$,
+where $I$ is the identity matrix and $\otimes$ is the Kronecker product.
 
 Equation \eqref{eqn:C_matrix} is linear in $G$ and $\gamma$,
-and so can be solved with straightforward linear algebra.
+and so after rearrangement can be solved with standard linear algebra.
 If $\gamma$ is given, then solving for $G$ is an example of a XXX equation \citep{matrixeqn}.
@@
 for the coalescence times:
+\plr{simplify these equations and remove matrix version?}
 \begin{align*}
-   -1 &= -C_{11} G_{12} + C_{21} G_{12} - C_{11} G_{12} + C_{12} G_{12} - C_{11} \gamma_1 \\
-   -1 &=  C_{11} G_{21} - C_{21} G_{21} - C_{21} G_{12} + C_{22} G_{12} \\
-   -1 &=  C_{12} G_{21} - C_{22} G_{21} + C_{21} G_{21} - C_{22} G_{21} - C_{22} \gamma_2 .
+   -C_{11} G_{12} + C_{21} G_{12} - C_{11} G_{12} + C_{12} G_{12} - C_{11} \gamma_1 &= -1 \\
+    C_{11} G_{21} - C_{21} G_{21} - C_{21} G_{12} + C_{22} G_{12}                   &= -1 \\
+    C_{12} G_{21} - C_{22} G_{21} + C_{21} G_{21} - C_{22} G_{21} - C_{22} \gamma_2 &= -1 .
 \end{align*}
@@
 and estimates uncertainty in the resulting estimates.
-Suppose that $\hat{C}_{ij} = C + \epsilon_{ij}$,
-where $\epsilon$ is Normally distributed
-with mean $\mathbf{0}$ and covariance $\Sigma$.
-For now, we will make a few assumptions about $\epsilon$.
-First, that all $\epsilon_{ij}$ are independent, identically distributed $\mathcal{N}(0,\sigma_{\epsilon}^2)$
-for $i \leq j$, and $\epsilon_{ij} = \epsilon_{ji}$.
-Additionally, we will assume that $C_{ij}$ is large compared to $\sigma_{\epsilon}$
-such that the probability that a value of $\hat{C}$ is ever negative is negligible.
-Writing $f(\mathbf{x})$ as the probability density function of 
-the multivariate normal distribution $\epsilon$,
-the likelihood function for the pair $(G,\gamma)$ is
-\begin{align*}
-\mathcal{L}(G,\gamma;\hat{C}) &= f(\mathcal{C}(G,\gamma) - \hat{C}) \\
-			      &= f(-(G^*-\text{diag$(\Gamma)$})^{-1} \mathbf{1} - \hat{C})
-\end{align*}
-recalling that $\mathcal{C}$ is the hitting time matrix for a given pair $(G,\gamma)$
-defined by equation \ref{eq:solve_H}. 
-If $\hat{C}$ is a valid coalescence time matrix,
-we can find the maximum likelihood estimate (MLE) of $G$;
-however, this is in practice unlikely to happen, 
-and the likelihood function is still useful 
-for providing information about the uncertainty of $G$.
-
-To complete the Bayesian model,
-we place priors on the parameters.
-Let's say that $i$ is said to be adjacent to $j$ if $i \neq j$ and $G_{ij}$ is one of 
-the values known to be non-zero.
-Then we could let the prior probability density for the movement parameters be that of an 
-exponential random variable with rate parameter $\lambda_{G}$. 
-That is, the prior likelihood of $G$ would be
-\begin{align*}
-\mathcal{L}_G(G) = 
-\begin{cases}
-	\lambda_{G}^{m_G} \prod_{i \sim j} e^{-\lambda_{G} G_{ij}} 
-		\quad & \text{for $G_{ij} \geq 0 \quad \forall \ i \neq j$} \\
-	0 & \text{otherwise}
-\end{cases}
-\end{align*}
-where $i \sim j$ means that $i$ and $j$ are adjacent and $m_G$ is the number of positive values of $G$. 
-Similarly, we can put a prior density on $\gamma$:
-\begin{align*}
-\mathcal{L}_{\gamma}(\gamma) = 
-\begin{cases}
-	\prod_{i}^{n} \lambda_{\gamma} e^{-\lambda_{\gamma}\gamma_i} 
-		\quad & \text{for $\gamma_i \geq 0 \quad \forall \ i$} \\
-	0 & \text{otherwise.}
-\end{cases}
-\end{align*}
-Recall that $n$ is the number of states in the single particle Markov chain.
-If we assume that all $\epsilon_{ij}$ are IID with $\epsilon_{ij} \sim \mathcal{N}(0,\sigma_{\epsilon}^2)$ 
-for $i \leq j$
-we can write the likelihood for $\hat{C}$:
-\begin{align*}
-\mathcal{L}(\hat{C};G,\gamma) = \frac{1}{(2 \pi \sigma_{\epsilon}^2)^{m/2}} \prod_{i \leq j}
-	\exp\left\{-\frac{(\mathcal{C}_{ij}(G,\gamma) - \hat{C}_{ij})^2}{2 \sigma_{\epsilon}^2}\right\},
-\end{align*}
-where $m$ is the number of nonmissing values on or above the diagonal in $\hat{C}$.
-Then the posterior distribution is
+To do this, we model genetic distances as equal to coalescence times plus noise:
+$D_{ij} = C_{ij} + \epsilon_{ij}$,
+where $\epsilon_{ij}$ are independent and Normally distributed
+with mean $\mathbf{0}$ and variance $\sigma_\epsilon^2$,
+except that $\epsilon_{ij} = \epsilon_{ji}$.
+(We imagine that $C_{ij}$ is large compared to $\sigma_{\epsilon}$,
+so that resulting values are still positive.)
+Using the calculations above, 
+given $G$ and $\gamma$ we can compute the corresponding coalescence times
+-- call these $\mathcal{C}(G, \gamma)$ --
+so our model is that $D$ has a Gaussian distribution 
+with mean $\mathcal{C}(G, \gamma)$.
+The maximum likelihood estimate of $G$ could be found directly;
+however, this in practice will quite likely contain negative movement rates,
+and does not allow us to impose constraints
+(such as only allowing a subset of movement rates to be nonzero).
+
+Therefore, we use Bayesian inference,
+placing independent exponential priors on the parameters --
+on those movement rates $G_{ij}$ that are not constrained to be zero,
+and on the coalescence rates, $\gamma$.
+Concretely, suppose that the spatial arrangement of populations is given by a graph,
+and only movement rates corresponding to edges in the graph are allowed to be nonzero.
+Write $i \sim j$ to mean that $i$ and $j$ are adjacent in the graph,
+that there are $m_G$ edges in the graph,
+and that each of the $n$ populations has its own coalescence rate.
+Then, the resulting log-posterior density is
 \begin{align} \label{eq:post}
-\mathcal{L}(\hat{C};G,\gamma) \mathcal{L}_{G}(G) \mathcal{L}_{\gamma}(\gamma) = 
-	\frac{\lambda_{G}^{m_G} \lambda_{\gamma}^n}{(2 \pi \sigma_{\epsilon}^2)^{m/2}}
-	\exp\{-\sum_{i \sim j} \lambda_G G_{ij} -\sum_{i}^n \lambda_{\gamma}\gamma_i
-		-\frac{1}{2 \sigma_{\epsilon}^2} \sum_{i \leq j} (\mathcal{C}_{ij}
-		(G,\gamma) - \hat{C}_{ij})^2 \}
+\mathcal{L}(D, G, \gamma) 
+    &=
+    2 m_G \log(\lambda_G) + n \log(\lambda_\gamma) 
+    - \frac{n(n+1)}{2} \log(2 \pi \sigma^2_\epsilon)
+	- \sum_{i \sim j} \lambda_G G_{ij}  \\
+    &\qquad
+    -\sum_{i=1}^n \lambda_{\gamma}\gamma_i
+		-\frac{1}{2 \sigma_{\epsilon}^2} \sum_{i \leq j} \left(
+            \mathcal{C}_{ij}(G,\gamma) - D_{ij}
+        \right)^2 
 \end{align}
-with the same support over $G$ and $\gamma$ as in their individual priors.
-To sample from this distribution, 
-we will use the Metropolis-Hastings algorithm \citep{brooks2011handbook}.
-
-\paragraph{Incomplete observations}
-With real world data, 
-unless the number of states is small or we have a large amount of data,
-our pairwise average coalescent time matrix will have a large number of unobserved values.
-For example, if our landscape is a $20 \times 20$ grid and we take $100$ samples, 
-each from a different grid square,
-we would have only $100(101)/2$ out of $400(401)/2$ of the unique values for $H$. 
-(Note that we are including the self-comparisons in this example;
-if the organisms were haploid or inbreeding were substantial,
-our $H$ matrix would have only $99(100)/2$ usable values.)
+Here $1/\lambda_\gamma$ and $1/\lambda_G$ are the prior means of $\gamma_i$ and $G_{ij}$,
+respectively.
+
+Inference with commute times is done in exactly the same way,
+except that $q$ replaces $\gamma$,
+and instead of $\mathcal{C}(G,\gamma)$ we use $\comdist(G, q)$
+as defined in equation \eqref{eq:commute_approx}.
+
+It is not required in this approach to have samples from every spatial location,
+allowing inference on incompletely sampled spatial discretizations.
 To carry out inference in this situation,
 we simply use Equation \ref{eq:post},
-but the rightmost sum will be only over observed entries of H.
-
-%%%%%%%%%%%%%%%%
+but only summing over observed entries of $D_{ij}$.
+
+\paragraph{MCMC methods}
+To sample from the posterior distribution of equation \eqref{eq:post}, 
+we used R \citep{R} 
+to implement a standard Metropolis-Hastings algorithm with Gaussian proposals \citep{brooks2011handbook}.
+Starting locations were chosen by sampling from the prior distribution.
+Before the usual ``burn-in'' phase before sampling for inference begins,
+we included a period of ``pre-burn-in'' which used the same MCMC procedure
+with a larger $\sigma_\epsilon$, to allow the chain to more quickly converge on the high-posterior
+portion of parameter space.
+Typically, we ran MCMC for $10^6$ iterations of pre-burn-in 
+and $3 \time 10^6$ iterations of burn-in,
+followed by $4 \times 10^6$ iterations which were used to estimate posterior distributions.
+We used standard methods to assess convergence and mixing.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection*{Test cases and simulations}
 
-\plr{describe graphs and slim sims}
-
-\plr{MCMC methods: pre-burnin, burnin, number of iterations, trace plots}
+We compared inference under coalescence and commute time models
+using both data ``from the model'' and more realistic simulations.
+For the first category, we designed a number of graphs to test particular aspects of inference,
+with migration rates specified on each directed edge
+(plotted below using igraph \citep{igraph}).
+Then we calculated exact coalescence and commute times on these graphs,
+and added Gaussian noise independently to each entry of the resulting matrices
+(while preserving symmetry)
+as described above.
+Particular parameters and levels of noise are given in the Results.
+
+These simulations are expected to be nearly equivalent to population-based simulations
+(either forwards-time or coalescent);
+the only difference being that noise terms about the theoretical mean should be somewhat correlated.
+
+\paragraph{Forwards-time simulations}
+To produce realistic data, we implemented forwards-time simulations using SLiM v3.0 \citep{haller2017},
+with individuals living across continuous, two-dimensional space (sometimes with barriers),
+from which we recorded genomic data.
+We performed three types of simulation:
+uniform migration on a continuous square landscape, 
+uniform migration on a rectangular landscape with barriers to migration, 
+and biased migration on a square landscape.
+Each simulation had $10^4$ diploid, hermaphroditic individuals with nonoverlapping generations.
+Mates were selected from nearby individuals using a Gaussian kernel with standard deviation $\sigma_d$,
+truncated at $3 \sigma_d$.
+Offspring are dispersed away from the mother using the same kernel,
+conditioned to lie within the habitable area.
+In each simulation, $\sigma_d$ was chosen 
+so that there were an average of 0.25 individuals per $\sigma_d^2$ units of area.
+In order to prevent excessive population clumpiness \citep{felsensten1975pain}, 
+we also model local competition. 
+To do this, we use the same truncated Gaussian kernel to define an ``interaction strength''
+between each pair of nearby individuals;
+then, the fitness of each individual is set to $\exp(-x/40)$, 
+where $x$ is the sum of their interaction strengths with other individuals.
+This reduces fitness for individuals in denser areas.
+Individuals are chosen to reproduce proportional to their fitness.
+Clumpiness is similar to what is seen in 
+Figure \ref{fig:ind_locs_5x3b_1}
+
+Individuals have 10 pairs of ``chromosomes'' with $2 \times 10^5$ loci each 
+and a recombination rate of $10^{-8}$ per generation per locus.
+Mutation rate was $10^{-8}$ per locus for the simulation with uniform migration and no barrier,
+and $10^{-7}$ for the remaining simulations.
+In all cases, the number of mutations was more than adequate to reliably estimate genetic distance.
+
+To compute genetic distances,
+geographic space was partitioned into rectangular regions as depicted below.
+within each region, 
+individuals are sampled uniformly for ``genotyping'' 
+from among those within the middle three quarters of each dimension of rectangle,
+as shown for one situation in Figure \ref{fig:ind_locs_5x3b_1}.
+This protocol was chosen as a compromise between sampling all individuals 
+from near the center of each location,
+which could result in a large number of close relatives being sampled, 
+leading to an underestimate of the mean genetic distance for individuals in that location
+and uniform sampling from the whole area of each grid, 
+which may result in many sampled individuals being very close to other grid squares.
+Mean genetic distances for each pair of regions is then computed as described above
+from genome sequence provided by SLiM.
+To estimate uncertainty, we computed the standard error of these mean distances
+using the minumum number of individuals in each location.
+
+Before being passed to the MCMC inference method,
+we rescaled genetic distances so that the movement rates resulting from the inference
+would be approximately of order one.
+We did this by multiplying genetic distances by a constant
+so that the rescaled mean of the entries of $D$
+is equal to the number of locations in the discretization.
 
@@
 all parameters moving right or up are 0.5.
-Differences between true mean coalescence times ($H$) and mean commute times ($\Dcom$) 
+Differences between true mean coalescence times ($H$) and mean commute times ($\comdist$) 
 are shown for each graph in Figure \ref{fig:4x4coalvcomH}. 
@@
 and 3.27 and 0.053 for the biased asymmetric graph, respectively.
-We can see that, as expected based on the differences between $H$ and $\Dcom$,
+We can see that, as expected based on the differences between $H$ and $\comdist$,
 the coalescence time inference performs very well in all four test cases
@@
 % \includegraphics[scale=.8]{4x4coalvcomH}
-\caption{Plots of the true values of $\Dcom$ vs $H$ 
+\caption{Plots of the true values of $\comdist$ vs $H$ 
 for the uniform graph, symmetric graph, the asymmetric graph,
@@
 
+other measures of genetic distance: what do people use in practice?
+
+\paragraph{Consequences for resistance methods}
+We have shown that commute time can show dramatically different patterns
+than coalescence time, especially if there is systematic bias in dispersal.
+How should this affect the results of resistance-based methods to genomic data,
+besides general inaccuracy?
+Suppose, for instance, that seed dispersal for a species of tree
+is biased downhill,
+and the landscape is dominated by a large valley.
+Since lineages tend to move uphill,
+genetic distance between locations on opposite slopes of the valley will be relatively large,
+so resistance methods will infer a barrier along the bottom of the valley
+when no barrier to dispersal exists.
+On the other hand, if there is a ridge instead of a valley,
+resistance methods will show no barrier (and relatively high movement rates),
+even if there is very little uphill dispersal.
+This could lead to a falsely high assessment of gene flow across the top of the ridge.
+
 With these methods, we provide a framework for
@@
 along with the benefit that you can also infer coalescence rates. 
-The commute time plus diversity estimate, $\Dcom$, 
+The commute time plus diversity estimate, $\comdist$, 
 provides a good estimate of the coalescence times, $H$,
@@
 
+\plr{pasted in: nonmarkov lineages}
+Even though the population dynamics in forwards time are Markov, the dynamics of lineages traced back
+in time (the coalescent process) may not be. For example, if fecundity is variable and population density
+is low, two lineages near each other are more likely to share a recent common origin, so knowing how one
+lineage moves gives you information about how the other lineage is likely to move. This effect becomes
+negligible as population density increases [Barton et al., 2002]), so it should not be an issue if the offspring
+of any one parent are typically interspersed with the offspring of many others. The methods I develop here
+model the coalescent process as Markov, which should be a good approximation for many systems.
+
 \bibliography{references}

### 2018-08-08 isolation_by_coalescence.tex
some figs
edited mostly until discussion

@@
 
+We apply the method to the desert tortoise \plr{say more here}.
+
 \plr{I didn't find myself wanting to explain isolation by distance here:
@@
 \centering
-    % \includegraphics[scale=.5]{conceptn}
+     \includegraphics[scale=.5]{figs/conceptn}
     \caption{
@@
 \centering
-% \includegraphics[scale=.7]{valid_range}
+         \includegraphics[scale=.7]{figs/valid_range}
     \caption{The green shaded region shows 
@@
 and $3 \time 10^6$ iterations of burn-in,
-followed by $4 \times 10^6$ iterations which were used to estimate posterior distributions.
+followed by $4 \times 10^6$ iterations which were used to estimate posterior distributions,
+although for small graphs fewer iterations were necessary.
 We used standard methods to assess convergence and mixing.
@@
 \plr{transition paragraph}
-% biased migration is bad for commute times: 
-%              figure 19 showing the two against each other, and the simple example
-% continuous space: can find big barriers but lots of between-replicate variability (life is still hard)
-% tortoises: results on small graph, showing strong asymmetry (?)
-
-\plr{Summarizing observations about inference being harder with more noise,
-more parameters,
-and lower coalescence rate.}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Model validation}
+
+We first tested the method using data generated under the model.
+To test the methods across a wide range of situations,
+we generated random movement rates for 
+a large number of $3 \times 2$ rectangular grids:
+twenty-five symmetric (i.e., with $G_{ij} = G_{ji}$) and twenty-five asymmetric.
+We generated movement parameters
+by rounding independent exponential random variables up to the nearest one tenth,
+and fixed coalescence rate at 1 for all locations.
+We then added independent noise to each matrix entry (preserving symmetry),
+and assume that we can estimate the amount of noise reasonably well, 
+so used the standard deviation of the noise for $\sigma_\epsilon$ in the likelihood function.
+To assess the impact on inference of the number of parameters,
+we performed inference both with and without the assumption 
+that coalescence rate is the same everywhere.
+For $3 \times 2$ graphs, there are 21 equations
+and 14 movement parameters,
+so that with a global coalescence rate there are 15 unknowns
+but with individual coalescence rates there are 20 unknowns,
+which is close to nonidentifiable.
+(For commute time inference, we analogously constrained all values of $q$ to be identical or not.)
+To quantify accuracy across these random graphs in different situations,
+we computed, for each model fit on a particular random graph,
+the absolute difference between the posterior median of each movement parameter
+and the true value.
+We then averaged these across movement parameters to produce a measure of accuracy for that model fit,
+reported below as ``mean absolute error''.
+
+\paragraph{Varying noise}
+Unsurprisingly, coalescence time inference became less accurate as the amount of noise increased.
+As we varied the standard deviation of the noise
+from 1/1000 to 1/50 of the mean value of $C$ for that grid,
+the mean absolute error 
+for coalescence time inference with a single coalescence rate
+increased from around 0.05 to 0.4.
+Posterior interquartile ranges increased proportionally.
+(although there was substantial variation in accuracy between replicates;
+see Figures \ref{fig:mult_noise} and \ref{fig:mult_noise_iqr}).
+Increasing the number of parameters by allowing multiple coalescence rates
+made the inference problem even harder --
+mean absolute errors no longer depended strongly on the amount of noise added,
+and hovered around 0.4.
+
+More surprisingly, the performance of commute time inference did not depend on the amount of noise --
+mean absolute error was around 0.6 across all levels of noise,
+both with and without more than one parameter for local diversity
+(see Figure \ref{fig:mult_noise}).
+Although symmetry did not affect coalescence time inference,
+commute time inference did substantially worse on asymmetric graphs,
+showing median absolute errors of around 0.8.
+
+\paragraph{Varying coalescence rates}
+Lower coalescence rates mean that lineages wander about the landscape for longer
+before they coalesce, 
+thus greatly reducing the geographic information we get from coalescence times.
+Indeed, coalescence rates had a strong effect on feasability of the inference problem
+(shown in Figures \ref{fig:mult_gam} and \ref{fig:mult_gam_iqr}):
+lowering the coalescence rate from 10 to 0.1
+(always adding noise with standard deviation 1/200th of the mean value)
+increased mean absolute error of coalescence time inference from around 0.05 to 0.5.
+Again, allowing multiple coalescence rates
+or using commute time inference were highly inaccurate in all cases,
+with mean absolute errors of above 0.5.
 
@@
 
-\plr{TODO: replace bottom two panels in figure 18 with same thing for commute time}
-
-\plr{pasted in}
-We now test the method's ability to locate barriers to migration
-on a $5 \times 3$ grid with asymmetric gene flow. 
-There are impassable boundaries on the bottom left and top right of the landscape,
-but the overall graph remains fairly well connected (see Figure \ref{fig:5x3b_grid}).
-Each value of $g$ in the true model is determined randomly as before.
-The coalescence rates are set to 1 everywhere, 
-and for the first case, we assume that we know it is the same everywhere, 
-although the specific value is not known.
-Gaussian noise with standard deviation equal to 1/1000 of the mean of $H$ is added to form $\hat{H}$, 
-and inference is performed using all values of $\hat{H}$.
-All inference in this section is done using
-$5 \times 10^6$ iterations of MCMC 
-after $4 \times 10^6$ iterations for burn-in and $1 \times 10^6$ iterations for pre-burn-in.
-In this case, the mean absolute error of $\hat{g}$ was 0.019, 
-while the error of $\hat{\gamma}$ was 0.0028. %\emph{talk more about results here?}
-This analysis is repeated in three more situations, 
-all with noise set to have standard deviation 1/100 of the mean of $H$.
-In the first case, there are no missing values of $\hat{H}$, 
-in the second case, values from rows and columns
-corresponding to populations 2, 5, 7, and 15 are missing from $\hat{H}$. 
-The third case is the same as the second except that
-there is a coalescence parameter for each location on the grid 
-instead of one parameter for all locations.
-With no missing values and noise of 1/100, 
-the mean absolute error of $\hat{g}$ was 0.17 and the error of $\hat{\gamma}$ was 0.0066.
-With the four missing values,
-the mean absolute error of $\hat{g}$ was 0.37 and the error of $\hat{\gamma}$ was 0.033.
-With missing values and multiple coalescence parameters, 
-the mean absolute error of $\hat{g}$ was 0.52 
-and the mean absolute error of $\hat{\gamma}$ was 0.42.
-Boxplots of the posterior distributions of $g$ can be seen in Figure \ref{fig:5x3boxplots}.
-We can see that the inference method does very well 
-in the case with noise of 1/1000.
-With each modification, the problem becomes less well defined, 
-so it is not surprising that the error increases as well.
-Notably, we are still able to identify the barrier 
-except when there is data missing from both locations on either side of the barrier, 
-as is the case for movement parameters $g_{5}$ and $g_{17}$ with locations 2 and 7.
-Additionally, inference of the coalescence rate is very accurate 
-when there is only one coalescence rate parameter,
-but is much harder with multiple parameters, 
-which in turn makes the movement parameters harder to estimate.
-
-%This is repeated with noise set to have standard deviation 1/100 of the mean of $H$ 
-%and with data missing from 4 of the 15 locations (\emph{also non-constant coalescence?}).
-%In this case, the mean absolute error of $\hat{g}$ was (insert) and (appropriate statement for $\gamma$)
-%after (insert) iterations of MCMC.
-%\emph{talk more about results here}
-
-\begin{figure}
-\centering
-% \includegraphics[scale=.6]{5x3b_grid}
+We then test the method's ability to locate barriers to migration
+on the $5 \times 3$ grid with asymmetric gene flow shown in Figure \ref{fig:5x3b_grid}. 
+Each nonzero movement rate is determined randomly as before.
+Coalescence rates are set to 1 everywhere.
+Figure \ref{fig:5x3boxplots} shows posterior distributions of the movement rates
+for both coalescence and commute time inference
+on this graph, with noise standard deviation equal to
+1/1000 of the mean coalescence time (left)
+and 1/100 of the mean coalescence time (right).
+Although increased noise in the data increased uncertainty in the estimated values,
+coalescence time inference correctly inferred the location of the barrier
+(as well as high-dispersal connections) in both cases.
+Commute time inference \plr{did what?}
+We also explored making the problem harder,
+by first (a) dropping rows and columns from the data corresponding to populations 2, 5, 7, and 15;
+and then (b) allowing a separate coalescence rate for each location,
+in both cases using a noise level of 1/100.
+The resulting posterior distributions are shown in Figure \ref{fig:5x3boxplots_mult_gam},
+and show substantially increased error,
+but only in movement rates to the removed locations.
+As before, uncertainty was increased when multiple coalescence rates were allowed,
+but not as strongly as before, likely because the problem is less close to being underconstrained,
+having 105 equations and 59 unknowns.
+
+\begin{figure}
+\centering
+     \includegraphics[scale=.6]{figs/5x3b_grid}
     \caption{Plot of the grid structure 
-for the 5 $\times$ 3 grid with barriers.
+    for the 5 $\times$ 3 grid with two barriers:
+    movement rates between $\{2,3\}$ and $\{7,8\}$ are zero,
+    as are those between $\{8,9\}$ and $\{13,14\}$.
     Values for the non-zero movement rates 
@@
 \centering
-% \includegraphics[scale=0.8]{5x3boxplots}
+     \includegraphics[scale=0.8]{figs/5x3boxplots}
     \caption{Posterior distributions for values of $g$ 
@@
     for each of the analysis cases.
-See Figure \ref{fig:5x3b_grid} to see movement parameter locations.}
+    See Figure \ref{fig:5x3b_grid} to see movement parameter locations.
+    \plr{TODO: replace bottom two panels with same thing for commute time}
+}
     \label{fig:5x3boxplots}
@@
 \subsection*{Biased migration}
-\label{sec:comvcoal}
-
-\plr{transition about how we expect biased migration to produce large differences}
-
-In order to further illustrate the differences between commute time and coalescence time inference,
-we test both methods on four different $4 \times 4$ graphs (shown in Figure \ref{fig:4x4_grids}:
-one where movement parameters are all equal to 1 (which we will call the uniform case),
-one where movement parameters are symmetric but randomly generated, 
-one asymmetric graph where all movement parameters are randomly generated,
-and one asymmetric graph where there is a bias in moving towards one direction in the graph.
-For the symmetric and unbiased asymmetric graphs, movement parameters are chosen randomly as before.
-In the biased asymmetric graph, all parameters for moving left or down are 2 while 
-all parameters moving right or up are 0.5.
-Differences between true mean coalescence times ($H$) and mean commute times ($\comdist$) 
-are shown for each graph in Figure \ref{fig:4x4coalvcomH}. 
-The differences are fairly small in the uniform and symmetric cases, moderate in the asymmetric case, 
+\label{sec:biased_migration}
+
+Both matrices of coalescence and commute times, $C$ and $R$, are symmetric,
+but they do not deal with asymmetric (biased) migration in the same way.
+Since commute time to $i$ requires paths both to and from $i$,
+but coalescence time to $i$ only reaquires lineages to leave $i$,
+if, for instance, there is a low rate of migration back into $i$,
+then commute times to $i$ could be quite long even while coalescence times are not.
+For instance, suppose there are three populations arranged in a line,
+and lineages currently in the outer populations are much more likely to come from the central island
+than the other way around.
+Since lineages quickly move to the center and coalesce regardless of starting position,
+coalescence time (and genetic diversity) will be relatively low between all individuals.
+Commute time between the two outer locations, on the other hand,
+will be much longer than other comparisons, since it requires a lineage to leave the center.
+
+In order to investigate this general situation,
+we tested both methods on four different $4 \times 4$ graphs (shown in Figure \ref{fig:4x4_grids}):
+\emph{uniform}, where all movement rates are 1.0;
+\emph{symmetric}, where movement rates are symmetric and randomly generated as before;
+\emph{asymmetric}, where all movement rates are random;
+and \emph{biased}, where movement rates down or to the left are equal to 2.0,
+and movement rates up or to the right are 0.5.
+
+Differences between true mean coalescence times ($C$) and mean commute times ($\comdist$) 
+are shown in Figure \ref{fig:4x4coalvcomH}:
+they are fairly small in the uniform and symmetric cases, moderate in the asymmetric case, 
 and extreme in the biased asymmetric case.
-%The structure of the graphs themselves can be seen in Figure \ref{fig:4x4_grids}.
-Inference is done using $\hat{H}$ with a noise level of 1/200 the mean value of $H$ for each grid.
-All trials with each inference method were done using $4 \times 10^6$ iterations of MCMC 
-after $1 \times 10^5$  iterations of pre-burn-in and $1 \times 10^6$ iterations of burn-in.
-Posterior distributions for each graph and and inference method are shown in Figure \ref{fig:4x4box}.
-The coalescence time inference was substantially more accurate in all cases.
-The commute time based inference and coalescence time based inference methods had 
-mean absolute errors in $\hat{g}$ of 0.30 and 0.12 respectively for the uniform graph,
-0.49 and 0.053 for the symmetric graph, 
-0.87 and 0.035 for the asymmetric graph,
-and 3.27 and 0.053 for the biased asymmetric graph, respectively.
-We can see that, as expected based on the differences between $H$ and $\comdist$,
-the coalescence time inference performs very well in all four test cases
-while the commute time inference performs fairly well in uniform and symmetric cases,
-poorly in the asymmetric case, and very poorly in the biased asymmetric case.
-The behavior of the biased asymmetric graph supports our previous thoughts about
-situations similar to the three state system seen in Figure \ref{fig:3_state},
-where commute times are much longer than coalescence times.
-In this case, the commute time inference predicts that many movement parameters
-are an order of magnitude larger than they are in truth. 
-%and there does not appear to be a correlation between 
-%the magnitude of the true and inferred values, 
-%indicating that we cannot infer much beyond that the population mixes quickly,
-%which was already clear from the data itself.
-
-%\emph{talk more about results here}
-
-\begin{figure}
-\centering
-% \includegraphics[scale=.8]{4x4_grids}
-\caption{Plots of the grid structure 
-for the uniform graph, symmetric graph, the asymmetric graph,
-and the biased asymmetric graph 
-using the igraph R package.}
-\label{fig:4x4_grids}
-\end{figure}
-
-\begin{figure}
-\centering
-% \includegraphics[scale=.8]{4x4coalvcomH}
+This difference suggests that commute time inference may be strongly misled 
+in the asymmetric and biased situations,
+but it still could be the case that the movement rates that give the best fit of commute time
+to coalescence time data could be close to the actual rates use to generate the data.
+
+Nonetheless, 
+coalescence time inference was substantially more accurate in all cases.
+Posterior distributions of movement parameters are shown in Figure \ref{fig:4x4box},
+produced by applying both inference methods
+after adding noise with standard deviation of 1/200 the mean value of $C$.
+The mean absolute errors for coalescence time inference are 0.12 or less in all cases.
+Commute time inference obtains roughly uniform migration rates
+(although varying by a factor of 2) in the uniform case;
+and migration rates noisily correlated with the truth in the symmetric graph.
+However, commute time inferences for the asymmetric and biased graphs are only weakly correlated, 
+if at all, to the truth.
+Commute times also drastically overestimate movement rates for the biased case,
+as we would expect based on the differences discussed above.
+
+\begin{figure}
+\centering
+     \includegraphics[scale=.8]{figs/4x4coalvcomH}
     \caption{Plots of the true values of $\comdist$ vs $H$ 
@@
 \centering
-% \includegraphics[scale=.8]{4x4box}
+     \includegraphics[scale=.8]{figs/4x4box}
     \caption{Boxplots of the posterior distributions for $g$ for commute time and coalescence time
@@
 
-\plr{transition about discretization and continuous space}
-
-In order to validate the methods in situations 
-that are more similar to those that we would see with actual populations,
-we performed simulations using SLiM (Selection on Linked Mutations) \citep{haller2016slim}.
-
-\plr{move to methods}
-We used SLiM to simulate discrete generations based on the Wright-Fisher model.
-Simulations were performed in three different situations:
-uniform migration on a continuous square landscape, 
-uniform migration on a rectangular landscape with barriers to migration, 
-and biased migration on a square landscape.
-The simulations performed in this section 
-were of populations of 10000 diploid hermaphroditic individuals.
-Mates were selected from nearby individuals using a Gaussian kernel 
-with standard deviation of $\sigma_{d}$,
-truncated at $3 \sigma_{d}$.
-In each simulation,
-$\sigma_{d}$ was chosen such that there would be an average of 
-0.25 individuals per $\sigma_{d}^2$ (2.25 per $(3 \sigma_{d})^2$),
-which is 0.005 for a square landscape with one unit of area.
-Offspring are dispersed according to a Gaussian distribution away from the mother
-using the same truncated Gaussian kernel as before.
-If the location of the offspring falls outside of the habitable area, 
-it is redrawn until it is in a valid location.
-In order to prevent excessive clumpiness of the population density distribution \citep{felsensten1975pain}, 
-we also model local competition. 
-To do this, we use the same truncated Gaussian kernel,
-with the strength of the interaction chosen 
-by visual inspection of the density distribution.
-Specifically, the new fitness of each individual after accounting for local competition
-is equal to $\exp{(-\text{totalStrength}/40)}$,
-where totalStrength is the sum of the strength of interations with all individuals
-within the cutoff distance according to the Gaussian kernel,
-resulting in reduced fitness for individuals in dense regions.
-Individuals are chosen to reproduce proportional to their fitness until the next generation.
-Clumpiness is similar to what is seen in 
-Figure \ref{fig:ind_locs_5x3b_1}
-
-In our simulations, individuals have 10 pairs of chromosomes with 200,000 base pairs each 
-with a recombination rate of $10^{-8}$ per generation per base pair.
-In the landscapes with barriers and biased migration,
-mutation rate was set to $10^{-7}$.
-In order to speed up simulation in the case of the landscape with uniform migration,
-the mutation rate was lowered to $10^{-8}$.
-This should not have a dramatic effect on the pattern of genetic relatedness between individuals
-aside from reducing the genetic distance for those simulations by a factor of ten.
-
-In each location, individuals are sampled uniformly from those 
-within the middle three quarters in each dimension of the grid space,
-as shown for one situation iin Figure \ref{fig:ind_locs_5x3b_1}.
-This protocol was chosen as a compromise between sampling all individuals 
-from near the center of each location,
-which could result in a large number of close relatives being sampled, 
-leading to an underestimate of the mean genetic distance for individuals in that location
-and uniform sampling from the whole area of each grid, 
-which may result in many sampled individuals being very close to other grid squares.
-An equal number of individuals are drawn from each location,
-but this need not be the case.
-We could even have some locations from which no individuals were drawn,
-but this lowers performance as discussed in Section \ref{sec:5x3b}.
-Mean genetic distances for each pair of locations 
-are calculated from the means of all of the pairwise genetic divergences 
-between genomes in the two locations, excluding self comparisons.
-Standard error of the mean is calculated conservatively using 
-the minumum number of independent samples, 
-in this case the smallest number of individuals from each location.
-
-Before being passed to the MCMC inference method,
-genetic distance values are scaled so that the resulting parameters 
-are approximately of order one.
-This is accomplished by multiplying the mean genetic distances by a constant
-such that the new mean of the mean genetic distances 
-is equal to the number of states in the discretization. 
-For example, if we discretize a landscape into 16 squares, 
-the mean of the mean genetic distances after scaling will be 16.
-
-For the $4 \times 4$ discretizations of the uniform and biased migration cases, 
-as well the barrier case, 
-inference was performed using $4 \times 10^6$ iterations of MCMC
-after $1 \times 10^6$ iterations of pre-burn-in and $3 \times 10^6$ iterations of burn-in. 
-For the $2 \times 2$ discretizations, $1 \times 10^6$ iterations were used 
-with $1 \times 10^5$ for pre-burn-in and $1 \times 10^6$ for burn-in.
-
-% {Continuous Landscape with a Barrier}
+The data we have used thus far are produced ``under the model'',
+and should provide an accurate depiction of our method applied to discrete, randomly mating populations
+whose connections by migration are known.
+However, this is nearly always a rough approximation to reality,
+in which organisms are distributed across continuous geography,
+and ``populations'' are constructed by necessity, often driven by sampling locations.
+
+\paragraph{Process noise}
+Talk about Figure \ref{fig:h_val_comp}.
+
+\begin{figure}
+\centering
+\includegraphics[scale=.8]{figs/h_val_comp}
+\caption{
+    Comparison of the values of the genetic distance matrix $C$
+    for a $4 \times 4$ discretization of a square landscape with uniform migration
+    calculated from various situations.
+    All values are plotted against the corresponding values from the $C$ matrix
+    calculated using the first set of individuals from the first simulation,
+    along with the $y=x$ line.
+    Black circles are from a second (non-overlapping) set of individuals from the first simulation,
+    red triangles are from the second simulation, 
+    and blue plusses are for the third simulation.
+    Mean absolute value relative differences between the new values of $C$ for each situation
+    and the original values from the first set of individuals from the first simulation are
+    1.8\% for the second set of indivuals from the first simulation,
+    13.2\% for the second simulation, 
+    and 7.3\% for the third simulation.
+    } \label{fig:h_val_comp}
+\end{figure}
+
+\paragraph{Identifying a barrier}
 We now revisit the earlier situation where a landscape has locations that are barriers to gene flow,
@@
 \centering
-% \includegraphics[scale=.8]{ind_locs_5x3b_1}
+     \includegraphics[scale=.8]{figs/ind_locs_5x3b_1}
     \caption{Locations of all and selected individuals on the $5 \times 3$ landscape with barriers
@@
 \centering
-% \includegraphics[scale=.8]{grid_5x3b_1}
-\caption{Medians of the posterior distributions of the inferred values of $g$ 
-for the first replicate of the $5 \times 3$ landscape with barriers.
-The gene flow rate across the barriers are small, 
-but there is no clear pattern amoung the rest of the values.}
-\label{fig:grid_5x3b_1}
-\end{figure}
-
-\begin{figure}
-\centering
-% \includegraphics[scale=.8]{post_dists_5x3b_1_new}
+     \includegraphics[scale=.8]{figs/post_dists_5x3b_1_new}
     \caption{Boxplots of the posterior distributions for $g$ 
@@
 This results in reverse time gene flow down and to the left 
-as in the biased asymmetric graph in Section \ref{sec:comvcoal}.
+as in the biased asymmetric graph in Section \ref{sec:biased_migration}.
 In terms of population genetics, 
@@
 \centering
-% \includegraphics[scale=.35]{ibd_comp}
+ \includegraphics[scale=.35]{figs/ibd_comp}
 \caption{Comparison of the isolation by distance plots for the square landscape with uniform migration
@@
 \centering
-% \includegraphics[scale=.5]{2x2grid_asb}
+ \includegraphics[scale=.5]{figs/2x2grid_asb}
 \caption{Medians of the posterior distributions of the inferred values of $g$ 
@@
 \centering
-% \includegraphics[scale=.8]{grid_bias_4x4_1}
+ \includegraphics[scale=.8]{figs/grid_bias_4x4_1}
 \caption{Medians of the posterior distributions of the inferred values of $g$ 
@@
 \centering
-% \includegraphics[scale=.6]{post_dists_bias_4x4_1}
+ \includegraphics[scale=.6]{figs/post_dists_bias_4x4_1}
 \caption{Boxplots of the posterior distributions for $g$ 
@@
 
-\section*{Application of Inference to Mojave Desert Tortoises}
-
-We also apply the method to a population of Mojave desert tortoises (\emph{Gopherus agassizii})
-living in California. 
-Location and low coverage whole genome sequencing data was collected from 271 individuals 
-\citep{shaffer2017desert}.
-Pairwise genetic distances between all individuals was obtained from the same study.
-The landscape was discretized into 13 regions based on watersheds and the locations of individuals.
-The mean coalescence time matrix and standard deviations were calculated as before. 
+\section*{Application to the Mojave Desert Tortoise}
+
+The Mojave desert tortoise (\textit{Gopherus agassizii})
+lives across much of the Mojave desert of California and Nevada,
+separated from its sister taxa, \textit{Gopherus morofkaii} by the Colorado river.
+We applied our method to genomic distances calculated from
+whole genome sequencing data of 271 individuals sampled from across much of the range,
+reported in \citet{shaffer2017desert}.
+We discretized the landscape into 13 regions based on watershed polygons
+from the Watershed Boundary Dataset \citep{WBD}
+(using WBD8 slightly modified by portions of WBD10 to improve contiguity of regions).
+The resulting discretization,
+overlaid on a map including tortoise sampling locations,
+is shown in Figure \ref{fig:tort_land}.
 Since the regions vary substantially in size, 
@@
 we do not assume that coalescence rates are the same everywhere, 
-which gives 13 values of $\gamma$. 
-Since allowing only nearest neighbor migration gives 50 values of $g$, 
-we have a total of 63 unknown parameters.
-Because there are $13(14)/2 = 91$ unique values of the mean coalescence time matrix, 
-we have more equations than unknowns.
-
-Posterior distributions of gene flow and coalescence rate parameters were inferred
-using $4 \times 10^7$ iterations of MCMC 
-after $1 \times 10^6$ iterations of pre-burn-in and $3 \times 10^6$ iterations of burn-in.
-The landscape and associated graph are shown in Figure \ref{fig:tort_land}. 
-Boxplots of the posterior distributions for the gene flow rates can be seen in Figure \ref{fig:tort_post_g} 
-and for the coalescence rates can be seen in Figure \ref{fig:tort_post_gam}.
+which gives 13 coalescence rates;
+combined with the 50 migration parameters between each pair of adjacent regions
+we have a total of 63 unknown parameters
+(and 91 equations).
+
+Posterior distributions of the parameters are shown in Figure \ref{fig:tort_post}.
 There is substantial variation in the gene flow rates 
@@
 This is not too surprising given its large area and central location.
-In general, the results are reasonable. 
-For example,  we observe relatively low gene flow between areas with mountains on the border, 
-%such as the Paiute and Castle Mountains between regions 1 and 4, 
-%the Providence and Granite Mountains between regions 4 and 12, 
-%the New York and McCollough Mountains between regions 1 and 9
-such as between regions 1 and 4, 1 and 9, 4 and 12, 4 and 5, and 9 and 10.
-On the other hand, there is an open connection between 9 and 13, 
+Gene flow rates are low across mountainous boundaries:
+for instance, regions 1 and 4 are separated by the Paiute and Castle mountains,
+and regions 4 and 12 are separated by the Providence and Granite mountains.
+% and regions 1 and 9 are separated by the New York and McCollough mountains.
+On the other hand, there is a large, mostly open bajada connecting areas 9 and 13, 
 and we infer high rates of migration between the two regions.
-%say something about regions that don't touch much?
-Overall, the model is a better fit than the resistance distance method of \citet{shaffer2017desert}
+The coalescence time model provides a good fit to the data,
+much better than was obtained by commute times, either \plr{in this paper??}
+or using the resistance distance method of \citet{shaffer2017desert}
 (Figure \ref{fig:tort_h_comp}).
 
-The strong asymmetries, such as those present between regions 9 and 11 or 9 and 12, 
-are probably the result of either source-sink dynamics or historical processes. 
-The presence of source-sink dynamics would be of great interest for tortoise conservation efforts; 
-however, it is too early in this analysis to use these numbers for that purpose.
-For example, we would want to check for the effects of the sample configuration 
-(e.g., are gene flow rates between regions 2 and 3 low 
-simply because individuals sampled in region 3 are geographically far from region 2) 
-as well as perform a sensitivity analysis, 
-potentially including simulations on the landscape.
-% sensitivity analysis would need to be performed, 
-%as well as checking for the effects of the sample configuration.
-
-\begin{figure}
-\centering
-% \includegraphics[scale=.34]{tort_land5}
+The most biologically interesting aspects
+are the strong asymmetries, for instance between region 9 and regions 11, 12, and 13.
+These indicate that resistance-based methods may be strongly misled.
+However, we cannot conclude from this analysis \emph{what} is leading to the observed asymmetries.
+One explanation would be source-sink dynamics:
+for instance, the observed bias in lineage movement could be observed 
+if the tortoise fecundity in the Ivanpah valley in region 9
+is higher than in the colder, higher-elevation regions to the north.
+Or, it could be the result of a historical range expansion out of the area,
+perhaps out of a more resticted range as climates changed after the last glacial maximum.
+The presence of source-sink dynamics would be of great interest for tortoise conservation efforts,
+but it is not clear from this analysis which explanation is more likely.
+Some aspects of the results may even be the produced by sample configuration:
+for instance, perhaps the inferred gene flow between regions 2 and 3 is low
+simply because individuals sampled in region 3 are geographically far from region 2.
+
+\begin{figure}
+\centering
+ \includegraphics[scale=.34]{figs/tort_land5}
 \caption{Left: Locations of the sampled tortoises.
-Right: Sample maximum likelihood estimates from the posterior distributions of the inferred values of $g$.}
-\label{fig:tort_land}
-\end{figure}
-
-\begin{figure}
-\centering
-% \includegraphics[scale=.6]{tort_post_g}
-\caption{Boxplots of the posterior distributions for $g$ 
-for the Mojave desert tortoise data set 
-on the thirteen region discretization of the landscape.}
-\label{fig:tort_post_g}
-\end{figure}
-
-\begin{figure}
-\centering
-% \includegraphics[scale=.6]{tort_post_gam}
-\caption{Boxplots of the posterior distributions for $\gamma$ 
-for the Mojave desert tortoise data set 
-on the thirteen region discretization of the landscape.}
-\label{fig:tort_post_gam}
-\end{figure}
-
-\begin{figure}
-\centering
-% \includegraphics[scale=.8]{tort_h_comp}
+    \plr{add mountain shadings and ocean; remove bits on the other side of the Colorado}
+    Right: Sample maximum posterior estimates 
+    of movement rates.
+    \plr{can these be maximum posterior estimates instead?}
+    } \label{fig:tort_land}
+\end{figure}
+
+\begin{figure}
+\centering
+ \includegraphics[width=0.8\textwidth]{figs/tort_post_g}
+ \includegraphics[width=0.8\textwidth]{figs/tort_post_gam}
+\caption{
+Boxplots of the posterior distributions for 
+\textbf{(top)} movement rates, and
+\textbf{(bottom)} coalescence rates,
+for the Mojave desert tortoise data set.
+Parameters are labeled as in Figure \ref{fig:tort_land}.
+\label{fig:tort_post}
+\end{figure}
+
+\begin{figure}
+\centering
+ \includegraphics[scale=.8]{figs/tort_h_comp}
 \caption{Mean coalescence times calculated from the sample maximum likelihood model parameters 
@@
 This could lead to a falsely high assessment of gene flow across the top of the ridge.
-
+\plr{Also applies to least-cost path analysis.}
+
+\paragraph{Difficulty of inference}
+What determines the feasability of inferring movement rates from genetic distances?
+Using coalescence times instead of commute times greatly improves accuracy in many situations,
+but there are still a number of factors that determine the tractability of the problem.
+Observation noise is clearly an issue;
+we found that estimation errors of a few percent was enough to seriously degrade accuracy.
+However, genetic distances can be estimated to much higher accuracy using modern genomic data.
+Perhaps more seriously,
+a rate of coalescence relative to movement can also make the problem essentially nonidentifiable.
+This is clear at least in the limit: with very low coalescence, 
+the population loses any isolation by distance and looks panmictic.
+In continuous landscapes, this balance is measured by Wright's local effective population size,
+which is proportional to the number of other individuals 
+within a circle of radius equal to the mean dispersal distance.
+\plr{more about continuous landscapes and discretization}
+
+\paragraph{Circuitscape}
+Could we use coalescence time in place of commute time
+for workflows that compute values on a given map of movement rates
+that are then compared to genetic distance?
+In principle this could be done,
+but not at the same (impressively fine) geographic resolution that is currently possible,
+e.g., using Circuitscape \citep{circuitscape}.
+This is because if we have $n$ sampled locations
+on a landscape discretized into $N$ regions,
+the computational complexity of finding commute times is $nN$,
+but coalescence time scales as $N^2$.
+This is because each $N$-vector of hitting times to a particular location can be found independently,
+but the entire $N \times N$ matrix of coalescence times must be found together.
+However, as we discuss above, this fine geographic resolution is in some sense illusory --
+for each fine-resolution map there should be coarser maps that give almost identical coalescence times.
+However, we are not aware of existing theory providing guidance on how to find these.
+
+\plr{from the thesis; merge in}
 With these methods, we provide a framework for
@@
 \appendix
+\renewcommand{\thefigure}{S\arabic{figure}}
+\setcounter{figure}{0}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Finding $G$ from $R$}
+\subsection{Finding $G$ from $R$}
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Further inference results}
+
+Here we put some more plots describing inference method accuracy.
+
+\begin{figure}
+\centering
+\includegraphics[scale=.6]{figs/mult_noise}
+\caption{Boxplots of the mean absolute error, %in $\hat{g}$.
+defined to be the absolute value difference between the true value and posterior median,
+averaged across movement parameters $g$ for each of the 25 graphs in each situation.
+Each box is labeled with the relative amount of noise added to $H$ to make $\hat{H}$.
+The top left shows the results for symmetric graphs 
+when there is a single coalescence parameter for all locations,
+the top right for symmetric graphs 
+when there is a separate coalescence parameter for each location,
+the bottom left for asymmetric graphs 
+when there is a single coalescence parameter for all locations,
+and the bottom right for asymmetric graphs 
+when there is a separate coalescence parameter for each location.}
+\label{fig:mult_noise}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[scale=.6]{figs/mult_noise_iqr}
+\caption{Boxplots of the mean interquartile ranges of the posterior distributions of $g$
+for each of the 25 graphs in each situation.
+Each box is labeled with the relative amount of noise added to $H$ to make $\hat{H}$.
+Subplot locations for each situation are the same as in Figure \ref{fig:mult_noise}.
+%The top left shows the results for symmetric graphs 
+%when there is a single coalescence parameter for all locations,
+%the top right for symmetric graphs 
+%when there is a separate coalescence parameter for each location,
+%the bottom left for asymmetric graphs 
+%when there is a single coalescence parameter for all locations,
+%and the bottom right for asymmetric graphs 
+%when there is a separate coalescence parameter for each location.
+}
+\label{fig:mult_noise_iqr}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[scale=.6]{figs/mult_gam}
+\caption{
+Boxplots of the mean absolute error, %in $\hat{g}$.
+defined to be the absolute value difference between the true value and posterior median,
+averaged across movement parameters $g$ for each of the 25 graphs in each situation.
+Each box is labeled with the coalescence rate in the underlying model for each situation.
+Subplot locations for each situation are the same as in Figure \ref{fig:mult_noise}.
+%The top left shows the results for symmetric graphs 
+%when there is a single coalescence parameter for all locations,
+%the top right for symmetric graphs 
+%when there is a separate coalescence parameter for each location,
+%the bottom left for asymmetric graphs 
+%when there is a single coalescence parameter for all locations,
+%and the bottom right for asymmetric graphs 
+%when there is a separate coalescence parameter for each location.
+}
+\label{fig:mult_gam}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[scale=.6]{figs/mult_gam_iqr}
+\caption{Boxplots of the mean interquartile ranges of the posterior distributions of $g$
+for each of the 25 graphs in each situation.
+Each box is labeled with the coalescence rate in the underlying model for each situation.
+Subplot locations for each situation are the same as in Figure \ref{fig:mult_noise}.
+%The top left shows the results for symmetric graphs 
+%when there is a single coalescence parameter for all locations,
+%the top right for symmetric graphs 
+%when there is a separate coalescence parameter for each location,
+%the bottom left for asymmetric graphs 
+%when there is a single coalescence parameter for all locations,
+%and the bottom right for asymmetric graphs 
+%when there is a separate coalescence parameter for each location.
+}
+\label{fig:mult_gam_iqr}
+\end{figure}
+
+\begin{figure}
+\centering
+     \includegraphics[scale=0.8]{figs/5x3boxplots}
+    \caption{Posterior distributions for values of $g$ 
+    for the 5 $\times$ 3 graph with barriers 
+    for each of the analysis cases.
+    See Figure \ref{fig:5x3b_grid} to see movement parameter locations.
+    \plr{TODO: put bottom two panels on top; on the bottom put the same thing for commute time}
+}
+    \label{fig:5x3boxplots_mult_gam}
+\end{figure}
+
+\begin{figure}
+\centering
+     \includegraphics[scale=.8]{figs/4x4_grids}
+    \caption{Plots of the grid structure 
+    for the uniform graph, symmetric graph, the asymmetric graph,
+    and the biased asymmetric graph 
+    using the igraph R package.}
+    \label{fig:4x4_grids}
+\end{figure}
+
+\begin{figure}
+\centering
+     \includegraphics[scale=.8]{figs/grid_5x3b_1}
+    \caption{Medians of the posterior distributions of the inferred values of $g$ 
+    for the first replicate of the $5 \times 3$ landscape with barriers.
+    The gene flow rate across the barriers are small, 
+    but there is no clear pattern amoung the rest of the values.}
+    \label{fig:grid_5x3b_1}
+\end{figure}
+
 \end{document}

### 2018-08-09 isolation_by_coalescence.tex
a few more edits

@@
 and $\circ$ is the componentwise product.
-This can also be seen by working with the product Markov chain $(X_t, Y_t)$,
+In practice, we solve this by working with the product Markov chain $(X_t, Y_t)$,
 whose generator matrix is $G \otimes I + I \otimes G$,
 where $I$ is the identity matrix and $\otimes$ is the Kronecker product.
+\plr{say a bit more about how we solve this?}
 
@@
 given the coalescence or commute times
-we must solve equations \eqref{eqn:C_matrix} or \eqref{eqn:R_from_H} and \eqref{G_from_H} for $G$,
-and that these do not hae unique solutions.
+we must solve equations \eqref{eqn:C_matrix} or \eqref{eqn:R_from_H} and \eqref{eqn:G_from_H} for $G$,
+and that these do not have unique solutions.
 The situation is worse when one needs to infer coalescence rates ($\gamma$) as well.
@@
     shown with the $y=x$ line.
+    \plr{TODO: make this a 2x2 grid instead of 4x1}
     See Figure \ref{fig:4x4_grids} to see overall graph structure.}
@@
 \paragraph{Process noise}
-Talk about Figure \ref{fig:h_val_comp}.
+The mean coalescence times of lineages that we compute analytically
+are exact for large, randomly mating populations,
+but the word ``mean'' indicates an average across several levels of randomness:
+and observed, empirical mean estimated using a sample of genotyeps
+will differ slightly depending on both the individuals sampled (``sampling noise'')
+and the stochasticity of the population history itself (``process noise'').
+For more discussion of this latter contribution in other contexts, see
+\citet{waples}, \citet{wakeley}, or \citet{ralph_empirical}.
+In populations living across continuous geography,
+both sources of noise can be substantially larger due to randomness in spatial locations.
+A hint of this is visible in Figure \ref{fig:ind_locs_5x3b_1},
+where we see that the distribution of individuals across the landscape
+is fairly patchy.
+This might make inference using a discretization of continuous space considerably more difficult.
+
+To quantify the relative contributions of the different sources of noise,
+we coompared mean genetic distances, calculated in the same way,
+between nonoverlapping sets of samples from the same simulation,
+and between different simulations.
+These simulations were done using SLiM as described in the Methods
+on a square, $4 \times 4$ landscape with no barriers.
+We found that process noise was much larger than sampling noise --
+as shown in Figure \ref{fig:h_val_comp},
+discrepancies between simulations are about five times that seen
+between different samples from the same simulation.
+An implication of this is that uncertainties in genetic distance
+estimated from the data will be (possibly large) underestimates,
+a fact also pointed out for effective population size by \citet{waples}.
 
@@
 \centering
-     \includegraphics[scale=.8]{figs/post_dists_5x3b_1_new}
-    \caption{Boxplots of the posterior distributions for $g$ 
-    for the first replicate of the $5 \times 3$ landscape with barriers.
-    Boxes for gene flow rates for edges that cross the barrier are colored in red.}
-    \label{fig:post_dists_5x3b_1}
+     \includegraphics[height=0.3\textheight]{figs/grid_5x3b_1}
+     \includegraphics[height=0.3\textheight]{figs/post_dists_5x3b_1_new}
+    \caption{
+    \textbf{(top)}
+    Posterior medians of movement rates for a replicate of the continuous landscape
+    with barriers depicted in Figure \ref{fig:ind_locs_5x3b_1}.
+    \textbf{(bottom)}
+    Boxplots of the same posterior distributions,
+    colored red for gene flow rates of edges that cross the barrier.
+    } \label{fig:post_dists_5x3b_1}
 \end{figure}
@@
 
-\subsection*{Continuous Landscape with Biased Migration}
-
+\paragraph{Continuous Landscape with Biased Migration}
 The third situation that we will discuss is where migration is biased.
@@
 \centering
- \includegraphics[scale=.34]{figs/tort_land5}
+ \includegraphics[width=0.45\textwidth]{figs/fancy_watershed_assignments}
+ \includegraphics[width=0.45\textwidth]{figs/tort_graph_results}
 \caption{Left: Locations of the sampled tortoises.
@@
     Parameters are labeled as in Figure \ref{fig:tort_land}.
-\label{fig:tort_post}
+    } \label{fig:tort_post}
 \end{figure}
@@
  \includegraphics[scale=.8]{figs/tort_h_comp}
-\caption{Mean coalescence times calculated from the sample maximum likelihood model parameters 
-plotted against the observed mean coalescence times calculated from the data, 
+\caption{
+    Mean coalescence times calculated from the sample maximum likelihood model parameters 
+    plotted against the observed mean coalescence times calculated from the tortoise data, 
     measured in the mean of pairwise $\pi$.
-The mean absolute value relative differences between the two is 0.4\%.}
-\label{fig:tort_h_comp}
+    The mean absolute value relative differences between the two is 0.4\%.
+    } \label{fig:tort_h_comp}
 \end{figure}
@@
 
-resistance distance can mislead, especially in the presence of biased migration
-
-coalescence time analogue is feasible but more computationally intensive
-
-practical example (tortoises) shows strong asymmetry
-
-but methods dealing with ill-posedness are necessary
-
-other measures of genetic distance: what do people use in practice?
-
-\paragraph{Consequences for resistance methods}
-We have shown that commute time can show dramatically different patterns
-than coalescence time, especially if there is systematic bias in dispersal.
-How should this affect the results of resistance-based methods to genomic data,
-besides general inaccuracy?
-Suppose, for instance, that seed dispersal for a species of tree
-is biased downhill,
-and the landscape is dominated by a large valley.
-Since lineages tend to move uphill,
-genetic distance between locations on opposite slopes of the valley will be relatively large,
-so resistance methods will infer a barrier along the bottom of the valley
-when no barrier to dispersal exists.
-On the other hand, if there is a ridge instead of a valley,
-resistance methods will show no barrier (and relatively high movement rates),
-even if there is very little uphill dispersal.
-This could lead to a falsely high assessment of gene flow across the top of the ridge.
-\plr{Also applies to least-cost path analysis.}
-
-\paragraph{Difficulty of inference}
-What determines the feasability of inferring movement rates from genetic distances?
-Using coalescence times instead of commute times greatly improves accuracy in many situations,
-but there are still a number of factors that determine the tractability of the problem.
-Observation noise is clearly an issue;
-we found that estimation errors of a few percent was enough to seriously degrade accuracy.
-However, genetic distances can be estimated to much higher accuracy using modern genomic data.
-Perhaps more seriously,
-a rate of coalescence relative to movement can also make the problem essentially nonidentifiable.
-This is clear at least in the limit: with very low coalescence, 
-the population loses any isolation by distance and looks panmictic.
-In continuous landscapes, this balance is measured by Wright's local effective population size,
-which is proportional to the number of other individuals 
-within a circle of radius equal to the mean dispersal distance.
-\plr{more about continuous landscapes and discretization}
-
-\paragraph{Circuitscape}
-Could we use coalescence time in place of commute time
-for workflows that compute values on a given map of movement rates
-that are then compared to genetic distance?
-In principle this could be done,
-but not at the same (impressively fine) geographic resolution that is currently possible,
-e.g., using Circuitscape \citep{circuitscape}.
-This is because if we have $n$ sampled locations
-on a landscape discretized into $N$ regions,
-the computational complexity of finding commute times is $nN$,
-but coalescence time scales as $N^2$.
-This is because each $N$-vector of hitting times to a particular location can be found independently,
-but the entire $N \times N$ matrix of coalescence times must be found together.
-However, as we discuss above, this fine geographic resolution is in some sense illusory --
-for each fine-resolution map there should be coarser maps that give almost identical coalescence times.
-However, we are not aware of existing theory providing guidance on how to find these.
-
-\plr{from the thesis; merge in}
 With these methods, we provide a framework for
@@
 The commute time plus diversity estimate, $\comdist$, 
-provides a good estimate of the coalescence times, $H$,
+provides a good estimate of the coalescence times, $C$,
 when gene flow rates are symmetric ($G_{ij} = G_{ji}$), 
@@
 
-\plr{pasted in: nonmarkov lineages}
+\paragraph{Consequences for resistance methods}
+We have shown that commute time can show dramatically different patterns
+than coalescence time, especially if there is systematic bias in dispersal.
+How should this affect the results of resistance-based methods to genomic data,
+besides general inaccuracy?
+Suppose, for instance, that seed dispersal for a species of tree
+is biased downhill,
+and the landscape is dominated by a large valley.
+Since lineages tend to move uphill,
+genetic distance between locations on opposite slopes of the valley will be relatively large,
+so resistance methods will infer a barrier along the bottom of the valley
+when no barrier to dispersal exists.
+On the other hand, if there is a ridge instead of a valley,
+resistance methods will show no barrier (and relatively high movement rates),
+even if there is very little uphill dispersal.
+This could lead to a falsely high assessment of gene flow across the top of the ridge.
+\plr{Also applies to least-cost path analysis.}
+
+\paragraph{Circuitscape}
+Could we use coalescence time in place of commute time
+for workflows that compute values on a given map of movement rates
+that are then compared to genetic distance?
+In principle this could be done,
+but not at the same (impressively fine) geographic resolution that is currently possible,
+e.g., using Circuitscape \citep{circuitscape}.
+This is because if we have $n$ sampled locations
+on a landscape discretized into $N$ regions,
+the computational complexity of finding commute times is $nN$,
+but coalescence time scales as $N^2$.
+This is because each $N$-vector of hitting times to a particular location can be found independently,
+but the entire $N \times N$ matrix of coalescence times must be found together.
+However, as we discuss above, this fine geographic resolution is in some sense illusory --
+for each fine-resolution map there should be coarser maps that give almost identical coalescence times.
+However, we are not aware of existing theory providing guidance on how to find these.
+
+\paragraph{Difficulty of inference}
+What determines the feasability of inferring movement rates from genetic distances?
+Using coalescence times instead of commute times greatly improves accuracy in many situations,
+but there are still a number of factors that determine the tractability of the problem.
+Observation noise is clearly an issue;
+we found that estimation errors of a few percent was enough to seriously degrade accuracy.
+However, genetic distances can be estimated to much higher accuracy using modern genomic data.
+Perhaps more seriously,
+a rate of coalescence relative to movement can also make the problem essentially nonidentifiable.
+This is clear at least in the limit: with very low coalescence, 
+the population loses any isolation by distance and looks panmictic.
+In continuous landscapes, this balance is measured by Wright's local effective population size,
+which is proportional to the number of other individuals 
+within a circle of radius equal to the mean dispersal distance.
+\plr{more about continuous landscapes and discretization}
+
+\paragraph{The effect of history}
+The models we discuss here assume that population sizes and migration rates
+have been constant on the time scale given by the within-species coalescent time.
+This is rarely true in practice \citep{avise,barton}.
+However, geographic differences in mean relatedness
+are established on a shorter time scale --
+the time scale over which a lineage, moving randomly across the landscape,
+``forgets'' where it started
+(i.e., the mixing time of the Markov chain).
+This is the time it takes for the uncertainty in lineage movement
+to reach the scale of the species range.
+For instance,
+the width of the tortoise range is roughly 500km,
+and if mean dispersal distance is 10km (a high estimate),
+the standard deviation of ancestor location $t$ generations ago is $10\sqrt{t}$,
+so the mixing time of tortoise lineages across the modern landscape,
+not accounting for barriers, is of order 2500 generations,
+or about 50,000 years.
+The landscape of the Mojave has changed substantially over tihs time \citep{pleistocene_mojave},
+which suggests that models incorporating change over time 
+may be required to model modern tortoise diversity.
+However, the effects of recent history are strongest,
+so landscapes estimated assuming constant populations
+may give a reasonable picture of the landscape averged over recent times.
+
+\paragraph{Assumptions}
+Modeling lineages as a Markov chain is nearly ubiquitous in population genetics today,
+but may not be appropriate.
 Even though the population dynamics in forwards time are Markov, the dynamics of lineages traced back
@@
 model the coalescent process as Markov, which should be a good approximation for many systems.
+
+\paragraph{other points:}
+Practical example (tortoises) shows strong asymmetry.
+Methods dealing with ill-posedness are necessary.
+Other measures of genetic distance: what do people use in practice?
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Finding $G$ from $R$}
-
-The resistance distance is given by
-$R = H + H^T$, so if we can modify $H$ by a skew-symmetric matrix,
-we won't change $R$.
-Can we do this and retain a valid $G$?
-I.e., does $R$ uniquely specify $G$?
-Let $Z + Z^T = 0$. 
-Since $G$ is a continuous function of $H$, if all entries of $G$ are nonzero
-we can find a small enough $\epsilon$ such that the $G'$ corresponding to $H' = H + \epsilon Z$
-is also a valid generator matrix.
-If $G$ has zero entries this may not be true.
-To find a direction that we can change $H$ in that doesn't mess this up,
-we could differentiate $dG/dH$.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Further inference results}
@@
 
-\begin{figure}
-\centering
-     \includegraphics[scale=.8]{figs/grid_5x3b_1}
-    \caption{Medians of the posterior distributions of the inferred values of $g$ 
-    for the first replicate of the $5 \times 3$ landscape with barriers.
-    The gene flow rate across the barriers are small, 
-    but there is no clear pattern amoung the rest of the values.}
-    \label{fig:grid_5x3b_1}
-\end{figure}
-
 \end{document}

### 2018-08-11 isolation_by_coalescence.tex
when are coal and commute equal

@@
 \begin{abstract}
+\plr{abstract needs editing}
+A number of methods commonly used in landscape genetics
+use analogies to electrical resistance on a network
+to describe and fit barriers to movement across the landscape
+using genetic distance data.
+These are motivated by a mathematical eqiuvalence of effective resistance
+to the mean commute time of a random walk.
+Here, we describe the differences between commute time
+and coalescence time, which more accurately models genetic distances,
+and explore the consequences for inference.
+We implement a Bayesian method to do inference under both these models,
+and find that inference using commute times
+when the data derive from coalescence times
+can dramatically mislead in the presence of biased gene flow.
+We apply the methods to empirical data,
+and find evidence for strongly biased gene flow.
 \end{abstract}
@@
 is equal to the effective resistance between $x$ and $y$ in an electrical network
-whose conductances are given by the movement rates of the Markov chain.
+whose conductances are given by the movement rates of the Markov chain \citep{levin}.
 Both electrical resistance and commute time between two nodes
@@
 
-We apply the method to the desert tortoise \plr{say more here}.
-
-\plr{I didn't find myself wanting to explain isolation by distance here:
-    I don't think we need to? Do you agree?}
+% \plr{I didn't find myself wanting to explain isolation by distance here:
+%     I don't think we need to? Do you agree?}
 
@@
 tells us that the stationary distribution of the Markov chain, denoted $\pi$,
-can be recovered from the hitting times by solving $\pi = H^{-1} \bone$.
+can be recovered from the hitting times by solving $\pi = H^{-1} \bone / \bone^T H^{-1} \bone$.
 As shown in Appendix \ref{apx::hitting_calcs},
@@
 Equation \eqref{eqn:C_matrix} is linear in $G$ and $\gamma$,
-and so after rearrangement can be solved with standard linear algebra.
-If $\gamma$ is given, then solving for $G$ is an example of a XXX equation \citep{matrixeqn}.
+and so after rearrangement can be solved with standard linear algebra
+(and is related to the Sylvester equation \citep{sylvester_eqn}).
 However, the solution is again not necessarily unique:
@@
 
+\paragraph{When are coalescence and commute times equal?}
+If in a given biological situation the two times were equal --
+i.e., if given $G$ and $\gamma$, there was a $q$ that made $\comdist = C$ --
+then inference with either model would be equivalent.
+We have already shown that this is not the case in general,
+but it turns out that it \emph{is} true
+under some restrictive assumptions not uncommonly found in population genetics.
+We show in Appendix \ref{sec:com_eq_coal} that
+if hitting times are symmetric ($H = H^T$),
+then this does occur,
+and so the two methods are equivalent for symmetric island models
+(where the populations are arranged on a ring 
+and migration rates only depend on the distance between them).
+Hitting times are not symmetric for a square grid with uniform migration
+(the mean time to hit the center from a corner is les than the reverse),
+but the square grid is quite close to the torus, which does have symmetric hitting times.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection*{The simplest example}
@@
 
+Although commute and coalescence times can be quite different,
+we show that they are equal (with a particular choice of local diversity)
+if populations are arranged on a ring or torus
+with symmetric migration.
+The similarity of these models to rectangular grids
+may explain why in other work, commute time methods have shown relatively good fit to data
+simulated with symmetric migration rates.
+
 We also apply the method to data from continuous space simulations. 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Acknowledgements}
+
+Thanks to David Levin for useful suggestions regarding hitting time calculations,
+to Paul Marjoram for useful comments,
+and to Brad Shaffer and Evan McCartney-Melstad for useful discussions
+and providing the genetic distance data for desert tortoises.
+We are also indebted to Chava Weissman, Bridgette Hagerty, Fran Sandmeier, Dick Tracy,
+and a large number of field workers for collecting the original tortoise tissue samples.
+\plr{also John, Hossein, Desi, ??}
+
 \bibliography{references}
@@
 i.e.,
-$$
+\begin{align} \label{eqn:GH}
     GH = - 1 1^T + \diag(x).
-$$
+\end{align}
 
@@
 the Random Target Lemma \citep{aldous} % https://www.stat.berkeley.edu/users/aldous/RWG/Book_Ralph/Ch2.S2.html
-tells us that $H \pi = 1$, so thus $H^{-1} 1 = \pi$, the stationary distribution of the chain.
-Therefore, $x_i = 1/\pi_i$,
-and in fact $1^T H^{-1} 1 = 1$.
-
-In summary,
-$$
-    G = (\diag(1/H^{-1} 1) - 1 1^T) H^{-1} .
-$$
+tells us that if $\pi$ is the stationary distribution of the chain, then 
+$(H \pi)_i$ does not depend on $i$,
+i.e., that $H \pi \propto 1$,
+and thus if $H^{-1} 1 = \nu$, 
+then $\pi = \nu / \sum_i \nu_i$.
+This says exactly that $x_i = 1/\pi_i$.
+
+In summary, $\pi = H^{-1} 1 / 1^T H^{-1} 1$, and 
+\begin{align}
+    G 
+    &= (\diag(1/\pi) - 1 1^T) H^{-1} .
+\end{align}
+Note that this equation only specifies $H$ up to a constant added to each column:
+i.e., if given $G$ one obtains a matrix $Y$ solving $GY = \diag(1/\pi) - 1 1^T$,
+then $H_{ij} = Y_{ij} - Y_{jj}$.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Equality of commute and coalescence times}
+\label{sec:com_eq_coal}
+
+Under what conditions are coalescence times and commute times (with some local diversity values) equal?
+In other words, for what choices of $G$, $\gamma$, and $q$ does
+\begin{align}
+    \comdist = (H + H^T)/4 + (q 1^T + 1 q^T)/2
+\end{align}
+solve equation \eqref{eqn:C_matrix}?
+Writing this out, this says that
+\begin{align} \label{eqn:R_coal}
+    ( GH + (GH)^T + GH^T + (GH^T)^T )/4 + (Gq 1^T + 1 (Gq)^T)/2
+    &=
+    \diag(q) \diag(\gamma) - 1 1^T .
+\end{align}
+(To simplify this we used the fact that $\diag(H) = 0$ and $G1 = 0$.)
+
+It is not clear what can be said about the general case because of the presence of $HG^T$,
+but if we assume that $H$ is symmetric, we can make progress,
+because then we have that $GH = GH^T$, 
+and so equation \eqref{eqn:GH} says that all four terms in the first group are equal,
+and equation \eqref{eqn:R_coal} simplifies to
+\begin{align} \label{eqn:R_eq_C}
+    \diag(1/\pi) + (Gq 1^T + 1 (Gq)^T)/2
+    &=
+    \diag(q) \diag(\gamma) ,
+\end{align}
+where the $-1 1^T$ terms on each side canceled.
+Because $G1=0$,
+the most obvious solution to this is if $q = c 1$ and $c \gamma = 1/\pi$, for some constant $c$
+(although there is a broader family of solutions).
+
+In summary, if hitting times are symmetric
+and coalescence rates are equal to the inverse of the stationary distribution,
+then coalescence times are equal to commute times plus one.
+This is fairly restrictive, 
+but does occur if the population configuration is isotropic
+(as for instance in an all-connected-to-all island model with equal migration rates)
+or in populations arranged around a ring
+with migration rates depending only on the distance between them.
+
+Can we solve equation \eqref{eqn:R_eq_C} more generally?
+For that equation to hold, we need $Gq 1^T +  (Gq)^T$ to be diagonal,
+i.e., that $(Gq)_i + (Gq)_j = 0$ for all $i \neq j$.
+For any $k \neq \ell$, there exists a vector $u^{(k\ell)}$ such that $(Gu^{(k\ell)})_i = \delta_{ik} - \delta_{i\ell}$;
+the only possible $q$ for which $Gq 1^T +  (Gq)^T$ is diagonal
+are of the form $q = \alpha 1 + \beta u^{(k,\ell)}$ for some $k \neq \ell$ and some constants $\alpha$ and $\beta$.
+We would then need
+\begin{align}
+    1/\pi_i + \delta_{ik} - \delta_{i\ell} = q_i \gamma_i \qquad \text{for all} i.
+\end{align}
+This implies that for any Markov chain with symmetric hitting times,
+we can find diversity values ($q$) and coalescence rates ($\gamma$) that make $\comdist = C$
+(and in fact there are many ways to do this).
+However, there is not a general solution for $q$ if coalescence rates are also given (as is the case in practice).
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Something else?}
+
+\citet{strobeck1987average} showed that $\diag(C)$ was constant and does not depend on movement rates
+for any isotropic conservative migration model.
+Can we explain this?
+(Also cite earlier.)
 

### 2018-08-11 sims/plot_locations.py
slim scripts

@@ -1,0 +1,19 @@
+import msprime
+import pyslim
+import numpy as np
+import matplotlib
+matplotlib.use('Agg')
+from matplotlib import pyplot as plt
+
+outdir = "test06/"
+treefiles = [outdir + "pop_{:03d}.trees".format(x) for x in range(1, 21)]
+
+for treefile in treefiles:
+    # treefile = outdir + "pop_100.trees"
+    outfile = treefile + ".locs.png"
+    ts = pyslim.load(treefile, slim_format=True)
+    locations = np.array([i.location for i in ts.individuals()])
+    fig = plt.figure(figsize=(5,5))
+    plt.scatter(locations[:,0], locations[:,1], marker='.')
+    plt.savefig(outfile, dpi=288)
+

### 2018-08-13 sims/plot_locations.py
run simulation machinery

@@ -5,7 +5,7 @@ import matplotlib
 matplotlib.use('Agg')
 from matplotlib import pyplot as plt
 
-outdir = "test06/"
+outdir = "run_003003/"
 treefiles = [outdir + "pop_{:03d}.trees".format(x) for x in range(1, 21)]
 
 for treefile in treefiles:
@@ -13,7 +13,7 @@ for treefile in treefiles:
     outfile = treefile + ".locs.png"
     ts = pyslim.load(treefile, slim_format=True)
     locations = np.array([i.location for i in ts.individuals()])
-    fig = plt.figure(figsize=(5,5))
+    fig = plt.figure(figsize=(max(locations[:,0]),max(locations[:,1])))
     plt.scatter(locations[:,0], locations[:,1], marker='.')
     plt.savefig(outfile, dpi=288)
 

### 2018-08-13 sims/submit_sim.sh

@@ -1,0 +1,27 @@
+#!/bin/bash
+
+USAGE="
+Usage:
+   $0 TYPE
+where TYPE should be 'flat', 'biased' or 'barrier'
+"
+
+if [ $# -lt 1 ]
+then
+    echo "$USAGE"
+    exit 0
+fi
+
+TYPE="$1"
+
+NAME="run"
+TAG=$(printf "%06d" $RANDOM); 
+OUTDIR="${NAME}_${TAG}"
+mkdir -p $OUTDIR
+echo "Directory: $OUTDIR"
+
+echo "defineConstant(\"TYPE\", \"${TYPE}\");" >> $OUTDIR/parameters.slim
+
+export OUTDIR
+sbatch -o $OUTDIR/slurm_${TAG}.out -e $OUTDIR/slurm_${TAG}.out ./run_sim.sbatch 
+# ./run_sim.sbatch 

### 2018-08-23 isolation_by_coalescence.tex
minor edits by erik

@@
 \begin{align} \label{eq:commute_approx}
-	D_{ij} \approx \comdist{ij} = R_{ij}/4 + q_{i}/2 + q_{j}/2 .
+	D_{ij} \approx \comdist_{ij} = R_{ij}/4 + q_{i}/2 + q_{j}/2 .
 \end{align}
@@
 and in half the time, because two particles are moving.
-Furthermore, coalescence time includes within-location diversity,
+Furthermore, coalescence time includes within-location diversity
 because particles do not coalesce instantly --
-commute time is most analogous to the first meeting time.
+therefore commute time is most analogous to the first meeting time.
 For a conceptual picture of the differences between the coalescence time and commute time,
@@
 and $G_{xx} = - \sum_{y \neq x} G_{xy}$.
-We will need to find the \emph{hitting times} of the chain,
+We will need to find the \emph{hitting times} of the chain --
 i.e., for each pair of states $x$ and $y$, 
@@
 However, the solution is again not necessarily unique:
-suppose that we have a matrix $Z$ such that $ZH$ is antisymmetric.
+suppose that we have a matrix $Z$ such that $ZC$ is antisymmetric.
 Then $(G + Z) C + C (G + Z)^T = GC + CG^T$,
@@
 although solutions may be in principle unique,
-the problems become \emph{ill-conditioned}, 
+the problems become \emph{ill-conditioned} 
 in the sense that arbitrarily small variations in the input times
@@
 Hitting times are not symmetric for a square grid with uniform migration
-(the mean time to hit the center from a corner is les than the reverse),
+(the mean time to hit the center from a corner is less than the reverse),
 but the square grid is quite close to the torus, which does have symmetric hitting times.
@@
 given $G$ and $\gamma$ we can compute the corresponding coalescence times
--- call these $\mathcal{C}(G, \gamma)$ --
+-- which we call $\mathcal{C}(G, \gamma)$ --
 so our model is that $D$ has a Gaussian distribution 
@@
 placing independent exponential priors on the parameters --
-on those movement rates $G_{ij}$ that are not constrained to be zero,
-and on the coalescence rates, $\gamma$.
+the movement rates, $G_{ij}$, that are not constrained to be zero
+and the coalescence rates, $\gamma$.
 Concretely, suppose that the spatial arrangement of populations is given by a graph,
@@
 we simply use Equation \ref{eq:post},
-but only summing over observed entries of $D_{ij}$.
+but only sum over observed entries of $D_{ij}$.
 
@@
 Typically, we ran MCMC for $10^6$ iterations of pre-burn-in 
-and $3 \time 10^6$ iterations of burn-in,
+and $3 \times 10^6$ iterations of burn-in,
 followed by $4 \times 10^6$ iterations which were used to estimate posterior distributions,
@@
 Clumpiness is similar to what is seen in 
-Figure \ref{fig:ind_locs_5x3b_1}
+Figure \ref{fig:ind_locs_5x3b_1}.
 

### 2018-08-23 isolation_by_coalescence.tex
edits from call

@@
 \newcommand{\comdist}{\widetilde R}
+\newcommand{\comdistfn}{\widetilde \mathcal{R}}
 
@@
 
-The idea of ``resistance'' is an important tool in the lanscape genomics toolbox.
+The idea of ``resistance distance'' is an important tool in the lanscape genomics toolbox.
 Introduced by \citet{mcrae},
@@
 if the conductance across a grid cell in a discretization of the landscape
-is set equal to its elevation,
+is higher in flatter areas,
 and this produces effective resistance values 
 that are significantly correlated with genetic distances,
-then one might conclude that the species in question tends to disperse more at higher elevations,
-perhaps along ridgelines.
+then one might conclude that the species in question tends to disperse more readily through flatter areas,
+perhaps along river valleys.
 \plr{reference some examples}
@@
 although the two concepts are equivalent.
+\plr{CHANGE TO mostly RESISTANCE STARTING AT SIMPLE EXAMPLE}
 
@@
 The \emph{commute time} of a Markov chain between locations $i$ and $j$
-is the mean time it takes a particle moving according to the rules of the chain, 
+is the expected time for a particle that moves according to the rules of the chain, 
 started at $i$, to first reach $j$ and then return.
@@
 between $i$ and $j$.
-The \emph{coalescence time} is the mean time it takes two particles started at $i$ and $j$
+The \emph{coalescence time} is the expected time for two particles started at $i$ and $j$
 to \emph{coalesce},
@@
 by introducing another set of model parameters, $q$, 
-that fill the same role as the coalescence rates, $\gamma$.
+that fill an analogous role to the coalescence rates, $\gamma$.
 Then, we use commute time to approximate genetic distance as
@@
 because particles do not coalesce instantly --
-therefore commute time is most analogous to the first meeting time.
+therefore, commute time is most analogous to the first meeting time.
 For a conceptual picture of the differences between the coalescence time and commute time,
@@
 Commute times only depend on the symmetric part of $H$ --
-given an antisymmetric matrix $Z$ (so that $Z + Z^T = 0$),
+given an skew-symmetric matrix $Z$ (so that $Z + Z^T = 0$),
 any Markov chain that has hitting times given by $H_\epsilon = H + \epsilon Z$ for some $\epsilon$
@@
 where $I$ is the identity matrix and $\otimes$ is the Kronecker product.
-\plr{say a bit more about how we solve this?}
 
@@
 (even numerical instability)
-produce large differences in the inferred rates.
+may produce large differences in the inferred rates.
 Very roughly speaking, this occurs because as the spatial resolution increases,
@@
 \end{align*}
-Writing out equation \ref{eqn:C_matrix}, we get three unique equations
-for the coalescence times:
-\plr{simplify these equations and remove matrix version?}
+Equation \ref{eqn:C_matrix} equates two $2 \times 2$ matrices,
+so provides four equations. Only three of these are unique;
+simplifying these and using that $C_{12} = C_{21}$, these are:
 \begin{align*}
-   -C_{11} G_{12} + C_{21} G_{12} - C_{11} G_{12} + C_{12} G_{12} - C_{11} \gamma_1 &= -1 \\
-    C_{11} G_{21} - C_{21} G_{21} - C_{21} G_{12} + C_{22} G_{12}                   &= -1 \\
-    C_{12} G_{21} - C_{22} G_{21} + C_{21} G_{21} - C_{22} G_{21} - C_{22} \gamma_2 &= -1 .
+    2 \left(C_{12} - C_{11} \right) G_{12} - C_{11} \gamma_1 &= -1 \\
+    \left(C_{11} - C_{12} \right) G_{21} + \left(C_{22} - C_{12}\right) G_{12} &= -1 \\
+    2 \left( C_{12} - C_{22} \right) G_{21} - C_{22} \gamma_2 &= -1 .
 \end{align*}
 Given $C$, we have three equations for the four unknowns, 
-$G_{12}$, $G_{21}$, $\gamma_{1}$, and $\gamma_{2}$.
-By grouping terms together, factoring back into matrix form, 
-and writing $C_{ji}$ instead of $C_{ij}$ when $i > j$,
-we have
-\begin{align*}
-\begin{bmatrix}
--2C_{11} + 2C_{12} & 0                 & -C_{11} & 0       \\
--C_{12} + C_{22}   & C_{11} - C_{12}   & 0       & 0       \\
-0                  & 2C_{12} - 2C_{22} & 0       & -C_{22}
-\end{bmatrix}
-\begin{bmatrix}
-G_{12} \\ G_{21} \\ \gamma(1) \\ \gamma(2)
-\end{bmatrix}
-=
-\begin{bmatrix}
--1 \\ -1 \\ -1
-\end{bmatrix}.
-\end{align*}
-This is small enough to be solved symbolically. 
+$G_{12}$, $G_{21}$, $\gamma_{1}$, and $\gamma_{2}$,
+which we can solve symbolically.
 The valid solutions to these equations are those with $G_{12}$, $G_{21}$, $\gamma_1$,
@@
 \begin{align} \label{eq:post}
+    \begin{split}
 \mathcal{L}(D, G, \gamma) 
@@
         \right)^2 
+    \end{split}
 \end{align}
@@
 except that $q$ replaces $\gamma$,
-and instead of $\mathcal{C}(G,\gamma)$ we use $\comdist(G, q)$
-as defined in equation \eqref{eq:commute_approx}.
+and instead of $\mathcal{C}(G,\gamma)$ we use $\comdistfn(G, q)$,
+which is the matrix $\comdist$ computed from $G$ and $q$ by solving equation \eqref{eq:commute_approx}.
 
@@
 (plotted below using igraph \citep{igraph}).
-Then we calculated exact coalescence and commute times on these graphs,
-and added Gaussian noise independently to each entry of the resulting matrices
-(while preserving symmetry)
+Given the edge weights ($G$) and the coalescence rates ($\gamma$) of a graph,
+we produced data by calculating exact coalescence times ($\mathcal{C}(G,\gamma)$),
+to whose entries we added independent Gaussian noise while preserving symmetry,
 as described above.
@@
 which is close to nonidentifiable.
-(For commute time inference, we analogously constrained all values of $q$ to be identical or not.)
+(We will not present results from an analogous version of resistance distance inference
+that constrains all values of $q$ to be identical
+because the true values are not identical, even if all coalescence rates are,
+and in practice the method infers true values of $q$ extremely reliably.)
 To quantify accuracy across these random graphs in different situations,
@@
 Although increased noise in the data increased uncertainty in the estimated values,
-coalescence time inference correctly inferred the location of the barrier
-(as well as high-dispersal connections) in both cases.
-Commute time inference \plr{did what?}
+coalescence time inference correctly inferred not only the location of the barrier
+but also the migration rates along all other edges within a reasonable margin of error.
+Commute time inference also identified the barrier,
+but was much less accurate for the other values.
 We also explored making the problem harder,
@@
     were chosen by rounding independent draws of an exponential random variable with mean 1 
-    up to the nearest tenth.}
-    \label{fig:5x3b_grid}
+    up to the nearest tenth.
+    } \label{fig:5x3b_grid}
 \end{figure}
@@
     See Figure \ref{fig:5x3b_grid} to see movement parameter locations.
-    \plr{TODO: replace bottom two panels with same thing for commute time}
+    \plr{TODO: replace bottom two panels with same thing for commute time (eg figs/5x3b100\_com.pdf);
+         put this figure in the supplement}
 }
@@
 For instance, suppose there are three populations arranged in a line,
-and lineages currently in the outer populations are much more likely to come from the central island
+and lineages currently in the outer populations are much more likely to come from the central population
 than the other way around.
@@
      \includegraphics[scale=.8]{figs/4x4coalvcomH}
-    \caption{Plots of the true values of $\comdist$ vs $H$ 
+    \caption{Plots of the true values of $\comdist$ vs $C$ 
     for the uniform graph, symmetric graph, the asymmetric graph,
@@
 
-While the primary focus is coalescence time based inference,
-we also compare the results with that of resistance distance (i.e., commute time) based inference.
+In addition to developing a method for inferring migration rates using pairwise coalescence times,
+we compare the results with that of resistance distance (i.e., commute time) based inference.
 Coalescence based inference typically has lower error than commute time based inference,

### 2018-08-27 sims/get_divergences.py
.

@@ -79,7 +79,7 @@ def grid_samples(ts, n, m, prob=1.0, shrink=0.75):
 for treefile in glob.glob(os.path.join(outdir, "*.trees")):
     logfile.write("Reading {}\n".format(treefile))
     logfile.flush()
-    decap = pyslim.load(treefile, slim_format=True)
+    decap = pyslim.load(treefile)
 
     sample_grid = grid_samples(decap, n=grid_width, m=grid_height, prob=1.0, shrink=0.75)
     samples = [a for b in sample_grid for a in b]
@@ -102,8 +102,7 @@ for treefile in glob.glob(os.path.join(outdir, "*.trees")):
         # we want individual divergences
         the_subsamples = [x for y in subsamples for x in y]
         # simplify first for speed
-        sub_ts = decap.simplify(the_subsamples)
-        sub_ts = pyslim.SlimTreeSequence.load_tables(sub_ts.tables)
+        sub_ts = pyslim.SlimTreeSequence(decap.simplify(the_subsamples))
         recap = sub_ts.recapitate(recombination_rate=1e-9, Ne=1e3)
         mut_ts = msprime.mutate(sub_ts, rate=mutation_rate)
         logfile.write("Recapitated; added {} mutations to {} trees. Now computing statistics.\n".format(mut_ts.num_mutations, mut_ts.num_trees))

### 2018-08-27 sims/submit_sim.sh

@@ -3,7 +3,7 @@
 USAGE="
 Usage:
    $0 TYPE
-where TYPE should be 'flat', 'biased' or 'barrier'
+where TYPE should be 'flat', 'biased', 'barrier', 'valley', or 'expansion'
 "
 
 if [ $# -lt 1 ]

### 2018-08-27 isolation_by_coalescence.tex
minor fix
minor latex fix
re-did figures 4,5,6

@@
 
-\newcommand{\comdist}{\widetilde R}
-\newcommand{\comdistfn}{\widetilde \mathcal{R}}
+\newcommand{\comdist}{\widetilde{R}}
+\newcommand{\comdistfn}{\widetilde{\mathcal{R}}}
 
@@
 Coalescence rates are set to 1 everywhere.
-Figure \ref{fig:5x3boxplots} shows posterior distributions of the movement rates
-for both coalescence and commute time inference
+Figure \ref{fig:5x3b_post_coalvcom} shows posterior distributions of the movement rates
+for both coalescence time and resistance distance inference
 on this graph, with noise standard deviation equal to
@@
 \centering
-     \includegraphics[scale=0.8]{figs/5x3boxplots}
+     \includegraphics[scale=1]{figs/5x3b_post_coalvcom}
     \caption{Posterior distributions for values of $g$ 
     for the 5 $\times$ 3 graph with barriers 
-    for each of the analysis cases.
+    with two different noise levels with both coalescence time and resistance distance inference.
     See Figure \ref{fig:5x3b_grid} to see movement parameter locations.
-    \plr{TODO: replace bottom two panels with same thing for commute time (eg figs/5x3b100\_com.pdf);
-         put this figure in the supplement}
+    %\plr{TODO: replace bottom two panels with same thing for commute time (eg figs/5x3b100\_com.pdf);
+    %     put this figure in the supplement}
 }
-    \label{fig:5x3boxplots}
+    \label{fig:5x3b_post_coalvcom}
 \end{figure}
@@
 and movement rates up or to the right are 0.5.
+For inference, we added noise with standard deviation $1/500^\text{th}$ of the mean value.
 
 Differences between true mean coalescence times ($C$) and mean commute times ($\comdist$) 
-are shown in Figure \ref{fig:4x4coalvcomH}:
+are shown in Figure \ref{fig:RvsC}:
 they are fairly small in the uniform and symmetric cases, moderate in the asymmetric case, 
@@
 \centering
-     \includegraphics[scale=.8]{figs/4x4coalvcomH}
+     \includegraphics[scale=1]{figs/RvsC}
     \caption{Plots of the true values of $\comdist$ vs $C$ 
@@
     shown with the $y=x$ line.
-    \plr{TODO: make this a 2x2 grid instead of 4x1}
+    %\plr{TODO: make this a 2x2 grid instead of 4x1}
     See Figure \ref{fig:4x4_grids} to see overall graph structure.}
-    \label{fig:4x4coalvcomH}
-\end{figure}
-
-\begin{figure}
-\centering
-     \includegraphics[scale=.8]{figs/4x4box}
+    \label{fig:RvsC}
+\end{figure}
+
+\begin{figure}
+\centering
+     \includegraphics[scale=1]{figs/4x4boxplots_paper}
     \caption{Boxplots of the posterior distributions for $g$ for commute time and coalescence time
@@
 when there is an overall bias in the direction of the asymmetry across the landscape 
-(Figure \ref{fig:4x4coalvcomH}).
+(Figure \ref{fig:RvsC}).
 When inference is performed on discrete landscapes 
@@
     for the 5 $\times$ 3 graph with barriers 
-    for each of the analysis cases.
+    for each of the analysis cases with coalescence time inference.
     See Figure \ref{fig:5x3b_grid} to see movement parameter locations.
-    \plr{TODO: put bottom two panels on top; on the bottom put the same thing for commute time}
+    %\plr{TODO: put bottom two panels on top; on the bottom put the same thing for commute time}
 }

### 2018-08-30 isolation_by_coalescence.tex

@@
 The idea of ``resistance distance'' is an important tool in the lanscape genomics toolbox.
-Introduced by \citet{mcrae},
+Introduced by \citet{mcrae2006isolation},
 it makes use of a mathematical equivalence between random walks and electrical resistance
@@
 whose conductances are given by the movement rates of the Markov chain \citep{levin}.
-Both electrical resistance and commute time between two nodes
-average over all possible paths through the network between the two.
-For this reason, it is thought that if genetic distances between samples
-from different parts of the landscape
-are significantly correlated with the effective resistances between those locations,
-with local conductances assigned based on the values of some landscape variable,
-then that landscape variable is likely a good indicator of the regions of the map 
-where gene flow occurs.
+This measure therefore averages over all possible paths through the network between the two.
+For this reason, 
+if assigning local conductances based on the values of some landscape variable
+results in a significant positive correlation between 
+genetic distances between samples from different parts of the landscape
+and the effective resistances between those locations,
+then this is taken as good evidence that the landscape variable is a good indicator of
+where gene flow occurs \citep{mcrae2007circuit,cushman2006complex,more_examples}.
 For instance,
@@
 perhaps along river valleys.
-\plr{reference some examples}
-Other methods \citep{eems} seek to build \emph{de novo} 
+More recent methods \citep{petkova2016visualizing,hanks2013circuit} seek to build \emph{de novo} 
 a map of conductance values that produces resistances 
@@
 
+As motivated by \citet{mcrae2006isolation},
+commute time (i.e., resistance distance) is a more computationally tractable approximation
+to \emph{coalescence time}, a more accurate model of genetic distance data.
 Genetic distances between two genomes sampled from the landscape
@@
 If so, would using coalescence time do better?
-To better contrast the two,
-we refer to ``commute time'' rather than ``resistance'' throughout,
-although the two concepts are equivalent.
+
+\citet{mcrae2006isolation} showed that coalescence time and resistance distance are equal
+(up to shift and scaling) in isotropic landscapes such as a ring \citep{matsen_wakeley},
+and found that the two were highly correlated in several test cases.
+However, all these test cases involved symmetric rates of gene flow,
+which is likely not the case in many situations, e.g.,
+in rivers \citep{morrissey2009maintenance,sundqvist2016directional,hanks2017modeling},
+source-sink dynamics \citep{source_sink},
+or after population expansions \citep{expansions}.
+
 \plr{CHANGE TO mostly RESISTANCE STARTING AT SIMPLE EXAMPLE}
@@
 In this paper, we contrast 
-coalescence time and commute time,
+coalescence time and resistance distance,
 and develop an inference method for movement rates on a landscape
@@
 commute time inference may get the wrong answer if data derive from a coalescent process,
-as does real genetic data.
+as does real genetic data, especially in the presence of asymmetry.
 We also explore the question of identifiability.
@@
 which allows very fine geographic resolution \citep{infer_resistance}.
-Others methods such as EEMS \citep{eems} infer conductances without such prior information;
+Others methods such as EEMS \citep{petkova2016visualizing} infer conductances without such prior information;
 however, the geographical resolution is much coarser.
@@
 implying the need for regularization to obtain reliable inference.
-Using maps for prior information, or restricting the geographical resolution of the results,
-are two ways to do this.
+This explains the results of \citet{graves2013current},
+who found a large area of resistance parameters
+that produced equivalently good fits to genetic distance
+(i.e., nonidentifiability, analogous to a flat likelihood surface).
 
@@
 for two genomes, it is calculated as the density of sites at which the two differ.
+It is more common in the literature to use $F_{ST}/(1-F_{ST})$,
+but the two are highly correlated for good theoretical reasons \citet{slatkin_fst,rousset1997genetic},
+especially for SNP data.
 The genomes differ at a site only if there has been a mutation in some ancestor on the lineages leading
@@
 \end{align}
+Similar recursions are common in the literature,
+going back at least to \citet{hill1972effective} (see also \citet{whitlock1997effective}).
 In matrix notation, this is
@@
 
+\paragraph{Asymmetric dispersal}
+(Review of asymmetric situations.)
+
+\citet{hanks2017modeling} developed a method that allows asymmetry
+by modeling genetic similarity as deriving from an underlying random field
+whose space-time covariance is given by the covariance of the (forwards-time) population fluctuations.
+
 \paragraph{Consequences for resistance methods}
@@
 
+\paragraph{Limitations of data}
+Something about microsats versus RADseq versus whole-genome.
+\plr{XXX}
+
+\paragraph{Landscapes or discrete demes?}
+Other inference methods \citep{wilson2003bayesian,beerli1999maximum,beerli2010unified,hey2007integration}
+work with coalescence inference, but only with a small number of loci and populations,
+because of computational efficiency.
+We're somewhere in between.
+\plr{XXX}
+
 \paragraph{Assumptions}

### 2018-08-30 sims/plot_locations.py
plotting

@@ -6,16 +6,17 @@ import glob
 matplotlib.use('Agg')
 from matplotlib import pyplot as plt
 
-outdir = "run_031087/"
-# treefiles = [outdir + "pop_{:03d}.trees".format(x) for x in range(1, 21)]
-treefiles = glob.glob(outdir + "pop_*.trees")
+treefiles = glob.glob("run_0*/*.trees")
 
 for treefile in treefiles:
-    # treefile = outdir + "pop_100.trees"
+    print(treefile)
     outfile = treefile + ".locs.png"
-    ts = pyslim.load(treefile, slim_format=True)
+    locfile = treefile + ".locs.tsv"
+    ts = msprime.load(treefile)
     locations = np.array([i.location for i in ts.individuals()])
+    np.savetxt(locfile, locations, header="x\ty\tz")
     fig = plt.figure(figsize=(max(locations[:,0]),max(locations[:,1])))
     plt.scatter(locations[:,0], locations[:,1], marker='.')
     plt.savefig(outfile, dpi=288)
+    print("... done!")
 

### 2018-08-31 sims/get_divergences.py
fixed divergence script

@@ -101,9 +101,8 @@ for treefile in glob.glob(os.path.join(outdir, "*.trees")):
         samplefile.close()
         # we want individual divergences
         the_subsamples = [x for y in subsamples for x in y]
-        # simplify first for speed
-        sub_ts = pyslim.SlimTreeSequence(decap.simplify(the_subsamples))
         recap = sub_ts.recapitate(recombination_rate=1e-9, Ne=1e3)
+        sub_ts = pyslim.SlimTreeSequence(recap.simplify(the_subsamples))
         mut_ts = msprime.mutate(sub_ts, rate=mutation_rate)
         logfile.write("Recapitated; added {} mutations to {} trees. Now computing statistics.\n".format(mut_ts.num_mutations, mut_ts.num_trees))
         logfile.flush()

### 2018-08-31 isolation_by_coalescence.tex
moar words

@@
 For more discussion of this latter contribution in other contexts, see
-\citet{waples}, \citet{wakeley}, or \citet{ralph_empirical}.
+\citet{wakeley2012genealogies}, \citet{waples2009modeling}, or \citet{ralph2015empirical}.
 In populations living across continuous geography,
@@
 Although commute and coalescence times can be quite different,
-we show that they are equal (with a particular choice of local diversity)
+as also noted by \citet{mcrae2006isolation}, 
+they are equal (with a particular choice of local diversity)
 if populations are arranged on a ring or torus
@@
 
+\paragraph{Other methods}
+Other methods that use resistance distance have substantial advantages over the method we present here.
+Circuitscape \citep{mcrae2008using} produces maps of much higher resolution 
+than is presently possible with coalescent times.
+\citet{hanks2013circuit} connects the problem to intrinsic conditional autoregressive models, 
+providing a Bayesian model of an underlying Gaussian Markov random field.
+EEMS \citep{petkova2014visualizing} integrates over possible tesselations of the region
+to obtain a landscape at the best possible resolution supported by the data.
+
+Although we present a new method, based on coalescence times,
+our focus is on understanding the underlying modeling issues
+rather than replacing these other tools.
+Most promisingly, a very recent method, MAPS \citep{alasadi2018estimating}
+is based on EEMS, but instead of resistance distances,
+uses pairwise sharing of long genomic segments (``IBD segments'', \citet{browning2013improving}),
+which have been shown to carry substantial information about recent demography
+\citep{ralph2013geography,palamara2013inference,browning2015accurate}.
+Furthermore, the likelihood model underlying MAPS is based on coalescent theory,
+and thus should not be subject to the drawbacks of resistance distance.
+However, application requires fairly dense genotyping and a good genetic map,
+something not available for many species.
+
+Other widely-used inference methods, such as
+MIGRATE \citep{beerli1999maximum,beerli2010unified},
+BayesAss \citep{wilson2003bayesian}, 
+and IMa \citet{hey2007integration}
+use likelihoods derived from a coalescent model --
+and use much more of the data than we do.
+However, because they use locus-specific data rather than an average across loci,
+these methods can only be applied to a relatively small number of populations and loci,
+because of computational efficiency.
+
 \paragraph{Asymmetric dispersal}
-(Review of asymmetric situations.)
+We have found that asymmetric rates of lineage movement
+are most likely to produce qualitative differences between coalescence times and resistance distances,
+and thereby mislead models based on resistance distance.
+This asymmetry occurs if one region receives a greater proportion of each new generation as migrants
+from another population than that other population does from it.
+This is expected, for instance,
+if organismal dispersal is biased by wind or water currents \citep{gaines2003avoiding,morrissey2009maintenance},
+or in the presence of source-sink dynamics \citep{dias1996sources,lenormand2002limits}.
+XXX MORE
 
@@
 
-\paragraph{Landscapes or discrete demes?}
-Other inference methods \citep{wilson2003bayesian,beerli1999maximum,beerli2010unified,hey2007integration}
-work with coalescence inference, but only with a small number of loci and populations,
-because of computational efficiency.
-We're somewhere in between.
-\plr{XXX}
-
 \paragraph{Assumptions}

### 2018-09-01 sims/get_divergences.py
minor fixes

@@ -84,7 +84,7 @@ for treefile in glob.glob(os.path.join(outdir, "*.trees")):
     sample_grid = grid_samples(decap, n=grid_width, m=grid_height, prob=1.0, shrink=0.75)
     samples = [a for b in sample_grid for a in b]
     # extract nonoverlapping subsamples
-    num_subs_per_cell = [min(samples_per_cell, int(len(a)/num_subsamples)) for a in samples]
+    num_subs_per_cell = [min(2*samples_per_cell, int(len(a)/num_subsamples)) for a in samples]
     for a in num_subs_per_cell:
         assert(a > 0)
 

### 2018-09-03 sims/get_divergences.py
.

@@ -101,7 +101,7 @@ for treefile in glob.glob(os.path.join(outdir, "*.trees")):
         samplefile.close()
         # we want individual divergences
         the_subsamples = [x for y in subsamples for x in y]
-        recap = sub_ts.recapitate(recombination_rate=1e-9, Ne=1e3)
+        recap = decap.recapitate(recombination_rate=1e-9, Ne=1e3)
         sub_ts = pyslim.SlimTreeSequence(recap.simplify(the_subsamples))
         mut_ts = msprime.mutate(sub_ts, rate=mutation_rate)
         logfile.write("Recapitated; added {} mutations to {} trees. Now computing statistics.\n".format(mut_ts.num_mutations, mut_ts.num_trees))

### 2018-10-04 sims/eemsify.R
run eems whatnot

@@ -1,0 +1,42 @@
+# convert our data to EEMS format
+
+usage <- "eemsify.R DISTANCE_FILE SAMPLE_FILE OUT_FILEBASE"
+
+args <- commandArgs(TRUE)
+cat("args: ", args, "\n")
+if (length(args) != 3) { stop(usage) }
+DISTFILE <- args[1]
+SAMPLEFILE <- args[2]
+OUTBASE <- args[3]
+
+DIFFFILE <- paste0(OUTBASE, ".diffs")
+COORDFILE <- paste0(OUTBASE, ".coord")
+OUTERFILE <- paste0(OUTBASE, ".outer")
+
+x <- scan(DISTFILE, skip=1)
+z <- read.table(SAMPLEFILE, header=TRUE, stringsAsFactors=FALSE)
+z$individual <- ordered(z$individual, levels=unique(z$individual))
+stopifnot(all(diff(as.numeric(z$individual)) >= 0))
+stopifnot(all(diff(as.numeric(z$individual)) <= 1))
+nr <- nrow(z)
+d <- matrix(NA, nrow=nr, ncol=nr)
+d[lower.tri(d, diag=TRUE)] <- x
+d[upper.tri(d)] <- t(d)[upper.tri(d)]
+dd <- do.call(rbind, tapply(1:nrow(d), z$individual, function (u) rowMeans(d[,u], na.rm=TRUE)))
+ddd <- do.call(cbind, tapply(1:nrow(d), z$individual, function (u) rowMeans(dd[,u], na.rm=TRUE)))
+
+write.table(ddd / mean(ddd), file=DIFFFILE, col.names=FALSE, row.names=FALSE)
+
+ii <- match(levels(z$individual), z$individual)
+write.table(z[ii,c("x","y")], file=COORDFILE, col.names=FALSE, row.names=FALSE)
+
+x0 <- min(z$x)/2
+x1 <- max(z$x)+x0
+y0 <- min(z$y)/2
+y1 <- max(z$y)+y0
+outers <- matrix(c(x0, y0,
+                   x0, y1,
+                   x1, y1,
+                   x1, y0,
+                   x0, y0), ncol=2, byrow=TRUE)
+write.table(outers, file=OUTERFILE, col.names=FALSE, row.names=FALSE)

### 2018-10-04 sims/submit_eems.sh

@@ -1,0 +1,47 @@
+#!/bin/bash
+
+USAGE="
+Usage:
+   $0 DATABASE
+where DATABASE is the file base of the data.
+"
+
+if [ $# -lt 1 ]
+then
+    echo "$USAGE"
+    exit 0
+fi
+
+DATABASE="$1"
+DATADIR="$(dirname $DATABASE)"
+
+NAME="$(basename $DATABASE)_eems"
+TAG=$(printf "%06d" $RANDOM); 
+OUTDIR="$DATADIR/${NAME}_${TAG}"
+mkdir -p $OUTDIR
+echo "Directory: $OUTDIR"
+
+OUTBASE="${OUTDIR}/eems_data"
+SAMPLEFILE="${DATABASE}.samples.tsv"
+DISTFILE="${DATABASE}.divs.tsv"
+
+# eemsify
+Rscript eemsify.R $DISTFILE $SAMPLEFILE $OUTBASE
+
+NINDIVS=$(wc -l ${OUTBASE}.coord | cut -f 1 -d ' ')
+
+INIFILE=$OUTDIR/eems.ini
+touch $INIFILE
+
+echo "datapath = $OUTBASE" >> $INIFILE
+echo "mcmcpath = $OUTDIR" >> $INIFILE
+echo "nIndiv = $NINDIVS" >> $INIFILE
+echo "nSites = 30000" >> $INIFILE
+echo "nDemes = 24" >> $INIFILE
+echo "diploid = false" >> $INIFILE
+echo "numMCMCIter = 500000" >> $INIFILE
+echo "numBurnIter = 100000" >> $INIFILE
+echo "numThinIter = 9999" >> $INIFILE
+
+export OUTDIR
+sbatch -o $OUTDIR/slurm_${TAG}.out -e $OUTDIR/slurm_${TAG}.out ./run_eems.sbatch 

### 2018-10-05 isolation_by_coalescence.tex
begun putting in other sims

@@
 \paragraph{Continuous Landscape with Biased Migration}
+We simulated several situations with an effective bias in migration on a continuous landscape.
+First, the ``biased'' situation, was a square landscape as above,
+but with mean offspring location one tenth of a standard deviation up and to the right of the parent's location.
+This produces reverse-time gene flow down and to the left.
+Second, we simulated a ``valley'' in a $3 \times 5$ landscape as above, 
+in which fecundity of individuals in the middle 1/4 of the landscape was reduced by 50\%.
+This produces gene flow ``uphill'', away from the ridge,
+and thus an effective barrier across the ridge.
+Third, we simulated a recent ``expansion'',
+in which the middle 4/5 of the landscape was uninhabitable 400,000 generations,
+until 500 generations before the present day.
+The resulting expansion produces gene flow back away from the center,
+similar to the ``valley''.
+\textbf{MAKE FIGURE OF THESE?}
+\textbf{Merge with berow.}
+
 The third situation that we will discuss is where migration is biased.
@@
 \label{fig:post_dists_bias_4x4_1}
-\end{figure}
-
-\section*{Application to the Mojave Desert Tortoise}
-
-The Mojave desert tortoise (\textit{Gopherus agassizii})
-lives across much of the Mojave desert of California and Nevada,
-separated from its sister taxa, \textit{Gopherus morofkaii} by the Colorado river.
-We applied our method to genomic distances calculated from
-whole genome sequencing data of 271 individuals sampled from across much of the range,
-reported in \citet{shaffer2017desert}.
-We discretized the landscape into 13 regions based on watershed polygons
-from the Watershed Boundary Dataset \citep{WBD}
-(using WBD8 slightly modified by portions of WBD10 to improve contiguity of regions).
-The resulting discretization,
-overlaid on a map including tortoise sampling locations,
-is shown in Figure \ref{fig:tort_land}.
-Since the regions vary substantially in size, 
-and may vary substantially in population density, 
-we do not assume that coalescence rates are the same everywhere, 
-which gives 13 coalescence rates;
-combined with the 50 migration parameters between each pair of adjacent regions
-we have a total of 63 unknown parameters
-(and 91 equations).
-
-Posterior distributions of the parameters are shown in Figure \ref{fig:tort_post}.
-There is substantial variation in the gene flow rates 
-but not the coalescence rates, 
-with the exception of region 4, which seems to have a relatively slow coalescence rate.
-This is not too surprising given its large area and central location.
-Gene flow rates are low across mountainous boundaries:
-for instance, regions 1 and 4 are separated by the Paiute and Castle mountains,
-and regions 4 and 12 are separated by the Providence and Granite mountains.
-% and regions 1 and 9 are separated by the New York and McCollough mountains.
-On the other hand, there is a large, mostly open bajada connecting areas 9 and 13, 
-and we infer high rates of migration between the two regions.
-The coalescence time model provides a good fit to the data,
-much better than was obtained by commute times, either \plr{in this paper??}
-or using the resistance distance method of \citet{shaffer2017desert}
-(Figure \ref{fig:tort_h_comp}).
-
-The most biologically interesting aspects
-are the strong asymmetries, for instance between region 9 and regions 11, 12, and 13.
-These indicate that resistance-based methods may be strongly misled.
-However, we cannot conclude from this analysis \emph{what} is leading to the observed asymmetries.
-One explanation would be source-sink dynamics:
-for instance, the observed bias in lineage movement could be observed 
-if the tortoise fecundity in the Ivanpah valley in region 9
-is higher than in the colder, higher-elevation regions to the north.
-Or, it could be the result of a historical range expansion out of the area,
-perhaps out of a more resticted range as climates changed after the last glacial maximum.
-The presence of source-sink dynamics would be of great interest for tortoise conservation efforts,
-but it is not clear from this analysis which explanation is more likely.
-Some aspects of the results may even be the produced by sample configuration:
-for instance, perhaps the inferred gene flow between regions 2 and 3 is low
-simply because individuals sampled in region 3 are geographically far from region 2.
-
-\begin{figure}
-\centering
- \includegraphics[width=0.45\textwidth]{figs/fancy_watershed_assignments}
- \includegraphics[width=0.45\textwidth]{figs/tort_graph_results}
-\caption{Left: Locations of the sampled tortoises.
-    \plr{add mountain shadings and ocean; remove bits on the other side of the Colorado}
-    Right: Sample maximum posterior estimates 
-    of movement rates.
-    \plr{can these be maximum posterior estimates instead?}
-    } \label{fig:tort_land}
-\end{figure}
-
-\begin{figure}
-\centering
- \includegraphics[width=0.8\textwidth]{figs/tort_post_g}
- \includegraphics[width=0.8\textwidth]{figs/tort_post_gam}
-\caption{
-    Boxplots of the posterior distributions for 
-    \textbf{(top)} movement rates, and
-    \textbf{(bottom)} coalescence rates,
-    for the Mojave desert tortoise data set.
-    Parameters are labeled as in Figure \ref{fig:tort_land}.
-    } \label{fig:tort_post}
-\end{figure}
-
-\begin{figure}
-\centering
- \includegraphics[scale=.8]{figs/tort_h_comp}
-\caption{
-    Mean coalescence times calculated from the sample maximum likelihood model parameters 
-    plotted against the observed mean coalescence times calculated from the tortoise data, 
-    measured in the mean of pairwise $\pi$.
-    The mean absolute value relative differences between the two is 0.4\%.
-    } \label{fig:tort_h_comp}
 \end{figure}
@@
 For instance,
-the width of the tortoise range is roughly 500km,
-and if mean dispersal distance is 10km (a high estimate),
+if the width of a species' range is roughly 500km,
+and if mean dispersal distance is 10km,
 the standard deviation of ancestor location $t$ generations ago is $10\sqrt{t}$,
-so the mixing time of tortoise lineages across the modern landscape,
+so the mixing time of a lineages across the landscape,
 not accounting for barriers, is of order 2500 generations,
 or about 50,000 years.
-The landscape of the Mojave has changed substantially over tihs time \citep{pleistocene_mojave},
+The landscape of many areas has changed substantially over this time, % \citep{pleistocene_mojave},
 which suggests that models incorporating change over time 
@@
 \end{document}
+
+\section*{Application to the Mojave Desert Tortoise}
+
+The Mojave desert tortoise (\textit{Gopherus agassizii})
+lives across much of the Mojave desert of California and Nevada,
+separated from its sister taxa, \textit{Gopherus morofkaii} by the Colorado river.
+We applied our method to genomic distances calculated from
+whole genome sequencing data of 271 individuals sampled from across much of the range,
+reported in \citet{shaffer2017desert}.
+We discretized the landscape into 13 regions based on watershed polygons
+from the Watershed Boundary Dataset \citep{WBD}
+(using WBD8 slightly modified by portions of WBD10 to improve contiguity of regions).
+The resulting discretization,
+overlaid on a map including tortoise sampling locations,
+is shown in Figure \ref{fig:tort_land}.
+Since the regions vary substantially in size, 
+and may vary substantially in population density, 
+we do not assume that coalescence rates are the same everywhere, 
+which gives 13 coalescence rates;
+combined with the 50 migration parameters between each pair of adjacent regions
+we have a total of 63 unknown parameters
+(and 91 equations).
+
+Posterior distributions of the parameters are shown in Figure \ref{fig:tort_post}.
+There is substantial variation in the gene flow rates 
+but not the coalescence rates, 
+with the exception of region 4, which seems to have a relatively slow coalescence rate.
+This is not too surprising given its large area and central location.
+Gene flow rates are low across mountainous boundaries:
+for instance, regions 1 and 4 are separated by the Paiute and Castle mountains,
+and regions 4 and 12 are separated by the Providence and Granite mountains.
+% and regions 1 and 9 are separated by the New York and McCollough mountains.
+On the other hand, there is a large, mostly open bajada connecting areas 9 and 13, 
+and we infer high rates of migration between the two regions.
+The coalescence time model provides a good fit to the data,
+much better than was obtained by commute times, either \plr{in this paper??}
+or using the resistance distance method of \citet{shaffer2017desert}
+(Figure \ref{fig:tort_h_comp}).
+
+The most biologically interesting aspects
+are the strong asymmetries, for instance between region 9 and regions 11, 12, and 13.
+These indicate that resistance-based methods may be strongly misled.
+However, we cannot conclude from this analysis \emph{what} is leading to the observed asymmetries.
+One explanation would be source-sink dynamics:
+for instance, the observed bias in lineage movement could be observed 
+if the tortoise fecundity in the Ivanpah valley in region 9
+is higher than in the colder, higher-elevation regions to the north.
+Or, it could be the result of a historical range expansion out of the area,
+perhaps out of a more resticted range as climates changed after the last glacial maximum.
+The presence of source-sink dynamics would be of great interest for tortoise conservation efforts,
+but it is not clear from this analysis which explanation is more likely.
+Some aspects of the results may even be the produced by sample configuration:
+for instance, perhaps the inferred gene flow between regions 2 and 3 is low
+simply because individuals sampled in region 3 are geographically far from region 2.
+
+\begin{figure}
+\centering
+ \includegraphics[width=0.45\textwidth]{figs/fancy_watershed_assignments}
+ \includegraphics[width=0.45\textwidth]{figs/tort_graph_results}
+\caption{Left: Locations of the sampled tortoises.
+    \plr{add mountain shadings and ocean; remove bits on the other side of the Colorado}
+    Right: Sample maximum posterior estimates 
+    of movement rates.
+    \plr{can these be maximum posterior estimates instead?}
+    } \label{fig:tort_land}
+\end{figure}
+
+\begin{figure}
+\centering
+ \includegraphics[width=0.8\textwidth]{figs/tort_post_g}
+ \includegraphics[width=0.8\textwidth]{figs/tort_post_gam}
+\caption{
+    Boxplots of the posterior distributions for 
+    \textbf{(top)} movement rates, and
+    \textbf{(bottom)} coalescence rates,
+    for the Mojave desert tortoise data set.
+    Parameters are labeled as in Figure \ref{fig:tort_land}.
+    } \label{fig:tort_post}
+\end{figure}
+
+\begin{figure}
+\centering
+ \includegraphics[scale=.8]{figs/tort_h_comp}
+\caption{
+    Mean coalescence times calculated from the sample maximum likelihood model parameters 
+    plotted against the observed mean coalescence times calculated from the tortoise data, 
+    measured in the mean of pairwise $\pi$.
+    The mean absolute value relative differences between the two is 0.4\%.
+    } \label{fig:tort_h_comp}
+\end{figure}

### 2018-10-12 isolation_by_coalescence.tex
more figs

@@
 \usepackage{natbib}
+% \usepackage{subcaption}
 
@@
 \paragraph{Continuous Landscape with Biased Migration}
-We simulated several situations with an effective bias in migration on a continuous landscape.
-First, the ``biased'' situation, was a square landscape as above,
+We also simulated three situations on a continuous landscape that had an effective bias in migration.
+First, the \textbf{biased} situation, was a square landscape as above,
 but with mean offspring location one tenth of a standard deviation up and to the right of the parent's location.
+% (or 0.05\% of the edge length of the landscape)
 This produces reverse-time gene flow down and to the left.
-Second, we simulated a ``valley'' in a $3 \times 5$ landscape as above, 
+Second, we simulated a \textbf{valley} in a landscape with a $3 \times 5$ aspect ratio as above, 
 in which fecundity of individuals in the middle 1/4 of the landscape was reduced by 50\%.
@@
 and thus an effective barrier across the ridge.
-Third, we simulated a recent ``expansion'',
+Third, we simulated a recent \textbf{expansion},
 in which the middle 4/5 of the landscape was uninhabitable 400,000 generations,
@@
 similar to the ``valley''.
-\textbf{MAKE FIGURE OF THESE?}
-\textbf{Merge with berow.}
-
-The third situation that we will discuss is where migration is biased.
-We will be using a square landscape as in the first case, 
-but the offspring locations of offspring are not centered around the parent.
-Instead, the mean of the distribution of the locations
-is one tenth of a standand deviation (or 0.05\% of the edge length of the landscape) 
-up and to the right of their parent's location.
+
 This results in reverse time gene flow down and to the left 
@@
 
+\begin{figure}
+\centering
+    \includegraphics[width=0.45\textwidth]{figs/valley_locations}
+    \includegraphics[width=0.45\textwidth]{figs/valley_eems}
+    \includegraphics[width=0.45\textwidth]{figs/valley_commute_graph}
+    \includegraphics[width=0.45\textwidth]{figs/valley_coal_graph}
+    \includegraphics[width=0.45\textwidth]{figs/valley_commute_dists}
+    \includegraphics[width=0.45\textwidth]{figs/valley_coal_dists}
+    \caption{
+        Results from the \textbf{valley} simulation:
+        \textbf{(top left)} Positions of all individuals at the end of simulation.
+        \textbf{(top right)} mean migration rates as estimated by EEMS.
+        Below are shown
+        posterior median migration rates 
+            as estimated using 
+            \textbf{(middle left)} the commute time method and
+            \textbf{(middle right)} the the commute time method.
+        Last are shown the posterior distributions, for
+            \textbf{(bottom left)} the commute time method and
+            \textbf{(bottom right)} the the commute time method.
+        \label{sfig:valley_results}
+    }
+\end{figure}
+
+\begin{figure}
+\centering
+    \includegraphics[width=0.45\textwidth]{figs/expansion_locations}
+    \includegraphics[width=0.45\textwidth]{figs/expansion_eems}
+    \includegraphics[width=0.45\textwidth]{figs/expansion_commute_graph}
+    \includegraphics[width=0.45\textwidth]{figs/expansion_coal_graph}
+    \includegraphics[width=0.45\textwidth]{figs/expansion_commute_dists}
+    \includegraphics[width=0.45\textwidth]{figs/expansion_coal_dists}
+    \caption{
+        Results from the \textbf{expansion} simulation:
+        \textbf{(top left)} Positions of all individuals at the end of simulation.
+        \textbf{(top right)} mean migration rates as estimated by EEMS.
+        Below are shown
+        posterior median migration rates 
+            as estimated using 
+            \textbf{(middle left)} the commute time method and
+            \textbf{(middle right)} the the commute time method.
+        Last are shown the posterior distributions, for
+            \textbf{(bottom left)} the commute time method and
+            \textbf{(bottom right)} the the commute time method.
+        \label{sfig:expansion_results}
+    }
+\end{figure}
+
+\begin{figure}
+\centering
+    \includegraphics[width=0.45\textwidth]{figs/barrier_locations}
+    \includegraphics[width=0.45\textwidth]{figs/barrier_eems}
+    \includegraphics[width=0.45\textwidth]{figs/barrier_commute_graph}
+    \includegraphics[width=0.45\textwidth]{figs/barrier_coal_graph}
+    \includegraphics[width=0.45\textwidth]{figs/barrier_commute_dists}
+    \includegraphics[width=0.45\textwidth]{figs/barrier_coal_dists}
+    \caption{
+        Results from the \textbf{barrier} simulation:
+        \textbf{(top left)} Positions of all individuals at the end of simulation.
+        \textbf{(top right)} mean migration rates as estimated by EEMS.
+        Below are shown
+        posterior median migration rates 
+            as estimated using 
+            \textbf{(middle left)} the commute time method and
+            \textbf{(middle right)} the the commute time method.
+        Last are shown the posterior distributions, for
+            \textbf{(bottom left)} the commute time method and
+            \textbf{(bottom right)} the the commute time method.
+        \label{sfig:barrier_results}
+    }
+\end{figure}
+
 \end{document}

### 2018-10-14 isolation_by_coalescence.tex
more rearrangement

@@
 \begin{abstract}
-\plr{abstract needs editing}
 A number of methods commonly used in landscape genetics
@@
 using genetic distance data.
-These are motivated by a mathematical eqiuvalence of effective resistance
-to the mean commute time of a random walk.
+These are motivated by a mathematical equivalence between electrical resistance between two nodes
+and the ``commute time'',
+which is the mean time for a random walk on that network to leave one node, visit the other, and return.
 Here, we describe the differences between commute time
@@
 and explore the consequences for inference.
-We implement a Bayesian method to do inference under both these models,
+We implement a Bayesian method to infer effective movement rates and population sizes 
+under both these models,
 and find that inference using commute times
-when the data derive from coalescence times
 can dramatically mislead in the presence of biased gene flow.
-We apply the methods to empirical data,
-and find evidence for strongly biased gene flow.
+Forwards-time simulation with continuous geography demonstrates that
+coalescence-based inference remains more accurate than resistance-based methods on realistic data,
+and highlights the need for methods that do not approximate continuous space
+by a discrete set of populations.
 \end{abstract}
@@
 which allows very fine geographic resolution \citep{infer_resistance}.
-Others methods such as EEMS \citep{petkova2016visualizing} infer conductances without such prior information;
+Other methods such as EEMS \citep{petkova2016visualizing} infer conductances without such prior information;
 however, the geographical resolution is much coarser.
@@
 (i.e., nonidentifiability, analogous to a flat likelihood surface).
+
+Talk about MAPS.
 
@@
 \paragraph{Forwards-time simulations}
-To produce realistic data, we implemented forwards-time simulations using SLiM v3.0 \citep{haller2017},
-with individuals living across continuous, two-dimensional space (sometimes with barriers),
+To produce realistic data, we implemented forwards-time simulations using SLiM v3.0 \citep{haller2017slim,haller2018treesequences},
+with individuals living across continuous, two-dimensional geography (sometimes with barriers),
 from which we recorded genomic data.
-We performed three types of simulation:
-uniform migration on a continuous square landscape, 
-uniform migration on a rectangular landscape with barriers to migration, 
-and biased migration on a square landscape.
-Each simulation had $10^4$ diploid, hermaphroditic individuals with nonoverlapping generations.
+The basic simulation, which we modified to produce several other situations,
+is as follows.
+Each simulation had around $10^4$ diploid, hermaphroditic individuals, 
+living across two-dimensional geography.
 Mates were selected from nearby individuals using a Gaussian kernel with standard deviation $\sigma_d$,
 truncated at $3 \sigma_d$.
+Simulations were done on either a square region with width equal to $80\sigma_d$,
+or a $3:5$ rectangular region with the same area.
 Offspring are dispersed away from the mother using the same kernel,
-conditioned to lie within the habitable area.
+reflected from the boundaries of the habitable area.
 In each simulation, $\sigma_d$ was chosen 
@@
 Clumpiness is similar to what is seen in 
-Figure \ref{fig:ind_locs_5x3b_1}.
+Figure \ref{fig:5x3b_continuous}.
 
@@
 and a recombination rate of $10^{-8}$ per generation per locus.
-Mutation rate was $10^{-8}$ per locus for the simulation with uniform migration and no barrier,
-and $10^{-7}$ for the remaining simulations.
+Neutral mutations were added at a rate of $10^{-9}$ per meiosis.
+To make these simulations computationally feasible,
+the mutation rate was set to 0.0 during the forwards simulation,
+while tree sequences were recorded \citep{haller2018treesequences},
+and mutations were added after the fact with msprime,
+which is equivalent to including mutations during the forwards simulation 
+but resulted in much faster run times \citep{kelleher2018xxx}.
 In all cases, the number of mutations was more than adequate to reliably estimate genetic distance.
@@
 from among those within the middle three quarters of each dimension of rectangle,
-as shown for one situation in Figure \ref{fig:ind_locs_5x3b_1}.
+as shown for one situation in Figure \ref{fig:5x3b_continuous}.
 This protocol was chosen as a compromise between sampling all individuals 
@@
 is equal to the number of locations in the discretization.
+
+\subsection*{Code availability}
+
+All code used for this work is available at \url{https://github.com/elundgre/gene-flow-inference}
+(including an R package to perform inference)
+and \url{https://github.com/petrelharp/isolation_by_coalescence} (including the SLiM simulations).
 
@@
 and ``populations'' are constructed by necessity, often driven by sampling locations.
+We also applied coalescence and resistance inference,
+as well as the popular resistance-based method EEMS \citep{petkova_eems},
+to data from simulations on continuous geography.
 
@@
 but the word ``mean'' indicates an average across several levels of randomness:
-and observed, empirical mean estimated using a sample of genotyeps
-will differ slightly depending on both the individuals sampled (``sampling noise'')
+each observed, empirical mean estimated using a sample of genotyeps
+will deviate from this depending on both the individuals sampled (``sampling noise'')
 and the stochasticity of the population history itself (``process noise'').
@@
 both sources of noise can be substantially larger due to randomness in spatial locations.
-A hint of this is visible in Figure \ref{fig:ind_locs_5x3b_1},
-where we see that the distribution of individuals across the landscape
-is fairly patchy.
+This is hinted at by Figure \ref{fig:5x3b_continuous},
+where we see that the distribution of individuals across the landscape is fairly patchy:
+a barrier might be inferred in a region that has not happened to have a local population recently
+only because of demographic stochasticity.
 This might make inference using a discretization of continuous space considerably more difficult.
@@
 \centering
-\includegraphics[scale=.8]{figs/h_val_comp}
+    \includegraphics[width=0.8\textwidth]{figs/h_val_comp}
 \caption{
@@
 
+Although we found substantial process noise
+with a large, flat landscape with uniform, unbiased migration,
+many biological processes are expected to increase between-replicate stochasticity.
+To demonstrate this, we simulated a square, flat landscape as above,
+but with \emph{biased migration}: 
+each offspring's location was chosen randomly as before,
+but with mean location $\sigma_d/10$
+up and to the right of the parent's location.
+% (or 0.05\% of the edge length of the landscape)
+This produces reverse-time gene flow down and to the left.
+As a population genetics model,
+this case is similar to a population expansion,
+replacing the leading edge of the expansion with the corner of landscape 
+that migration bias leads away from.
+In this situation, as in an expansion, 
+individuals at the edge may have disproportionately more offspring,
+leading to low genetic diversity and strong drift \citep{surfing}.
+This results in substantially more noise in the data,
+clearly visible in the isolation by distance plots of Figure \ref{fig:ibd_comp}. 
+Despite this large contribution of process noise,
+coalescence time inference correctly infers the bias on a broad scale:
+using both a $2 \times 2$ and a $4 \times 4$ grid,
+inferred gene flow rates down and to the left are consistently much larger
+than those up and to the right 
+(Supplemental Figures \ref{fig:2x2grid_asb}, \ref{fig:grid_bias_4x4_1}, 
+and \ref{fig:post_dists_bias_4x4_1}).
+However, inferred rates show substantial heterogeneity,
+likely due to stochasticity of recent demography.
+
+\begin{figure}
+\centering
+     \includegraphics[width=0.75\textwidth]{figs/ibd_comp}
+    \caption{
+        Genetic distance against geographic distance,
+        for the square landscape with 
+        \textbf{(above)} uniform migration
+        and \textbf{(below)} biased migration.
+        The horizontal lines in the latter plot
+        are likely from comparing genomes of individuals from
+        families that have recently increased in size and geographical spread, 
+        resulting in many pairs of genomes with similar genetic distance 
+        but varying geographic distance.
+    \label{fig:ibd_comp}
+    }
+\end{figure}
+
 \paragraph{Identifying a barrier}
-We now revisit the earlier situation where a landscape has locations that are barriers to gene flow,
-as in Section \ref{sec:5x3b}.
-The landscape is designed to be a continuous version of this situation.
+We now revisit the earlier situation where a landscape has locations that are barriers to gene flow.
+The landscape is a continuous version of this situation,
+shown in Figure \ref{fig:5x3b_continuous},
+where red bars depict uninhabitable regions that block migration.
 Since the distribution of offspring locations is truncated at three standard deviations
-away from the mother, a sufficiently thick uninhabitable area can completely stop migration
+away from the mother, a sufficiently thick uninhabitable area completely stops migration
 directly across it.
-As in \ref{sec:5x3b}, though, the landscape is still relatively well connected.
-The layout of the landscape can be seen in Figure \ref{fig:ind_locs_5x3b_1},
-with the red bars being the uninhabitable regions that serve as barriers to migration.
-Unlike the case of a square landscape with no barriers, 
-there is not a natual way to discretize it more coarsely 
-with grid squares of equal size and shape,
-so a $5 \times 3$ grid will be the only discretization used in this section.
 We calculate mean genetic distances using 50 individuals 
@@
 here giving us a total of 750 individuals.
+The gene flow rates across the barriers are reliably estimated to be very small,
+while other values have substantial variation:
+in particular, gene flow rates are estimated to be low
+across some edges that do not cross barriers;
+this may be because of recent demographic stochasticity.
+The barriers were also identified with comparable accuracy by both our resistance distance method
+and EEMS, as shown in Supplemental Figure \ref{sfig:barrier_results}.
+Note also that this analysis began by grouping samples into discrete ``populations'' appropriately;
+in absence of a good \textit{a priori} idea of where barriers are,
+this step might not be straightfoward.
+% For instance, we may wonder whether a river or other geographical feature
+% restricts gene flow.
+% We would then discretize the landscape such that individuals on opposite sides of the 
+% potential obstruction are in separate grid spaces.
 
@@
      \includegraphics[scale=.8]{figs/ind_locs_5x3b_1}
-    \caption{Locations of all and selected individuals on the $5 \times 3$ landscape with barriers
-    in the first replicate.
-    Locations of individuals used to compute the mean genetic distance matrix are shown in bold points
-    whereas the locations of other individuals are shown in smaller points.
-    The red bars show the locations of uninhabitable regions of the landscape.
-    They are sufficiently thick so that no migration can occur directly across them.}
-    \label{fig:ind_locs_5x3b_1}
-\end{figure}
-
-\begin{figure}
-\centering
      \includegraphics[height=0.3\textheight]{figs/grid_5x3b_1}
@@
         \textbf{(top)}
-    Posterior medians of movement rates for a replicate of the continuous landscape
-    with barriers depicted in Figure \ref{fig:ind_locs_5x3b_1}.
-    \textbf{(bottom)}
-    Boxplots of the same posterior distributions,
-    colored red for gene flow rates of edges that cross the barrier.
-    } \label{fig:post_dists_5x3b_1}
-\end{figure}
-
-The gene flow rates across the barriers are reliably estimated to be very small
-%(\emph{maybe reference all 6 supplemental figures here?}),
-but there is considerable variation in the other values.
-Note, however, that this analysis requires that we guess a reasonable place
-to put the divisions between the locations.
-For instance, we may wonder whether a river or other geographical feature
-restricts gene flow.
-We would then discretize the landscape such that individuals on opposite sides of the 
-potential obstruction are in separate grid spaces.
-
-\paragraph{Continuous Landscape with Biased Migration}
-We also simulated three situations on a continuous landscape that had an effective bias in migration.
-First, the \textbf{biased} situation, was a square landscape as above,
-but with mean offspring location one tenth of a standard deviation up and to the right of the parent's location.
-% (or 0.05\% of the edge length of the landscape)
-This produces reverse-time gene flow down and to the left.
-Second, we simulated a \textbf{valley} in a landscape with a $3 \times 5$ aspect ratio as above, 
-in which fecundity of individuals in the middle 1/4 of the landscape was reduced by 50\%.
-This produces gene flow ``uphill'', away from the ridge,
-and thus an effective barrier across the ridge.
-Third, we simulated a recent \textbf{expansion},
-in which the middle 4/5 of the landscape was uninhabitable 400,000 generations,
-until 500 generations before the present day.
+        Locations of all and selected individuals on the $5 \times 3$ landscape with barriers,
+        with individuals used to compute the mean genetic distance matrix marked in bold.
+        The red bars show the locations of uninhabitable regions of the landscape.
+        They are sufficiently thick so that no migration can occur directly across them.
+        \textbf{(bottom left)}
+        Posterior medians of movement rates inferred using coalescence time inference,
+        and \textbf{(bottom right)}
+        boxplots of corresponding posterior distributions,
+        with movement rates crossing the barriers colored in red.
+        The grouping of individuals into a $5 \times 3$ grid is shown above,
+        and labels are the same as in Figure \ref{fig:5x3b_grid}.
+    \label{fig:5x3b_continuous}
+    }
+\end{figure}
+
+\paragraph{Other barriers: population sinks and recent expansions}
+We simulated two other situations on a continuous landscape that produced an bias in effective migration.
+First, we simulated a \textbf{valley} in a landscape with a $5 \times 3$ aspect ratio as above, 
+in which fecundity of individuals in the middle 1/4 of the landscape was reduced by 50\%
+(a ``source-sink'' landscape).
+Since individuals in the valley are more likely to have parents outside of the valley,
+this produces gene flow ``uphill'', away from the ridge.
+Next, we simulated a recent \textbf{expansion},
+in which the middle 4/5 of the landscape was uninhabitable for 400,000 generations,
+until 500 generations before the present day
+(a model of ``secondary contact'').
 The resulting expansion produces gene flow back away from the center,
@@
 
-This results in reverse time gene flow down and to the left 
-as in the biased asymmetric graph in Section \ref{sec:biased_migration}.
-In terms of population genetics, 
-this is very similar to the case where a population is expanding into new territory 
-since in both cases, 
-offspring tend to be further from the edge of the expansion than their parents.
-The difference is that, in our case, 
-the population is constrained to be a bounded distance 
-away from the edge of the ``expansion''.
-In both situations, individuals at the edge may have disproportionately more offspring,
-leading to low genetic diversity and strong drift.
-This results in substantially more noise in the isolation by distance plots 
-as seen in Figure \ref{fig:ibd_comp}. 
-The horizontal lines in that figure are likely from comparing genomes of individuals from
-two families that have recently increased in size and geographical spread, 
-resulting in many pairs of genomes with similar genetic distance 
-but varying geographic distance.
-
-As in Section \ref{sec:slim_unif}, we discretize the landscape into
-both a $2 \times 2$ grid and a $4 \times 4$ grid for inference.
-Looking at the medians of the inferred posteriors of the gene flow rates in the $2 \times 2$ case
-(Figure \ref{fig:2x2grid_asb}),
-it is immediately clear how strong the bias 
-against reverse time gene flow out of the lower left corner is.
-That is, hardly any individuals in the lower left area have ancestors
-from other regions of the landscape.
-This is to be expected as even though the single generation migration bias is small,
-having a descendant in a location against the bias after a large number of generations
-becomes less and less likely as the number of generations increases.
-In the $4 \times 4$ discretization, 
-the bottom left also shows very strongly biased gene flow (Figure \ref{fig:grid_bias_4x4_1}).
-Most (but not all) of the other pairs also show bias in the expected direction.
-Boxplots of the posterior distributions are shown in Figure \ref{fig:post_dists_bias_4x4_1}.
-%\emph{maybe talk about/cite some biased random walk stuff here?}
-
-%\emph{talk about some kind of log ratio test for the 4x4 case?}
-
-\begin{figure}
-\centering
- \includegraphics[scale=.35]{figs/ibd_comp}
-\caption{Comparison of the isolation by distance plots for the square landscape with uniform migration
-and the square landscape with biased migration.
-Faster coalescence in the biased migration case results in more noticeable horizontal lines,
-indicative of families that recently increased in size and geographical spread.}
-\label{fig:ibd_comp}
-\end{figure}
-
-\begin{figure}
-\centering
- \includegraphics[scale=.5]{figs/2x2grid_asb}
-\caption{Medians of the posterior distributions of the inferred values of $g$ 
-for the first replicate of the $2 \times 2$ discretization of the square landscape with biased migration.
-The reverse time gene flow rates out of the bottom left section are very small,
-indicating that very few individuals living in the bottom left of the landscape
-have ancestors from other regions of the habitat,
-consistent with migration biased up and to the right.}
-\label{fig:2x2grid_asb}
-\end{figure}
-
-\begin{figure}
-\centering
- \includegraphics[scale=.8]{figs/grid_bias_4x4_1}
-\caption{Medians of the posterior distributions of the inferred values of $g$ 
-for the first replicate of the $4 \times 4$ discretization of the square landscape with biased migration.
-As in the $2 \times 2$ case, 
-the reverse time gene flow rates out of the bottom left section are very small,
-indicating that very few individuals living in that area of the landscape
-have ancestors from other regions of the graph.
-Most, but not all of the other gene flow rates are biased 
-in the direction we would expect.}
-\label{fig:grid_bias_4x4_1}
-\end{figure}
-
-\begin{figure}
-\centering
- \includegraphics[scale=.6]{figs/post_dists_bias_4x4_1}
-\caption{Boxplots of the posterior distributions for $g$ 
-for the first replicate of the $4 \times 4$ discretization of the square landscape with biased migration.
-Black boxes are gene flow rates down and to the left 
-with each green box being the corresponding gene flow rate up and to the right.}
-\label{fig:post_dists_bias_4x4_1}
+Both coalescence time inference and the resistance landscape inferred by EEMS
+show a barrier to gene flow in the middle of the landscape (Figure \ref{fig:more_barriers}):
+this is arguably correct, 
+although there is in fact no current barrier to migration in either simulation.
+Coalescence time inference also correctly infers 
+the general bias in recent gene flow away from the center,
+although as in other continuous-space simulations, 
+there was substantial noise in the inferred patterns.
+The resistance landscapes as inferred by EEMS show patchy, vertical bands of high resistance:
+these patterns are commonly seen when attempting to use EEMS, 
+whose underlying model is inherently symmetric, 
+to describe asymmetric gene flow.
+Our analysis above shows that this is an unavoidable consequence of the resistance approximation,
+and our implementation of commute time inference,
+which allows asymmetric gene flow, shows simliar patterns
+(Supplemental Figures \ref{sfig:valley_results} and \ref{sfig:expansion_results}).
+
+\begin{figure}
+\centering
+    \includegraphics[width=0.45\textwidth]{figs/valley_eems}
+    \includegraphics[width=0.45\textwidth]{figs/valley_coal_graph}
+    \includegraphics[width=0.45\textwidth]{figs/expansion_eems}
+    \includegraphics[width=0.45\textwidth]{figs/expansion_coal_graph}
+    \caption{
+        \textbf{(Left:)}
+        resistance landscapes inferred by EEMS
+        and 
+        \textbf{(right:)}
+        gene flow rates inferred using coalescence time inference,
+        from individual-based simulations on a continuous landscape,
+        with \textbf{(top)} a \emph{valley} of reduced fecundity in the middle, and
+        \textbf{(bottom)} a recent \emph{population expansion} from both edges of the range
+        that met in the middle.
+        The color scale shows the log migration rates inferred by EEMS.
+        \label{fig:more_barriers}
+    }
 \end{figure}
@@
 
-With these methods, we provide a framework for
-inferring demographic parameters of 
-populations living on a habitat spread over a two dimensional landscape 
-by using present day location and genetic sequencing information.
+In this paper, we have studied the problem of
+inferring demographic parameters of populations spread over heterogeneous two-dimensional landscapes
+by using genetic and geographic distances between present-day samples.
+In particular, we have shown that the resistance distance approximation
+-- which underlies several of the most commonly-used tools of landscape genetics --
+can produce misleading results, especially in the presence of biased gene flow.
+We implement an alternative method, 
+which uses coalescence times instead of the resistance distance approximation,
+which provides good estimates of gene flow in a wider range of situations.
+This weakness of resistance-based methods is not surprising --
+the original papers describing Circuitscape \citep{mcrae_circuitscape} and EEMS \citep{petkova_eems}
+present the resistance approximation as a necessary step for computational feasibility --
+but perhaps not widely appreciated.
+
+Our new method infers effective population sizes and gene flow rates quite well 
+-- and, being Bayesian, provides estimates of uncertainty --
+given data from discrete populations connected by migration.
+Why do we not advocate our method as an immediate replacement for resistance-based methods?
+The substantial uncertainty we saw on graphs with only tens of nodes
+is indicative of a larger problem we face for realistic models.
+There is a high degree of unavoidable uncertainty,
+and process noise can lead to overfitting of models.
+To produce reproducible, reliable results we need to address these problems.
+
 We are able to do inference of reverse time gene flow and coalescence rates 
@@
 \centering
-    \includegraphics[width=0.45\textwidth]{figs/valley_locations}
-    \includegraphics[width=0.45\textwidth]{figs/valley_eems}
-    \includegraphics[width=0.45\textwidth]{figs/valley_commute_graph}
-    \includegraphics[width=0.45\textwidth]{figs/valley_coal_graph}
-    \includegraphics[width=0.45\textwidth]{figs/valley_commute_dists}
-    \includegraphics[width=0.45\textwidth]{figs/valley_coal_dists}
+ \includegraphics[scale=.5]{figs/2x2grid_asb}
 \caption{
-        Results from the \textbf{valley} simulation:
-        \textbf{(top left)} Positions of all individuals at the end of simulation.
-        \textbf{(top right)} mean migration rates as estimated by EEMS.
-        Below are shown
-        posterior median migration rates 
-            as estimated using 
-            \textbf{(middle left)} the commute time method and
-            \textbf{(middle right)} the the commute time method.
-        Last are shown the posterior distributions, for
-            \textbf{(bottom left)} the commute time method and
-            \textbf{(bottom right)} the the commute time method.
-        \label{sfig:valley_results}
+    Medians of the posterior distributions of the inferred values of $g$ 
+    for the first replicate of the $2 \times 2$ discretization of the square landscape with biased migration.
+    The reverse time gene flow rates out of the bottom left section are very small,
+    indicating that very few individuals living in the bottom left of the landscape
+    have ancestors from other regions of the habitat,
+    consistent with migration biased up and to the right.
+    These inferred rates make it clear how strong the bias 
+    against reverse time gene flow out of the lower left corner is.
+    Hardly any individuals in the lower left area have ancestors
+    from other regions of the landscape.
+    This is to be expected as even though the single generation migration bias is small,
+    having a descendant in a location against the bias after a large number of generations
+    becomes less and less likely as the number of generations increases.
+    \label{fig:2x2grid_asb}
     }
@@
 \centering
-    \includegraphics[width=0.45\textwidth]{figs/expansion_locations}
-    \includegraphics[width=0.45\textwidth]{figs/expansion_eems}
-    \includegraphics[width=0.45\textwidth]{figs/expansion_commute_graph}
-    \includegraphics[width=0.45\textwidth]{figs/expansion_coal_graph}
-    \includegraphics[width=0.45\textwidth]{figs/expansion_commute_dists}
-    \includegraphics[width=0.45\textwidth]{figs/expansion_coal_dists}
+ \includegraphics[scale=.8]{figs/grid_bias_4x4_1}
 \caption{
-        Results from the \textbf{expansion} simulation:
-        \textbf{(top left)} Positions of all individuals at the end of simulation.
-        \textbf{(top right)} mean migration rates as estimated by EEMS.
-        Below are shown
-        posterior median migration rates 
-            as estimated using 
-            \textbf{(middle left)} the commute time method and
-            \textbf{(middle right)} the the commute time method.
-        Last are shown the posterior distributions, for
-            \textbf{(bottom left)} the commute time method and
-            \textbf{(bottom right)} the the commute time method.
-        \label{sfig:expansion_results}
+    Medians of the posterior distributions of the inferred values of $g$ 
+    for the first replicate of the $4 \times 4$ discretization 
+    of the square landscape with biased migration.
+    As in the $2 \times 2$ case, 
+    the reverse time gene flow rates out of the bottom left section are very small,
+    indicating that very few individuals living in that area of the landscape
+    have ancestors from other regions of the graph.
+    Most, but not all of the other gene flow rates are biased 
+    in the direction we would expect.
+    Boxplots of the posterior distributions are shown in 
+    Supplemental Figure \ref{fig:post_dists_bias_4x4_1}.
     }
+\label{fig:grid_bias_4x4_1}
+\end{figure}
+
+\begin{figure}
+\centering
+ \includegraphics[scale=.6]{figs/post_dists_bias_4x4_1}
+\caption{Boxplots of the posterior distributions for $g$ 
+for the first replicate of the $4 \times 4$ discretization of the square landscape with biased migration.
+Black boxes are gene flow rates down and to the left 
+with each green box being the corresponding gene flow rate up and to the right.}
+\label{fig:post_dists_bias_4x4_1}
 \end{figure}
@@
 
+\begin{figure}
+\centering
+    \includegraphics[width=0.45\textwidth]{figs/valley_locations}
+    \includegraphics[width=0.45\textwidth]{figs/valley_eems}
+    \includegraphics[width=0.45\textwidth]{figs/valley_commute_graph}
+    \includegraphics[width=0.45\textwidth]{figs/valley_coal_graph}
+    \includegraphics[width=0.45\textwidth]{figs/valley_commute_dists}
+    \includegraphics[width=0.45\textwidth]{figs/valley_coal_dists}
+    \caption{
+        Results from the \textbf{valley} simulation:
+        \textbf{(top left)} Positions of all individuals at the end of simulation.
+        \textbf{(top right)} mean migration rates as estimated by EEMS.
+        Below are shown
+        posterior median migration rates 
+            as estimated using 
+            \textbf{(middle left)} the commute time method and
+            \textbf{(middle right)} the coalescent time method.
+        Last are shown the posterior distributions, for
+            \textbf{(bottom left)} the commute time method and
+            \textbf{(bottom right)} the coalescent time method.
+        \label{sfig:valley_results}
+    }
+\end{figure}
+
+\begin{figure}
+\centering
+    \includegraphics[width=0.45\textwidth]{figs/expansion_locations}
+    \includegraphics[width=0.45\textwidth]{figs/expansion_eems}
+    \includegraphics[width=0.45\textwidth]{figs/expansion_commute_graph}
+    \includegraphics[width=0.45\textwidth]{figs/expansion_coal_graph}
+    \includegraphics[width=0.45\textwidth]{figs/expansion_commute_dists}
+    \includegraphics[width=0.45\textwidth]{figs/expansion_coal_dists}
+    \caption{
+        Results from the \textbf{expansion} simulation:
+        \textbf{(top left)} Positions of all individuals at the end of simulation.
+        \textbf{(top right)} mean migration rates as estimated by EEMS.
+        Below are shown
+        posterior median migration rates 
+            as estimated using 
+            \textbf{(middle left)} the commute time method and
+            \textbf{(middle right)} the coalescence time method.
+        Last are shown the posterior distributions, for
+            \textbf{(bottom left)} the commute time method and
+            \textbf{(bottom right)} the coalescence time method.
+        \label{sfig:expansion_results}
+    }
+\end{figure}
+
 \end{document}

### 2018-10-16 isolation_by_coalescence.tex
last big edit?

@@
 A number of methods commonly used in landscape genetics
-use analogies to electrical resistance on a network
+use an analogy to electrical resistance on a network
 to describe and fit barriers to movement across the landscape
 using genetic distance data.
-These are motivated by a mathematical equivalence between electrical resistance between two nodes
+These are motivated by a mathematical equivalence between electrical resistance 
+between two nodes of a network
 and the ``commute time'',
@@
 can dramatically mislead in the presence of biased gene flow.
-Forwards-time simulation with continuous geography demonstrates that
+We then use forwards-time simulation with continuous geography to demonstrate that
 coalescence-based inference remains more accurate than resistance-based methods on realistic data,
-and highlights the need for methods that do not approximate continuous space
-by a discrete set of populations.
+but difficulties highlight the need for methods 
+that explicitly model continuous, heterogeneous geography.
 \end{abstract}
@@
 This is a potentially important source of information,
-as direct observation can be difficult \citep[especially of long-distance movement]{dispersal_estimation},
+as direct observation can be difficult 
+(especially of long-distance movement, \citep{cayuelademographic,levin_et_al_seed_dispersal}),
 or even impossible, if some of the populations in question no longer exist.
 
-The idea of ``resistance distance'' is an important tool in the lanscape genomics toolbox.
+The idea of ``resistance distance'' is an important tool in the landscape genomics toolbox.
 Introduced by \citet{mcrae2006isolation},
 it makes use of a mathematical equivalence between random walks and electrical resistance
-\citep{resistance}:
+\citep{nashwilliams1959random}:
 the expected time for a continuous-time Markov chain that starts at node $x$
@@
 is equal to the effective resistance between $x$ and $y$ in an electrical network
-whose conductances are given by the movement rates of the Markov chain \citep{levin}.
+whose conductances are given by the movement rates of the Markov chain \citep{levin2008markov,doyle2006random}.
 This measure therefore averages over all possible paths through the network between the two.
@@
 then this is taken as good evidence that the landscape variable is a good indicator of
-where gene flow occurs \citep{mcrae2007circuit,cushman2006complex,more_examples}.
+where gene flow occurs \citep{mcrae2007circuit,cushman2006complex}.
 For instance,
@@
 \citet{mcrae2006isolation} showed that coalescence time and resistance distance are equal
-(up to shift and scaling) in isotropic landscapes such as a ring \citep{matsen_wakeley},
+(up to shift and scaling) in isotropic landscapes such as a ring \citep{matsen2006convergence},
 and found that the two were highly correlated in several test cases.
@@
 in rivers \citep{morrissey2009maintenance,sundqvist2016directional,hanks2017modeling},
-source-sink dynamics \citep{source_sink},
-or after population expansions \citep{expansions}.
-
-\plr{CHANGE TO mostly RESISTANCE STARTING AT SIMPLE EXAMPLE}
+source-sink dynamics \citep{dias1996sources},
+or after population expansions \citep{with2002landscape}.
 
@@
 coalescence time and resistance distance,
-and develop an inference method for movement rates on a landscape
+and develop an inference method for movement rates on a discrete landscape
 using both as estimated from genetic distances.
@@
 commute time inference may get the wrong answer if data derive from a coalescent process,
-as does real genetic data, especially in the presence of asymmetry.
+especially in the presence of asymmetry.
 We also explore the question of identifiability.
@@
 Some methods find the combination of landscape layers to best explain genetic distance,
-which allows very fine geographic resolution \citep{infer_resistance}.
+which allows very fine geographic resolution \citep{shaffer2017desert}.
 Other methods such as EEMS \citep{petkova2016visualizing} infer conductances without such prior information;
 however, the geographical resolution is much coarser.
+A recently-released method 
+that is based on EEMS but uses haplotype length data and an underlying model of coalescence time
+\citep{alasadi2018estimating}
+is promising, but still produces similarly coarse maps.
 One reason for this difference in resolution
 is that EEMS uses more computationally intensive Bayesian methods.
-However, we show that this is a more general problem --
-inference of movement rates, in either a coalescence or commute time framework,
-is an ill-conditioned problem,
-implying the need for regularization to obtain reliable inference.
-This explains the results of \citet{graves2013current},
-who found a large area of resistance parameters
-that produced equivalently good fits to genetic distance
-(i.e., nonidentifiability, analogous to a flat likelihood surface).
-
-Talk about MAPS.
-
-% \plr{I didn't find myself wanting to explain isolation by distance here:
-%     I don't think we need to? Do you agree?}
+However, we show that this is a more general problem,
+both because of unmodeled ``process noise'' involved in discretization of geographical populations,
+and the mathematical structure of the underlying problem.
 
@@
 
-There are a number of ways to calculate measures of dissimilarity using genetic data \citep{distances}.
+There are a number of ways to calculate measures of dissimilarity using genetic data.
 In this paper we use mean pairwise divergence:
@@
 It is more common in the literature to use $F_{ST}/(1-F_{ST})$,
-but the two are highly correlated for good theoretical reasons \citet{slatkin_fst,rousset1997genetic},
+but the two are highly correlated for good theoretical reasons 
+\citet{slatkin1991inbreeding,rousset1997genetic},
 especially for SNP data.
-The genomes differ at a site only if there has been a mutation in some ancestor on the lineages leading
+The two genomes differ at a site only if there has been a mutation in some ancestor on the lineages leading
 from the two genomes back to their most recent common ancestor at that site;
 under an infinite-sites model with average mutation rate $\mu$ per site and per meiosis,
-pairwise divergence is therefore an unbiased estimate of the mean time to most recent common ancestor,
-averaged across the genome \citep{hudson2007variance,ralphXX}.
+pairwise divergence divided by $\mu$
+is therefore an unbiased estimate of the mean time to most recent common ancestor,
+averaged across the genome \citep{hudson2007variance,ralph2015empirical}.
 It is natural to model this process by following the two lineages back through time
@@
 i.e., until they coalesce.
-It is common to model the geographic distribution of a species
+Suppose we model the geographic distribution of the species
 as a collection of randomly mating populations
@@
 thus discretizing geographic space.
-If the genome is selectively netural,
+If the genome is selectively neutral,
 then the motion of these lineages back through time
@@
 This means that if the Markov chain is at location $X_t$ at time $t$,
-then for $x \neq y$, the probability the chain jumps from $x$ to $y$ is proportional to $G_{xy}$,
+then for $x \neq y$, the probability the chain jumps from $x$ to $y$ in the next unit of time
+is proportional to $G_{xy}$,
 i.e., $\PP{X_{t+\epsilon} = y \given X_t = y} = \epsilon G_{xy} + O(\epsilon^2)$,
@@
 but it is related.
-The Random Target Lemma \citep{aldous}
+The Random Target Lemma \citep{aldous-fill-2014}
 tells us that the stationary distribution of the Markov chain, denoted $\pi$,
@@
 The \emph{commute times} are a symmetrized version of the hitting times
-(which need not be symmetric):
+(which need not be symmetric),
+defining $R = H + H^T$, i.e.:
 \begin{align} \label{eqn:R_from_H}
 \text{\bf (commute time)} \qquad
-    R_{xy} = H_{xy} + H_{yx},
+    R_{xy} = H_{xy} + H_{yx}.
 \end{align}
-or $R = H + H^T$.
-We use $R$ to refer to the commute time as a reminder that commute times are also effetive resistances,
-and to reserve $C$ for the coalescence time.
 
@@
 The resulting hitting times may not be valid --
-if we let $G_\epsilon$ be the matrix constructed by applying equation \eqref{eqn:G_from_H} to $H_\epsilon$
-(assuming this is invertible),
+if we let $G_\epsilon$ be the matrix constructed by applying equation \eqref{eqn:G_from_H} to $H_\epsilon$,
+% (assuming this is invertible),
 then it is not guaranteed that the offdiagonal elements of $G_\epsilon$ are all nonnegative, as required.
 (It is always true that rows of $G_\epsilon$ sum to zero.)
+We provide a small concrete example in Appendix \ref{ss:simple_example}.
 However, since $G$ and $H$ are continuous functions of each other,
@@
 and so after rearrangement can be solved with standard linear algebra
-(and is related to the Sylvester equation \citep{sylvester_eqn}).
+(and is related to the Sylvester equation \citep{bhatia1997solve}).
 However, the solution is again not necessarily unique:
-suppose that we have a matrix $Z$ such that $ZC$ is antisymmetric.
+suppose that we have a matrix $Z$ such that $ZC$ is skew-symmetric.
 Then $(G + Z) C + C (G + Z)^T = GC + CG^T$,
@@
 As before, if all entries of $G$ are nonzero, 
-it is always possible to find sufficiently small $Z$
-so that $G + Z$ remains the generator of a valid Markov chain,
-but this is not guaranteed in general.
+it is always possible to find sufficiently small $\epsilon$
+so that $G + \epsilon Z$ remains the generator of a valid Markov chain.
 
@@
 so that many of the entries of $G$ can be assumed to be zero.
+This reduces the number of unknowns to solve for,
+and so can render the solution unique.
 We can get an idea of this by simply counting equations and unknowns.
@@
 Since movement rates in each direction can be different,
-there are $4n$ free parameters, each corresponding to an offdiagonal entry of $G$.
+there are $4n$ free parameters, each corresponding to an off-diagonal entry of $G$.
 Coalescence rates provide another $n$ parameters.
@@
 Since commute times are symmetric, and $R_{zz} = 0$, 
-equation \eqref{eqn:R_from_H} only provides $n (n-1)/2$ informative equations.
+equation \eqref{eqn:R_from_H} provides $n (n-1)/2$ informative equations.
 This is larger than $4n$, the number of parameters,
-for a grid with at least nine nodes (i.e., $3 \times 3$).
+for a grid with at least nine nodes (i.e., $3 \times 3$),
+so the system of equations will usually have a unique solution
+for grids larger than this, except in special cases.
 The same calculation holds for commute times, 
@@
 may produce large differences in the inferred rates.
-Very roughly speaking, this occurs because as the spatial resolution increases,
-the matrix $G$ converges to a second-order elliptic differential operator \citep{pde};
-such operators are deformations of the Laplacian \citep{laplacian},
-which is well-known to have eigenvalues that decay exponentially \citep{laplacian_eigenvals}.
+Roughly speaking, this occurs because as the spatial resolution increases,
+the matrix $G$ converges to a second-order elliptic differential operator \citep{stroock1997multidimensional};
+such operators are deformations of the Laplacian \citep{feller4},
+which is well-known to have eigenvalues that decay rapidly 
+\citep{hpmckean1967curvature,kuttler1984eigenvalues}.
 Information about the landscape contained in higher-order modes
@@
 making the inverse problem possible in theory but not in practice.
-For more discussion of related problems, see \citet{badtruth,shape_of_a_pop,inverse_prob}.
+For more discussion of related problems, see 
+\citet{epstein2008badtruth,myers2008learn,terhorst2015fundamental}.
 
 \paragraph{When are coalescence and commute times equal?}
-If in a given biological situation the two times were equal --
+If in a given biological situation commute and coalescence times are equal --
 i.e., if given $G$ and $\gamma$, there was a $q$ that made $\comdist = C$ --
 then inference with either model would be equivalent.
+This was demonstrated in some situations by \citet{mcrae2006isolation}.
 We have already shown that this is not the case in general,
 but it turns out that it \emph{is} true
-under some restrictive assumptions not uncommonly found in population genetics.
+under some restrictive assumptions not uncommonly found in abstract population genetics models.
 We show in Appendix \ref{sec:com_eq_coal} that
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Bayesian inference of movement rates}
+
+Mean coalescence times estimated from real data are subject to a number of sources of noise
+(which we discuss in more detail later).
+Here, we use
+the exact solutions based on linear algebra developed above
+to develop a Bayesian inference method that accounts for noise in the data
+and estimates uncertainty in the resulting estimates.
+To do this, we model genetic distances as equal to coalescence times plus noise:
+$D_{ij} = C_{ij} + \epsilon_{ij}$,
+where $\epsilon_{ij}$ are independent and Normally distributed
+with mean $\mathbf{0}$ and variance $\sigma_\epsilon^2$,
+except that $\epsilon_{ij} = \epsilon_{ji}$.
+(We imagine that $C_{ij}$ is large compared to $\sigma_{\epsilon}$,
+so that resulting values are still positive.)
+Using the calculations above, 
+given $G$ and $\gamma$ we can compute the corresponding coalescence times
+-- which we call $\mathcal{C}(G, \gamma)$ --
+so our model is that $D$ has a Gaussian distribution 
+with mean $\mathcal{C}(G, \gamma)$.
+The maximum likelihood estimate of $G$ could be found directly;
+however, this in practice will quite likely contain negative movement rates,
+and does not allow us to impose constraints
+(such as only allowing a subset of movement rates to be nonzero).
+
+Therefore, we use Bayesian inference,
+placing independent exponential priors on the parameters --
+the movement rates, $G_{ij}$, that are not constrained to be zero
+and the coalescence rates, $\gamma$.
+Concretely, suppose that the spatial arrangement of populations is given by a graph,
+and only movement rates corresponding to edges in the graph are allowed to be nonzero.
+Write $i \sim j$ to mean that $i$ and $j$ are adjacent in the graph,
+that there are $m_G$ edges in the graph,
+and that each of the $n$ populations has its own coalescence rate.
+Then, the resulting log-posterior density is
+\begin{align} \label{eq:post}
+    \begin{split}
+\mathcal{L}(D, G, \gamma) 
+    &=
+    2 m_G \log(\lambda_G) + n \log(\lambda_\gamma) 
+    - \frac{n(n+1)}{2} \log(2 \pi \sigma^2_\epsilon)
+	- \sum_{i \sim j} \lambda_G G_{ij}  \\
+    &\qquad
+    -\sum_{i=1}^n \lambda_{\gamma}\gamma_i
+		-\frac{1}{2 \sigma_{\epsilon}^2} \sum_{i \leq j} \left(
+            \mathcal{C}_{ij}(G,\gamma) - D_{ij}
+        \right)^2 
+    \end{split}
+\end{align}
+Here $1/\lambda_\gamma$ and $1/\lambda_G$ are the prior means of $\gamma_i$ and $G_{ij}$,
+respectively.
+
+Inference with resistance distance is done in exactly the same way,
+except that $q$ replaces $\gamma$,
+and instead of $\mathcal{C}(G,\gamma)$ we use $\comdistfn(G, q)$,
+which is the matrix $\comdist$ computed from $G$ and $q$ by solving equation \eqref{eq:commute_approx}.
+
+It is not required in this approach to have samples from every spatial location,
+allowing inference on incompletely sampled spatial discretizations.
+To carry out inference in this situation,
+we simply use Equation \ref{eq:post},
+but only sum over observed entries of $D_{ij}$.
+
+\paragraph{MCMC methods}
+To sample from the posterior distribution of equation \eqref{eq:post}, 
+we used R \citep{Rmanual} 
+to implement a standard Metropolis-Hastings algorithm with Gaussian proposals \citep{brooks2011handbook}.
+Starting locations were chosen by sampling from the prior distribution.
+Before the usual ``burn-in'' phase before sampling for inference begins,
+we included a period of ``pre-burn-in'' which used the same MCMC procedure
+with a larger $\sigma_\epsilon$, to allow the chain to more quickly converge on the high-posterior
+portion of parameter space.
+Typically, we ran MCMC for $10^6$ iterations of pre-burn-in 
+and $3 \times 10^6$ iterations of burn-in,
+followed by $4 \times 10^6$ iterations which were used to estimate posterior distributions,
+although for small graphs fewer iterations were necessary.
+We used standard methods to assess convergence and mixing.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Test cases and simulations}
+
+We compared inference under coalescence and commute time models
+using both data ``from the model'' and more realistic simulations.
+For the first category, we designed a number of graphs to test particular aspects of inference,
+with migration rates specified on each directed edge
+(plotted below using igraph \citep{igraph}).
+Given the edge weights ($G$) and the coalescence rates ($\gamma$) of a graph,
+we produced data by calculating exact coalescence times ($\mathcal{C}(G,\gamma)$),
+to whose entries we added independent Gaussian noise while preserving symmetry,
+as described above.
+Particular parameters and levels of noise are given in the Results.
+
+These simulations are expected to be nearly equivalent to neutral, discrete population-based simulations
+(either forwards-time or coalescent);
+the only difference being that noise terms about the theoretical mean should be somewhat correlated.
+
+\paragraph{Forwards-time simulations}
+To produce realistic data, we implemented forwards-time simulations using SLiM v3.1 
+\citep{haller2018forward,haller2018treesequence},
+with individuals living across continuous, two-dimensional geography (sometimes with barriers),
+from which we recorded genomic data.
+The basic simulation, which we modified to produce several other situations,
+is as follows.
+Each simulation had around $10^4$ diploid, hermaphroditic individuals, 
+living across two-dimensional geography.
+Mates were selected from nearby individuals using a Gaussian kernel with standard deviation $\sigma_d$,
+truncated at $3 \sigma_d$.
+Simulations were done on either a square region with width equal to $80\sigma_d$,
+or a $5:3$ rectangular region with the same area.
+Offspring are dispersed away from the mother using the same kernel,
+reflected from the boundaries of the habitable area.
+In each simulation, $\sigma_d$ was chosen 
+so that there were an average of 0.25 individuals per $\sigma_d^2$ units of area.
+In order to prevent excessive population clumpiness \citep{felsensten1975pain}, 
+we also model local competition. 
+To do this, we use the same truncated Gaussian kernel to define an ``interaction strength''
+between each pair of nearby individuals;
+then, the fitness of each individual is set to $\exp(-x/40)$, 
+where $x$ is the sum of their interaction strengths with other individuals.
+This reduces fitness for individuals in denser areas.
+Individuals are chosen to reproduce proportional to their fitness.
+Clumpiness is similar to what is seen in 
+Figure \ref{fig:5x3b_continuous}.
+
+Individuals have 10 pairs of ``chromosomes'' with $2 \times 10^5$ loci each 
+and a recombination rate of $10^{-8}$ per generation per locus.
+Neutral mutations were added at a rate of $10^{-9}$ per meiosis.
+To make these simulations computationally feasible,
+the mutation rate was set to 0.0 during the forwards simulation,
+while tree sequences were recorded \citep{haller2018treesequence},
+and mutations were added after the fact with msprime,
+which is equivalent to including mutations during the forwards simulation 
+but resulted in much faster run times \citep{kelleher2018efficient}.
+
+To compute genetic distances,
+geographic space was partitioned into square regions,
+and within each region, 
+individuals are sampled uniformly for ``genotyping'' 
+from among those within the middle three quarters of each dimension of the square,
+as shown for one situation in Figure \ref{fig:5x3b_continuous}.
+This protocol was chosen as a compromise between sampling all individuals 
+from near the center of each location,
+which could result in a large number of close relatives being sampled, 
+leading to an underestimate of the mean genetic distance for individuals in that location,
+and uniform sampling from the whole area of each grid, 
+which may result in many sampled individuals being very close to other grid squares.
+Mean genetic distances for each pair of regions are then computed as described above.
+To estimate uncertainty, we computed the standard error of these mean distances
+using the minimum number of individuals in each location.
+
+Before being passed to the MCMC inference method,
+we rescaled genetic distances so that the movement rates resulting from the inference
+would be approximately of order one.
+We did this by multiplying genetic distances by a constant
+so that the rescaled mean of the entries of $D$
+is equal to the number of locations in the discretization.
+
+\subsection*{Code availability}
+
+All code used for this work is available at \url{https://github.com/elundgre/gene-flow-inference}
+(including an R package that implements the MCMC method)
+and \url{https://github.com/petrelharp/isolation_by_coalescence} (including the SLiM simulations).
+
+%%%%%%%%%%%%%%%%%%
+\section*{Results}
+%%%%%%%%%%%%%%%%%%
+
+\plr{transition paragraph}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Model validation}
+
+We first tested the method using data generated under the model.
+To test the methods across a wide range of situations,
+we generated fifty sets of random movement rates on $3 \times 2$ rectangular grids:
+twenty-five symmetric (i.e., with $G_{ij} = G_{ji}$) and twenty-five asymmetric.
+We generated movement parameters
+by rounding independent exponential random variables up to the nearest one tenth,
+and fixed coalescence rate at 1 for all locations.
+We then added independent noise to each matrix entry (preserving symmetry),
+and assume that we can estimate the amount of noise reasonably well, 
+so used the standard deviation of the noise for $\sigma_\epsilon$ in the likelihood function.
+To assess the impact on inference of the number of parameters,
+we performed inference both with and without the assumption 
+that coalescence rate is the same everywhere.
+For $3 \times 2$ graphs, there are 21 equations
+and 14 movement parameters,
+so that with a global coalescence rate there are 15 unknowns
+but with individual coalescence rates there are 20 unknowns,
+which is close to nonidentifiable.
+To quantify accuracy across these random graphs in different situations,
+we computed, for each model fit on a particular random graph,
+the absolute difference between the posterior median of each movement parameter
+and the true value.
+We then averaged these across movement parameters to produce a measure of accuracy for that model fit,
+reported below as ``mean absolute error''.
+
+\paragraph{Varying noise}
+Unsurprisingly, coalescence time inference became less accurate as the amount of noise increased.
+As we varied the standard deviation of the noise
+from 1/1000 to 1/50 of the mean value of $C$ for that grid,
+the mean absolute error 
+for coalescence time inference with a single coalescence rate
+increased from around 0.05 to 0.4.
+Posterior interquartile ranges increased proportionally.
+(although there was substantial variation in accuracy between replicates;
+see Figures \ref{fig:mult_noise} and \ref{fig:mult_noise_iqr}).
+
+Increasing the number of parameters by allowing multiple coalescence rates
+made the inference problem even harder --
+mean absolute errors no longer depended strongly on the amount of noise added,
+and hovered around 0.4.
+
+More surprisingly, the performance of commute time inference did not depend on the amount of noise --
+mean absolute error was around 0.6 across all levels of noise,
+both with and without more than one parameter for local diversity
+(see Figure \ref{fig:mult_noise}).
+Although symmetry did not affect coalescence time inference,
+commute time inference did substantially worse on asymmetric graphs,
+showing median absolute errors of around 0.8.
+
+\paragraph{Varying coalescence rates}
+Lower coalescence rates mean that lineages wander about the landscape for longer
+before they coalesce, 
+thus greatly reducing the geographic information we get from coalescence times.
+Indeed, coalescence rates had a strong effect on feasability of the inference problem
+(shown in Figures \ref{fig:mult_gam} and \ref{fig:mult_gam_iqr}):
+lowering the coalescence rate from 10 to 0.1
+(always adding noise with standard deviation 1/200th of the mean value)
+increased mean absolute error of coalescence time inference from around 0.05 to 0.5.
+Again, allowing multiple coalescence rates
+or using commute time inference were highly inaccurate in all cases,
+with mean absolute errors of above 0.5.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Identifying a barrier to gene flow}
+\label{sec:5x3b}
+
+We then test the method's ability to locate barriers to migration
+on the $5 \times 3$ grid with asymmetric gene flow shown in Figure \ref{fig:5x3b_grid}. 
+Each nonzero movement rate is determined randomly as before,
+and coalescence rates are set to 1 everywhere.
+Figure \ref{fig:5x3b_post_coalvcom} shows posterior distributions of the movement rates
+for both coalescence time and resistance distance inference
+on this graph, with noise standard deviation equal to
+1/1000 of the mean coalescence time (left)
+and 1/100 of the mean coalescence time (right).
+Although increased noise in the data increased uncertainty in the estimated values,
+coalescence time inference correctly inferred not only the location of the barrier
+but also the migration rates along all other edges within a reasonable margin of error.
+Commute time inference also identified the barrier,
+but was much less accurate for the other values.
+We also explored making the problem harder,
+by first (a) dropping data corresponding to populations 2, 5, 7, and 15;
+and then (b) allowing a separate coalescence rate for each location,
+in both cases using a noise level of 1/100.
+The resulting posterior distributions are shown in Figure \ref{fig:5x3boxplots_mult_gam},
+and show substantially increased error,
+but only in movement rates to the removed locations.
+As before, uncertainty was increased when multiple coalescence rates were allowed,
+but not as strongly as before, likely because the problem is less close to being underconstrained,
+having 105 equations and 59 unknowns.
+
+\begin{figure}
+\centering
+     \includegraphics[width=0.6\textwidth]{figs/5x3b_grid}
+    \caption{The grid structure and parameter labels
+    for the 5 $\times$ 3 grid with two barriers:
+    movement rates between $\{2,3\}$ and $\{7,8\}$ are zero,
+    as are those between $\{8,9\}$ and $\{13,14\}$.
+    Values for the non-zero movement rates 
+    were chosen by rounding independent draws of an exponential random variable with mean 1 
+    up to the nearest tenth.
+    } \label{fig:5x3b_grid}
+\end{figure}
+
+\begin{figure}
+\centering
+     \includegraphics[scale=1]{figs/5x3b_post_coalvcom}
+    \caption{
+        Coalescence time inference is much more accurate than resistance inference:
+        posterior distributions for gene flow rates in the 5 $\times$ 3 graph with barriers,
+        compared to true values (in green),
+        using \textbf{(top)} coalescence time and \textbf{(bottom)} resistance distance inference,
+        with \textbf{(left)} low noise and \textbf{(right)} high noise,
+        See Figure \ref{fig:5x3b_grid} to match parameter labels to positions on the graph.
+    \label{fig:5x3b_post_coalvcom}
+}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Biased migration}
+\label{sec:biased_migration}
+
+Both matrices of coalescence and commute times, $C$ and $R$, are symmetric,
+but they do not deal with asymmetric (biased) migration in the same way.
+Since commute time to location $i$ requires paths both to and from $i$,
+but coalescence time to $i$ only requires lineages to leave $i$,
+if there is a low rate of migration back into $i$,
+then commute times to $i$ could be quite long even while coalescence times are not.
+For instance, suppose there are three populations arranged in a line,
+and lineages currently in the outer populations are much more likely to come from the central population
+than the other way around.
+Since lineages quickly move to the center and coalesce regardless of starting position,
+coalescence time (and genetic diversity) will be relatively low between all individuals.
+Commute time between the two outer locations, on the other hand,
+will be much longer than other comparisons, since it requires a lineage to leave the center.
+
+In order to investigate this general situation,
+we tested both methods on four different $4 \times 4$ graphs (shown in Figure \ref{fig:4x4_grids}):
+\emph{uniform}, where all movement rates are 1.0;
+\emph{symmetric}, where movement rates are symmetric and randomly generated as before;
+\emph{asymmetric}, where all movement rates are random;
+and \emph{biased}, where movement rates down or to the left are equal to 2.0,
+and movement rates up or to the right are 0.5.
+For inference, we added noise with standard deviation $1/500^\text{th}$ of the mean value.
+
+Differences between true mean coalescence times ($C$) and mean resistance distances ($\comdist$) 
+are shown in Figure \ref{fig:RvsC}:
+they are fairly small in the uniform and symmetric cases, moderate in the asymmetric case, 
+and extreme in the biased asymmetric case.
+This difference suggests that commute time inference may be strongly misled 
+in the asymmetric and biased situations,
+but in principle it still could be the case that the movement rates that give the best fit of commute time
+to coalescence time data could be close to the actual rates use to generate the data.
+This does not seem to occur in practice:
+coalescence time inference was substantially more accurate in all cases.
+Posterior distributions of movement parameters are shown in Figure \ref{fig:4x4box},
+produced by applying both inference methods with a noise level of 1/200.
+The mean absolute errors for coalescence time inference are 0.12 or less in all cases.
+Commute time inference obtains roughly uniform migration rates
+(although varying by a factor of 2) in the uniform case;
+and migration rates noisily correlated with the truth in the symmetric graph.
+However, commute time inferences for the asymmetric and biased graphs are only weakly correlated, 
+if at all, to the truth.
+Commute times also drastically overestimate movement rates for the biased case,
+as we would expect based on the differences discussed above.
+
+\begin{figure}
+\centering
+     \includegraphics[scale=1]{figs/RvsC}
+    \caption{
+        Resistance distances ($\comdist$) compared to coalescence times ($C$)
+        on the $4 \times 4$ graph, with 
+        \textbf{(top left)} uniform movement rates,
+        \textbf{(bottom left)} random, symmetric rates,
+        \textbf{(top right)} random, asymmetric rates,
+        and \textbf{(bottom right)} rates with a small diagonal bias (see text).
+        The line shows $y=x$.
+        See Figure \ref{fig:4x4_grids} for specific values and graph structure.
+    \label{fig:RvsC}
+}
+\end{figure}
+
+\begin{figure}
+\centering
+     \includegraphics[scale=1]{figs/4x4boxplots_paper}
+    \caption{
+        Boxplots of the posterior distributions of movement rates
+        inferred using the
+        \textbf{(left)} resistance distances
+        and \textbf{(right)} coalescence times
+        shown in Figure \ref{fig:RvsC};
+        i.e., on the $4 \times 4$ graphs with
+        \textbf{(top)} uniform,
+        \textbf{(middle top)} symmetric,
+        \textbf{(middle bottom)} asymmetric,
+        and \textbf{(bottom)} biased
+        movement rates.
+        See Figure \ref{fig:4x4_grids} for specific values and graph structure.
+    \label{fig:4x4box}
+}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Continuous geographical space}
+
+The data we have used thus far are produced ``under the model'',
+and should provide an accurate depiction of our method applied to discrete, randomly mating populations
+whose connections by migration are known.
+However, this is nearly always a rough approximation to reality,
+in which organisms are distributed across continuous geography,
+and ``populations'' are constructed by necessity, often driven by sampling locations.
+For this reason, we applied coalescence and resistance inference,
+as well as the popular resistance-based method EEMS \citep{petkova2016visualizing},
+to data from simulations on continuous geography.
+
+\paragraph{Process noise}
+The mean coalescence times of lineages that we compute analytically
+are exact for large, randomly mating populations,
+but the word ``mean'' indicates an average across several levels of randomness:
+each observed, empirical mean estimated using a sample of genotypes
+will deviate from this depending on both the individuals sampled (``sampling noise'')
+and the stochasticity of the population history itself (``process noise'').
+For more discussion of this latter contribution in other contexts, see
+\citet{wakeley2012genealogies}, \citet{waples2009modelling}, or \citet{ralph2015empirical}.
+Both sources of noise can be substantially larger 
+in populations living across continuous geography
+than in discrete populations
+due to randomness in spatial locations.
+This is hinted at by Figure \ref{fig:5x3b_continuous},
+where we see that the distribution of individuals across the landscape is fairly patchy:
+a barrier might be inferred in a region that has not happened to have a local population recently
+only because of demographic stochasticity.
+This might make inference using a discretization of continuous space considerably more difficult.
+
+To quantify the relative contributions of the different sources of noise,
+we compared mean genetic distances, calculated in the same way,
+between nonoverlapping sets of samples from the same simulation,
+and between different simulations.
+These simulations were done using SLiM as described in the Methods.
+We found that process noise was much larger than sampling noise --
+as shown in Figure \ref{fig:h_val_comp},
+discrepancies between simulations are about five times that seen
+between different samples from the same simulation.
+An implication of this is that uncertainties in genetic distance
+estimated from the data will be (possibly large) underestimates,
+a fact also pointed out for effective population size by \citet{waples2009modelling}.
+
+\begin{figure}
+\centering
+    \includegraphics[width=0.8\textwidth]{figs/h_val_comp}
+\caption{
+    Comparison of mean coalescence times between samples and between replicates
+    for a $4 \times 4$ discretization of a square landscape with uniform migration.
+    All values are plotted against 
+    mean coalescence times calculated using the first set of individuals from the first simulation.
+    Circles show values calculated from a second (non-overlapping) set of individuals 
+    from the same simulation:
+    without sampling noise, all would fall on the $y=x$ line.
+    The other symbols are calculated from two independent simulations:
+    without process noise, these should deviate from $y=x$ no more than the circles.
+    The mean percent absolute differences between each set of coalescence times
+    and the reference values are
+    1.8\% for the second set of individuals from the first simulation,
+    13.2\% for the second simulation, 
+    and 7.3\% for the third simulation.
+    } \label{fig:h_val_comp}
+\end{figure}
+
+Although we found substantial process noise
+with a large, flat landscape with uniform, unbiased migration,
+many biological processes are expected to increase between-replicate stochasticity.
+To demonstrate this, we simulated a square, flat landscape as above,
+but with \emph{biased migration}: 
+each offspring's location was chosen randomly as before,
+but with mean location $\sigma_d/10$
+up and to the right of the parent's location.
+% (or 0.05\% of the edge length of the landscape)
+This produces reverse-time gene flow down and to the left.
+As a population genetics model,
+this case is similar to a population expansion,
+replacing the leading edge of the expansion with the corner of landscape 
+that migration bias leads away from.
+In this situation, as in an expansion, 
+individuals at the edge may have disproportionately more offspring,
+leading to low genetic diversity and strong drift \citep{edmonds2004mutations,hallatschek2010life}.
+This results in substantially more noise in the data,
+clearly visible in the isolation by distance plots of Figure \ref{fig:ibd_comp}. 
+Despite this large contribution of process noise,
+coalescence time inference correctly infers the bias on a broad scale:
+using both a $2 \times 2$ and a $4 \times 4$ grid,
+inferred gene flow rates down and to the left are consistently much larger
+than those up and to the right 
+(Supplemental Figures \ref{fig:2x2grid_asb}, \ref{fig:grid_bias_4x4_1}, 
+and \ref{fig:post_dists_bias_4x4_1}).
+However, inferred rates show substantial heterogeneity,
+likely due to stochasticity of recent demography.
+
+\begin{figure}
+\centering
+     \includegraphics[width=0.75\textwidth]{figs/ibd_comp}
+    \caption{
+        Genetic distance against geographic distance
+        between all pairs of sampled individuals
+        on a square landscape with 
+        \textbf{(above)} uniform migration
+        and \textbf{(below)} biased migration.
+        The horizontal lines in the latter plot
+        are likely from comparing genomes of individuals from
+        families that have recently increased in size and geographical spread, 
+        resulting in many pairs of genomes with similar genetic distance 
+        but varying geographic distance.
+    \label{fig:ibd_comp}
+    }
+\end{figure}
+
+\paragraph{Identifying a barrier}
+We now revisit the earlier situation where a landscape has locations that are barriers to gene flow.
+The landscape is shown in Figure \ref{fig:5x3b_continuous},
+where red bars depict uninhabitable regions that block migration.
+Since the distribution of offspring locations is truncated at three standard deviations
+away from the mother, a sufficiently thick uninhabitable area completely stops migration
+directly across it.
+We calculate mean genetic distances using 50 individuals 
+from each grid location,
+here giving us a total of 750 individuals.
+The gene flow rates across the barriers are reliably estimated to be very small,
+while other values have substantial variation:
+in particular, gene flow rates are estimated to be low
+across some edges that do not cross barriers;
+this may be because of recent demographic stochasticity.
+The barriers were also identified with comparable accuracy by both our resistance distance method
+and EEMS, as shown in Supplemental Figure \ref{sfig:barrier_results}.
+Note also that this analysis began by grouping samples into discrete ``populations'' appropriately;
+in absence of a good \textit{a priori} idea of where barriers are,
+this step might not be straightforward.
+% For instance, we may wonder whether a river or other geographical feature
+% restricts gene flow.
+% We would then discretize the landscape such that individuals on opposite sides of the 
+% potential obstruction are in separate grid spaces.
+
+\begin{figure}
+\centering
+     \includegraphics[scale=.8]{figs/ind_locs_5x3b_1}
+     \includegraphics[height=0.3\textheight]{figs/grid_5x3b_1}
+     \includegraphics[height=0.3\textheight]{figs/post_dists_5x3b_1_new}
+    \caption{
+        \textbf{(top)}
+        Locations of all and selected individuals on the $5 \times 3$ landscape with barriers,
+        with individuals used to compute the mean genetic distance matrix marked in bold.
+        The red bars show the locations of uninhabitable regions of the landscape,
+        which are sufficiently thick so that no migration can occur directly across them.
+        \textbf{(bottom left)}
+        Posterior medians of movement rates inferred using coalescence time inference,
+        and \textbf{(bottom right)}
+        boxplots of corresponding posterior distributions,
+        with movement rates crossing the barriers colored in red.
+        The grouping of individuals into a $5 \times 3$ grid is shown above,
+        and labels are the same as in Figure \ref{fig:5x3b_grid}.
+    \label{fig:5x3b_continuous}
+    }
+\end{figure}
+
+\paragraph{Other barriers: population sinks and recent expansions}
+We simulated two other situations on a continuous landscape that produced an bias in effective migration.
+First, we simulated a \textbf{valley} in a landscape with a $5 \times 3$ aspect ratio as above, 
+in which fecundity of individuals in the middle 1/4 of the landscape was reduced by 50\%.
+The valley is a population ``sink'',
+although it would not be detectable with mortality rates,
+and shows only a slightly lower population density (Supplemental Figure \ref{sfig:valley_results}).
+Since individuals in the valley are more likely to have parents outside of the valley,
+this produces gene flow ``uphill'', away from the ridge.
+Next, we simulated a recent \textbf{expansion},
+in which the middle 4/5 of the landscape was uninhabitable for 400,000 generations,
+until 500 generations before the present day
+(a model of ``secondary contact'').
+The resulting expansion produces gene flow back away from the center,
+similar to the ``valley''.
+
+Both coalescence time inference and the resistance landscape inferred by EEMS
+show a barrier to gene flow in the middle of the landscape (Figure \ref{fig:more_barriers}).
+This is arguably correct, 
+although there is in fact no current barrier to migration in either simulation.
+Coalescence time inference also correctly infers 
+the general bias in recent gene flow away from the center,
+although as in other continuous-space simulations, 
+there was substantial noise in the inferred patterns.
+The resistance landscapes as inferred by EEMS show patchy, vertical bands of high resistance:
+these patterns are commonly seen when attempting to use EEMS, 
+whose underlying model is inherently symmetric, 
+to describe asymmetric gene flow.
+Our analysis above shows that this is an unavoidable consequence of the resistance approximation,
+and our implementation of commute time inference,
+which allows asymmetric gene flow, shows similar patterns
+(Supplemental Figures \ref{sfig:valley_results} and \ref{sfig:expansion_results}).
+
+\begin{figure}
+\centering
+    \includegraphics[width=0.45\textwidth]{figs/valley_eems}
+    \includegraphics[width=0.45\textwidth]{figs/valley_coal_graph}
+    \includegraphics[width=0.45\textwidth]{figs/expansion_eems}
+    \includegraphics[width=0.45\textwidth]{figs/expansion_coal_graph}
+    \caption{
+        \textbf{(Left:)}
+        resistance landscapes inferred by EEMS
+        and 
+        \textbf{(right:)}
+        gene flow rates inferred using coalescence time inference,
+        from individual-based simulations on a continuous landscape,
+        with \textbf{(top)} a \emph{valley} of reduced fecundity in the middle, and
+        \textbf{(bottom)} a recent \emph{population expansion} from both edges of the range
+        that met in the middle.
+        The color scale shows the posterior mean log migration rates inferred by EEMS.
+        \label{fig:more_barriers}
+    }
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%
+\section*{Discussion}
+
+In this paper, we study how to 
+use genetic and geographic distances between present-day samples
+to infer the demographic parameters of a population that lives across
+a heterogeneous two-dimensional landscape.
+In particular, we have shown that the resistance distance approximation
+-- which underlies several of the most commonly-used tools of landscape genetics --
+can produce erroneous estimates, especially in the presence of biased gene flow.
+We implement an alternative method, 
+which uses coalescence times instead of the resistance distance approximation,
+which provides good estimates of gene flow in a wider range of situations.
+This weakness of resistance-based methods is not surprising --
+the original papers describing Circuitscape \citep{mcrae2006isolation} and EEMS \citep{petkova2016visualizing}
+present the resistance approximation as a necessary step for computational feasibility --
+but perhaps not widely appreciated.
+Results of resistance-based methods should therefore be interpreted with caution:
+for example, high inferred gene flow between two areas
+may actually mean the areas are being fed by a common source.
+
+Our new method infers effective population sizes and gene flow rates quite well 
+-- and, being Bayesian, provides estimates of uncertainty --
+given data from discrete populations connected by migration.
+Can our method replace resistance-based methods?
+Perhaps, but
+the substantial uncertainty we saw on graphs with only tens of nodes
+is indicative of a larger problem we face for realistic models.
+We have seen that discretization of space results in substantial modeling error,
+in part because of randomness of geographic sampling
+and unmodeled process noise can lead to overfitting.
+Partitioning space into a finer grid should help with these problems,
+but tends to make the inference problem itself more ill-conditioned:
+with more connections,
+changing the value of one connection affects coalescence time less,
+and so inferences about that value must necessarily be less certain.
+This tradeoff implies a high degree of unavoidable uncertainty.
+Reproducible, reliable inference will likely require development of new inference methods
+explicitly incorporating continuous geography.
+
+Although commute and coalescence times can be quite different,
+as also noted by \citet{mcrae2006isolation}, 
+they are equal (with a particular choice of local diversity)
+if populations are arranged on a ring or torus
+with symmetric migration.
+The similarity of these models to rectangular grids
+may explain why in other work, commute time methods have shown relatively good fit to data
+simulated with symmetric migration rates.
+
+\paragraph{Other methods}
+For comparison on equal footing, we have implemented a resistance-based inference method.
+However, methods that use resistance distance have substantial advantages over the method we present here.
+Circuitscape \citep{mcrae2008using} produces maps of much higher resolution 
+than is presently possible with coalescent times.
+\citet{hanks2013circuit} connects the problem to intrinsic conditional autoregressive models, 
+providing a Bayesian model of an underlying Gaussian Markov random field.
+EEMS \citep{petkova2014visualizing} integrates over possible tessellations of the region
+as a way of regularizing the inferred landscape,
+hopefully obtaining a map at the best possible resolution supported by the data.
+
+Most promisingly, the recent method, MAPS \citep{alasadi2018estimating}
+is based on EEMS, but instead of resistance distances,
+uses pairwise sharing of long haplotype segments,
+which have been shown to carry substantial information about recent demography
+\citep{ralph2013geography,palamara2013inference,browning2015accurate,ringbauer2017inferring}.
+Furthermore, the likelihood model underlying MAPS is based on coalescent theory,
+and thus should not be subject to the drawbacks of resistance distance.
+\citet{ringbauer2018estimating} also uses shared haplotype lengths
+to identify a barrier in a coalescence-based framework.
+However, these methods requires fairly dense genotyping and a genetic map,
+something not available for many species.
+
+Other widely-used inference methods, such as
+MIGRATE \citep{beerli1999maximumlikelihood,beerli2010unified},
+BayesAss \citep{wilson2003bayesian}, 
+and IMa \citep{hey2007integration}
+use likelihoods derived from a coalescent model --
+and use much more of the data than we do.
+However, because they use locus-specific data rather than an average across loci,
+these methods can only be applied to a relatively small number of populations and loci,
+because of computational efficiency.
+We have found asymmetry in migration rates to be important:
+\citet{hanks2017modeling} developed a method that allows asymmetry
+by modeling genetic similarity as deriving from an underlying random field
+whose space-time covariance is given by the covariance of the (forwards-time) population fluctuations.
+However, this is not motivated by a generative model for genetic data.
+
+\paragraph{Asymmetric dispersal and consequences for resistance methods}
+We have found that asymmetric rates of lineage movement
+can produce qualitative differences between coalescence times and resistance distances,
+and thereby mislead models based on resistance distance.
+How should this affect the results of resistance-based methods to genomic data,
+besides general inaccuracy?
+This asymmetry occurs if one region receives a greater proportion of each new generation as migrants
+from another population than that other population does from it.
+This is expected, for instance,
+if organismal dispersal is biased by wind or water currents \citep{gaines2003avoiding,morrissey2009maintenance},
+or in the presence of source-sink dynamics \citep{dias1996sources,lenormand2002limits}.
+Suppose, for instance, that seed dispersal for a species of tree
+is biased downhill,
+and the landscape is dominated by a large valley.
+Since lineages tend to move uphill,
+genetic distance between locations on opposite slopes of the valley will be relatively large,
+so resistance methods will infer a barrier along the bottom of the valley
+when no barrier to dispersal exists.
+On the other hand, if there is a ridge instead of a valley,
+resistance methods will show no barrier (and relatively high movement rates),
+even if there is very little uphill dispersal.
+This could lead to a falsely high assessment of gene flow across the top of the ridge.
+
+\paragraph{Circuitscape}
+Could we use coalescence time in place of commute time
+for workflows that compute values on a given map of movement rates
+that are then compared to genetic distance?
+In principle this could be done,
+but there is no clear way to do this 
+at the same (impressively fine) geographic resolution that is currently possible,
+e.g., using Circuitscape \citep{mcrae2006isolation}.
+This is because if we have $n$ sampled locations
+on a landscape discretized into $N$ regions,
+the computational complexity of finding commute times is $nN$,
+but coalescence time scales as $N^2$.
+This is because each $N$-vector of hitting times to a particular location can be found independently,
+but the entire $N \times N$ matrix of coalescence times must be found together.
+However, as we discuss above, this fine geographic resolution is in some sense illusory --
+for each fine-resolution map there should be coarser maps that give almost identical coalescence times.
+However, we are not aware of existing theory providing guidance on how to find these.
+
+\paragraph{Difficulty of inference}
+What determines the feasibility of inferring movement rates from genetic distances?
+Using coalescence times instead of commute times greatly improves accuracy in many situations,
+but there are still a number of factors that determine the tractability of the problem.
+Observation noise is clearly an issue;
+we found that estimation errors of a few percent was enough to seriously degrade accuracy.
+However, genetic distances can be estimated to much higher accuracy using modern genomic data.
+Perhaps more seriously,
+a low rate of coalescence relative to movement can also make the problem essentially nonidentifiable.
+This is clear at least in the limit: with very low coalescence, 
+the population loses any isolation by distance and is indistinguishable from a randomly mating population.
+In continuous landscapes, this balance is measured by Wright's local effective population size,
+which is proportional to the number of other individuals 
+within a circle of radius equal to the mean dispersal distance.
+
+Inference of movement rates, in either a coalescence or commute time framework,
+is an ill-conditioned problem,
+implying the need for regularization to obtain reliable inference.
+This explains the results of \citet{graves2013current},
+who found a large area of resistance parameters
+that produced equivalently good fits to genetic distance
+(i.e., nonidentifiability, analogous to a flat likelihood surface).
+
+\paragraph{The effect of history}
+The models we discuss here assume that population sizes and migration rates
+have been constant on the time scale given by the within-species coalescent time.
+This is rarely true in practice \citep{neigel1991estimation,barton1995genealogies}.
+However, geographic differences in mean relatedness
+are established on a shorter time scale --
+the time scale over which a lineage, moving randomly across the landscape,
+``forgets'' where it started
+(i.e., the mixing time of the Markov chain).
+This is the time it takes for the uncertainty in lineage movement
+to reach the scale of the species range.
+For instance,
+if the width of a species' range is roughly 500km,
+and if mean dispersal distance is 10km,
+the standard deviation of ancestor location $t$ generations ago is $10\sqrt{t}$,
+so the mixing time of a lineages across the landscape,
+not accounting for barriers, is of order 2500 generations,
+or about 50,000 years.
+The landscape of many areas has changed substantially over this time, % \citep{pleistocene_mojave},
+which suggests that models incorporating change over time 
+may be required to model modern tortoise diversity.
+However, the effects of recent history are strongest,
+so landscapes estimated assuming constant populations
+may give a reasonable picture of the landscape averaged over recent times.
+
+\paragraph{Assumptions}
+Modeling lineages as a Markov chain is nearly ubiquitous in population genetics today,
+but may not be appropriate.
+Even though the population dynamics in forwards time are Markov, 
+the dynamics of lineages traced back in time (the coalescent process) may not be. 
+For example, if fecundity is variable and population density is low, 
+two lineages near each other are more likely to share a recent common origin, 
+so knowing how one lineage moves is informative about how the other lineage is likely to move. 
+This effect becomes small as population density increases, 
+so it should be a relatively minor point if the offspring
+of any one parent are typically interspersed with the offspring of many others.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Acknowledgements}
+
+Thanks to David Levin for useful suggestions regarding hitting time calculations,
+to Paul Marjoram for useful comments,
+and to Brad Shaffer and Evan McCartney-Melstad for useful discussions
+and providing the genetic distance data for desert tortoises.
+\plr{ERIK: also John, Hossein, Desi, ???}
+
+\bibliography{references}
+
+\appendix
+\renewcommand{\thefigure}{S\arabic{figure}}
+\setcounter{figure}{0}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{The simplest example}
-
+\subsection{The simplest example}
+\label{ss:simple_example}
+
+To demonstrate the main theoretical ideas,
+we provide a short example.
 Consider a Markov chain with two states, 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Bayesian inference of movement rates}
-
-Mean coalescence times estimated from real data are subject to a number of sources of noise
-(which we discuss in more detail later).
-Here, we use
-the exact solutions based on linear algebra developed above
-to develop a Bayesian inference method that accounts for noise in the data
-and estimates uncertainty in the resulting estimates.
-To do this, we model genetic distances as equal to coalescence times plus noise:
-$D_{ij} = C_{ij} + \epsilon_{ij}$,
-where $\epsilon_{ij}$ are independent and Normally distributed
-with mean $\mathbf{0}$ and variance $\sigma_\epsilon^2$,
-except that $\epsilon_{ij} = \epsilon_{ji}$.
-(We imagine that $C_{ij}$ is large compared to $\sigma_{\epsilon}$,
-so that resulting values are still positive.)
-Using the calculations above, 
-given $G$ and $\gamma$ we can compute the corresponding coalescence times
--- which we call $\mathcal{C}(G, \gamma)$ --
-so our model is that $D$ has a Gaussian distribution 
-with mean $\mathcal{C}(G, \gamma)$.
-The maximum likelihood estimate of $G$ could be found directly;
-however, this in practice will quite likely contain negative movement rates,
-and does not allow us to impose constraints
-(such as only allowing a subset of movement rates to be nonzero).
-
-Therefore, we use Bayesian inference,
-placing independent exponential priors on the parameters --
-the movement rates, $G_{ij}$, that are not constrained to be zero
-and the coalescence rates, $\gamma$.
-Concretely, suppose that the spatial arrangement of populations is given by a graph,
-and only movement rates corresponding to edges in the graph are allowed to be nonzero.
-Write $i \sim j$ to mean that $i$ and $j$ are adjacent in the graph,
-that there are $m_G$ edges in the graph,
-and that each of the $n$ populations has its own coalescence rate.
-Then, the resulting log-posterior density is
-\begin{align} \label{eq:post}
-    \begin{split}
-\mathcal{L}(D, G, \gamma) 
-    &=
-    2 m_G \log(\lambda_G) + n \log(\lambda_\gamma) 
-    - \frac{n(n+1)}{2} \log(2 \pi \sigma^2_\epsilon)
-	- \sum_{i \sim j} \lambda_G G_{ij}  \\
-    &\qquad
-    -\sum_{i=1}^n \lambda_{\gamma}\gamma_i
-		-\frac{1}{2 \sigma_{\epsilon}^2} \sum_{i \leq j} \left(
-            \mathcal{C}_{ij}(G,\gamma) - D_{ij}
-        \right)^2 
-    \end{split}
-\end{align}
-Here $1/\lambda_\gamma$ and $1/\lambda_G$ are the prior means of $\gamma_i$ and $G_{ij}$,
-respectively.
-
-Inference with commute times is done in exactly the same way,
-except that $q$ replaces $\gamma$,
-and instead of $\mathcal{C}(G,\gamma)$ we use $\comdistfn(G, q)$,
-which is the matrix $\comdist$ computed from $G$ and $q$ by solving equation \eqref{eq:commute_approx}.
-
-It is not required in this approach to have samples from every spatial location,
-allowing inference on incompletely sampled spatial discretizations.
-To carry out inference in this situation,
-we simply use Equation \ref{eq:post},
-but only sum over observed entries of $D_{ij}$.
-
-\paragraph{MCMC methods}
-To sample from the posterior distribution of equation \eqref{eq:post}, 
-we used R \citep{R} 
-to implement a standard Metropolis-Hastings algorithm with Gaussian proposals \citep{brooks2011handbook}.
-Starting locations were chosen by sampling from the prior distribution.
-Before the usual ``burn-in'' phase before sampling for inference begins,
-we included a period of ``pre-burn-in'' which used the same MCMC procedure
-with a larger $\sigma_\epsilon$, to allow the chain to more quickly converge on the high-posterior
-portion of parameter space.
-Typically, we ran MCMC for $10^6$ iterations of pre-burn-in 
-and $3 \times 10^6$ iterations of burn-in,
-followed by $4 \times 10^6$ iterations which were used to estimate posterior distributions,
-although for small graphs fewer iterations were necessary.
-We used standard methods to assess convergence and mixing.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Test cases and simulations}
-
-We compared inference under coalescence and commute time models
-using both data ``from the model'' and more realistic simulations.
-For the first category, we designed a number of graphs to test particular aspects of inference,
-with migration rates specified on each directed edge
-(plotted below using igraph \citep{igraph}).
-Given the edge weights ($G$) and the coalescence rates ($\gamma$) of a graph,
-we produced data by calculating exact coalescence times ($\mathcal{C}(G,\gamma)$),
-to whose entries we added independent Gaussian noise while preserving symmetry,
-as described above.
-Particular parameters and levels of noise are given in the Results.
-
-These simulations are expected to be nearly equivalent to population-based simulations
-(either forwards-time or coalescent);
-the only difference being that noise terms about the theoretical mean should be somewhat correlated.
-
-\paragraph{Forwards-time simulations}
-To produce realistic data, we implemented forwards-time simulations using SLiM v3.0 \citep{haller2017slim,haller2018treesequences},
-with individuals living across continuous, two-dimensional geography (sometimes with barriers),
-from which we recorded genomic data.
-The basic simulation, which we modified to produce several other situations,
-is as follows.
-Each simulation had around $10^4$ diploid, hermaphroditic individuals, 
-living across two-dimensional geography.
-Mates were selected from nearby individuals using a Gaussian kernel with standard deviation $\sigma_d$,
-truncated at $3 \sigma_d$.
-Simulations were done on either a square region with width equal to $80\sigma_d$,
-or a $3:5$ rectangular region with the same area.
-Offspring are dispersed away from the mother using the same kernel,
-reflected from the boundaries of the habitable area.
-In each simulation, $\sigma_d$ was chosen 
-so that there were an average of 0.25 individuals per $\sigma_d^2$ units of area.
-In order to prevent excessive population clumpiness \citep{felsensten1975pain}, 
-we also model local competition. 
-To do this, we use the same truncated Gaussian kernel to define an ``interaction strength''
-between each pair of nearby individuals;
-then, the fitness of each individual is set to $\exp(-x/40)$, 
-where $x$ is the sum of their interaction strengths with other individuals.
-This reduces fitness for individuals in denser areas.
-Individuals are chosen to reproduce proportional to their fitness.
-Clumpiness is similar to what is seen in 
-Figure \ref{fig:5x3b_continuous}.
-
-Individuals have 10 pairs of ``chromosomes'' with $2 \times 10^5$ loci each 
-and a recombination rate of $10^{-8}$ per generation per locus.
-Neutral mutations were added at a rate of $10^{-9}$ per meiosis.
-To make these simulations computationally feasible,
-the mutation rate was set to 0.0 during the forwards simulation,
-while tree sequences were recorded \citep{haller2018treesequences},
-and mutations were added after the fact with msprime,
-which is equivalent to including mutations during the forwards simulation 
-but resulted in much faster run times \citep{kelleher2018xxx}.
-In all cases, the number of mutations was more than adequate to reliably estimate genetic distance.
-
-To compute genetic distances,
-geographic space was partitioned into rectangular regions as depicted below.
-within each region, 
-individuals are sampled uniformly for ``genotyping'' 
-from among those within the middle three quarters of each dimension of rectangle,
-as shown for one situation in Figure \ref{fig:5x3b_continuous}.
-This protocol was chosen as a compromise between sampling all individuals 
-from near the center of each location,
-which could result in a large number of close relatives being sampled, 
-leading to an underestimate of the mean genetic distance for individuals in that location
-and uniform sampling from the whole area of each grid, 
-which may result in many sampled individuals being very close to other grid squares.
-Mean genetic distances for each pair of regions is then computed as described above
-from genome sequence provided by SLiM.
-To estimate uncertainty, we computed the standard error of these mean distances
-using the minumum number of individuals in each location.
-
-Before being passed to the MCMC inference method,
-we rescaled genetic distances so that the movement rates resulting from the inference
-would be approximately of order one.
-We did this by multiplying genetic distances by a constant
-so that the rescaled mean of the entries of $D$
-is equal to the number of locations in the discretization.
-
-\subsection*{Code availability}
-
-All code used for this work is available at \url{https://github.com/elundgre/gene-flow-inference}
-(including an R package to perform inference)
-and \url{https://github.com/petrelharp/isolation_by_coalescence} (including the SLiM simulations).
-
-%%%%%%%%%%%%%%%%%%
-\section*{Results}
-%%%%%%%%%%%%%%%%%%
-
-\plr{transition paragraph}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Model validation}
-
-We first tested the method using data generated under the model.
-To test the methods across a wide range of situations,
-we generated random movement rates for 
-a large number of $3 \times 2$ rectangular grids:
-twenty-five symmetric (i.e., with $G_{ij} = G_{ji}$) and twenty-five asymmetric.
-We generated movement parameters
-by rounding independent exponential random variables up to the nearest one tenth,
-and fixed coalescence rate at 1 for all locations.
-We then added independent noise to each matrix entry (preserving symmetry),
-and assume that we can estimate the amount of noise reasonably well, 
-so used the standard deviation of the noise for $\sigma_\epsilon$ in the likelihood function.
-To assess the impact on inference of the number of parameters,
-we performed inference both with and without the assumption 
-that coalescence rate is the same everywhere.
-For $3 \times 2$ graphs, there are 21 equations
-and 14 movement parameters,
-so that with a global coalescence rate there are 15 unknowns
-but with individual coalescence rates there are 20 unknowns,
-which is close to nonidentifiable.
-(We will not present results from an analogous version of resistance distance inference
-that constrains all values of $q$ to be identical
-because the true values are not identical, even if all coalescence rates are,
-and in practice the method infers true values of $q$ extremely reliably.)
-To quantify accuracy across these random graphs in different situations,
-we computed, for each model fit on a particular random graph,
-the absolute difference between the posterior median of each movement parameter
-and the true value.
-We then averaged these across movement parameters to produce a measure of accuracy for that model fit,
-reported below as ``mean absolute error''.
-
-\paragraph{Varying noise}
-Unsurprisingly, coalescence time inference became less accurate as the amount of noise increased.
-As we varied the standard deviation of the noise
-from 1/1000 to 1/50 of the mean value of $C$ for that grid,
-the mean absolute error 
-for coalescence time inference with a single coalescence rate
-increased from around 0.05 to 0.4.
-Posterior interquartile ranges increased proportionally.
-(although there was substantial variation in accuracy between replicates;
-see Figures \ref{fig:mult_noise} and \ref{fig:mult_noise_iqr}).
-Increasing the number of parameters by allowing multiple coalescence rates
-made the inference problem even harder --
-mean absolute errors no longer depended strongly on the amount of noise added,
-and hovered around 0.4.
-
-More surprisingly, the performance of commute time inference did not depend on the amount of noise --
-mean absolute error was around 0.6 across all levels of noise,
-both with and without more than one parameter for local diversity
-(see Figure \ref{fig:mult_noise}).
-Although symmetry did not affect coalescence time inference,
-commute time inference did substantially worse on asymmetric graphs,
-showing median absolute errors of around 0.8.
-
-\paragraph{Varying coalescence rates}
-Lower coalescence rates mean that lineages wander about the landscape for longer
-before they coalesce, 
-thus greatly reducing the geographic information we get from coalescence times.
-Indeed, coalescence rates had a strong effect on feasability of the inference problem
-(shown in Figures \ref{fig:mult_gam} and \ref{fig:mult_gam_iqr}):
-lowering the coalescence rate from 10 to 0.1
-(always adding noise with standard deviation 1/200th of the mean value)
-increased mean absolute error of coalescence time inference from around 0.05 to 0.5.
-Again, allowing multiple coalescence rates
-or using commute time inference were highly inaccurate in all cases,
-with mean absolute errors of above 0.5.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Identifying a barrier to gene flow}
-\label{sec:5x3b}
-
-We then test the method's ability to locate barriers to migration
-on the $5 \times 3$ grid with asymmetric gene flow shown in Figure \ref{fig:5x3b_grid}. 
-Each nonzero movement rate is determined randomly as before.
-Coalescence rates are set to 1 everywhere.
-Figure \ref{fig:5x3b_post_coalvcom} shows posterior distributions of the movement rates
-for both coalescence time and resistance distance inference
-on this graph, with noise standard deviation equal to
-1/1000 of the mean coalescence time (left)
-and 1/100 of the mean coalescence time (right).
-Although increased noise in the data increased uncertainty in the estimated values,
-coalescence time inference correctly inferred not only the location of the barrier
-but also the migration rates along all other edges within a reasonable margin of error.
-Commute time inference also identified the barrier,
-but was much less accurate for the other values.
-We also explored making the problem harder,
-by first (a) dropping rows and columns from the data corresponding to populations 2, 5, 7, and 15;
-and then (b) allowing a separate coalescence rate for each location,
-in both cases using a noise level of 1/100.
-The resulting posterior distributions are shown in Figure \ref{fig:5x3boxplots_mult_gam},
-and show substantially increased error,
-but only in movement rates to the removed locations.
-As before, uncertainty was increased when multiple coalescence rates were allowed,
-but not as strongly as before, likely because the problem is less close to being underconstrained,
-having 105 equations and 59 unknowns.
-
-\begin{figure}
-\centering
-     \includegraphics[scale=.6]{figs/5x3b_grid}
-    \caption{Plot of the grid structure 
-    for the 5 $\times$ 3 grid with two barriers:
-    movement rates between $\{2,3\}$ and $\{7,8\}$ are zero,
-    as are those between $\{8,9\}$ and $\{13,14\}$.
-    Values for the non-zero movement rates 
-    were chosen by rounding independent draws of an exponential random variable with mean 1 
-    up to the nearest tenth.
-    } \label{fig:5x3b_grid}
-\end{figure}
-
-\begin{figure}
-\centering
-     \includegraphics[scale=1]{figs/5x3b_post_coalvcom}
-    \caption{Posterior distributions for values of $g$ 
-    for the 5 $\times$ 3 graph with barriers 
-    with two different noise levels with both coalescence time and resistance distance inference.
-    See Figure \ref{fig:5x3b_grid} to see movement parameter locations.
-    %\plr{TODO: replace bottom two panels with same thing for commute time (eg figs/5x3b100\_com.pdf);
-    %     put this figure in the supplement}
-}
-    \label{fig:5x3b_post_coalvcom}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Biased migration}
-\label{sec:biased_migration}
-
-Both matrices of coalescence and commute times, $C$ and $R$, are symmetric,
-but they do not deal with asymmetric (biased) migration in the same way.
-Since commute time to $i$ requires paths both to and from $i$,
-but coalescence time to $i$ only reaquires lineages to leave $i$,
-if, for instance, there is a low rate of migration back into $i$,
-then commute times to $i$ could be quite long even while coalescence times are not.
-For instance, suppose there are three populations arranged in a line,
-and lineages currently in the outer populations are much more likely to come from the central population
-than the other way around.
-Since lineages quickly move to the center and coalesce regardless of starting position,
-coalescence time (and genetic diversity) will be relatively low between all individuals.
-Commute time between the two outer locations, on the other hand,
-will be much longer than other comparisons, since it requires a lineage to leave the center.
-
-In order to investigate this general situation,
-we tested both methods on four different $4 \times 4$ graphs (shown in Figure \ref{fig:4x4_grids}):
-\emph{uniform}, where all movement rates are 1.0;
-\emph{symmetric}, where movement rates are symmetric and randomly generated as before;
-\emph{asymmetric}, where all movement rates are random;
-and \emph{biased}, where movement rates down or to the left are equal to 2.0,
-and movement rates up or to the right are 0.5.
-For inference, we added noise with standard deviation $1/500^\text{th}$ of the mean value.
-
-Differences between true mean coalescence times ($C$) and mean commute times ($\comdist$) 
-are shown in Figure \ref{fig:RvsC}:
-they are fairly small in the uniform and symmetric cases, moderate in the asymmetric case, 
-and extreme in the biased asymmetric case.
-This difference suggests that commute time inference may be strongly misled 
-in the asymmetric and biased situations,
-but it still could be the case that the movement rates that give the best fit of commute time
-to coalescence time data could be close to the actual rates use to generate the data.
-
-Nonetheless, 
-coalescence time inference was substantially more accurate in all cases.
-Posterior distributions of movement parameters are shown in Figure \ref{fig:4x4box},
-produced by applying both inference methods
-after adding noise with standard deviation of 1/200 the mean value of $C$.
-The mean absolute errors for coalescence time inference are 0.12 or less in all cases.
-Commute time inference obtains roughly uniform migration rates
-(although varying by a factor of 2) in the uniform case;
-and migration rates noisily correlated with the truth in the symmetric graph.
-However, commute time inferences for the asymmetric and biased graphs are only weakly correlated, 
-if at all, to the truth.
-Commute times also drastically overestimate movement rates for the biased case,
-as we would expect based on the differences discussed above.
-
-\begin{figure}
-\centering
-     \includegraphics[scale=1]{figs/RvsC}
-    \caption{Plots of the true values of $\comdist$ vs $C$ 
-    for the uniform graph, symmetric graph, the asymmetric graph,
-    and the biased asymmetric graph, 
-    shown with the $y=x$ line.
-    %\plr{TODO: make this a 2x2 grid instead of 4x1}
-    See Figure \ref{fig:4x4_grids} to see overall graph structure.}
-    \label{fig:RvsC}
-\end{figure}
-
-\begin{figure}
-\centering
-     \includegraphics[scale=1]{figs/4x4boxplots_paper}
-    \caption{Boxplots of the posterior distributions for $g$ for commute time and coalescence time
-    based inference for the uniform graph, symmetric graph, 
-    the asymmetric graph, and the biased symmetric graph.
-    See Figure \ref{fig:4x4_grids} to see movement parameter locations.}
-    \label{fig:4x4box}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Continuous geographical space}
-
-The data we have used thus far are produced ``under the model'',
-and should provide an accurate depiction of our method applied to discrete, randomly mating populations
-whose connections by migration are known.
-However, this is nearly always a rough approximation to reality,
-in which organisms are distributed across continuous geography,
-and ``populations'' are constructed by necessity, often driven by sampling locations.
-We also applied coalescence and resistance inference,
-as well as the popular resistance-based method EEMS \citep{petkova_eems},
-to data from simulations on continuous geography.
-
-\paragraph{Process noise}
-The mean coalescence times of lineages that we compute analytically
-are exact for large, randomly mating populations,
-but the word ``mean'' indicates an average across several levels of randomness:
-each observed, empirical mean estimated using a sample of genotyeps
-will deviate from this depending on both the individuals sampled (``sampling noise'')
-and the stochasticity of the population history itself (``process noise'').
-For more discussion of this latter contribution in other contexts, see
-\citet{wakeley2012genealogies}, \citet{waples2009modeling}, or \citet{ralph2015empirical}.
-In populations living across continuous geography,
-both sources of noise can be substantially larger due to randomness in spatial locations.
-This is hinted at by Figure \ref{fig:5x3b_continuous},
-where we see that the distribution of individuals across the landscape is fairly patchy:
-a barrier might be inferred in a region that has not happened to have a local population recently
-only because of demographic stochasticity.
-This might make inference using a discretization of continuous space considerably more difficult.
-
-To quantify the relative contributions of the different sources of noise,
-we coompared mean genetic distances, calculated in the same way,
-between nonoverlapping sets of samples from the same simulation,
-and between different simulations.
-These simulations were done using SLiM as described in the Methods
-on a square, $4 \times 4$ landscape with no barriers.
-We found that process noise was much larger than sampling noise --
-as shown in Figure \ref{fig:h_val_comp},
-discrepancies between simulations are about five times that seen
-between different samples from the same simulation.
-An implication of this is that uncertainties in genetic distance
-estimated from the data will be (possibly large) underestimates,
-a fact also pointed out for effective population size by \citet{waples}.
-
-\begin{figure}
-\centering
-    \includegraphics[width=0.8\textwidth]{figs/h_val_comp}
-\caption{
-    Comparison of the values of the genetic distance matrix $C$
-    for a $4 \times 4$ discretization of a square landscape with uniform migration
-    calculated from various situations.
-    All values are plotted against the corresponding values from the $C$ matrix
-    calculated using the first set of individuals from the first simulation,
-    along with the $y=x$ line.
-    Black circles are from a second (non-overlapping) set of individuals from the first simulation,
-    red triangles are from the second simulation, 
-    and blue plusses are for the third simulation.
-    Mean absolute value relative differences between the new values of $C$ for each situation
-    and the original values from the first set of individuals from the first simulation are
-    1.8\% for the second set of indivuals from the first simulation,
-    13.2\% for the second simulation, 
-    and 7.3\% for the third simulation.
-    } \label{fig:h_val_comp}
-\end{figure}
-
-Although we found substantial process noise
-with a large, flat landscape with uniform, unbiased migration,
-many biological processes are expected to increase between-replicate stochasticity.
-To demonstrate this, we simulated a square, flat landscape as above,
-but with \emph{biased migration}: 
-each offspring's location was chosen randomly as before,
-but with mean location $\sigma_d/10$
-up and to the right of the parent's location.
-% (or 0.05\% of the edge length of the landscape)
-This produces reverse-time gene flow down and to the left.
-As a population genetics model,
-this case is similar to a population expansion,
-replacing the leading edge of the expansion with the corner of landscape 
-that migration bias leads away from.
-In this situation, as in an expansion, 
-individuals at the edge may have disproportionately more offspring,
-leading to low genetic diversity and strong drift \citep{surfing}.
-This results in substantially more noise in the data,
-clearly visible in the isolation by distance plots of Figure \ref{fig:ibd_comp}. 
-Despite this large contribution of process noise,
-coalescence time inference correctly infers the bias on a broad scale:
-using both a $2 \times 2$ and a $4 \times 4$ grid,
-inferred gene flow rates down and to the left are consistently much larger
-than those up and to the right 
-(Supplemental Figures \ref{fig:2x2grid_asb}, \ref{fig:grid_bias_4x4_1}, 
-and \ref{fig:post_dists_bias_4x4_1}).
-However, inferred rates show substantial heterogeneity,
-likely due to stochasticity of recent demography.
-
-\begin{figure}
-\centering
-     \includegraphics[width=0.75\textwidth]{figs/ibd_comp}
-    \caption{
-        Genetic distance against geographic distance,
-        for the square landscape with 
-        \textbf{(above)} uniform migration
-        and \textbf{(below)} biased migration.
-        The horizontal lines in the latter plot
-        are likely from comparing genomes of individuals from
-        families that have recently increased in size and geographical spread, 
-        resulting in many pairs of genomes with similar genetic distance 
-        but varying geographic distance.
-    \label{fig:ibd_comp}
-    }
-\end{figure}
-
-\paragraph{Identifying a barrier}
-We now revisit the earlier situation where a landscape has locations that are barriers to gene flow.
-The landscape is a continuous version of this situation,
-shown in Figure \ref{fig:5x3b_continuous},
-where red bars depict uninhabitable regions that block migration.
-Since the distribution of offspring locations is truncated at three standard deviations
-away from the mother, a sufficiently thick uninhabitable area completely stops migration
-directly across it.
-We calculate mean genetic distances using 50 individuals 
-from each grid location,
-here giving us a total of 750 individuals.
-The gene flow rates across the barriers are reliably estimated to be very small,
-while other values have substantial variation:
-in particular, gene flow rates are estimated to be low
-across some edges that do not cross barriers;
-this may be because of recent demographic stochasticity.
-The barriers were also identified with comparable accuracy by both our resistance distance method
-and EEMS, as shown in Supplemental Figure \ref{sfig:barrier_results}.
-Note also that this analysis began by grouping samples into discrete ``populations'' appropriately;
-in absence of a good \textit{a priori} idea of where barriers are,
-this step might not be straightfoward.
-% For instance, we may wonder whether a river or other geographical feature
-% restricts gene flow.
-% We would then discretize the landscape such that individuals on opposite sides of the 
-% potential obstruction are in separate grid spaces.
-
-\begin{figure}
-\centering
-     \includegraphics[scale=.8]{figs/ind_locs_5x3b_1}
-     \includegraphics[height=0.3\textheight]{figs/grid_5x3b_1}
-     \includegraphics[height=0.3\textheight]{figs/post_dists_5x3b_1_new}
-    \caption{
-        \textbf{(top)}
-        Locations of all and selected individuals on the $5 \times 3$ landscape with barriers,
-        with individuals used to compute the mean genetic distance matrix marked in bold.
-        The red bars show the locations of uninhabitable regions of the landscape.
-        They are sufficiently thick so that no migration can occur directly across them.
-        \textbf{(bottom left)}
-        Posterior medians of movement rates inferred using coalescence time inference,
-        and \textbf{(bottom right)}
-        boxplots of corresponding posterior distributions,
-        with movement rates crossing the barriers colored in red.
-        The grouping of individuals into a $5 \times 3$ grid is shown above,
-        and labels are the same as in Figure \ref{fig:5x3b_grid}.
-    \label{fig:5x3b_continuous}
-    }
-\end{figure}
-
-\paragraph{Other barriers: population sinks and recent expansions}
-We simulated two other situations on a continuous landscape that produced an bias in effective migration.
-First, we simulated a \textbf{valley} in a landscape with a $5 \times 3$ aspect ratio as above, 
-in which fecundity of individuals in the middle 1/4 of the landscape was reduced by 50\%
-(a ``source-sink'' landscape).
-Since individuals in the valley are more likely to have parents outside of the valley,
-this produces gene flow ``uphill'', away from the ridge.
-Next, we simulated a recent \textbf{expansion},
-in which the middle 4/5 of the landscape was uninhabitable for 400,000 generations,
-until 500 generations before the present day
-(a model of ``secondary contact'').
-The resulting expansion produces gene flow back away from the center,
-similar to the ``valley''.
-
-Both coalescence time inference and the resistance landscape inferred by EEMS
-show a barrier to gene flow in the middle of the landscape (Figure \ref{fig:more_barriers}):
-this is arguably correct, 
-although there is in fact no current barrier to migration in either simulation.
-Coalescence time inference also correctly infers 
-the general bias in recent gene flow away from the center,
-although as in other continuous-space simulations, 
-there was substantial noise in the inferred patterns.
-The resistance landscapes as inferred by EEMS show patchy, vertical bands of high resistance:
-these patterns are commonly seen when attempting to use EEMS, 
-whose underlying model is inherently symmetric, 
-to describe asymmetric gene flow.
-Our analysis above shows that this is an unavoidable consequence of the resistance approximation,
-and our implementation of commute time inference,
-which allows asymmetric gene flow, shows simliar patterns
-(Supplemental Figures \ref{sfig:valley_results} and \ref{sfig:expansion_results}).
-
-\begin{figure}
-\centering
-    \includegraphics[width=0.45\textwidth]{figs/valley_eems}
-    \includegraphics[width=0.45\textwidth]{figs/valley_coal_graph}
-    \includegraphics[width=0.45\textwidth]{figs/expansion_eems}
-    \includegraphics[width=0.45\textwidth]{figs/expansion_coal_graph}
-    \caption{
-        \textbf{(Left:)}
-        resistance landscapes inferred by EEMS
-        and 
-        \textbf{(right:)}
-        gene flow rates inferred using coalescence time inference,
-        from individual-based simulations on a continuous landscape,
-        with \textbf{(top)} a \emph{valley} of reduced fecundity in the middle, and
-        \textbf{(bottom)} a recent \emph{population expansion} from both edges of the range
-        that met in the middle.
-        The color scale shows the log migration rates inferred by EEMS.
-        \label{fig:more_barriers}
-    }
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%
-\section*{Discussion}
-
-In this paper, we have studied the problem of
-inferring demographic parameters of populations spread over heterogeneous two-dimensional landscapes
-by using genetic and geographic distances between present-day samples.
-In particular, we have shown that the resistance distance approximation
--- which underlies several of the most commonly-used tools of landscape genetics --
-can produce misleading results, especially in the presence of biased gene flow.
-We implement an alternative method, 
-which uses coalescence times instead of the resistance distance approximation,
-which provides good estimates of gene flow in a wider range of situations.
-This weakness of resistance-based methods is not surprising --
-the original papers describing Circuitscape \citep{mcrae_circuitscape} and EEMS \citep{petkova_eems}
-present the resistance approximation as a necessary step for computational feasibility --
-but perhaps not widely appreciated.
-
-Our new method infers effective population sizes and gene flow rates quite well 
--- and, being Bayesian, provides estimates of uncertainty --
-given data from discrete populations connected by migration.
-Why do we not advocate our method as an immediate replacement for resistance-based methods?
-The substantial uncertainty we saw on graphs with only tens of nodes
-is indicative of a larger problem we face for realistic models.
-There is a high degree of unavoidable uncertainty,
-and process noise can lead to overfitting of models.
-To produce reproducible, reliable results we need to address these problems.
-
-We are able to do inference of reverse time gene flow and coalescence rates 
-with the matrix of mean coalescence times
-in cases where the matrix is noise-free, noisy, or incomplete, 
-or when it made after discretization of a simulated or real landscape.
-To do this, we assume that the reverse time movements of lineages are Markov
-and the population is near stationarity.
-Typically, there is more error and uncertainty in the posterior distributions of the parameters
-when there is more noise, slower coalescence rate, more missing values,
-and multiple parameters for coalescence rate.
-
-In addition to developing a method for inferring migration rates using pairwise coalescence times,
-we compare the results with that of resistance distance (i.e., commute time) based inference.
-Coalescence based inference typically has lower error than commute time based inference,
-along with the benefit that you can also infer coalescence rates. 
-The commute time plus diversity estimate, $\comdist$, 
-provides a good estimate of the coalescence times, $C$,
-when gene flow rates are symmetric ($G_{ij} = G_{ji}$), 
-but can be very wrong when they are asymmetric, 
-including extreme discrepencies 
-when there is an overall bias in the direction of the asymmetry across the landscape 
-(Figure \ref{fig:RvsC}).
-When inference is performed on discrete landscapes 
-with varying degrees of migrational asymmetry, 
-coalescence time based inference is more accurate than commute time inference in all tested cases
-(Figure \ref{fig:4x4box}).
-Although performance of the commute time based method is reasonable 
-when gene flow rates in the underlying model are symmetric, 
-it becomes worse as the gene flow rates become more asymmetric 
-to the point where there is little correlation at all with the true values 
-when there is biased asymmetric migration.
-On the other hand, commute time inference is much less computationally intensive, 
-so much larger spatial systems can be used.
-If the gene flow rates are symmetric, the higher resolution allowed by commute time inference
-may outweigh the loss in accuracy.
-The implication of these results for the body of literature using resistance distance methods
-is that one may need to use caution when interpreting results.
-For example, inferring high gene flow rates between two areas by using resistance distance
-could potentially mean that those were areas were actually being fed by a common source.
-Therefore, it may be worthwhile to apply a coarse resolution coalescence time based method 
-alongside a finer resolution resistance distance method 
-in order to check for evidence of asymmetric gene flow patterns.
-
-Although commute and coalescence times can be quite different,
-as also noted by \citet{mcrae2006isolation}, 
-they are equal (with a particular choice of local diversity)
-if populations are arranged on a ring or torus
-with symmetric migration.
-The similarity of these models to rectangular grids
-may explain why in other work, commute time methods have shown relatively good fit to data
-simulated with symmetric migration rates.
-
-We also apply the method to data from continuous space simulations. 
-While there is substantial variation between the inferred values 
-for different simulations with the same underlying demographic parameters, 
-we are able to reliably find evidence of biased asymmetric migration (if it is present)
-and verify the exitence of a barrier to migration, 
-even when the landscape is otherwise relatively well connected.
-Further work could involve more precise study of the relationship
-between reverse time gene flow rates and forwards time migration rates.
-
-\paragraph{Other methods}
-Other methods that use resistance distance have substantial advantages over the method we present here.
-Circuitscape \citep{mcrae2008using} produces maps of much higher resolution 
-than is presently possible with coalescent times.
-\citet{hanks2013circuit} connects the problem to intrinsic conditional autoregressive models, 
-providing a Bayesian model of an underlying Gaussian Markov random field.
-EEMS \citep{petkova2014visualizing} integrates over possible tesselations of the region
-to obtain a landscape at the best possible resolution supported by the data.
-
-Although we present a new method, based on coalescence times,
-our focus is on understanding the underlying modeling issues
-rather than replacing these other tools.
-Most promisingly, a very recent method, MAPS \citep{alasadi2018estimating}
-is based on EEMS, but instead of resistance distances,
-uses pairwise sharing of long genomic segments (``IBD segments'', \citet{browning2013improving}),
-which have been shown to carry substantial information about recent demography
-\citep{ralph2013geography,palamara2013inference,browning2015accurate}.
-Furthermore, the likelihood model underlying MAPS is based on coalescent theory,
-and thus should not be subject to the drawbacks of resistance distance.
-However, application requires fairly dense genotyping and a good genetic map,
-something not available for many species.
-
-Other widely-used inference methods, such as
-MIGRATE \citep{beerli1999maximum,beerli2010unified},
-BayesAss \citep{wilson2003bayesian}, 
-and IMa \citet{hey2007integration}
-use likelihoods derived from a coalescent model --
-and use much more of the data than we do.
-However, because they use locus-specific data rather than an average across loci,
-these methods can only be applied to a relatively small number of populations and loci,
-because of computational efficiency.
-
-\paragraph{Asymmetric dispersal}
-We have found that asymmetric rates of lineage movement
-are most likely to produce qualitative differences between coalescence times and resistance distances,
-and thereby mislead models based on resistance distance.
-This asymmetry occurs if one region receives a greater proportion of each new generation as migrants
-from another population than that other population does from it.
-This is expected, for instance,
-if organismal dispersal is biased by wind or water currents \citep{gaines2003avoiding,morrissey2009maintenance},
-or in the presence of source-sink dynamics \citep{dias1996sources,lenormand2002limits}.
-XXX MORE
-
-\citet{hanks2017modeling} developed a method that allows asymmetry
-by modeling genetic similarity as deriving from an underlying random field
-whose space-time covariance is given by the covariance of the (forwards-time) population fluctuations.
-
-\paragraph{Consequences for resistance methods}
-We have shown that commute time can show dramatically different patterns
-than coalescence time, especially if there is systematic bias in dispersal.
-How should this affect the results of resistance-based methods to genomic data,
-besides general inaccuracy?
-Suppose, for instance, that seed dispersal for a species of tree
-is biased downhill,
-and the landscape is dominated by a large valley.
-Since lineages tend to move uphill,
-genetic distance between locations on opposite slopes of the valley will be relatively large,
-so resistance methods will infer a barrier along the bottom of the valley
-when no barrier to dispersal exists.
-On the other hand, if there is a ridge instead of a valley,
-resistance methods will show no barrier (and relatively high movement rates),
-even if there is very little uphill dispersal.
-This could lead to a falsely high assessment of gene flow across the top of the ridge.
-\plr{Also applies to least-cost path analysis.}
-
-\paragraph{Circuitscape}
-Could we use coalescence time in place of commute time
-for workflows that compute values on a given map of movement rates
-that are then compared to genetic distance?
-In principle this could be done,
-but not at the same (impressively fine) geographic resolution that is currently possible,
-e.g., using Circuitscape \citep{circuitscape}.
-This is because if we have $n$ sampled locations
-on a landscape discretized into $N$ regions,
-the computational complexity of finding commute times is $nN$,
-but coalescence time scales as $N^2$.
-This is because each $N$-vector of hitting times to a particular location can be found independently,
-but the entire $N \times N$ matrix of coalescence times must be found together.
-However, as we discuss above, this fine geographic resolution is in some sense illusory --
-for each fine-resolution map there should be coarser maps that give almost identical coalescence times.
-However, we are not aware of existing theory providing guidance on how to find these.
-
-\paragraph{Difficulty of inference}
-What determines the feasability of inferring movement rates from genetic distances?
-Using coalescence times instead of commute times greatly improves accuracy in many situations,
-but there are still a number of factors that determine the tractability of the problem.
-Observation noise is clearly an issue;
-we found that estimation errors of a few percent was enough to seriously degrade accuracy.
-However, genetic distances can be estimated to much higher accuracy using modern genomic data.
-Perhaps more seriously,
-a rate of coalescence relative to movement can also make the problem essentially nonidentifiable.
-This is clear at least in the limit: with very low coalescence, 
-the population loses any isolation by distance and looks panmictic.
-In continuous landscapes, this balance is measured by Wright's local effective population size,
-which is proportional to the number of other individuals 
-within a circle of radius equal to the mean dispersal distance.
-\plr{more about continuous landscapes and discretization}
-
-\paragraph{The effect of history}
-The models we discuss here assume that population sizes and migration rates
-have been constant on the time scale given by the within-species coalescent time.
-This is rarely true in practice \citep{avise,barton}.
-However, geographic differences in mean relatedness
-are established on a shorter time scale --
-the time scale over which a lineage, moving randomly across the landscape,
-``forgets'' where it started
-(i.e., the mixing time of the Markov chain).
-This is the time it takes for the uncertainty in lineage movement
-to reach the scale of the species range.
-For instance,
-if the width of a species' range is roughly 500km,
-and if mean dispersal distance is 10km,
-the standard deviation of ancestor location $t$ generations ago is $10\sqrt{t}$,
-so the mixing time of a lineages across the landscape,
-not accounting for barriers, is of order 2500 generations,
-or about 50,000 years.
-The landscape of many areas has changed substantially over this time, % \citep{pleistocene_mojave},
-which suggests that models incorporating change over time 
-may be required to model modern tortoise diversity.
-However, the effects of recent history are strongest,
-so landscapes estimated assuming constant populations
-may give a reasonable picture of the landscape averged over recent times.
-
-\paragraph{Limitations of data}
-Something about microsats versus RADseq versus whole-genome.
-\plr{XXX}
-
-\paragraph{Assumptions}
-Modeling lineages as a Markov chain is nearly ubiquitous in population genetics today,
-but may not be appropriate.
-Even though the population dynamics in forwards time are Markov, the dynamics of lineages traced back
-in time (the coalescent process) may not be. For example, if fecundity is variable and population density
-is low, two lineages near each other are more likely to share a recent common origin, so knowing how one
-lineage moves gives you information about how the other lineage is likely to move. This effect becomes
-negligible as population density increases [Barton et al., 2002]), so it should not be an issue if the offspring
-of any one parent are typically interspersed with the offspring of many others. The methods I develop here
-model the coalescent process as Markov, which should be a good approximation for many systems.
-
-\paragraph{other points:}
-Practical example (tortoises) shows strong asymmetry.
-Methods dealing with ill-posedness are necessary.
-Other measures of genetic distance: what do people use in practice?
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Acknowledgements}
-
-Thanks to David Levin for useful suggestions regarding hitting time calculations,
-to Paul Marjoram for useful comments,
-and to Brad Shaffer and Evan McCartney-Melstad for useful discussions
-and providing the genetic distance data for desert tortoises.
-We are also indebted to Chava Weissman, Bridgette Hagerty, Fran Sandmeier, Dick Tracy,
-and a large number of field workers for collecting the original tortoise tissue samples.
-\plr{also John, Hossein, Desi, ??}
-
-\bibliography{references}
-
-\appendix
-\renewcommand{\thefigure}{S\arabic{figure}}
-\setcounter{figure}{0}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\plr{edit}
-
 Suppose that $G$ is the generator matrix of an irreducible continuous-time Markov chain $X$,
-so that $G_{ij} \ge 0$ for $i \neq j$ and
-$$ G 1 = 0 $$.
-
+so that $G_{ij} \ge 0$ for $i \neq j$ and $G \bone = 0$.
 Let $\tau_{j} = \inf\{t \ge 0 : X_t = j\}$ and $H_{ij} = \E[\tau_j \,|\, X_0 = i]$.
 Then we know that
-$$
+\begin{align*}
     (G H)_{ij} = -1 \qquad \text{for} \; i \neq j ,
-$$
-i.e.,
+\end{align*}
+and hence that for some vector $x$,
 \begin{align} \label{eqn:GH}
-    GH = - 1 1^T + \diag(x).
+    GH = - \bone \bone^T + \diag(x).
 \end{align}
-
 What is $x$?  Well, note that
-$$
-    G = (\diag(x) - 1 1^T) H^{-1}
-$$
+\begin{align*}
+    G = (\diag(x) - \bone \bone^T) H^{-1}
+\end{align*}
 and so
-$$ \begin{aligned}
-    0 &= G1 \\
-    &= (\diag(x) - 1 1^T) H^{-1} x 
-\end{aligned} $$
-and so
-$$
-    x_i = \frac{ 1^T H^{-1} 1 }{ (H^{-1} 1)_i } .
-$$
+\begin{align*}
+    0 &= G\bone \\
+    &= (\diag(x) - \bone \bone^T) H^{-1} x  .
+\end{align*}
+Rearranging, we find that
+\begin{align*}
+    x_i = \frac{ \bone^T H^{-1} \bone }{ (H^{-1} \bone)_i } .
+\end{align*}
 Furthermore,
-the Random Target Lemma \citep{aldous} % https://www.stat.berkeley.edu/users/aldous/RWG/Book_Ralph/Ch2.S2.html
+the Random Target Lemma \citep{aldous-fill-2014}
+% https://www.stat.berkeley.edu/users/aldous/RWG/Book_Ralph/Ch2.S2.html
 tells us that if $\pi$ is the stationary distribution of the chain, then 
@@
 i.e., that $H \pi \propto 1$,
-and thus if $H^{-1} 1 = \nu$, 
+and thus if $H^{-1} \bone = \nu$, 
 then $\pi = \nu / \sum_i \nu_i$.
@@
 
-In summary, $\pi = H^{-1} 1 / 1^T H^{-1} 1$, and 
+In summary, $\pi = H^{-1} \bone / \bone^T H^{-1} \bone$, and 
 \begin{align}
     G 
-    &= (\diag(1/\pi) - 1 1^T) H^{-1} .
+    &= (\diag(1/\pi) - \bone \bone^T) H^{-1} .
 \end{align}
 Note that this equation only specifies $H$ up to a constant added to each column:
-i.e., if given $G$ one obtains a matrix $Y$ solving $GY = \diag(1/\pi) - 1 1^T$,
+i.e., if given $G$ one obtains a matrix $Y$ solving $GY = \diag(1/\pi) - \bone \bone^T$,
 then $H_{ij} = Y_{ij} - Y_{jj}$.
@@
 \begin{align}
-    \comdist = (H + H^T)/4 + (q 1^T + 1 q^T)/2
+    \comdist = (H + H^T)/4 + (q \bone^T + \bone q^T)/2
 \end{align}
@@
 \begin{align} \label{eqn:R_coal}
-    ( GH + (GH)^T + GH^T + (GH^T)^T )/4 + (Gq 1^T + 1 (Gq)^T)/2
+    ( GH + (GH)^T + GH^T + (GH^T)^T )/4 + (Gq \bone^T + \bone (Gq)^T)/2
     &=
-    \diag(q) \diag(\gamma) - 1 1^T .
+    \diag(q) \diag(\gamma) - \bone \bone^T .
 \end{align}
-(To simplify this we used the fact that $\diag(H) = 0$ and $G1 = 0$.)
+(To simplify this we used the fact that $\diag(H) = 0$ and $G\bone = 0$.)
 
@@
 \begin{align} \label{eqn:R_eq_C}
-    \diag(1/\pi) + (Gq 1^T + 1 (Gq)^T)/2
+    \diag(1/\pi) + (Gq \bone^T + \bone (Gq)^T)/2
     &=
@@
 \end{align}
-where the $-1 1^T$ terms on each side canceled.
-Because $G1=0$,
-the most obvious solution to this is if $q = c 1$ and $c \gamma = 1/\pi$, for some constant $c$
+where the $-\bone \bone^T$ terms on each side canceled.
+Because $G\bone=0$,
+the most obvious solution to this is if $q = c \bone$ and $c \gamma = 1/\pi$, for some constant $c$
 (although there is a broader family of solutions).
@@
 Can we solve equation \eqref{eqn:R_eq_C} more generally?
-For that equation to hold, we need $Gq 1^T +  (Gq)^T$ to be diagonal,
+For that equation to hold, we need $Gq \bone^T + \bone (Gq)^T$ to be diagonal,
 i.e., that $(Gq)_i + (Gq)_j = 0$ for all $i \neq j$.
 For any $k \neq \ell$, there exists a vector $u^{(k\ell)}$ such that $(Gu^{(k\ell)})_i = \delta_{ik} - \delta_{i\ell}$;
-the only possible $q$ for which $Gq 1^T +  (Gq)^T$ is diagonal
-are of the form $q = \alpha 1 + \beta u^{(k,\ell)}$ for some $k \neq \ell$ and some constants $\alpha$ and $\beta$.
+the only possible $q$ for which $Gq \bone^T + \bone (Gq)^T$ is diagonal
+are of the form $q = \alpha \bone + \beta u^{(k,\ell)}$ for some $k \neq \ell$ and some constants $\alpha$ and $\beta$.
 We would then need
 \begin{align}
-    1/\pi_i + \delta_{ik} - \delta_{i\ell} = q_i \gamma_i \qquad \text{for all} i.
+    1/\pi_i + \delta_{ik} - \delta_{i\ell} = q_i \gamma_i \qquad \text{for all} \quad i.
 \end{align}
@@
 However, there is not a general solution for $q$ if coalescence rates are also given (as is the case in practice).
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Something else?}
-
-\citet{strobeck1987average} showed that $\diag(C)$ was constant and does not depend on movement rates
+This is related to the fact proved by \citet{strobeck1987average} 
+that $\diag(C)$ is constant and does not depend on movement rates
 for any isotropic conservative migration model.
-Can we explain this?
-(Also cite earlier.)
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Further inference results}
-
-Here we put some more plots describing inference method accuracy.
 
@@
 \includegraphics[scale=.6]{figs/mult_noise}
-\caption{Boxplots of the mean absolute error, %in $\hat{g}$.
+\caption{Boxplots of the mean absolute error
 defined to be the absolute value difference between the true value and posterior median,

### 2018-10-19 isolation_by_coalescence.tex
added note abt mcrae
fixed typos, added figure

@@
 Commute times only depend on the symmetric part of $H$ --
-given an skew-symmetric matrix $Z$ (so that $Z + Z^T = 0$),
+given a skew-symmetric matrix $Z$ (so that $Z + Z^T = 0$),
 any Markov chain that has hitting times given by $H_\epsilon = H + \epsilon Z$ for some $\epsilon$
@@
 and providing the genetic distance data for desert tortoises.
+We would also like to gratefully acknowledge the immense positive impact that Brad McRae \citep[1966--2017;][]{lawler2018tribute}
+has had on landscape genetics and conservation biology through the introduction of landscape resistance.
+We hope this paper helps to carry forward this work.
 \plr{ERIK: also John, Hossein, Desi, ???}
@@
 averaged across movement parameters $g$ for each of the 25 graphs in each situation.
-Each box is labeled with the relative amount of noise added to $H$ to make $\hat{H}$.
+Each box is labeled with the relative amount of noise added to $C$ to produce the values used for inference.
 The top left shows the results for symmetric graphs 
@@
 for each of the 25 graphs in each situation.
-Each box is labeled with the relative amount of noise added to $H$ to make $\hat{H}$.
+Each box is labeled with the relative amount of noise added to $C$ to produce the values used for inference.
 Subplot locations for each situation are the same as in Figure \ref{fig:mult_noise}.

### 2018-10-19 isolation_by_coalescence.tex
proposed rewording of uniqueness section

@@
 
-Since commute times are symmetric, and $R_{zz} = 0$, 
-equation \eqref{eqn:R_from_H} provides $n (n-1)/2$ informative equations.
-This is larger than $4n$, the number of parameters,
+Since coalescence times are symmetric, 
+equation \eqref{eqn:C_matrix} provides $n (n+1)/2$ informative equations.
+This is larger than $4n+n$, the number of parameters,
 for a grid with at least nine nodes (i.e., $3 \times 3$),
@@
 for grids larger than this, except in special cases.
-The same calculation holds for commute times, 
-except that (a) we get an additional $n$ equations from the diagonal entries,
-but (b) this is counteracted by the additional $n$ coalescence parameters.
+The same calculation holds for commute times 
+except that there are $n$ local diversity parameters instead of $n$ coalescence parameters.
 

### 2018-10-19 isolation_by_coalescence.tex
addressed comments mostly

@@
 although the two methods are conceptually similar,
-commute time inference may get the wrong answer if data derive from a coalescent process,
+commute time inference may get the wrong answer if data are derived from a coalescent process,
 especially in the presence of asymmetry.
@@
 given the coalescence or commute times
-we must solve equations \eqref{eqn:C_matrix} or \eqref{eqn:R_from_H} and \eqref{eqn:G_from_H} for $G$,
+we must solve either equation \eqref{eqn:C_matrix} 
+or equations \eqref{eqn:G_from_H} and \eqref{eqn:R_from_H} for $G$,
 and that these do not have unique solutions.
@@
 If in a given biological situation commute and coalescence times are equal --
-i.e., if given $G$ and $\gamma$, there was a $q$ that made $\comdist = C$ --
+i.e., if given $G$ and $\gamma$, there was a $q$ that made $C = \comdist$ --
 then inference with either model would be equivalent.
@@
 %%%%%%%%%%%%%%%%%%
-
-\plr{transition paragraph}
 
@@
 so the mixing time of a lineages across the landscape,
-not accounting for barriers, is of order 2500 generations,
-or about 50,000 years.
-The landscape of many areas has changed substantially over this time, % \citep{pleistocene_mojave},
+not accounting for barriers, is of order 2500 generations.
+The habitat of many species would have changed substantially over this time, % \citep{pleistocene_mojave}
 which suggests that models incorporating change over time 
-may be required to model modern tortoise diversity.
+may be required to model modern diversity.
 However, the effects of recent history are strongest,
@@
 to Paul Marjoram for useful comments,
-and to Brad Shaffer and Evan McCartney-Melstad for useful discussions
-and providing the genetic distance data for desert tortoises.
+to Brad Shaffer and Evan McCartney-Melstad for discussions about landscape modeling,
+and to John Novembre, Hussein Al-Asadi and Benjamin Peter for help with EEMS
+and general input.
 We would also like to gratefully acknowledge the immense positive impact that Brad McRae \citep[1966--2017;][]{lawler2018tribute}
@@
 We hope this paper helps to carry forward this work.
-\plr{ERIK: also John, Hossein, Desi, ???}
 

### 2018-10-20 isolation_by_coalescence.tex
another pass through

@@
 which is the mean time for a random walk on that network to leave one node, visit the other, and return.
-Here, we describe the differences between commute time
-and coalescence time, which more accurately models genetic distances,
+However, genetic data are more accurately modeled by a different quantity,
+the coalescence time.
+Here, we describe the differences between resistance distance and coalescence time,
 and explore the consequences for inference.
@@
 As motivated by \citet{mcrae2006isolation},
-commute time (i.e., resistance distance) is a more computationally tractable approximation
-to \emph{coalescence time}, a more accurate model of genetic distance data.
+commute time (i.e., resistance distance) is a computationally tractable approximation
+to the more accurate random walk-based model of genetic distance,
+the \emph{coalescence time}.
 Genetic distances between two genomes sampled from the landscape
@@
 This naturally raises the question: 
-Are methods that depend on effective resistances being misled?
+Are methods that depend on effective resistances being misled by model misspecification?
 If so, would using coalescence time do better?
@@
 and found that the two were highly correlated in several test cases.
-However, all these test cases involved symmetric rates of gene flow,
+However, all test cases had symmetric migration
+(the same rate of gene flow in both directions along each edge),
 which is likely not the case in many situations, e.g.,
@@
 although the two methods are conceptually similar,
-commute time inference may get the wrong answer if data are derived from a coalescent process,
-especially in the presence of asymmetry.
+resistance-based inference may get the wrong answer if data are derived from a coalescent process,
+especially in the presence of gene flow asymmetry.
 We also explore the question of identifiability.
@@
 however, the geographical resolution is much coarser.
-A recently-released method 
+(A recently-released method 
 that is based on EEMS but uses haplotype length data and an underlying model of coalescence time
 \citep{alasadi2018estimating}
-is promising, but still produces similarly coarse maps.
+is promising, but still produces similarly coarse maps.)
 One reason for this difference in resolution
@@
 but the two are highly correlated for good theoretical reasons 
-\citet{slatkin1991inbreeding,rousset1997genetic},
+\citep{slatkin1991inbreeding,rousset1997genetic},
 especially for SNP data.
 The two genomes differ at a site only if there has been a mutation in some ancestor on the lineages leading
-from the two genomes back to their most recent common ancestor at that site;
+from the two genomes back to their most recent common ancestor at that site.
+Therefore,
 under an infinite-sites model with average mutation rate $\mu$ per site and per meiosis,
 pairwise divergence divided by $\mu$
-is therefore an unbiased estimate of the mean time to most recent common ancestor,
+is an unbiased estimate of the mean time to most recent common ancestor,
 averaged across the genome \citep{hudson2007variance,ralph2015empirical}.
@@
 We do this in the same way as \citet{petkova2016visualizing},
-by introducing another set of model parameters, $q$, 
-that fill an analogous role to the coalescence rates, $\gamma$.
-Then, we use commute time to approximate genetic distance as
+by introducing the \emph{local diversity} parameters $q_i$,
+and using the model
 \begin{align} \label{eq:commute_approx}
@@
 \end{align}
-The factor of four appears to make commute time more similar to coalescence time:
+The factor of four is necessary to make commute time work on the same time scale as coalescence time:
 the commute time covers the path from $i$ to $j$ twice
@@
 and in half the time, because two particles are moving.
-Furthermore, coalescence time includes within-location diversity
-because particles do not coalesce instantly --
-therefore, commute time is most analogous to the first meeting time.
+% Furthermore, coalescence time includes within-location diversity
+% because particles do not coalesce instantly --
+% therefore, commute time is most analogous to the first meeting time.
 For a conceptual picture of the differences between the coalescence time and commute time,
@@
 then it is not guaranteed that the offdiagonal elements of $G_\epsilon$ are all nonnegative, as required.
-(It is always true that rows of $G_\epsilon$ sum to zero.)
+% (It is always true that rows of $G_\epsilon$ sum to zero.)
 We provide a small concrete example in Appendix \ref{ss:simple_example}.
@@
 
-This suggests that although nonuniqueness of solutions may not be a problem in practice,
-as long as geography is discretized into sufficiently many locations.
+This suggests that nonuniqueness of solutions may not be a problem in practice,
+as long as geography is discretized into sufficiently many locations
+and direct, long-distance migration is disallowed.
 However, another problem appears at finer geographic resolutions:
@@
 For more discussion of related problems, see 
-\citet{epstein2008badtruth,myers2008learn,terhorst2015fundamental}.
+\citet{epstein2008badtruth}, \citet{myers2008learn}, or \citet{terhorst2015fundamental}.
 
@@
 Here, we use
-the exact solutions based on linear algebra developed above
+exact solutions of the equations above
 to develop a Bayesian inference method that accounts for noise in the data
@@
 except that $\epsilon_{ij} = \epsilon_{ji}$.
-(We imagine that $C_{ij}$ is large compared to $\sigma_{\epsilon}$,
+(We assume that $C_{ij}$ is large compared to $\sigma_{\epsilon}$,
 so that resulting values are still positive.)
-Using the calculations above, 
-given $G$ and $\gamma$ we can compute the corresponding coalescence times
+Given $G$ and $\gamma$ we can solve equation \eqref{eqn:C_matrix} 
+to find the corresponding coalescence times
 -- which we call $\mathcal{C}(G, \gamma)$ --
@@
 and instead of $\mathcal{C}(G,\gamma)$ we use $\comdistfn(G, q)$,
-which is the matrix $\comdist$ computed from $G$ and $q$ by solving equation \eqref{eq:commute_approx}.
+which is the matrix $\comdist$ computed from $G$ and $q$ 
+by putting the solution to equation \eqref{eqn:hitting_sum}
+through equations \eqref{eqn:R_from_H} and \eqref{eq:commute_approx}.
 
@@
 Starting locations were chosen by sampling from the prior distribution.
-Before the usual ``burn-in'' phase before sampling for inference begins,
+Before the usual ``burn-in'' phase,
 we included a period of ``pre-burn-in'' which used the same MCMC procedure
@@
 
-We compared inference under coalescence and commute time models
-using both data ``from the model'' and more realistic simulations.
+We compared inference under coalescence time and resistance distance models
+using both data ``from the model'' and from more realistic simulations.
 For the first category, we designed a number of graphs to test particular aspects of inference,
@@
 we produced data by calculating exact coalescence times ($\mathcal{C}(G,\gamma)$),
-to whose entries we added independent Gaussian noise while preserving symmetry,
-as described above.
+to whose entries we added independent Gaussian noise while preserving symmetry.
+We scaled the noise to the mean value of the times themselves,
+so that a ``noise level'' of $\alpha$ refers to data in which the standard deviation of the noise
+was set to $\alpha$ multiplied by the average coalescence time across all pairs of populations.
 Particular parameters and levels of noise are given in the Results.
-
 These simulations are expected to be nearly equivalent to neutral, discrete population-based simulations
-(either forwards-time or coalescent);
-the only difference being that noise terms about the theoretical mean should be somewhat correlated.
+(either forwards-time or coalescent),
+except that noise terms about the theoretical mean should be somewhat correlated.
 
@@
 is as follows.
-Each simulation had around $10^4$ diploid, hermaphroditic individuals, 
+Each simulation had around 6400 diploid, hermaphroditic individuals, 
 living across two-dimensional geography.
@@
 reflected from the boundaries of the habitable area.
-In each simulation, $\sigma_d$ was chosen 
-so that there were an average of 0.25 individuals per $\sigma_d^2$ units of area.
-In order to prevent excessive population clumpiness \citep{felsensten1975pain}, 
-we also model local competition. 
-To do this, we use the same truncated Gaussian kernel to define an ``interaction strength''
-between each pair of nearby individuals;
-then, the fitness of each individual is set to $\exp(-x/40)$, 
-where $x$ is the sum of their interaction strengths with other individuals.
-This reduces fitness for individuals in denser areas.
-Individuals are chosen to reproduce proportional to their fitness.
-Clumpiness is similar to what is seen in 
+To produce a realistic model and prevent excessive population clumpiness \citep{felsensten1975pain}, 
+we model local competition instead of a globally constant population size.
+To do this, 
+we use the same truncated Gaussian kernel to define an ``interaction strength''
+between each pair of nearby individuals, 
+and compute the probability of survival of each individual to the next time step as
+$\min(0.9, 2 / (1 +  5x/12 \pi))$,
+where $x$ is the sum of the interaction strengths of that individual with all other individuals
+within distance $3 \sigma_d$.
+(The value $x$ is an estimate of local population density.)
+This reduces fitness for individuals in denser areas,
+and produces an equilibrium population density
+of close to one individual per $\sigma_d^2$ units of area.
+Clumpiness of the resulting simulations is similar to what is seen in 
 Figure \ref{fig:5x3b_continuous}.
 
-Individuals have 10 pairs of ``chromosomes'' with $2 \times 10^5$ loci each 
+Each diploid individual has one pair of homologous chromosomes with $10^8$ loci each 
 and a recombination rate of $10^{-8}$ per generation per locus.
@@
 which could result in a large number of close relatives being sampled,
-leading to an underestimate of the mean genetic distance for individuals in that location,
 and uniform sampling from the whole area of each grid, 
@@
 by rounding independent exponential random variables up to the nearest one tenth,
-and fixed coalescence rate at 1 for all locations.
-We then added independent noise to each matrix entry (preserving symmetry),
-and assume that we can estimate the amount of noise reasonably well, 
-so used the standard deviation of the noise for $\sigma_\epsilon$ in the likelihood function.
+and fixed coalescence rates at 1 for all locations.
+We then analytically computed mean coalescence times,
+and added independent noise to each pairwise coalescence time.
+We assume that we can estimate the amount of noise reasonably well, 
+so used the true standard deviation of the noise for $\sigma_\epsilon$ in the likelihood function.
 To assess the impact on inference of the number of parameters,
@@
 from 1/1000 to 1/50 of the mean value of $C$ for that grid,
-the mean absolute error 
+the mean absolute error of inferred gene flow rates ($G_{ij}$) 
 for coalescence time inference with a single coalescence rate
 increased from around 0.05 to 0.4.
-Posterior interquartile ranges increased proportionally.
-(although there was substantial variation in accuracy between replicates;
-see Figures \ref{fig:mult_noise} and \ref{fig:mult_noise_iqr}).
-
+Since true values of $G_{ij}$ were of order 1,
+this is a transition from precise to very rough (but still informative) inference.
+Posterior interquartile ranges increased proportionally,
+although there was substantial variation in accuracy between replicates;
+see Figures \ref{fig:mult_noise} and \ref{fig:mult_noise_iqr}.
 Increasing the number of parameters by allowing multiple coalescence rates
@@
 
-More surprisingly, the performance of commute time inference did not depend on the amount of noise --
+More surprisingly, the performance of resistance distance inference 
+did not depend on the amount of noise --
 mean absolute error was around 0.6 across all levels of noise,
@@
 Although symmetry did not affect coalescence time inference,
-commute time inference did substantially worse on asymmetric graphs,
-showing median absolute errors of around 0.8.
+resistance distance inference did substantially worse on asymmetric graphs,
+showing median absolute errors of around 0.8 --
+inferred gene flow rates were almost uncorrelated with true values.
 
@@
 thus greatly reducing the geographic information we get from coalescence times.
-Indeed, coalescence rates had a strong effect on feasability of the inference problem
+Indeed, coalescence rates had a strong effect on feasibility of the inference problem
 (shown in Figures \ref{fig:mult_gam} and \ref{fig:mult_gam_iqr}):
@@
 Again, allowing multiple coalescence rates
-or using commute time inference were highly inaccurate in all cases,
+or using resistance distance inference were highly inaccurate in all cases,
 with mean absolute errors of above 0.5.
@@
 but also the migration rates along all other edges within a reasonable margin of error.
-Commute time inference also identified the barrier,
+Resistance distance inference also identified the barrier,
 but was much less accurate for the other values.
@@
 \emph{uniform}, where all movement rates are 1.0;
-\emph{symmetric}, where movement rates are symmetric and randomly generated as before;
+\emph{symmetric}, where movement rates are symmetric ($G_{ij} = G_{ji}$)
+and randomly generated as before;
 \emph{asymmetric}, where all movement rates are random;
@@
 and extreme in the biased asymmetric case.
-This difference suggests that commute time inference may be strongly misled 
+This difference suggests that resistance distance inference may be strongly misled 
 in the asymmetric and biased situations,
-but in principle it still could be the case that the movement rates that give the best fit of commute time
+but in principle it still could be the case that 
+the movement rates that give the best fit of resistance distance
 to coalescence time data could be close to the actual rates use to generate the data.
@@
 The mean absolute errors for coalescence time inference are 0.12 or less in all cases.
-Commute time inference obtains roughly uniform migration rates
+Resistance distance inference obtains roughly uniform migration rates
 (although varying by a factor of 2) in the uniform case;
 and migration rates noisily correlated with the truth in the symmetric graph.
-However, commute time inferences for the asymmetric and biased graphs are only weakly correlated, 
+However, resistance distance inferences for the asymmetric and biased graphs are only weakly correlated, 
 if at all, to the truth.
-Commute times also drastically overestimate movement rates for the biased case,
+Resistance distance inference also drastically overestimates movement rates for the biased case,
 as we would expect based on the differences discussed above.
@@
 there was substantial noise in the inferred patterns.
-The resistance landscapes as inferred by EEMS show patchy, vertical bands of high resistance:
+However, the resistance landscapes as inferred by EEMS show patchy, vertical bands of high resistance:
 these patterns are commonly seen when attempting to use EEMS, 
@@
 Our analysis above shows that this is an unavoidable consequence of the resistance approximation,
-and our implementation of commute time inference,
+and our implementation of resistance distance inference,
 which allows asymmetric gene flow, shows similar patterns
@@
 which uses coalescence times instead of the resistance distance approximation,
-which provides good estimates of gene flow in a wider range of situations.
+which provides good estimates of gene flow in a wider range of situations
+(Figures \ref{fig:5x3b_post_coalvcom} and \ref{fig:4x4box}).
 This weakness of resistance-based methods is not surprising --
@@
 The similarity of these models to rectangular grids
-may explain why in other work, commute time methods have shown relatively good fit to data
+may explain why in other work, resistance distance methods have shown relatively good fit to data
 simulated with symmetric migration rates.
@@
 as a way of regularizing the inferred landscape,
-hopefully obtaining a map at the best possible resolution supported by the data.
+aiming to obtain a map at the best possible resolution supported by the data.
 
@@
 to identify a barrier in a coalescence-based framework.
-However, these methods requires fairly dense genotyping and a genetic map,
+However, the underlying migration model is symmetric,
+and haplotype length-based methods require fairly dense genotyping and a genetic map,
 something not available for many species.
@@
 and IMa \citep{hey2007integration}
-use likelihoods derived from a coalescent model --
-and use much more of the data than we do.
+use likelihoods for the joint frequency spectrum derived from a coalescent model --
+and so in principle could use much more of the data than our method,
+which only uses pairwise divergences.
 However, because they use locus-specific data rather than an average across loci,
@@
 However, as we discuss above, this fine geographic resolution is in some sense illusory --
-for each fine-resolution map there should be coarser maps that give almost identical coalescence times.
-However, we are not aware of existing theory providing guidance on how to find these.
+for each fine-resolution map there should be coarser maps that give almost identical coalescence times --
+but we are not aware of existing theory providing guidance on how to find these.
 
@@
 but there are still a number of factors that determine the tractability of the problem.
-Observation noise is clearly an issue;
+Observation noise is clearly an issue:
 we found that estimation errors of a few percent was enough to seriously degrade accuracy.
@@
 
-Inference of movement rates, in either a coalescence or commute time framework,
+Inference of movement rates, in either a coalescence time or resistance distance framework,
 is an ill-conditioned problem,
 implying the need for regularization to obtain reliable inference.
-This explains the results of \citet{graves2013current},
+This fact could explain the results of \citet{graves2013current},
 who found a large area of resistance parameters
@@
 of any one parent are typically interspersed with the offspring of many others. 
+However, more investigation with continuous-space models is needed.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{The simplest example}
+\section{The simplest example}
 \label{ss:simple_example}
@@
 \begin{align*}
-    0 &= G\bone \\
-    &= (\diag(x) - \bone \bone^T) H^{-1} x  .
+    0 = G\bone 
+    = (\diag(x) - \bone \bone^T) H^{-1} \bone  .
 \end{align*}

### 2018-10-23 isolation_by_coalescence.tex
update to all new sims

@@
 of close to one individual per $\sigma_d^2$ units of area.
-Clumpiness of the resulting simulations is similar to what is seen in 
+The probability of survival for individuals within $10\sigma_d$ of the edge of the habitat
+was multiplied by a linear factor that scaled from 0 to 1 with distance from the edge.
+Configuration the resulting simulations was similar to what is seen in 
 Figure \ref{fig:5x3b_continuous}.
@@
 will deviate from this depending on both the individuals sampled (``sampling noise'')
-and the stochasticity of the population history itself (``process noise'').
-For more discussion of this latter contribution in other contexts, see
-\citet{wakeley2012genealogies}, \citet{waples2009modelling}, or \citet{ralph2015empirical}.
-Both sources of noise can be substantially larger 
-in populations living across continuous geography
-than in discrete populations
-due to randomness in spatial locations.
-This is hinted at by Figure \ref{fig:5x3b_continuous},
-where we see that the distribution of individuals across the landscape is fairly patchy:
-a barrier might be inferred in a region that has not happened to have a local population recently
-only because of demographic stochasticity.
-This might make inference using a discretization of continuous space considerably more difficult.
+\citep{ashander2018demographic},
+and the stochasticity of the population history itself (``process noise'')
+\citep{wakeley2012genealogies,waples2009modelling}.
+% Both sources of noise can be substantially larger 
+% in populations living across continuous geography
+% than in discrete populations
+% due to randomness in spatial locations.
+% This is hinted at by Figure \ref{fig:5x3b_continuous},
+% where we see that the distribution of individuals across the landscape is fairly patchy:
+% a barrier might be inferred in a region that has not happened to have a local population recently
+% only because of demographic stochasticity.
+% This might make inference using a discretization of continuous space considerably more difficult.
 
@@
 we compared mean genetic distances, calculated in the same way,
-between nonoverlapping sets of samples from the same simulation,
-and between different simulations.
+between two nonoverlapping sets of samples from each of three identical simulations.
 These simulations were done using SLiM as described in the Methods.
-We found that process noise was much larger than sampling noise --
-as shown in Figure \ref{fig:h_val_comp},
-discrepancies between simulations are about five times that seen
-between different samples from the same simulation.
-An implication of this is that uncertainties in genetic distance
-estimated from the data will be (possibly large) underestimates,
-a fact also pointed out for effective population size by \citet{waples2009modelling}.
-
-\begin{figure}
-\centering
-    \includegraphics[width=0.8\textwidth]{figs/h_val_comp}
-\caption{
-    Comparison of mean coalescence times between samples and between replicates
-    for a $4 \times 4$ discretization of a square landscape with uniform migration.
-    All values are plotted against 
-    mean coalescence times calculated using the first set of individuals from the first simulation.
-    Circles show values calculated from a second (non-overlapping) set of individuals 
-    from the same simulation:
-    without sampling noise, all would fall on the $y=x$ line.
-    The other symbols are calculated from two independent simulations:
-    without process noise, these should deviate from $y=x$ no more than the circles.
-    The mean percent absolute differences between each set of coalescence times
-    and the reference values are
-    1.8\% for the second set of individuals from the first simulation,
-    13.2\% for the second simulation, 
-    and 7.3\% for the third simulation.
-    } \label{fig:h_val_comp}
-\end{figure}
-
-Although we found substantial process noise
-with a large, flat landscape with uniform, unbiased migration,
-many biological processes are expected to increase between-replicate stochasticity.
-To demonstrate this, we simulated a square, flat landscape as above,
+We found that with these parameter values, process and sampling noise were of similar magnitude,
+contributing 6.7\% and 13.8\% of the variance, respectively
+(Supplemental Figure \ref{fig:h_val_comp}).
+% An implication of this is that uncertainties in genetic distance
+% estimated from the data will be (possibly large) underestimates,
+% a fact also pointed out for effective population size by \citet{waples2009modelling}.
+
+\paragraph{Biased migration}
+% Although we found substantial process noise
+% with a large, flat landscape with uniform, unbiased migration,
+% many biological processes are expected to increase between-replicate stochasticity.
+To test inference on a simple, continuous landscape with bias, 
+we simulated a square, flat landscape as above,
 but with \emph{biased migration}: 
@@
 % (or 0.05\% of the edge length of the landscape)
-This produces reverse-time gene flow down and to the left.
-As a population genetics model,
-this case is similar to a population expansion,
-replacing the leading edge of the expansion with the corner of landscape 
-that migration bias leads away from.
-In this situation, as in an expansion, 
-individuals at the edge may have disproportionately more offspring,
-leading to low genetic diversity and strong drift \citep{edmonds2004mutations,hallatschek2010life}.
-This results in substantially more noise in the data,
-clearly visible in the isolation by distance plots of Figure \ref{fig:ibd_comp}. 
-Despite this large contribution of process noise,
-coalescence time inference correctly infers the bias on a broad scale:
-using both a $2 \times 2$ and a $4 \times 4$ grid,
+This again produces reverse-time gene flow down and to the left.
+% As a population genetics model,
+% this case is similar to a population expansion,
+% replacing the leading edge of the expansion with the corner of landscape 
+% that migration bias leads away from.
+% In this situation, as in an expansion, 
+% individuals at the edge may have disproportionately more offspring,
+% leading to low genetic diversity and strong drift \citep{edmonds2004mutations,hallatschek2010life}.
+% This results in substantially more noise in the data,
+% clearly visible in the isolation by distance plots of Figure \ref{fig:ibd_comp}. 
+% Despite this large contribution of process noise,
+Coalescence time inference correctly infers the bias on a broad scale:
+using a $4 \times 4$ grid,
 inferred gene flow rates down and to the left are consistently much larger
-than those up and to the right 
-(Supplemental Figures \ref{fig:2x2grid_asb}, \ref{fig:grid_bias_4x4_1}, 
-and \ref{fig:post_dists_bias_4x4_1}).
-However, inferred rates show substantial heterogeneity,
+than those up and to the right;
+resistance distance-based results show no clear pattern
+(Supplemental Figure \ref{fig:grid_bias_4x4_1}).
+However, we do not obtain the clear picture of the landscape
+we saw for the corresponding discrete-population model
+(bottom-right plot, Figure \ref{fig:4x4box}):
+inferred rates show substantial heterogeneity,
 likely due to stochasticity of recent demography.
-
-\begin{figure}
-\centering
-     \includegraphics[width=0.75\textwidth]{figs/ibd_comp}
-    \caption{
-        Genetic distance against geographic distance
-        between all pairs of sampled individuals
-        on a square landscape with 
-        \textbf{(above)} uniform migration
-        and \textbf{(below)} biased migration.
-        The horizontal lines in the latter plot
-        are likely from comparing genomes of individuals from
-        families that have recently increased in size and geographical spread, 
-        resulting in many pairs of genomes with similar genetic distance 
-        but varying geographic distance.
-    \label{fig:ibd_comp}
-    }
-\end{figure}
 
@@
 \centering
-     \includegraphics[scale=.8]{figs/ind_locs_5x3b_1}
-     \includegraphics[height=0.3\textheight]{figs/grid_5x3b_1}
-     \includegraphics[height=0.3\textheight]{figs/post_dists_5x3b_1_new}
+     \includegraphics{figs/barrier_sample_locations_pretty}
+     \includegraphics[height=0.3\textheight]{figs/barrier_coal_grid}
+     \includegraphics[height=0.3\textheight]{figs/barrier_coal_boxplots}
     \caption{
         \textbf{(top)}
-        Locations of all and selected individuals on the $5 \times 3$ landscape with barriers,
+        Locations of individuals on the $5 \times 3$ landscape with barriers,
         with individuals used to compute the mean genetic distance matrix marked in bold.
-        The red bars show the locations of uninhabitable regions of the landscape,
+        Red bars show uninhabitable regions,
         which are sufficiently thick so that no migration can occur directly across them.
@@
 This weakness of resistance-based methods is not surprising --
-the original papers describing Circuitscape \citep{mcrae2006isolation} and EEMS \citep{petkova2016visualizing}
-present the resistance approximation as a necessary step for computational feasibility --
-but perhaps not widely appreciated.
+the original papers 
+describing Circuitscape \citep{mcrae2006isolation} and EEMS \citep{petkova2016visualizing}
+present the resistance approximation as a necessary step for computational feasibility.
 Results of resistance-based methods should therefore be interpreted with caution:
@@
 Although commute and coalescence times can be quite different,
-as also noted by \citet{mcrae2006isolation}, 
 they are equal (with a particular choice of local diversity)
 if populations are arranged on a ring or torus
-with symmetric migration.
+with symmetric migration
+(as also noted by \citet{mcrae2006isolation}).
 The similarity of these models to rectangular grids
@@
     See Figure \ref{fig:5x3b_grid} to see movement parameter locations.
-    %\plr{TODO: put bottom two panels on top; on the bottom put the same thing for commute time}
 }
@@
 \centering
-     \includegraphics[scale=.8]{figs/4x4_grids}
-    \caption{Plots of the grid structure 
+     \includegraphics[scale=.7]{figs/4x4_grids}
+    \caption{
+        Grid structure and gene flow rates
         for the uniform graph, symmetric graph, the asymmetric graph,
         and the biased asymmetric graph 
-    using the igraph R package.}
+        using the igraph R package.
+    }
     \label{fig:4x4_grids}
@@
 \centering
- \includegraphics[scale=.5]{figs/2x2grid_asb}
+    \includegraphics{figs/all_flat_divergences}
 \caption{
-    Medians of the posterior distributions of the inferred values of $g$ 
-    for the first replicate of the $2 \times 2$ discretization of the square landscape with biased migration.
-    The reverse time gene flow rates out of the bottom left section are very small,
-    indicating that very few individuals living in the bottom left of the landscape
-    have ancestors from other regions of the habitat,
-    consistent with migration biased up and to the right.
-    These inferred rates make it clear how strong the bias 
-    against reverse time gene flow out of the lower left corner is.
-    Hardly any individuals in the lower left area have ancestors
-    from other regions of the landscape.
-    This is to be expected as even though the single generation migration bias is small,
-    having a descendant in a location against the bias after a large number of generations
-    becomes less and less likely as the number of generations increases.
-    \label{fig:2x2grid_asb}
+    Comparison of mean coalescence times between samples and between replicates
+    for a $4 \times 4$ discretization of a square landscape with uniform migration.
+    Between the 16 geographic regions, there are 136 comparisons;
+    these are shown for two nonverlapping samples taken from three independent replicate simulations,
+    each plotted agains the the mean value across all replicates.
+    Colors denote simulation replicate,
+    and point type (circle/triangle) denotes the sample.
+    Without sampling or process noise, all points would fall on the $y=x$ line.
+    Among the six sets of observations,
+    simulation replicate explains 6.7\% of the variance,
+    and spatial sampling explains 13.8\% of the variance.
+    } \label{fig:h_val_comp}
+\end{figure}
+
+\begin{figure}
+\centering
+     \includegraphics{figs/example_flat_ibd}
+    \caption{
+        Genetic distance against geographic distance
+        between all pairs of sampled individuals
+        on a square landscape with uniform migration.
+        Red dots denote mean genetic distances
+        between sampled individuals from a $4 \times 4$ discretization of the landscape,
+        as used for instance in Figure \ref{fig:more_barriers}.
+        % and \textbf{(below)} biased migration.
+        % The horizontal lines in the latter plot
+        % are likely from comparing genomes of individuals from
+        % families that have recently increased in size and geographical spread, 
+        % resulting in many pairs of genomes with similar genetic distance 
+        % but varying geographic distance.
+    \label{fig:ibd_comp}
     }
@@
 \centering
- \includegraphics[scale=.8]{figs/grid_bias_4x4_1}
+    \includegraphics[width=0.45\textwidth]{figs/bias_coal_graph}
+    \includegraphics[width=0.45\textwidth]{figs/bias_commute_graph}
+    \includegraphics[width=0.45\textwidth]{figs/bias_coal_dists}
+    \includegraphics[width=0.45\textwidth]{figs/bias_commute_dists}
 \caption{
-    Medians of the posterior distributions of the inferred values of $g$ 
-    for the first replicate of the $4 \times 4$ discretization 
-    of the square landscape with biased migration.
-    As in the $2 \times 2$ case, 
-    the reverse time gene flow rates out of the bottom left section are very small,
+    \textbf{(Above:)} Posterior medians and \textbf{(below)} posterior distributions
+    of of the inferred values of $g$ 
+    for the $4 \times 4$ discretization 
+    of the square landscape with biased migration,
+    inferred with the
+    \textbf{(left)} coalescence time-based,
+    and \textbf{(right)} resistance distance-based method.
+    Reverse time gene flow rates out of the bottom left section are very small,
     indicating that very few individuals living in that area of the landscape
@@
     in the direction we would expect.
-    Boxplots of the posterior distributions are shown in 
-    Supplemental Figure \ref{fig:post_dists_bias_4x4_1}.
+    In the boxplots,
+    black boxes show posterior distributions of gene flow rates down and to the left,
+    and the adjacent green box corresponds to gene flow rate along the same edge
+    in the opposite direction. 
     }
 \label{fig:grid_bias_4x4_1}
-\end{figure}
-
-\begin{figure}
-\centering
- \includegraphics[scale=.6]{figs/post_dists_bias_4x4_1}
-\caption{Boxplots of the posterior distributions for $g$ 
-for the first replicate of the $4 \times 4$ discretization of the square landscape with biased migration.
-Black boxes are gene flow rates down and to the left 
-with each green box being the corresponding gene flow rate up and to the right.}
-\label{fig:post_dists_bias_4x4_1}
 \end{figure}

### 2018-10-23 sims/plot_barrier_layout.R

@@ -1,0 +1,40 @@
+locs <- read.table("run_032600/pop_500000.trees.locs.tsv", skip=1, header=FALSE)
+names(locs) <- c("x", "y", "z")
+locs <- subset(locs, locs$x > 0.0)
+samples <- read.table("run_032600/pop_500000.trees.0.samples.tsv", header=TRUE)
+
+
+W <- 8.0
+aspectRatio <- 5/3
+wh <- W * sqrt(c(aspectRatio, 1/aspectRatio))
+LOWER_Y <- (W / sqrt(aspectRatio) ) / 3
+UPPER_Y <- 2 * (W / sqrt(aspectRatio) ) / 3
+LOWER_LEFT <- 1 * (W * sqrt(aspectRatio) ) / 5
+LOWER_RIGHT <- 3 * (W * sqrt(aspectRatio) ) / 5
+UPPER_LEFT <- 2 * (W * sqrt(aspectRatio) ) / 5
+UPPER_RIGHT <- 4 * (W * sqrt(aspectRatio) ) / 5
+BARRIER_HALFWIDTH <- 0.2 * 3 / 2
+
+
+pdf(file="run_032600/sample_locations_pretty.pdf", width=5, height=3.5, pointsize=10)
+par(mar=c(2, 2, 2, 0)+.2, mgp=c(1,1,0))
+    plot(locs$x, locs$y, asp=1, type='n',
+         xlab='Position (x)', ylab='Position (y)',
+         main='Individual locations',
+         xaxt='n', yaxt='n')
+    polygon(x=c(0, 0, wh[1], wh[1]), 
+            y=c(0, wh[2], wh[2], 0),
+            col=adjustcolor("yellow", 0.25))
+    polygon(x=c(LOWER_LEFT, LOWER_LEFT, LOWER_RIGHT, LOWER_RIGHT),
+            y=LOWER_Y + c(-1, +1, +1, -1) * BARRIER_HALFWIDTH,
+            col=adjustcolor("red", 0.5),
+            border=NA)
+    polygon(x=c(UPPER_LEFT, UPPER_LEFT, UPPER_RIGHT, UPPER_RIGHT),
+            y=UPPER_Y + c(-1, +1, +1, -1) * BARRIER_HALFWIDTH,
+            col=adjustcolor("red", 0.5),
+            border=NA)
+    segments(x0=rep(0,4), x1=rep(wh[1], 4), y0=seq(0, wh[2], length.out=4))
[…]

### 2018-10-23 sims/plot_divergences.R

@@ -1,0 +1,66 @@
+flat_dirs <- c("run_013233", "run_032104", "run_023222")
+
+read_divs <- function (dir, rep) {
+    divs <- matrix(NA, nrow=640, ncol=640)
+    divs[lower.tri(divs, diag=TRUE)] <- scan(file.path(dir, sprintf("pop_500000.trees.%d.divs.tsv", rep)), skip=1)
+    divs[upper.tri(divs)] <- t(divs)[upper.tri(divs)]
+    samps <- read.table(file.path(dir, sprintf("pop_500000.trees.%d.samples.tsv", rep)), header=TRUE)
+    dists <- sqrt(outer(samps$x, samps$x, "-")^2 + outer(samps$y, samps$y, "-")^2)
+    samps$pop <- as.numeric(cut(samps$x, 4)) + 4 * (as.numeric(cut(samps$y, 4)) - 1)
+    sub_divs <- matrix(NA, nrow=16, ncol=16)
+    for (i in 1:16) {
+        for (j in (i:16)) {
+            sub_divs[i,j] <- sub_divs[j,i] <- mean(divs[samps$pop == i, samps$pop == j], na.rm=TRUE)
+        }
+    }
+    sub_xy <- expand.grid(x=2*(1:4), y=2*(1:4))
+    sub_dists <- sqrt(outer(sub_xy$x, sub_xy$x, "-")^2 + outer(sub_xy$y, sub_xy$y, "-")^2)
+    return(list(divs=divs, samples=samps, dists=dists, sub_divs=sub_divs, sub_dists=sub_dists))
+}
+
+all_subs <- list()
+for (fdir in c(flat_dirs, "run_018268")) {
+    for (repnum in c(0,1)) {
+        x <- read_divs(fdir, repnum)
+        ut <- upper.tri(x$divs, diag=TRUE)
+        sub_ut <- upper.tri(x$sub_divs, diag=TRUE)
+        all_subs <- c(all_subs, list(x$sub_divs[sub_ut]))
+            png(file.path(fdir, sprintf("ibd.%d.png", repnum)), width=6*300, height=4*300, res=300)
+            par(mar=c(4,4,1,1)+.1)
+            plot(x$dists[ut], x$divs[ut] * 1e5, pch=20, 
+                 xlab='geographic distance',
+                 ylab='genetic distance (per 100Kb)',
+                 cex=0.2, col=adjustcolor('black', 0.5))
+            points(x$sub_dists[sub_ut], x$sub_divs[sub_ut] * 1e5, 
[…]

### 2018-10-23 sims/plot_ibd_comparison.R

@@ -1,0 +1,29 @@
+flat_dir <- "run_013233"
+biased_dir <- "run_018268"
+
+read_divs <- function (dir, rep) {
+    divs <- matrix(NA, nrow=640, ncol=640)
+    divs[lower.tri(divs, diag=TRUE)] <- scan(file.path(dir, sprintf("pop_500000.trees.%d.divs.tsv", rep)), skip=1)
+    divs[upper.tri(divs)] <- t(divs)[upper.tri(divs)]
+    samps <- read.table(file.path(dir, sprintf("pop_500000.trees.%d.samples.tsv", rep)), header=TRUE)
+    dists <- sqrt(outer(samps$x, samps$x, "-")^2 + outer(samps$y, samps$y, "-")^2)
+    samps$pop <- as.numeric(cut(samps$x, 4)) + 4 * (as.numeric(cut(samps$y, 4)) - 1)
+    sub_divs <- matrix(NA, nrow=16, ncol=16)
+    for (i in 1:16) {
+        for (j in (i:16)) {
+            sub_divs[i,j] <- sub_divs[j,i] <- mean(divs[samps$pop == i, samps$pop == j], na.rm=TRUE)
+        }
+    }
+    sub_xy <- expand.grid(x=2*(1:4), y=2*(1:4))
+    sub_dists <- sqrt(outer(sub_xy$x, sub_xy$x, "-")^2 + outer(sub_xy$y, sub_xy$y, "-")^2)
+    return(list(divs=divs, samples=samps, dists=dists, sub_divs=sub_divs, sub_dists=sub_dists))
+}
+
+flat <- read_divs(flat_dir, 0)
+bias <- read_divs(biased_dir, 0)
+
+ut <- upper.tri(flat$divs)
+
+layout((1:2))
+plot(flat$dists[ut], flat$divs[ut], pch=20, cex=0.5, col=adjustcolor('black', 0.5))
+plot(bias$dists[ut], bias$divs[ut], pch=20, cex=0.5, col=adjustcolor('black', 0.5))

### 2018-10-23 isolation_by_coalescence.tex
final!! pass

@@
 and find that inference using commute times
-can dramatically mislead in the presence of biased gene flow.
+can produce misleading results in the presence of biased gene flow.
 We then use forwards-time simulation with continuous geography to demonstrate that
@@
 from their most recent common ancestor.
-The more recent this ancestor is,
-the smaller the genetic distance will be.
 Thanks to recombination, there are a large number of such paths, 
 and genetic distance averages over these.
+The more recent these common ancestors tend to be,
+the smaller the genetic distance is.
 It is reasonable in some situations to model these lineages as random walks across the landscape;
@@
 in rivers \citep{morrissey2009maintenance,sundqvist2016directional,hanks2017modeling},
-source-sink dynamics \citep{dias1996sources},
+with source-sink dynamics \citep{dias1996sources},
 or after population expansions \citep{with2002landscape}.
@@
 coalescence time and resistance distance,
-and develop an inference method for movement rates on a discrete landscape
-using both as estimated from genetic distances.
+and develop a method to infer movement rates on a discrete landscape
+using genetic distances.
 As we will see,
@@
     Illustration of the conceptual differences
-    betweeen coalescence and commute time between states 1 and 3 
+    between coalescence and commute time between states 1 and 3 
     for a continuous time Markov chain with four linearly connected states. 
@@
 This, along with $H_{zz} = 0$, allows computation of $H$
-from the movement rates $G$.
+from the movement rates $G$ \citep{kemeny1983finite}.
 
@@
 making the inverse problem possible in theory but not in practice.
+This implies a fundamental limit to the geographical resolution of inferred maps.
 For more discussion of related problems, see 
@@
 and only movement rates corresponding to edges in the graph are allowed to be nonzero.
-Write $i \sim j$ to mean that $i$ and $j$ are adjacent in the graph,
-that there are $m_G$ edges in the graph,
-and that each of the $n$ populations has its own coalescence rate.
+Suppose there are $m_G$ edges in the graph,
+that each of the $n$ populations has its own coalescence rate,
+and write $i \sim j$ to mean that $i$ and $j$ are adjacent in the graph.
 Then, the resulting log-posterior density is
@@
 was multiplied by a linear factor that scaled from 0 to 1 with distance from the edge.
-Configuration the resulting simulations was similar to what is seen in 
+Patchiness of the resulting simulations was similar to what is seen in 
 Figure \ref{fig:5x3b_continuous}.
@@
 and 14 movement parameters,
-so that with a global coalescence rate there are 15 unknowns
-but with individual coalescence rates there are 20 unknowns,
+so that with a global coalescence rate there are 15 unknowns.
+However, with individual coalescence rates there are 20 unknowns,
 which is close to nonidentifiable.
@@
 As we varied the standard deviation of the noise
-from 1/1000 to 1/50 of the mean value of $C$ for that grid,
+from 1/1000${}^\text{th}$ to 1/50${}^\text{th}$ of the mean value of $C$ for that grid,
 the mean absolute error of inferred gene flow rates ($G_{ij}$) 
@@
 Increasing the number of parameters by allowing multiple coalescence rates
-made the inference problem even harder --
+made the inference problem much harder --
 mean absolute errors no longer depended strongly on the amount of noise added,
@@
 before they coalesce, 
-thus greatly reducing the geographic information we get from coalescence times.
+thus greatly reducing the geographic information we get from coalescence times 
+\citep{wilkins2004separationoftimescales}.
 Indeed, coalescence rates had a strong effect on feasibility of the inference problem
@@
 lowering the coalescence rate from 10 to 0.1
-(always adding noise with standard deviation 1/200th of the mean value)
+(always adding noise with standard deviation 1/200$^\text{th}$ of the mean value)
 increased mean absolute error of coalescence time inference from around 0.05 to 0.5.
@@
 on this graph, with noise standard deviation equal to
-1/1000 of the mean coalescence time (left)
-and 1/100 of the mean coalescence time (right).
+1/1000${}^\text{th}$ of the mean coalescence time (left)
+and 1/100${}^\text{th}$ of the mean coalescence time (right).
 Although increased noise in the data increased uncertainty in the estimated values,
@@
 Since commute time to location $i$ requires paths both to and from $i$,
-but coalescence time to $i$ only requires lineages to leave $i$,
+but coalescence time to $i$ does not,
 if there is a low rate of migration back into $i$,
@@
 coalescence time inference was substantially more accurate in all cases.
-Posterior distributions of movement parameters are shown in Figure \ref{fig:4x4box},
-produced by applying both inference methods with a noise level of 1/200.
+Posterior distributions of inferred movement parameters are shown in Figure \ref{fig:4x4box}.
 The mean absolute errors for coalescence time inference are 0.12 or less in all cases.
@@
 For this reason, we applied coalescence and resistance inference,
-as well as the popular resistance-based method EEMS \citep{petkova2016visualizing},
+as well as the resistance-based method EEMS \citep{petkova2016visualizing},
 to data from simulations on continuous geography.
@@
 % This might make inference using a discretization of continuous space considerably more difficult.
-
 To quantify the relative contributions of the different sources of noise,
@@
 until 500 generations before the present day
-(a model of ``secondary contact'').
+(a model of secondary contact).
 The resulting expansion produces gene flow back away from the center,
@@
 there was substantial noise in the inferred patterns.
-However, the resistance landscapes as inferred by EEMS show patchy, vertical bands of high resistance:
-these patterns are commonly seen when attempting to use EEMS, 
-whose underlying model is inherently symmetric, 
-to describe asymmetric gene flow.
+However, the resistance landscapes as inferred by EEMS show patchy, vertical bands of high resistance.
+These patterns are commonly seen when attempting to use EEMS, 
+whose underlying model of gene flow is symmetric,
+to describe anisotropic gene flow.
 Our analysis above shows that this is an unavoidable consequence of the resistance approximation,
@@
 and so inferences about that value must necessarily be less certain.
-This tradeoff implies a high degree of unavoidable uncertainty.
+This tradeoff implies some degree of unavoidable uncertainty.
 Reproducible, reliable inference will likely require development of new inference methods
-explicitly incorporating continuous geography.
+that explicitly model continuous geography.
 
@@
 the computational complexity of finding commute times is $nN$,
-but coalescence time scales as $N^2$.
-This is because each $N$-vector of hitting times to a particular location can be found independently,
-but the entire $N \times N$ matrix of coalescence times must be found together.
+but coalescence time scales as $N^2$
+(Each $N$-vector of hitting times to a particular location can be found independently,
+but the entire $N \times N$ matrix of coalescence times must be found together.)
 However, as we discuss above, this fine geographic resolution is in some sense illusory --
@@
 ``forgets'' where it started
-(i.e., the mixing time of the Markov chain).
+(i.e., the mixing time of the Markov chain \citep{wilkins2004separationoftimescales}).
 This is the time it takes for the uncertainty in lineage movement
@@
 of any one parent are typically interspersed with the offspring of many others. 
+Natural selection also makes the coalescent process non-Markovian.
+Differences in the strength of linked selection along the genome
+could even cause the statistical behavior of lineages to depend
+on the region of the genome being studied \citep{wang2014isolation,li2016local}.
 However, more investigation with continuous-space models is needed.
@@
 
-To produce a comcrete example, suppose that
+To produce a concrete example, suppose that
 $C_{11} = 1$, $C_{12} = 2$, and $C_{22} = 1.5$ in arbitrary time units.
@@
     Between the 16 geographic regions, there are 136 comparisons;
-    these are shown for two nonverlapping samples taken from three independent replicate simulations,
-    each plotted agains the the mean value across all replicates.
+    these are shown for two nonoverlapping samples taken from three independent replicate simulations,
+    each plotted against the mean value across all replicates.
     Colors denote simulation replicate,
@@
             \textbf{(middle left)} the commute time method and
-            \textbf{(middle right)} the the commute time method.
+            \textbf{(middle right)} the commute time method.
         Last are shown the posterior distributions, for
             \textbf{(bottom left)} the commute time method and
-            \textbf{(bottom right)} the the commute time method.
+            \textbf{(bottom right)} the commute time method.
         \label{sfig:barrier_results}
@@
 Or, it could be the result of a historical range expansion out of the area,
-perhaps out of a more resticted range as climates changed after the last glacial maximum.
+perhaps out of a more restricted range as climates changed after the last glacial maximum.
 The presence of source-sink dynamics would be of great interest for tortoise conservation efforts,

### 2018-10-23 isolation_by_coalescence.tex
added .tex changes

--- 2018-10-23 isolation_by_coalescence.tex
@@
 \centering
-     \includegraphics[scale=1]{figs/5x3b_post_coalvcom}
+     \includegraphics[scale=1]{figs/5x3b_post_coalvcom2}
     \caption{
@@
         compared to true values (in green),
-        using \textbf{(top)} coalescence time and \textbf{(bottom)} resistance distance inference,
-        with \textbf{(left)} low noise and \textbf{(right)} high noise,
+        using \textbf{(left)} coalescence time and \textbf{(right)} resistance distance inference,
+        with \textbf{(top)} low noise and \textbf{(bottom)} high noise,
         See Figure \ref{fig:5x3b_grid} to match parameter labels to positions on the graph.
@@
 \centering
-     \includegraphics[scale=1]{figs/4x4boxplots_paper}
+     \includegraphics[scale=1]{figs/4x4boxplots_paper2}
     \caption{
@@
         inferred using the
-        \textbf{(left)} resistance distances
-        and \textbf{(right)} coalescence times
+        \textbf{(left)} coalescence times
+        and \textbf{(right)} resistance distances
         shown in Figure \ref{fig:RvsC};
@@
     these are shown for two nonverlapping samples taken from three independent replicate simulations,
-    each plotted agains the the mean value across all replicates.
+    each plotted against the the mean value across all replicates.
     Colors denote simulation replicate,
@@
     \textbf{(Above:)} Posterior medians and \textbf{(below)} posterior distributions
-    of of the inferred values of $g$ 
+    of the inferred values of $g$ 
     for the $4 \times 4$ discretization 
@@
     and \textbf{(right)} resistance distance-based method. 
-    Reverse time gene flow rates out of the bottom left section are very small,
-    indicating that very few individuals living in that area of the landscape
-    have ancestors from other regions of the graph.
-    Most, but not all of the other gene flow rates are biased 
-    in the direction we would expect.
     In the boxplots,
-    black boxes show posterior distributions of gene flow rates down and to the left,
+    black boxes show posterior distributions of gene flow rates up and to the right,
     and the adjacent green box corresponds to gene flow rate along the same edge
@@
     \includegraphics[width=0.45\textwidth]{figs/barrier_eems}
+    \includegraphics[width=0.45\textwidth]{figs/barrier_coal_graph}
     \includegraphics[width=0.45\textwidth]{figs/barrier_commute_graph}
-    \includegraphics[width=0.45\textwidth]{figs/barrier_coal_graph}
+    \includegraphics[width=0.45\textwidth]{figs/barrier_coal_dists}
     \includegraphics[width=0.45\textwidth]{figs/barrier_commute_dists}
-    \includegraphics[width=0.45\textwidth]{figs/barrier_coal_dists}
     \caption{
@@
             as estimated using 
-            \textbf{(middle left)} the commute time method and
-            \textbf{(middle right)} the the commute time method.
+            \textbf{(middle left)} the coalescence time method and
+            \textbf{(middle right)} the the resistance distance method.
         Last are shown the posterior distributions, for
-            \textbf{(bottom left)} the commute time method and
-            \textbf{(bottom right)} the the commute time method.
+            \textbf{(bottom left)} the coalescence time method and
+            \textbf{(bottom right)} the the resistance distance method.
         \label{sfig:barrier_results}
@@
     \includegraphics[width=0.45\textwidth]{figs/valley_eems}
+    \includegraphics[width=0.45\textwidth]{figs/valley_coal_graph}
     \includegraphics[width=0.45\textwidth]{figs/valley_commute_graph}
-    \includegraphics[width=0.45\textwidth]{figs/valley_coal_graph}
+    \includegraphics[width=0.45\textwidth]{figs/valley_coal_dists}
     \includegraphics[width=0.45\textwidth]{figs/valley_commute_dists}
-    \includegraphics[width=0.45\textwidth]{figs/valley_coal_dists}
     \caption{
@@
             as estimated using 
-            \textbf{(middle left)} the commute time method and
-            \textbf{(middle right)} the coalescent time method.
+            \textbf{(middle left)} the coalescence time method and
+            \textbf{(middle right)} the the resistance distance method.
         Last are shown the posterior distributions, for
-            \textbf{(bottom left)} the commute time method and
-            \textbf{(bottom right)} the coalescent time method.
+            \textbf{(bottom left)} the coalescence time method and
+            \textbf{(bottom right)} the the resistance distance method.
         \label{sfig:valley_results}
@@
     \includegraphics[width=0.45\textwidth]{figs/expansion_eems}
+    \includegraphics[width=0.45\textwidth]{figs/expansion_coal_graph}
     \includegraphics[width=0.45\textwidth]{figs/expansion_commute_graph}
-    \includegraphics[width=0.45\textwidth]{figs/expansion_coal_graph}
+    \includegraphics[width=0.45\textwidth]{figs/expansion_coal_dists}
     \includegraphics[width=0.45\textwidth]{figs/expansion_commute_dists}
-    \includegraphics[width=0.45\textwidth]{figs/expansion_coal_dists}
     \caption{
@@
             as estimated using 
-            \textbf{(middle left)} the commute time method and
-            \textbf{(middle right)} the coalescence time method.
+            \textbf{(middle left)} the coalescence time method and
+            \textbf{(middle right)} the the resistance distance method.
         Last are shown the posterior distributions, for
-            \textbf{(bottom left)} the commute time method and
-            \textbf{(bottom right)} the coalescence time method.
+            \textbf{(bottom left)} the coalescence time method and
+            \textbf{(bottom right)} the the resistance distance method.
         \label{sfig:expansion_results}

### 2018-10-23 isolation_by_coalescence.tex
fixed merge repeats

--- 2018-10-23 isolation_by_coalescence.tex
@@
     Between the 16 geographic regions, there are 136 comparisons;
-    these are shown for two nonoverlapping samples taken from three independent replicate simulations,
-    each plotted against the mean value across all replicates.
+    these are shown for two nonverlapping samples taken from three independent replicate simulations,
+    each plotted against the the mean value across all replicates.
     Colors denote simulation replicate,
@@
             as estimated using
-            \textbf{(middle left)} the commute time method and
-            \textbf{(middle right)} the commute time method.
+            \textbf{(middle left)} the coalescence time method and
+            \textbf{(middle right)} the the resistance distance method.
         Last are shown the posterior distributions, for
-            \textbf{(bottom left)} the commute time method and
-            \textbf{(bottom right)} the commute time method.
+            \textbf{(bottom left)} the coalescence time method and
+            \textbf{(bottom right)} the the resistance distance method.
         \label{sfig:barrier_results}

### 2018-10-23 isolation_by_coalescence.tex
minor fixen

@@
 We hope this paper helps to carry forward this work.
+Work on this project was supported by funding from
+the Sloan Foundation and the NSF (under DBI-1262645) to PR.
 
@@
     Between the 16 geographic regions, there are 136 comparisons;
-    these are shown for two nonverlapping samples taken from three independent replicate simulations,
-    each plotted against the the mean value across all replicates.
+    these are shown for two nonoverlapping samples taken from three independent replicate simulations,
+    each plotted against the mean value across all replicates.
     Colors denote simulation replicate,
@@
             \textbf{(middle left)} the coalescence time method and
-            \textbf{(middle right)} the the resistance distance method.
+            \textbf{(middle right)} the resistance distance method.
         Last are shown the posterior distributions, for
             \textbf{(bottom left)} the coalescence time method and
-            \textbf{(bottom right)} the the resistance distance method.
+            \textbf{(bottom right)} the resistance distance method.
         \label{sfig:barrier_results}
@@
             \textbf{(middle left)} the coalescence time method and
-            \textbf{(middle right)} the the resistance distance method.
+            \textbf{(middle right)} the resistance distance method.
         Last are shown the posterior distributions, for
             \textbf{(bottom left)} the coalescence time method and
-            \textbf{(bottom right)} the the resistance distance method.
+            \textbf{(bottom right)} the resistance distance method.
         \label{sfig:valley_results}
@@
             \textbf{(middle left)} the coalescence time method and
-            \textbf{(middle right)} the the resistance distance method.
+            \textbf{(middle right)} the resistance distance method.
         Last are shown the posterior distributions, for
             \textbf{(bottom left)} the coalescence time method and
-            \textbf{(bottom right)} the the resistance distance method.
+            \textbf{(bottom right)} the resistance distance method.
         \label{sfig:expansion_results}

### 2019-01-05 review-responses.tex
review responses

%%%%%%
%%
%%  Don't reorder the reviewer points; that'll mess up the automatic referencing!
%%
%%%%%

\begin{minipage}[b]{2.5in}
  Resubmission Cover Letter \\
  {\it Molecular Ecology Resources}
\end{minipage}
\hfill
\begin{minipage}[b]{2.5in}
    Erik Lundgren \\
    \emph{and} Peter Ralph \\
  \today
\end{minipage}

\vskip 2em

\noindent
{\bf To the Editor(s) -- }

\vskip 1em

We are writing to submit a revised version of our manuscript, 
``Are populations like a circuit? The relationship between isolation by distance and isolation by resistance''.

...

\vspace{5em}

\noindent \hspace{4em}
\begin{minipage}{3in}
\noindent
{\bf Sincerely,}

\vskip 2em

{\bf 
Erik Lundgren and
Peter Ralph
}\\
\end{minipage}

\vskip 4em

\pagebreak
\setcounter{page}{1}

%%%%%%%%%%%%%%
\reviewersection{AE}

\begin{quote}
We received five reviews for your submitted article. While three are quite positive, two others are more critical. In the balance of comments, all reviewers felt that your article should be published in MER, though significant improvements were requested. Therefore, my decision is to request a revision before accepting your article.

Please respond to all major comments raised by the reviewers. In particular, try to address the following points: 1) Revise figures for clarity and legibility. 2) Include an empirical dataset, as suggested by REV 2. 3) Rewrite for a broader audience of biologists (translating some difficult concepts when possible and as appropriate). 4) Clarify aspects of the statistical methods as indicated by several reviewers.
\end{quote}

%%%%%%%%%%%%%%
\reviewersection{1}

\begin{quote}
    The authors present a new Bayesian method to fit movement rates between a
    set of demes to pairwise genetic differentiation, utilizing a MCMC. They first
    state the basic equations, which closely resemble existing theory of stepping
    stone models. They extensively compare their computational method to resistance
    based methods using an own implementation of the latter, both "under the model"
    as well as using various quite realistic forward in time simulations.
    Resistance based methods are much faster and can handle much more demes than
    the method presented here (no case with more than a 5x5 grid is shown).
    Therefore, as the authors point out, this work will likely not contest the use
    of resistance distances for visualizing spatial genetic patterns (which is
    often done in an exploratory way). But importantly, the method here can infer
    asymmetric and biased rates, unlike almost all existing methods, which presents
    a major advance -- and certainly fills a niche.  
\end{quote}

\begin{quote}
    Despite the computational limitations, I believe that this paper will provide
    researchers a good starting point for future development, by both giving a
    review about difficulties and also a first idea and example for an
    implementation.

    Overall, I think this article makes a very valid contribution for Molecular
    Ecology Ressources. The authors put their work into the right context and
    show good knowledge of the relevant literature. The writing is both clear
    and concise. Importantly, they do not shy away from pointing out the limits
    of their method, and the very stringent (often unrealistic) assumptions
    such methods usually make.

    I have several comments, but none of them are ``make-or-break'' or fatal flaws.
    The manuscript is already in a good state. But I hope the authors can use my
    comments to further improve their work.  
\end{quote}

\begin{point}{}
    Clearer point out the current computational limitations of your coalescence
time method. It seems it gets tricky with more than about 30 demes? What is the
run-time then, does the whole thing blow up? Give an idea how big is too big!
\end{point}

\reply{
}

\begin{point}{}
    What is the exact scaling? The bottleneck seems to be calculating the
    coalescence times with equation (7). You only say that it ``can be solved
    with standard linear algebra''. Are you doing Gaussian elimination or
    something else worse than $N^2$? Do you use any speed-ups to
    deal with the sparse nature of G for neighbor-only migration? Is it faster
    than getting the coalescence times by iterating time-discretized
    recursions?  It would be helpful for further progress to describe the exact
    scaling and the exact algorithm you use.
\end{point}

\reply{
}

\begin{point}{}
    The title of the manuscript is a bit misleading. One would not guess that
    the introduction of a method to infer biased and asymmetric gene flow is the
    main point.  ``Isolation by distance'' does not appear often in the manuscript,
    neither its classic literature (Wright, Maryuama, Malecot, Barton etc.) nor
    direct investigation of genetic differentiation with geographic distance. But
    these things are also not the scope of this paper.  
\end{point}

% Citation:
% The Effects of Genetic and Geographic Structure on Neutral Variation
% Brian Charlesworth, Deborah Charlesworth and Nicholas H. Barton
% Annual Review of Ecology, Evolution, and Systematics
% Vol. 34 (2003), pp. 99-125

\reply{
}

\begin{point}{}
    You could write an outlook stating that your method could be used to fit
    parameterized models with few parameters. For instance, in case the putative
    location of a barrier or an area of biased gene flow is known, one could try to
    infer it without fitting ALL movement parameters, but a simple parametric
    model, which would implicitly take care of the regularization.  
\end{point}

\reply{
}

\begin{point}{}
    You use mean absolute error in movement rate to describe the model fit. I
    believe this measure can be quite misleading for exponential movement rates.
    Large rates can be notoriously hard to infer (are you super or super-super well
    connected is almost the same), so it is easy to be off quite a bit, which
    drives the average absolute error in rate. Also in practice, one does not care
    so much about absolute rates but ``relative'' ones. So using a correlation
    measure R**2  of inferred movement rates with actual movement rates would be a
    very useful and natural fit parameter.  
\end{point}

\reply{
}

\begin{point}{}
    You write you use standard methods to check convergence and mixing of the
    MCMC chain (line 182). Given the degeneracy and the high dimensionality,
    where there any problems? E.g., local optima you were stuck sometimes? Or
    was it always pretty straightforward? Maybe write a few sentences on that!
\end{point}

\reply{
}

\begin{point}{\revref}
    $F_ST \neq$ Absolute divergence!!! . $F_ST$ is more a
    co-variance measure, and for instance heavily influence by sampling a batch
    of related individuals (whereas the mean of absolute divergence will be
    unchanged).
\end{point}

\reply{
}

\begin{point}{}
    You have your own implementation of resistance distance methods, by first
    extracting hitting times. However, you allow non-symmetric rates, something
    no existing resistance method would do. Given your observation that there
    is a lot degeneracy for resistance times (adding skew stuff to hitting
    times), this introduces uncertainty that likely cannot be broken such as
    with coalescence methods. This could drive the high uncertainty in Fig 5
    upper right. Given your good results for Fig. 4 for the symmetric case,
    this is a bit surprising. Maybe run one scheme where you FIX to a single
    symmetric movement rate between each pair of demes. This setting would be a
    fairer comparison (and be much closer to what EEMS etc. actually do -- they
    do not even try to infer asymmetric rates).
\end{point}

\reply{
}

\begin{point}{}
    Maybe describe the model you use for you calculation and inference, a
    continuous backward in time Markov model with coalescence rates in demes
    before the actual recursions, and not bit by bit in there.
\end{point}

\reply{
}

\begin{point}{\revref}
    Do you actually use this tensor product for the movement
    rates anywhere? And seems Eq. 7 (which has no Tensor product) is all you
    need.
\end{point}

\reply{
    We included this because we think it's a nice compact way to write, code, 
    and think about the equation,
    and the tensor product equation is actually the one used in the R code.
    It is correct that the equations are equivalent.
}

\begin{point}{Line 115:}
    Not all $ZC$ skew-symmetric would translate to valid Generators 
    $G + Z$. The diagonal elements of G have another condition (they have to be minus
    the sum of all other jumping rates). Would be interesting to explore this a
    bit more -- overall what the dimensionality this degeneracy is (implicit
    function theorem), and also if there are simple toy examples where they pop
    up.
\end{point}

\reply{
}

\begin{point}{Eq. 4:}
    $H$ may not be always invertible ($H$ symmetric and positive off-diagonal
    and 0 on the diagonal is NOT enough). So there might be some pathological
    cases where multiple G link to one $H$, maybe even in some simple toy
    examples.
\end{point}

\reply{
    Good point; we had not justified this.
    Under the standing assumption that the chain is irreducible, 
    $H$ is always invertible.
    We now prove this in the Appendix \llname{rr:H_invertible},
    and have made reference to this fact. \revref
}

\begin{point}{}
    Dimensionality arguments (Line 129 onward) can be a bit misleading when you
    already now the equations have a non-trivial kernel. Set all
    non-neighboring elements of $G$ to 0 might not get rid of parts of the
    kernel. So maybe flag it even more as ``be cautious'' as you already have.
\end{point}

\reply{
}

\begin{point}{}
    You could just call the thing by its name: ``Stepping stone'' model. This
    term has such a rich history in population genetics!
\end{point}

\reply{
}

\begin{point}{}
    In the main text, you never mention that the mean of the exponential
    movement rates for simulations is one, only in a figure. One can also kind
    of see it from the model figures, but it would be nice to have it
    explicitly stated in the main text.
\end{point}

\reply{
}

\begin{point}{}
    In the figures, probably keep model figure and inference results closely
    together. This helps the reader, and makes the figures stand more for
    themselves.
\end{point}

\reply{
}

\begin{point}{}
    Useful result to know: Maryuama’s invariance principle (Nagylaki 1998):
    That the correctly weighted within coalescence time within demes only
    depends on the total population size  (given conservative migration) -- so
    it cannot fluctuate too wildly across demes and movement rates! This fact
    motivates why not fitting too wildly differing within-deme coalescence
    times/diversities.
    % Nagylaki, T., 1998 The expected number of heterozygous sites in a subdivided
    % population. Genetics 149: 1599
\end{point}

\reply{
}

\begin{point}{\revref}
    400 000 generations will look a bit lunatic for someone not in
    popgen who does not know about rescaling. You could give re-scaled, more
    realistic values!
\end{point}

\reply{
}

\begin{point}{Paragraph beginning \revref:}
    Maybe move this section in front of the inference part, as it is useful to
    know the most ``fundamental'' limit of resistance distances.
\end{point}

\reply{
}

\begin{point}{Fig. 7:}
    With EEMS one can see very suspiciously the 4x4 sampling structure - there
    are red dots in almost all 3x3 centers of the sampling squares. Since EEMS
    is quite fast, could you increase the resolution of sampling?
\end{point}

\reply{
}

\begin{point}{}
    You could experiment with visualizations a bit. EEMS got quite popular as
    exploratory visualization tool because it produces visually pleasing plots.
    Do lines with two arrowheads of different sizes near the middle, maybe even
    colored (blue to red) work?
\end{point}

\reply{
}

\reviewersection{2}

\begin{quote}
    This work revisits McRae’s seminal isolation-by-resistance (IBR) model through
    the lens of coalescent theory. The original idea of IBR was to compare patterns of
    genetic distance observed on a landscape with the so-called ``resistance distance''.
    The resistance distance between states i and j in a Markov chain is defined to be
    the ``commute time'': the expected amount of time it takes for the chain to go
    from i to j and back, integrated over all possible paths of the chain. Resistance
    distance comes from electrical circuit theory and is not a model of genetic
    variation across a landscape. Also, it is inherently symmetric in i and j, and
    cannot accomodate asymmetric migration between demes. A more biologically
    realistic model is the structured coalescent. Hence, if data are generated under
    the something resembling the coalescent, then there is scope for improving IBR
    by reducing model misspecification.

    This paper implements such a method and conducts comparative analyses on a
    number of simulated data sets. It seems that in general, this method and the
    resistance-based approaches (as exemplified by EEMS) produce somewhat com-
    parable results so long as migration rates are symmetric. The largest differences
    emerge when gene flow is asymmetric, because this fundamentally violates the
    resistance distance model used by EEMS. Here, the authors show convincingly
    that inferring migration rates using the resistance distance approximation can
    be wildly inaccurate compared to the same estimates obtained using coalescent
    methods. To the extent that gene flow is more likely to be asymmetric in many
    natural populations, this seems like an important finding.

    My overall impression of this work is fairly positive. The paper is very clearly
    written, and the idea of replacing IBR with a method that is more faithful to
    the underlying biology is appealing. The methods section (as well as appendices)
    feature a very nice and readable summary of the mathematical innovations of
    this work, as well as some interesting discussion of issues of identifiability and
    power. Similarly, discussion section in very well done and really strengthens
    the manuscript. It includes a thorough survey of the related work, as well as a
    forthright discussion of advantages and disadvantages of the proposed method,
    and it raises a number of interesting questions and directions for future research.
    Overall, the paper was a pleasure to read.

    I find the main deficiency to be the lack of any analysis of real data. All
    experiments are carried out on simulated data only. This is a puzzling omission.
    The authors went to the trouble of creating an R package (which is good), so
    clearly they intend for the method to be used in practice. Personally, I would
    hesitate to publish the results obtained from a method which had not previously
    been verified to produce reasonable answers in real data. This really feels like
    something the authors ought to have included. It is not necessary that the data
    be novel -- reanalyzing (say) the African elephant dataset in the EEMS paper
    would provide for an interesting comparison (even if it cannot be done at an
    equivalent spatial resolution) and would also go long way towards reassuring
    practitioners.
\end{quote}

\begin{point}{}
    ``Bayesian inference of movement rates'': In their model the authors assume
    that many entries of the migration rate matrix are zero, and place an
    exponential prior on the rest. Is it safe to assume that we will always
    want to constrain rates between discontiguous cells to zero? Also, given
    the valid concerns that they raise elsewhere about identifiability and
    power, I wonder if the authors considered placing a sparse prior on the
    entire G matrix, enabling the model to adapt to the amount of signal while
    shrinking other migration rates to zero. This might have the effect of
    making the resulting non-zero estimates more stable. (I realize that this
    raises potential problems for inference if the underlying graph becomes
    disconnected.)
\end{point}

\reply{
}

\begin{point}{}
    The discussion of ill-conditioning is welcome. From experience, it often
    feels like coalescent-based methods have little power to accurately
    estimate migration unless the rates are large. It might be interesting to
    shift emphasis from estimation to the slightly easier problem of testing:
    how good are these methods at distinguishing between a given null and an
    alternative hypothesis (say, barrier vs. no barrier) for the rate matrix G?
    I think the gain of using a coalescent model might be more apparent in this
    case.
\end{point}

\reply{
}

\begin{point}{}
    This method is computationally limited by equation (7), which has a
    complexity of $O(d^6)$ for the product Markov chain on d demes mentioned a
    few lines below. At least, that is the cost for solving (7) exactly -- but
    given the noisiness of the matrix C, there may be little point in doing
    that anyways. I wonder if the authors have thought at all about obtaining
    an approximate solution to (7), where the error in approximation is chosen
    to in some way to balance out the statistical error inherent in C / D. It
    may also be possible to leverage the fact that the solution G is sparse
    here.  (This is more of a suggestion for future research rather than a
    suggestion for revision.)
\end{point}

\reply{
}

\begin{point}{\revref}
     Do the authors have any other guidance on how to discretize the landscape?
\end{point}

\reply{
}

\begin{point}{}
    Does the issue of having negative entries of G (lines 112-117, Appendix A)
    arise during the course of MCMC sampling, and if so, how is it dealt with?
\end{point}

\reply{
}

\begin{point}{\revref}
     suggest ``commonly'' in place of ``not uncommonly''
\end{point}

\reply{
    Changed.
}

\begin{point}{\revref}
     strongly misleading?
\end{point}

\reply{
    We think this is correct (and reads reasonably well?) as it stands.
}

\reviewersection{3}

\begin{quote}
    This is an interesting paper on a topic that should be of wide interest to
    readers of MER.  The main results are useful: 
    \begin{enumerate}
        \item      landscape genetics models based on resistance theory can be
            misleading if gene flow is asymmetrical
        \item      coalescent-based models deal better with asymmetry but are
            not a panacea for a variety of reasons.
    \end{enumerate}
    On the other hand, I found myself with quite a few questions after reading
    the manuscript.  Raising new questions is not bad in and of itself, but
    most of these were questions that could/should have been clarified in the
    text.  As a consequence, the manuscript is not as effective or useful as it
    might be.  The problem is not in the writing itself, which generally is
    careful and precise.  Instead, the main problem involves translating a very
    technical exercise into something that is accessible to typical readers of
    the journal.  Some examples are described below.

    Although the theme of the paper is relevant to the journal, I think most
    readers of MER would find the manuscript hard going in many places.  For
    example, the text at 102-109 and 136-146 would benefit from translation
    into insights more meaningful to biologists.  The authors do a good job of
    this in discussing results for asymmetric gene flow at lines 296-299.
\end{quote}

\begin{quote}
    The reader also needs to understand the biological consequences of various
    assumptions.  Examples  where this could be improved include:
\end{quote}

\begin{point}{86:}
      if the genome is selectively neutral ...
\end{point}

\reply{
}

\begin{point}{\revref}
     as long as long distance migration is disallowed
\end{point}

\reply{
}

\begin{point}{\revref}
     $C$ is large compared to $\sigma$
\end{point}

\reply{
}

\begin{point}{\revref}
     we can estimate amount of noise well
\end{point}

\reply{
}

\begin{point}{}
    The text at 468ff relating to the assumption of constant pop size and migration
    rates does a good job of explaining these consequences.  Perhaps all
    assumptions could be listed in a single place and their biological
    consequences explained.
\end{point}

\reply{
}

\begin{point}{\revref}
    How exactly is the pairwise genetic distance calculated?  What does ``density
    of sites'' mean?  Is it just the fraction of sites at which two genomes
    differ?  If so it is just pairwise theta as used by Wakeley and others.  
    Eq 1:  what is $R_{ij}$ and how is $q$ calculated?
\end{point}

\reply{
    That's correct; we've reworded this to be more clear.
}

\begin{point}{}
    Eq 1:  what is $R_{ij}$ and how is $q$ calculated?
\end{point}

\reply{
}

\begin{point}{}
    $G$, the movement rate, is clearly a key parameter but is not clearly defined (or
    if it was I missed it).  What are the units of G?  Geographic distance per
    unit time?  What are the time units?  Years?  Generations?  If the latter,
    then does G represent parent-offspring dispersal distance as defined by
    Wright?  Are generations assumed to be discrete?  If not, can a migrant
    leave offspring at various different geographic points?
\end{point}

\reply{
}

\begin{point}{}
    As described, the model deals with groups of individuals (populations) that
    sometimes exchange individuals.  But much of landscape genetics/genomics
    deals with continuously-distributed species that don’t have recognizable
    populations.  Are the results also applicable to scenarios like this?  Why
    or why not?
\end{point}

\reply{
}

\begin{point}{}
    The authors identify two sources of noise: Sampling noise (arising from
    sampling of individuals) and Process noise (stochasticity of population
    history).  Using SLiM, which I gather controls for random variation in the
    realized pedigree and hence eliminates (or only reduces?) Process noise,
    the authors obtain estimates for the relative magnitude of the two sources
    of noise.  This is useful but seems incomplete for two reasons:
    \begin{enumerate}
        \item       A third source of noise is sampling a finite number of
            genes or SNPs.  To some extent one can trade off sampling
            individuals vs loci, but there are limits to how effective that is
        \item       The relative magnitudes of the three sources of error are
            not constant and can vary dramatically depending on parameter
            values.  Process error is inversely related to Ne but otherwise
            fixed in magnitude; it does not change when sampling errors change.
            This means that Process error becomes relatively more important
            when there is a lot of replication in sampling.  In particular,
            genetic sampling error gets very small as the number of SNPs
            becomes arbitrarily large, in which case it is dwarfed by Process
            error (e.g., see Figure 6 in our 2009 Mol Ecol paper cited by the
            authors).
    \end{enumerate}
\end{point}

\reply{
}

\begin{point}{}
    Also in relation to genomics-scale datasets, two general potential
    advantages can be identified:
    \begin{enumerate}
        \item       Arbitrarily large numbers of loci can increase precision
            and statistical power for existing analyses;
        \item       Ability to access information related to genome structure
            and function can provide novel insights and make possible
            qualitatively new types of analyses
    \end{enumerate}
    Lots of people have genomics-scale datasets these days so some discussion of
    this issue seems important.
\end{point}

\reply{
}

\begin{point}{}
    ``Can our method replace resistance-based methods? Perhaps, but the substantial
    uncertainty we saw on graphs with only tens of nodes is indicative of a
    larger problem we face for realistic models. We have seen that
    discretization of space results in substantial modeling error, in part
    because of randomness of geographic sampling and unmodeled process noise
    can lead to overfitting. Partitioning space into a finer grid should help
    with these problems, but tends to make the inference problem itself more
    ill-conditioned: with more connections, changing the value of one
    connection affects coalescence time less, and so inferences about that
    value must necessarily be less certain. This tradeoff implies some degree
    of unavoidable uncertainty.  Reproducible, reliable inference will likely
    require development of new inference methods that explicitly model
    continuous geography. `` This seems like a fairly crucial paragraph.  But
    what exactly is the take-home message for biologists interested in
    practical application of the methodology?
\end{point}

\reply{
}

\begin{point}{}
    Coalescent rates are critical to model performance, which can be poor if the
    rate is too low.  It is obvious that the coalescent rate is inversely
    related to Ne, but in practical terms, what does it mean when the
    coalescent rate is lowered from 10 to 0.1?
\end{point}

\reply{
}

\begin{point}{\revref}
    ``we varied the standard deviation of the noise from 1/1000th to 1/50th of
    the mean value of C for that grid.''  This sounds like a CV of 0.001 to
    0.02 for noise, both of which would be exceedingly small.  If that is not
    what is meant, please clarify.
\end{point}

\reply{
}

\begin{point}{\revref}
    ``if fecundity is variable'' -- on what scale?  Among populations?  Among
    individuals?  Among ages within the same individual? 
\end{point}

\reply{
}

\begin{point}{\revref}
    ``It is not required in this approach to have samples from every spatial
    location ...''  Required for what?  Are they saying that unsampled but real
    (ghost) populations have no effect on precision or bias?
\end{point}

\reply{
}

\begin{point}{}
    Many of the figures are too small to read easily without a magnifying glass.
    Figures 3-5 are large enough but they include miniscule fonts that make it
    impossible to tell exactly what is plotted.  Green and black do not provide
    enough contrast.
\end{point}

\reply{
}

\begin{point}{\revref}
    another limitation of direct observations is that they only provide
    information about movement of individuals but not gene flow, which requires
    successful breeding
\end{point}

\reply{
}

\begin{point}{\revref}
    the ``wrong answer'' compared to what?
\end{point}

\reply{
}

\begin{point}{\revref}
    What is EEMS?
\end{point}

\reply{
}

\begin{point}{\revref}
    What exactly is a more general problem?
\end{point}

\reply{
}

\begin{point}{\revref}
    need a semicolon, not a comma, after ``two genomes''
\end{point}

\reply{
    We reworded this.
}

\begin{point}{184:}
    why quotes for ``from the model''?
\end{point}

\reply{
}

\begin{point}{}
    ``suppose that the spatial arrangement of populations is given by a graph
    ...''  ``Graph'' is a very general term and not clear exactly what is meant
    here.  If the word has a more specific meaning in this context, please
    define it.
\end{point}

\reply{
}

\begin{point}{Fig 6:}
    the barrier is ``sufficiently thick'' to prohibit migration.  ``Thick''
    implies distance, but is that a requirement?  Couldn’t the barrier be
    higher or deeper or less penetrable for a different reason than thickness?
\end{point}

\reply{
}

\reviewersection{4}

\begin{quote}
    This is a polished theoretical and methodological contribution towards
    understanding the genetic processes that affect genetic distance and thus
    connectivity of species across the landscape. The authors make a
    little-appreciated point about considering the coalescent time as a
    predictor of genetic distance as opposed to resistance distance which can
    result in spurious results  where genetic distance is shaped by
    asymmetrical gene flow and coalescent processes.

    There are some points of clarification however that I think need to be made
    so that the paper is accessible and clear to its intended audience, and to
    ensure the paper can be adopted and influence analyses in this area. The
    differences in the highly spatial resistance distance approach versus the
    genetics-based coalescent time-based approach need to be distinguished more
    clearly, as in some ways this may be seen as comparing apples and oranges.
    Issues around timescales of genetic change (effects of landscape vs
    coalescent times) need to be further addressed and elucidated.  The
    differences in the results gained from the approaches are clearly defined
    and presented however.

    The types of questions and the type of demographic scenarios under which
    resistance vs. coalescense based approaches may be an improvement, or
    combined in the same study should be made more clearly. I suggest a
    summarising table or flow chart to this effect, that addresses time scales
    also.
\end{quote}

\begin{point}{Line 7 of abstract and generally:}
This statement is conditional on the characteristics of the data you have, and
    your study aim. If the study aim is to  ask how landscape features affect
    genetic distance, then the coalescent time is not going to explicitly test
    this, or? It needs to be made clear how different study objectives will
    benefit in landscape genetics and under what circumstances a 'more
    accurate' predictor of genetic distance can be gained. It seems appropriate
    to mention that this approach would greatly complement existing landscape
    genetic analysis approaches to see the relative effects of resistance and
    coalescent time. Implications for management decisions based on resistance
    models could be mentioned, as resistance models are intended to inform
    conservation decisions in some cases.
\end{point}

\reply{
}

\begin{point}{Line 26-38:}
    Extremely long sentence, please divide.
\end{point}

\reply{
    Not sure what they mean.
}

\begin{point}{\revref}
    Would it make sense to select genetic markers that are more informative for
    coalescence rates by narrowing the level of variation accepted in the
    model? Would this help error rates justifiably or just bias interpretation?
\end{point}

\reply{
}

\begin{point}{\revref}
    I think it  might be helpful if this point was made earlier and reiterated
\end{point}

\reply{
}

\reviewersection{5}

\begin{quote}
    This article focuses on the inference of gene flows between populations from
    the analysis of georeferenced genetic data. In particular, a new
    coalescent-based technique, which applies to the analysis of pairs of
    sampled individuals, is compared to an approach based on "resistance
    distances" which aim at approximating coalescent distances. This work is
    interesting because resistance distances are at the core of several popular
    techniques in landscape genetics. It is thus important to gain a good
    understanding of how good this phenomenological approximation is compared
    to more mechanistic techniques such as the structured coalescent. The
    article clarifies the connections between the two approaches in a
    convincing manner. It thereby fully deserves to be published in Molecular
    Ecology Resources. There are however a few points (see below) that I think
    should be improved should the authors see fit.
\end{quote}

% Section "Hitting times of Markov chains"

\begin{point}{}
    $Pr(X_{t+e}=y | X_t=y)$ should read $Pr(X_{t+e}=y | X_t=x)$.
\end{point}

\reply{
}

\begin{point}{}
    Regarding Equation 2, I would recommend giving more detail explaining where
    this comes from. Here is a suggestion:
    \begin{align*}
        H_xz &= dt + \sum_y Pr(X_{dt} = y | X_0 = x) H_{yz} \\
             &= dt + \sum_{y != x} Q_{xy} dt H_{yz} + (Q_{xx} dt + 1) H_{xz}
    \end{align*}
\end{point}

\reply{
}

\begin{point}{}
    On a related note, it might be helpful to give more explanations about Equation 6. In
    fact, I am not quite sure I agree with it. Indeed, for $x \neq z$, we have
    \begin{align*}
        C_xz &= dt + {} \\
                & \qquad
             1/2 \sum_y Pr(X_{dt} = y | X_0 = x) C_{yz} + {} \\
                & \qquad
             1/2 \sum_y Pr(X_{dt} = y | X_0 = z) C_{xy}, \\
           &= dt + {} \\
                & \qquad
             1/2 \sum_{y != x} Q_{xy} dt C_{yz} + (Q_{xx} dt + 1) C_{xz} + \\
                & \qquad
             1/2 \sum_{y != x} Q_{zy} dt C_{xy} + (Q_{xx} dt + 1) C_{xz},
    \end{align*}
    where the 1/2 terms correspond to the probability of "selecting" the lineage found in x
    (resp. z) at time 0 to move to y after dt. The last equation above then gives
    \begin{align*}
        -2 = \sum_y Q_{xy} C_{yz} + Q_{zy} C_{xy}
    \end{align*}
    which differs from the one given in the ms. I have probably missed something in the derivation
    used by the authors. Giving a few more explanations here might help avoiding any confusion.
\end{point}

\reply{
}

% Section "Bayesian inference of movement rates"

\begin{point}{}
    I was also a bit puzzled by the likelihood function that is at the core of the
    Bayesian inference proposed here (see Equation 8). The spatial component of
    the data seems completely hidden. It is in fact conveyed by the coalescent
    times $C_{ij}(G,\gamma)$ but a more detailed description of the likelihood
    function where both genetic and spatial information appear explicitly would
    facilitate the understanding of the proposed inference technique.
\end{point}

\begin{point}{}
    Moreover, my understanding is that the joint probability of the pairwise genetic distances
    ($D$) and spatial locations ($L$) given $G$ and $\gamma$, i.e.,
    $p(D,L|G,\gamma)$, is expressed as the product 
    $p(D|L,G,\gamma) \times p(L|G,\gamma)$.  Thanks to Equation (7) the first density is in fact
    $p(D|C)$, where $C$ are the coalescent times. The authors use a normal distribution for $D|C$ but
    nothing is said  about the distribution of $L|G,\gamma$. It is paramount to explicitly define
    every probabilistic distribution in the likelihood of the model used for the inference so as
    to clarify what are the underlying simplifying hypotheses the proposed approach relies on.
\end{point}

\begin{point}{}
    On a related note, my understanding is that the uncertainty affecting
    coalescent times conditional on the movement rates is not taken into
    account in the proposed approach (i.e., knowing $G$, one gets $C$ through
    Equation 7 in a deterministic fashion). I would recommend that the authors
    discuss this point.
\end{point}

\reply{
}

% Section "Identifying a barrier to gene flow"

\begin{point}{}
    The sentence "with noise standard deviation equal to 1/1000th of the mean
    coalescence time (left) and 1/100th of the mean coalescence time (right)"
    should read "with noise standard deviation equal to 1/1000th of the mean
    coalescence time (top) and 1/100th of the mean coalescence time (bottom)"
\end{point}

\reply{
}

% Section "Discussion"

\begin{point}{}
    The authors call for new models to be developed that would properly accommodate
    for individuals distributed in space along a continuum as opposed to
    discrete demes, as is the case with the structured coalescent. The spatial
    Lambda-Fleming Viot model is one of these. It has been studied extensively
    from a mathematical standpoint (see the body of work by Etheridge, Barton
    and Veber amongst others) and has been used for the inference using a
    Bayesian approach (see Guindon, Guo, Welch, Theoretical Population Biology,
    2016; but also Koskela's PhD thesis
    \url{http://wrap.warwick.ac.uk/88065/1/WRAP_Theses_Koskela_2016.pdf}).  The
    spatial Lambda-Fleming-Viot model is mechanistic and enjoys many good
    statistical and biological properties (e.g., it does not suffer from the
    ``clumping'' issue that affects the Wright-Malecot model).
\end{point}

\reply{
}

\begin{point}{}
    Assuming that space is discrete in arguably the biggest weakness of the structured
    coalescent. It is quite often not biologically realistic and breaks down
    mathematically when considering larger numbers of smaller demes. The
    spatial Lambda-Fleming-Viot avoids this important shortcoming and thus
    probably deserves more attention.
\end{point}

\reply{
}

### 2019-01-05 isolation_by_coalescence.tex
submission tweaks
review responses

@@
 % \usepackage{subcaption}
+\usepackage{lineno}
 
@@
 \DeclareMathOperator{\diag}{\mathop{\mbox{diag}}}
-
 \renewcommand{\P}{\mathbb{P}}
@@
 \newcommand{\sgn}{\mathop{\mbox{sgn}}}
-\newcommand{\EE}[1]{\E\!\left[#1\right]}
 \newcommand{\PP}[1]{\P\!\left\{#1\right\}}
-\newcommand{\PPP}[2]{\P_{#1}\!\left\{#2\right\}}
-\newcommand{\EEE}[2]{\E_{#1}\!\left[#2\right]}
-\newcommand{\EEsup}[2]{\E^{#1}\!\left[#2\right]}
-
 \newcommand{\bone}{\mathbf{1}}
@@
 \newcommand{\plr}[1]{{\em \color{blue} #1}}
+
+\newif\ifsubmission
+% \submissiontrue  % for submission
+\submissionfalse % for actual reading
+
+%%% WARNING: right now, with submissiontrue, the supplemental figures dissappear
+%%% and are not referred to!!
+\ifsubmission
+    % \renewcommand{\includegraphics}[2][]{\relax}
+    \usepackage{endfloat}
+    % \renewcommand{\efloatseparator}{\mbox{}} 
+\fi
+
+% Responses to reviews:
+% set this to show line numbers and include responses to reviews or not
+\newif\ifreviewresponses
+\reviewresponsestrue  % include them
+% \reviewresponsesfalse  % don't include them
+\input{review-response-commands}
+\newcommand{\responsefile}{review-responses.tex}  % name of the review reponses file
 
@@
 \maketitle
+
+\linenumbers
 
@@
 (especially of long-distance movement, \citep{cayuelademographic,levin_et_al_seed_dispersal}),
-or even impossible, if some of the populations in question no longer exist.
+or even impossible, if some of the populations in question no longer exist. \revpoint{3}{18}
 
@@
 resistance-based inference may get the wrong answer if data are derived from a coalescent process,
-especially in the presence of gene flow asymmetry.
+especially in the presence of gene flow asymmetry. \revpoint{3}{19}
 We also explore the question of identifiability.
@@
 which allows very fine geographic resolution \citep{shaffer2017desert}.
-Other methods such as EEMS \citep{petkova2016visualizing} infer conductances without such prior information;
+Other methods such as EEMS \citep{petkova2016visualizing} infer conductances without such prior information; \revpoint{3}{20}
 however, the geographical resolution is much coarser.
@@
 is that EEMS uses more computationally intensive Bayesian methods.
-However, we show that this is a more general problem,
+However, we show that this is a more general problem, \revpoint{3}{21}
 both because of unmodeled ``process noise'' involved in discretization of geographical populations,
@@
 There are a number of ways to calculate measures of dissimilarity using genetic data.
-In this paper we use mean pairwise divergence:
-for two genomes, it is calculated as the density of sites at which the two differ.
-It is more common in the literature to use $F_{ST}/(1-F_{ST})$,
+In this paper we use mean pairwise divergence, \revpoint{3}{6}
+which for two genomes is calculated as the proportion of sites in the genome
+at which the two genomes differ. \revpoint{3}{22}
+It is more common in the literature to use $F_{ST}/(1-F_{ST})$, \revpoint{1}{7}
 but the two are highly correlated for good theoretical reasons 
@@
 \end{align}
+(We prove that $H$ is invertible in Appendix~\ref{apx::hitting_calcs}.)
 Therefore, $H$ uniquely determines $G$.
@@
 In practice, we solve this by working with the product Markov chain $(X_t, Y_t)$,
-whose generator matrix is $G \otimes I + I \otimes G$,
+whose generator matrix is $G \otimes I + I \otimes G$, \revpoint{1}{10}
 where $I$ is the identity matrix and $\otimes$ is the Kronecker product.
@@
 as long as geography is discretized into sufficiently many locations
-and direct, long-distance migration is disallowed.
+and direct, long-distance migration is disallowed. \revpoint{3}{2}
 However, another problem appears at finer geographic resolutions:
@@
 but it turns out that it \emph{is} true
-under some restrictive assumptions not uncommonly found in abstract population genetics models.
+under some restrictive assumptions commonly found in abstract population genetics models. \revpoint{2}{6}
 We show in Appendix \ref{sec:com_eq_coal} that
@@
 except that $\epsilon_{ij} = \epsilon_{ji}$.
-(We assume that $C_{ij}$ is large compared to $\sigma_{\epsilon}$,
+(We assume that $C_{ij}$ is large compared to $\sigma_{\epsilon}$, \revpoint{3}{3}
 so that resulting values are still positive.)
@@
 and the coalescence rates, $\gamma$.
-Concretely, suppose that the spatial arrangement of populations is given by a graph,
+Concretely, suppose that the spatial arrangement of populations is given by a graph 
+(i.e., a collection of ``nodes'' connected by ``edges''),
 and only movement rates corresponding to edges in the graph are allowed to be nonzero.
@@
 
-It is not required in this approach to have samples from every spatial location,
+It is not required in this approach to have samples from every spatial location, \revpoint{3}{16}
 allowing inference on incompletely sampled spatial discretizations.
@@
 and added independent noise to each pairwise coalescence time.
-We assume that we can estimate the amount of noise reasonably well, 
+We assume that we can estimate the amount of noise reasonably well,  \revpoint{3}{4}
 so used the true standard deviation of the noise for $\sigma_\epsilon$ in the likelihood function.
@@
 Unsurprisingly, coalescence time inference became less accurate as the amount of noise increased.
-As we varied the standard deviation of the noise
+As we varied the standard deviation of the noise \revpoint{3}{14}
 from 1/1000${}^\text{th}$ to 1/50${}^\text{th}$ of the mean value of $C$ for that grid,
@@
 Lower coalescence rates mean that lineages wander about the landscape for longer
-before they coalesce, 
+before they coalesce,  \revpoint{4}{3}
 thus greatly reducing the geographic information we get from coalescence times 
@@
 Since lineages quickly move to the center and coalesce regardless of starting position,
-coalescence time (and genetic diversity) will be relatively low between all individuals.
+coalescence time (and genetic diversity) will be relatively low between all individuals. \revpoint{4}{4}
 Commute time between the two outer locations, on the other hand,
@@
 Differences between true mean coalescence times ($C$) and mean resistance distances ($\comdist$)
-are shown in Figure \ref{fig:RvsC}:
+are shown in Figure \ref{fig:RvsC}: \revpoint{1}{19}
 they are fairly small in the uniform and symmetric cases, moderate in the asymmetric case, 
 and extreme in the biased asymmetric case.
-This difference suggests that resistance distance inference may be strongly misled 
+This difference suggests that resistance distance inference may be strongly misled \revpoint{2}{7}
 in the asymmetric and biased situations,
@@
 in absence of a good \textit{a priori} idea of where barriers are,
-this step might not be straightforward.
+this step might not be straightforward. \revpoint{2}{4}
 % For instance, we may wonder whether a river or other geographical feature
@@
 \centering
-     \includegraphics{figs/barrier_sample_locations_pretty}
+     \includegraphics{{figs/barrier_sample_locations_pretty.fonts}.pdf}
      \includegraphics[height=0.3\textheight]{figs/barrier_coal_grid}
@@
 Next, we simulated a recent \textbf{expansion},
-in which the middle 4/5 of the landscape was uninhabitable for 400,000 generations,
+in which the middle 4/5 of the landscape was uninhabitable for 400,000 generations, \revpoint{1}{18}
 until 500 generations before the present day
@@
 the dynamics of lineages traced back in time (the coalescent process) may not be. 
-For example, if fecundity is variable and population density is low, 
+For example, if fecundity is variable and population density is low,  \revpoint{3}{15}
 two lineages near each other are more likely to share a recent common origin, 
@@
 
+\ifsubmission
+\processdelayedfloats
+\fi
+
+\newpage
+
 \appendix
@@
 \setcounter{figure}{0}
+\ifsubmission
+    % for endfloat
+    \renewcommand{\thepostfigure}{S\arabic{postfigure}}
+    \setcounter{postfigure}{0}
+\fi
 
@@
 
-Suppose that $G$ is the generator matrix of an irreducible continuous-time Markov chain $X$,
+Suppose that $G$ is the generator matrix of an irreducible continuous-time Markov chain $X$
+on at least two states,
 so that $G_{ij} \ge 0$ for $i \neq j$ and $G \bone = 0$.
+Under these assumptions, there is a unique stationary distribution, $\pi$,
+that satisfies $\pi^T \bone = 1$ and $\pi^T G = 0$.
 Let $\tau_{j} = \inf\{t \ge 0 : X_t = j\}$ and $H_{ij} = \E[\tau_j \,|\, X_0 = i]$.
@@
 \end{align}
-What is $x$?  Well, note that
-\begin{align*}
-    G = (\diag(x) - \bone \bone^T) H^{-1}
-\end{align*}
-and so
-\begin{align*}
-    0 = G\bone 
-    = (\diag(x) - \bone \bone^T) H^{-1} \bone  .
-\end{align*}
-Rearranging, we find that
-\begin{align*}
-    x_i = \frac{ \bone^T H^{-1} \bone }{ (H^{-1} \bone)_i } .
-\end{align*}
-Furthermore,
-the Random Target Lemma \citep{aldous-fill-2014}
+What is $x$?  
+First note that the Random Target Lemma \citep{aldous-fill-2014}
 % https://www.stat.berkeley.edu/users/aldous/RWG/Book_Ralph/Ch2.S2.html
@@
 $(H \pi)_i$ does not depend on $i$,
-i.e., that $H \pi \propto 1$,
-and thus if $H^{-1} \bone = \nu$, 
-then $\pi = \nu / \sum_i \nu_i$.
-This says exactly that $x_i = 1/\pi_i$.
+and so there exists a vector $\nu \propto \pi$
+such that $H \nu = \bone$.
+Multiplying equation \eqref{eqn:GH} by $\nu$, we get that
+\begin{align*}
+    0 = G\bone 
+    = (\diag(x) - \bone \bone^T) \nu ,
+\end{align*}
+which rearranges to
+\begin{align*}
+    x_i = \frac{ \bone^T \nu }{ \nu_i } .
+\end{align*}
+But, since $\nu \propto \pi$,
+it must be that $\pi = \nu / \bone^T \nu$,
+and so $x_i = 1/\pi_i$.
+
+Now, we prove that $H$ is in fact invertible, \llabel{rr:H_invertible}
+by showing that $z^T H \neq 0$ for every nonzero vector $z$.
+First note that $\bone^T H \neq 0$, since the entries of $H$ are nonnegative.
+Now take $z$ orthogonal to $\bone$.
+Since the chain is irreducible,
+the eigenspace associated with the eigenvalue 0 for $G$ has multiplicity one,
+and so any vector $y$ that satisfies $y^T G = 0$ is proportional to $\pi$.
+This implies that there is a vector $y$ such that $y^T G = z^T$.
+Multiplying equation \eqref{eqn:GH} on the left by $y^T$ obtains
+$z^T H = y^T (\diag(1/\pi) - \bone \bone^T)$.
+If $H$ is not invertible, then there is a $z$ for which this is zero,
+i.e., $y_i / \pi_i = \sum_j y_j$ for every $i$.
+But, this only holds if $y = \pi$, which was disallowed,
+because then we would have $z = 0$.
+Therefore, $H$ is invertible.
 
@@
 \centering
-    \includegraphics{figs/all_flat_divergences}
+    \includegraphics{{figs/all_flat_divergences.fonts}.pdf}
 \caption{
@@
 
+\ifsubmission
+\processdelayedfloats
+\fi
+
+\newpage
+
+\includereviews
+
 \end{document}

### 2019-01-11 review-responses.tex
review-responses draft1

@@
 \reply{
+    This is something we were considering doing. 
+    We decided to allow symmetric migration in the resistance distance case 
+    in order to make the two methods as similar as possible, 
+    although it could arguably be more useful to compare them
+    when they are implemented how they would be in practice.
 }
@@
 \reply{
+    Thank you for catching this.
+    You are correct that in addition to $ZC$ being skew-symmetric,
+    the rows of $Z$ must also sum to zero.
+    The second part is also a good point, 
+    but probably beyond the scope of this paper.
 }
@@
 \reply{
+    More samples could be added for the EEMS runs, 
+    but doing so would have the drawback 
+    that the two methods would no longer be using the same data for inference. 
+    We could add a note pointing out that the $3 \times 3$ grid of slow 
+    moving spaces is an artifact.
 }
@@
 \reply{
+    Whether or not it is reasonable to assume there is no long distance migration
+    depends on the population under study. 
+    There are cases where it might be a good idea to allow it.
+    Putting a sparse prior on all of $G$ is a good idea for future improvements, 
+    although it would probably be appropriate for the prior probability of a value
+    to be non-zero to be much higher for adjacent demes as compared to non-adjacent ones, 
+    which would also likely prevent the graph from becoming disconnected very often.
+    Since disconnecting the graph would result in some infinite coalescence times, 
+    these steps would be rejected anyway.
 }
@@
 \reply{
+    Solving (7) directly for $G$ and $\gamma$ with a noisy $C$ is extremely difficult
+    even in cases small enough to just use least squares; 
+    it typically resulted in some negative values unless there is very little noise in $C$.
+    In order to solve for $C$ from $G$ and $\gamma$, 
+    we briefly experimented with the Jacobi method. 
+    While it seemed to converge fairly reliably, 
+    at the grid sizes where it was computationally advantageous to use it over solving exactly, 
+    it would still take a very long time for the MCMC to converge
+    due to the large number of parameters.
+    Devoloping a way to reduce the number of parameters in a way similar to EEMS
+    while still allowing asymmetric migration is a good direction for future implementations.
 }
@@
 \reply{
+    Aside from making decisions based on prior knowledge of the geography, 
+    the sampling locations might lend themselves to being grouped together in a certain way.
 }
@@
 \reply{
+    No. The proposal distribution includes taking the absolute value of parameters.
 }

### 2019-02-16 isolation_by_coalescence.tex
ack cronk

@@
 to Brad Shaffer and Evan McCartney-Melstad for discussions about landscape modeling,
-and to John Novembre, Hussein Al-Asadi and Benjamin Peter for help with EEMS
-and general input.
+to John Novembre, Hussein Al-Asadi and Benjamin Peter for help with EEMS
+and general input,
+and to Quentin Cronk and Armando Geraldes for consultations about the \emph{Populus} results.
 We would also like to gratefully acknowledge the immense positive impact that Brad McRae \citep[1966--2017;][]{lawler2018tribute}

### 2019-02-28 review-responses.tex
more review responses

@@
 \reply{
+    The runs with 16 demes take around 10 hours with the number of iterations we used.
+    30 is probably a good limit for the current implementation. 
+    There are changes that could be made to make it more efficient for larger systems, 
+    but at that point the ill conditioning problem would be more severe 
+    because there would be more parameters, 
+    so we would probably have to do something similar to EEMS, 
+    where movement rates are the same within regions, 
+    but in a way that still allowed asymmetric migration.
 }
@@
 \reply{
+    In the current implementation, 
+    equation (7) is solved using the solve(A,b) in R. 
+    The actual method varies depending on the specific linear algebra library used, 
+    but it is typically Gaussian elimination or LU decomposition, 
+    both of which are $\mathcal{O}(N^3)$ 
+    (unfortunately the decomposition is different each step;
+    if it were the same it would be $\mathcal{O}(N^2)$).
+    Since $n$ demes gives you order $n^2$ equations and unknowns, 
+    the time complexity is 
 }
@@
 \reply{
+    Convergence could be slow, and the likelihood surface is often ridgy, 
+    but we did not see too many problems with getting stuck in completely wrong
+    areas of state space.
+    This is partly how ``pre-burn-in'' helps.
 }
@@
 \reply{
+    added parenthetical clarification
 }
@@
 \reply{
+    Yes, this is one of the reasons that continuous space was used 
+    in the forwards time simulations. 
+    As long as the landscape behaves as if it is fairly continuous
+    and changing the boundary does not move a large number of individuals 
+    from one side to the other, 
+    the exact boundary in the discretization should not have a large effect.
+    Putting two isolated subpopulations into a single group would also be problematic.
 }
@@
 \reply{
+    Assuming that the movement parameters are of order one, 
+    a coalescence rate of 10 means that 
+    lineages would be more likely to coalesce the first time they are in the same location
+    and a coalescence rate of 0.1 means that lineages would likely meet many times
+    before finally coalescing. 
+    For example, if the gene flow rates out of a deme for both lineages sum to 8
+    (rate of 1 for each direction for each lineage), 
+    the chance of the lineages coalescing before one of them leaves 
+    would be 10/18 (~56\%) with a coalescence rate of 10 
+    but only 0.1/8.1 (~1.2\%) with a coalescence rate of 0.1.
 }
@@
 \reply{
+    In practice, the more locations with samples you have, the better the result will be.
+    We would not typically recommend doing inference that includes unsampled demes
+    if it could be avoided.
+    However, in the theoretical case of a population fitting the model perfectly
+    and knowing the other genetic distance values perfectly, 
+    it is possible to solve exactly for paramters relating to the unsampled populations
+    as long as there are as many unique equations as unknowns.
+    In a real situation, if data is very good for the sampled locations, 
+    inference should be reasonable for the unsampled ones.
 }

### 2019-02-28 isolation_by_coalescence.tex
more review responses

@@
 as long as geography is discretized into sufficiently many locations
-and direct, long-distance migration is disallowed. \revpoint{3}{2}
+and direct, long-distance migration is disallowed \revpoint{3}{2}
+(i.e., an individual is unlikely to live many demes away from their parents).
 However, another problem appears at finer geographic resolutions:

### 2019-03-02 review-responses.tex
responses during meeting

@@
 We are writing to submit a revised version of our manuscript, 
-``Are populations like a circuit? The relationship between isolation by distance and isolation by resistance''.
+``Are populations like a circuit? 
+Comparing isolation by resistance to a new coalescent-based method.''
+(formerly titled ``Are populations like a circuit? The relationship between isolation by distance and isolation by resistance'').
 
@@
 \reply{
+    We do think a big point of the manuscript is discussion of the drawbacks of resistance-based methods,
+    but we agree with the point about isolation by distance.
+    We've retitled it: 
+    ``Are populations like a circuit? 
+    Comparing isolation by resistance to a new coalescent-based method.''
 }
@@
 \reply{
+    This is a good thing to emphasize. We've added a bit to this effect. \revref
 }
@@
 \reply{
+    We've added a bit more to our big-picture overview that we hope achieves
+    the goal of setting the stage appropriately. \revref
 }
@@
     Do you actually use this tensor product for the movement
-    rates anywhere? And seems Eq. 7 (which has no Tensor product) is all you
+    rates anywhere? And seems Eq.\ 7 (which has no tensor product) is all you
     need.
@@
     Dimensionality arguments (Line 129 onward) can be a bit misleading when you
-    already now the equations have a non-trivial kernel. Set all
+    already know the equations have a non-trivial kernel. Setting all
     non-neighboring elements of $G$ to 0 might not get rid of parts of the
@@
 \reply{
+    Good point; we've changed the wording here. \revref
 }
@@
 \reply{
+    In our experience, the term ``stepping stone model'' is used in a variety of ways in the literature
+    (e.g., also for all-migrate-to-all island models),
+    so we prefer to describe explicitly the grid that we use.
 }
@@
 \reply{
+    Good point; we've added this. \revref
 }
@@
 \reply{
+    This is a good idea,
+    but we haven't been able to see how to implement in a way that would improve things.
+    Maybe we don't see your idea,
+    but the additional graphs that describe population structure in the Appendix
+    won't be legible if added to existing figures,
+    and we think we already have more than enough figures.
 }
@@
 \reply{
-}
-
-\begin{point}{\revref}
-    400 000 generations will look a bit lunatic for someone not in
+    Good reference! We'd referred to the simpler version of \citet{strobeck1987average},
+    and added this reference. \revref
+}
+
+\begin{point}{\revref}
+    400,000 generations will look a bit lunatic for someone not in
     popgen who does not know about rescaling. You could give re-scaled, more
@@
 \reply{
+    We didn't really intend that number to sound so outlandish,
+    but we think it's more ``quite long'' than ``absurdly long''?
+    We're not sure that mentioning rescaling will help,
+    but have added a reassuring note. \revref
+    (Note that we \emph{did} actually simulate for that many generations;
+    we didn't use rescaling to run a shorter or smaller simulation.)
 }
@@
 \reply{
+    This is a good point, but we'd rather keep the different results having to do with the same model
+    together in the same place.
 }
@@
     that the two methods would no longer be using the same data for inference. 
-    We could add a note pointing out that the $3 \times 3$ grid of slow 
-    moving spaces is an artifact.
+    We've been told that similar patterns show up when EEMS tries to approximate
+    anisotropic migration,
+    so it may not be the sampling pattern.
+    We've added a note about this. \revref
+    (Also, note that these samples are more spread out that people often have in practice:
+    they do occur in the middle three-quarters of each grid square,
+    but they aren't all located at the center.)
 }
@@
     how good are these methods at distinguishing between a given null and an
-    alternative hypothesis (say, barrier vs. no barrier) for the rate matrix G?
-    I think the gain of using a coalescent model might be more apparent in this
-    case.
-\end{point}
-
-\reply{
+    alternative hypothesis (say, barrier vs.\ no barrier) for the rate matrix $G$?
+    I think the gain of using a coalescent model might be more apparent in this case.
+\end{point}
+
+\reply{
+    This is a good point; we've added it to the discussion
+    (and credited you with it). \revref
 }

### 2019-03-02 isolation_by_coalescence.tex
responses during meeting

@@
 
-\title{Are populations like a circuit? The relationship between isolation by distance and isolation by resistance}
+\title{Are populations like a circuit? 
+Comparing isolation by resistance to a new coalescent-based method.}
 \author{Erik Lundgren\footnote{Computational Biology, University of Southern California}, and
-        Peter L. Ralph\footnote{Institute for Ecology and Evolution, University of Oregon}}
+        Peter L.\ Ralph\footnote{Institute for Ecology and Evolution, University of Oregon}}
 \maketitle
@@
 First, we explain this connection,
-and then discuss their computation and application.
+and then discuss their computation,
+and finally develop a method that infers movement rates of the Markov chain
+(which are reverse-time migration probabilities)
+from observations of coalescence time. \revpoint{1}{9}
 
@@
 \end{align}
-(We prove that $H$ is invertible in Appendix~\ref{apx::hitting_calcs}.)
+(We prove that $H$ is invertible in Appendix~\ref{apx::hitting_calcs}.) \revpoint{1}{12}
 Therefore, $H$ uniquely determines $G$.
@@
 for a grid with at least nine nodes (i.e., $3 \times 3$),
-so the system of equations will usually have a unique solution
-for grids larger than this, except in special cases.
+so we would expect the system of equations to usually have a unique solution
+for grids larger than this,
+although some special cases will still have free parameters. \revpoint{1}{13}
 The same calculation holds for commute times 
@@
 \citet{epstein2008badtruth}, \citet{myers2008learn}, or \citet{terhorst2015fundamental}.
+
+One strategy for circumventing ill-conditioning is to impose additional constraints --
+for instance, constraining all population sizes to be the same (something we do below),
+or allowing only two distinct migration rates: one rate across a barrier, 
+and another across ``non-barriers''.
+Such constraints amount to a form of regularization,
+since they look for solutions to the original problem with certain desireable properties
+(e.g., most of the values are the same). \revpoint{1}{4}
 
@@
 We generated movement parameters
-by rounding independent exponential random variables up to the nearest one tenth,
+by rounding independent exponential random variables with mean 1 up to the nearest one tenth, \revpoint{1}{15}
 and fixed coalescence rates at 1 for all locations.
@@
 in which the middle 4/5 of the landscape was uninhabitable for 400,000 generations, \revpoint{1}{18}
-until 500 generations before the present day
-(a model of secondary contact).
+until 500 generations before the present day.
+(This is a model of secondary contact with a long period of isolation,
+but not outside the realm of possibility.)
 The resulting expansion produces gene flow back away from the center,
@@
 to describe anisotropic gene flow.
+The grid-like pattern visible in this and other EEMS results
+may be due to the sampling scheme
+(recall that samples are drawn from the middle three-quarters of each grid square). \revpoint{1}{20}
 Our analysis above shows that this is an unavoidable consequence of the resistance approximation,
@@
 (i.e., nonidentifiability, analogous to a flat likelihood surface).
+An anonymous reviewer suggested
+that although inference of landscapes of movement rates may carry a degree of unavoidable uncertainty,
+the problem of \emph{model comparison} 
+(e.g., between models with and without migration across a certain barrier)
+may still produce more certain answers. \revpoint{2}{2}
 
@@
 This is related to the fact proved by \citet{strobeck1987average} 
+(and generalized by \citet{nagylaki1998expected}) \revpoint{1}{17}
 that $\diag(C)$ is constant and does not depend on movement rates

### 2019-04-03 isolation_by_coalescence.tex
split out torts

@@
 \end{document}
-
-\section*{Application to the Mojave Desert Tortoise}
-
-The Mojave desert tortoise (\textit{Gopherus agassizii})
-lives across much of the Mojave desert of California and Nevada,
-separated from its sister taxa, \textit{Gopherus morofkaii} by the Colorado river.
-We applied our method to genomic distances calculated from
-whole genome sequencing data of 271 individuals sampled from across much of the range,
-reported in \citet{shaffer2017desert}.
-We discretized the landscape into 13 regions based on watershed polygons
-from the Watershed Boundary Dataset \citep{WBD}
-(using WBD8 slightly modified by portions of WBD10 to improve contiguity of regions).
-The resulting discretization,
-overlaid on a map including tortoise sampling locations,
-is shown in Figure \ref{fig:tort_land}.
-Since the regions vary substantially in size, 
-and may vary substantially in population density, 
-we do not assume that coalescence rates are the same everywhere, 
-which gives 13 coalescence rates;
-combined with the 50 migration parameters between each pair of adjacent regions
-we have a total of 63 unknown parameters
-(and 91 equations).
-
-Posterior distributions of the parameters are shown in Figure \ref{fig:tort_post}.
-There is substantial variation in the gene flow rates 
-but not the coalescence rates, 
-with the exception of region 4, which seems to have a relatively slow coalescence rate.
-This is not too surprising given its large area and central location.
-Gene flow rates are low across mountainous boundaries:
-for instance, regions 1 and 4 are separated by the Paiute and Castle mountains,
-and regions 4 and 12 are separated by the Providence and Granite mountains.
-% and regions 1 and 9 are separated by the New York and McCollough mountains.
-On the other hand, there is a large, mostly open bajada connecting areas 9 and 13, 
-and we infer high rates of migration between the two regions.
-The coalescence time model provides a good fit to the data,
-much better than was obtained by commute times, either \plr{in this paper??}
-or using the resistance distance method of \citet{shaffer2017desert}
-(Figure \ref{fig:tort_h_comp}).
-
-The most biologically interesting aspects
-are the strong asymmetries, for instance between region 9 and regions 11, 12, and 13.
-These indicate that resistance-based methods may be strongly misled.
-However, we cannot conclude from this analysis \emph{what} is leading to the observed asymmetries.
-One explanation would be source-sink dynamics:
-for instance, the observed bias in lineage movement could be observed 
-if the tortoise fecundity in the Ivanpah valley in region 9
-is higher than in the colder, higher-elevation regions to the north.
-Or, it could be the result of a historical range expansion out of the area,
-perhaps out of a more restricted range as climates changed after the last glacial maximum.
-The presence of source-sink dynamics would be of great interest for tortoise conservation efforts,
-but it is not clear from this analysis which explanation is more likely.
-Some aspects of the results may even be the produced by sample configuration:
-for instance, perhaps the inferred gene flow between regions 2 and 3 is low
-simply because individuals sampled in region 3 are geographically far from region 2.
-
-\begin{figure}
-\centering
- \includegraphics[width=0.45\textwidth]{figs/fancy_watershed_assignments}
- \includegraphics[width=0.45\textwidth]{figs/tort_graph_results}
-\caption{Left: Locations of the sampled tortoises.
-    \plr{add mountain shadings and ocean; remove bits on the other side of the Colorado}
-    Right: Sample maximum posterior estimates 
-    of movement rates.
-    \plr{can these be maximum posterior estimates instead?}
-    } \label{fig:tort_land}
-\end{figure}
-
-\begin{figure}
-\centering
- \includegraphics[width=0.8\textwidth]{figs/tort_post_g}
- \includegraphics[width=0.8\textwidth]{figs/tort_post_gam}
-\caption{
-    Boxplots of the posterior distributions for 
-    \textbf{(top)} movement rates, and
-    \textbf{(bottom)} coalescence rates,
-    for the Mojave desert tortoise data set.
-    Parameters are labeled as in Figure \ref{fig:tort_land}.
-    } \label{fig:tort_post}
-\end{figure}
-
-\begin{figure}
-\centering
- \includegraphics[scale=.8]{figs/tort_h_comp}
-\caption{
-    Mean coalescence times calculated from the sample maximum likelihood model parameters 
-    plotted against the observed mean coalescence times calculated from the tortoise data, 
-    measured in the mean of pairwise $\pi$.
-    The mean absolute value relative differences between the two is 0.4\%.
-    } \label{fig:tort_h_comp}
-\end{figure}

### 2019-04-17 review-responses.tex
more responses

@@
 \reply{
-    The runs with 16 demes take around 10 hours with the number of iterations we used.
-    30 is probably a good limit for the current implementation. 
+    The runs with 16 demes take around 10 hours with the number of iterations we used
+    (noted in the text, \revref).
+    Thirty demes is probably a good limit for the current implementation. 
     There are changes that could be made to make it more efficient for larger systems, 
     but at that point the ill conditioning problem would be more severe 
-    because there would be more parameters, 
-    so we would probably have to do something similar to EEMS, 
-    where movement rates are the same within regions, 
-    but in a way that still allowed asymmetric migration.
+    because there would be more parameters.
+    We've added a note about this to the discussion. \llname{ll:computation}
 }
@@
     In the current implementation, 
-    equation (7) is solved using the solve(A,b) in R. 
+    equation (7) is solved using \texttt{solve(A,b)} in R. 
     The actual method varies depending on the specific linear algebra library used, 
@@
     both of which are $\mathcal{O}(N^3)$ 
-    (unfortunately the decomposition is different each step;
+    (unfortunately the decomposition is different in each step;
     if it were the same it would be $\mathcal{O}(N^2)$).
-    Since $n$ demes gives you order $n^2$ equations and unknowns, 
-    the time complexity is 
+    For fine discretizations of space, the matrices are sparse,
+    but in our testing, using sparse matrix methods didn't help at this scale.
+    We've added a point about this last thing to the discussion \revref
+    but haven't put in more detail (e.g., plots of scaling with $N$)
+    because it depends on the details (sparsity of the adjacency matrix),
+    and as we discuss above, we think that nonidentifiability is a more serious barrier
+    to finer resolutions.
 }
@@
 \reply{
+    This is a good point, and we did experiment with other measures of model fit.
+    However, in practice the mean absolute error captured patterns as well as other methods.
+    In the interest of not increasing complexity of the paper, we haven't included results
+    from the other methods.
 }
@@
     You write you use standard methods to check convergence and mixing of the
-    MCMC chain (line 182). Given the degeneracy and the high dimensionality,
+    MCMC chain \revreffull{1}{6}. Given the degeneracy and the high dimensionality,
     where there any problems? E.g., local optima you were stuck sometimes? Or
@@
     areas of state space.
-    This is partly how ``pre-burn-in'' helps.
-}
-
-\begin{point}{\revref}
-    $F_ST \neq$ Absolute divergence!!! . $F_ST$ is more a
+    This the reason we introduced the non-standard ``pre-burn-in'' phase.
+}
+
+\begin{point}{\revref}
+    $F_ST \neq$ Absolute divergence!!!  $F_ST$ is more a
     co-variance measure, and for instance heavily influence by sampling a batch
@@
 \reply{
+    We can't agree more!!!!
+    We've added a note about this, although not a lot of detail,
+    because it seems outside the scope of this paper. \revref
+    And, we thought we should mention $F_{ST}$ because it is commonly used
+    in these situations.
 }
@@
     times), this introduces uncertainty that likely cannot be broken such as
-    with coalescence methods. This could drive the high uncertainty in Fig 5
-    upper right. Given your good results for Fig. 4 for the symmetric case,
+    with coalescence methods. This could drive the high uncertainty in Fig.\ 5
+    upper right. Given your good results for Fig.\ 4 for the symmetric case,
     this is a bit surprising. Maybe run one scheme where you FIX to a single
     symmetric movement rate between each pair of demes. This setting would be a
-    fairer comparison (and be much closer to what EEMS etc. actually do -- they
+    fairer comparison (and be much closer to what EEMS etc.\ actually do -- they
     do not even try to infer asymmetric rates).
@@
 \reply{
-    This is something we were considering doing. 
-    We decided to allow symmetric migration in the resistance distance case 
-    in order to make the two methods as similar as possible, 
-    although it could arguably be more useful to compare them
-    when they are implemented how they would be in practice.
+    We considered doing this, but
+    decided to allow asymmetric migration for resistance distance
+    in order to make the two methods more directly comparable.
+    For instance, how would we compare to the truth if the truth is asymmetric?
+    Comparing methods with both restricted to be symmetric 
+    certainly could be useful and informative,
+    but we chose not to include it because it seems not to be the most important comparison
+    (as reality is probably often asymmetric)
+    and we didn't want the paper to get overly complex.
 }
@@
 \begin{point}{Line 115:}
-    Not all $ZC$ skew-symmetric would translate to valid Generators 
+    Not all $ZC$ skew-symmetric would translate to valid generators 
     $G + Z$. The diagonal elements of G have another condition (they have to be minus
@@
     You are correct that in addition to $ZC$ being skew-symmetric,
-    the rows of $Z$ must also sum to zero.
-    The second part is also a good point, 
-    but probably beyond the scope of this paper.
+    the rows of $Z$ must also sum to zero, which we now say. \revref
+    Further exploration of the constraints would be interesting,
+    but we think is beyond the scope of this paper.
 }
@@
 \reply{
-    Good reference! We'd referred to the simpler version of \citet{strobeck1987average},
-    and added this reference. \revref
+    Good reference! We'd already referred to the simpler version of \citet{strobeck1987average},
+    and have now added this reference. \revref
 }
@@
     We're not sure that mentioning rescaling will help,
-    but have added a reassuring note. \revref
-    (Note that we \emph{did} actually simulate for that many generations;
-    we didn't use rescaling to run a shorter or smaller simulation.)
+    but have added a reassuring note. {\revref} 
+    Note that we \emph{did} actually simulate for that many generations;
+    we didn't use rescaling to run a shorter or smaller simulation.
 }
@@
 \reply{
-    This is a good point, but we'd rather keep the different results having to do with the same model
+    This is a good suggestion, 
+    but we'd rather keep the different results having to do with the same model
     together in the same place.
@@
     so it may not be the sampling pattern.
-    We've added a note about this. \revref
-    (Also, note that these samples are more spread out that people often have in practice:
+    We've added a note about this. {\revref}
+    Also, note that these samples are more spread out that people often have in practice:
     they do occur in the middle three-quarters of each grid square,
-    but they aren't all located at the center.)
+    but they aren't all located at the center.
 }
@@
 \reply{
+    We agree -- there's a lot of directions we could move in to improve our package,
+    and hope to work more on this.
 }
@@
     genetic distance observed on a landscape with the so-called ``resistance distance''.
-    The resistance distance between states i and j in a Markov chain is defined to be
+    The resistance distance between states $i$ and $j$ in a Markov chain is defined to be
     the ``commute time'': the expected amount of time it takes for the chain to go
-    from i to j and back, integrated over all possible paths of the chain. Resistance
+    from $i$ to $j$ and back, integrated over all possible paths of the chain. Resistance
     distance comes from electrical circuit theory and is not a model of genetic
-    variation across a landscape. Also, it is inherently symmetric in i and j, and
+    variation across a landscape. Also, it is inherently symmetric in $i$ and $j$, and
     cannot accomodate asymmetric migration between demes. A more biologically
@@
     Overall, the paper was a pleasure to read.
-
+\end{quote}
+
+Thanks very much for the kind words.
+One minor comment: resistance distances are not strictly based in a population genetics model,
+but we think that in practice it is thought of as (an approximation to) such a model
+-- for instance, it is common to see the explanation that it
+``integrates over all possible paths that the organisms might take''.
+I think it is clear that \citet{mcrae2006isolation} saw resistance distance as an approximation
+to the more-accurate coalescence models; we're looking at the effects of this approximation.
+We hope this point comes through in the paper.
+
+\begin{quote}
     I find the main deficiency to be the lack of any analysis of real data. All
@@
 \end{quote}
+
+We've now added analysis of a real dataset, from \citet{geraldes2014landscape}.
+We're quite pleased with the results.
 

### 2019-04-17 isolation_by_coalescence.tex
more responses

@@
 especially for SNP data.
+(However, note that $F_{ST}$ depends on the sampling scheme, while divergence does not.) \revpoint{1}{7}
 The two genomes differ at a site only if there has been a mutation in some ancestor on the lineages leading
@@
 as the original chain.
+For $G + Z$ to be a valid generator, the rows of $Z$ must sum to zero,
+and all offdiagonal entries of $G + Z$ must be nonnegative. \revpoint{1}{11}
 As before, if all entries of $G$ are nonzero, 
@@
 followed by $4 \times 10^6$ iterations which were used to estimate posterior distributions,
-although for small graphs fewer iterations were necessary.
-We used standard methods to assess convergence and mixing.
+which took around 10 hours on a single core of a modern computer with 16 demes. \revpoint{1}{1}
+For small graphs fewer iterations were necessary.
+We used standard methods to assess convergence and mixing. \revpoint{1}{6}
 
@@
 
+%%%%%%%%%%%%%%%
+\subsection*{Data from \textit{Populus trichocarpa}}
+
+We also applied our method to genotyping data from 423 samples of \textit{Populus trichocarpa}
+\citet{geraldes2014landscape}
+and 8 samples of the closely related \textit{Populus balsamifera} (with which it hybridizes),
+from a large region of northwestern North America,
+described in \citet{geraldes2014landscape}.
+These data \citep{geraldes2014data},
+include geographical coordinates and genotypes from a genotyping chip targeting 34K SNPs.
+In addition to preliminary filtering described in \citet{geraldes2014landscape},
+we additionally removed 2,314 SNPs with more than 5\% missing data, retaining 30,756 SNPs,
+and two individuals, one with more than 15\% missing data and one that is much further east
+than all the other samples.
+We then computed pairwise divergence between the remaining samples.
+We discretized the landscape manually into nine regions,
+preserving species boundaries and choosing divisions along gaps in sampling
+to keep geographic areas roughly equal.
+The numbers of samples included in each discretized region varied from 1 to 268,
+but our method does not assume equal sample numbers
+(however, an $F_{ST}$-based method would likely be affected by such large differences in sampling).
+
+%%%%%%%%%%%%%%%%
 \subsection*{Code availability}
@@
 
+%%%%%%%%%%%%%
+\subsection*{Application to \textit{Populus trichocarpa}}
+
+Our discretization of the sampling area for the two poplar species
+is shown in Figure~\ref{fig:populus_results},
+along with
+
 %%%%%%%%%%%%%%%%%%%%
@@
 may still produce more certain answers. \revpoint{2}{2}
+
+Computation time currently limits this method to a discretization of less than 30 populations,
+but nonidentifiability is probably a more serious barrier that must be concurrently addressed
+before scaling the method to finer discretizations of space. \llabel{ll:computation}
+Finer discretizations of space should benefit from the use of sparse matrix methods,
+but in our testing these did not speed up computation at this scale. \revpoint{1}{2}
 

### 2019-04-17 tort_results.tex
more responses

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath,amssymb,amsthm}
\usepackage[hidelinks]{hyperref}
\usepackage{graphicx}
\usepackage[utf8]{inputenc}
\usepackage{xcolor}
\usepackage{natbib}
% \usepackage{subcaption}
\usepackage{lineno}

\bibliographystyle{plainnat}

\newcommand{\comdist}{\widetilde{R}}
\newcommand{\comdistfn}{\widetilde{\mathcal{R}}}

\newcommand{\R}{\mathbb{R}}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\DeclareMathOperator{\var}{\mathop{\mbox{Var}}}
\DeclareMathOperator{\cov}{\mathop{\mbox{Cov}}}
\DeclareMathOperator{\diag}{\mathop{\mbox{diag}}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\given}{\,\vert\,}
\newcommand{\supp}{\mathop{\mbox{supp}}}
\newcommand{\sgn}{\mathop{\mbox{sgn}}}
\newcommand{\PP}[1]{\P\!\left\{#1\right\}}
\newcommand{\bone}{\mathbf{1}}

\newcommand{\plr}[1]{{\em \color{blue} #1}}

\begin{document}

\section*{Application to the Mojave Desert Tortoise}

The Mojave desert tortoise (\textit{Gopherus agassizii})
lives across much of the Mojave desert of California and Nevada,
separated from its sister taxa, \textit{Gopherus morofkaii} by the Colorado river.
We applied our method to genomic distances calculated from
whole genome sequencing data of 271 individuals sampled from across much of the range,
reported in \citet{shaffer2017desert}.
We discretized the landscape into 13 regions based on watershed polygons
from the Watershed Boundary Dataset \citep{WBD}
(using WBD8 slightly modified by portions of WBD10 to improve contiguity of regions).
The resulting discretization,
overlaid on a map including tortoise sampling locations,
is shown in Figure \ref{fig:tort_land}.
Since the regions vary substantially in size, 
and may vary substantially in population density, 
we do not assume that coalescence rates are the same everywhere, 
which gives 13 coalescence rates;
combined with the 50 migration parameters between each pair of adjacent regions
we have a total of 63 unknown parameters
(and 91 equations).

Posterior distributions of the parameters are shown in Figure \ref{fig:tort_post}.
There is substantial variation in the gene flow rates 
but not the coalescence rates, 
with the exception of region 4, which seems to have a relatively slow coalescence rate.
This is not too surprising given its large area and central location.
Gene flow rates are low across mountainous boundaries:
for instance, regions 1 and 4 are separated by the Paiute and Castle mountains,
and regions 4 and 12 are separated by the Providence and Granite mountains.
% and regions 1 and 9 are separated by the New York and McCollough mountains.
On the other hand, there is a large, mostly open bajada connecting areas 9 and 13, 
and we infer high rates of migration between the two regions.
The coalescence time model provides a good fit to the data,
much better than was obtained by commute times, either \plr{in this paper??}
or using the resistance distance method of \citet{shaffer2017desert}
(Figure \ref{fig:tort_h_comp}).

The most biologically interesting aspects
are the strong asymmetries, for instance between region 9 and regions 11, 12, and 13.
These indicate that resistance-based methods may be strongly misled.
However, we cannot conclude from this analysis \emph{what} is leading to the observed asymmetries.
One explanation would be source-sink dynamics:
for instance, the observed bias in lineage movement could be observed 
if the tortoise fecundity in the Ivanpah valley in region 9
is higher than in the colder, higher-elevation regions to the north.
Or, it could be the result of a historical range expansion out of the area,
perhaps out of a more restricted range as climates changed after the last glacial maximum.
The presence of source-sink dynamics would be of great interest for tortoise conservation efforts,
but it is not clear from this analysis which explanation is more likely.
Some aspects of the results may even be the produced by sample configuration:
for instance, perhaps the inferred gene flow between regions 2 and 3 is low
simply because individuals sampled in region 3 are geographically far from region 2.

\begin{figure}
\centering
 \includegraphics[width=0.45\textwidth]{figs/fancy_watershed_assignments}
 \includegraphics[width=0.45\textwidth]{figs/tort_graph_results}
\caption{Left: Locations of the sampled tortoises.
    \plr{remove bits on the other side of the Colorado}
    Right: Sample maximum posterior estimates 
    of movement rates.
    \plr{Note these are \emph{reverse-time} movement rates; we should reverse the arrows.}
    } \label{fig:tort_land}
\end{figure}

\begin{figure}
\centering
 \includegraphics[width=0.8\textwidth]{figs/tort_post_g}
 \includegraphics[width=0.8\textwidth]{figs/tort_post_gam}
\caption{
    Boxplots of the posterior distributions for 
    \textbf{(top)} movement rates, and
    \textbf{(bottom)} coalescence rates,
    for the Mojave desert tortoise data set.
    Parameters are labeled as in Figure \ref{fig:tort_land}.
    } \label{fig:tort_post}
\end{figure}

\begin{figure}
\centering
 \includegraphics[scale=.8]{figs/tort_h_comp}
\caption{
    Mean coalescence times calculated from the sample maximum likelihood model parameters 
    plotted against the observed mean coalescence times calculated from the tortoise data, 
    measured in the mean of pairwise $\pi$.
    The mean absolute value relative differences between the two is 0.4\%.
    } \label{fig:tort_h_comp}
\end{figure}

\end{document}

### 2019-04-18 review-responses.tex
more responses

@@
     You are correct that in addition to $ZC$ being skew-symmetric,
-    the rows of $Z$ must also sum to zero, which we now say. \revref
+    the rows of $Z$ must also sum to zero, which we now say. {\revref}
     Further exploration of the constraints would be interesting,
@@
 ``integrates over all possible paths that the organisms might take''.
-I think it is clear that \citet{mcrae2006isolation} saw resistance distance as an approximation
+We think it is clear that \citet{mcrae2006isolation} saw resistance distance as an approximation
 to the more-accurate coalescence models; we're looking at the effects of this approximation.
@@
 
-We've now added analysis of a real dataset, from \citet{geraldes2014landscape}.
-We're quite pleased with the results.
+We've now added analysis of a real dataset, from \citet{geraldes2014landscape},
+and are quite pleased with the results!
+This has also given us a chance to spend more time talking through how to interpret the results.
+(We didn't get to say this in the paper,
+but initially were worried by the southward bias in gene flow for the inland poplar species,
+since we would expect the opposite if poplars had expanded to the north following the glaciers;
+but then we learned that this species likely expanded from Alaksa, rather than from the south.)
 
@@
     depends on the population under study. 
-    There are cases where it might be a good idea to allow it.
-    Putting a sparse prior on all of $G$ is a good idea for future improvements, 
-    although it would probably be appropriate for the prior probability of a value
-    to be non-zero to be much higher for adjacent demes as compared to non-adjacent ones, 
-    which would also likely prevent the graph from becoming disconnected very often.
-    Since disconnecting the graph would result in some infinite coalescence times, 
-    these steps would be rejected anyway.
+    Putting a sparse prior on all of $G$ is a good idea for future improvement.
+    % although it would probably be appropriate for the prior probability of a value
+    % to be non-zero to be much higher for adjacent demes as compared to non-adjacent ones, 
+    % which would also likely prevent the graph from becoming disconnected very often.
+    (Since disconnecting the graph would result in some infinite coalescence times, 
+    such proposals would be rejected by the MCMC algorithm.)
 }
@@
 \reply{
+    Thanks for the suggestions.
     Solving (7) directly for $G$ and $\gamma$ with a noisy $C$ is extremely difficult
@@
     In order to solve for $C$ from $G$ and $\gamma$, 
-    we briefly experimented with the Jacobi method. 
+    we experimented with the Jacobi method. 
     While it seemed to converge fairly reliably, 
@@
     due to the large number of parameters.
-    Devoloping a way to reduce the number of parameters in a way similar to EEMS
-    while still allowing asymmetric migration is a good direction for future implementations.
+    % Devoloping a way to reduce the number of parameters in a way similar to EEMS
+    % while still allowing asymmetric migration is a good direction for future implementations.
 }
@@
 \reply{
-    Aside from making decisions based on prior knowledge of the geography, 
-    the sampling locations might lend themselves to being grouped together in a certain way.
+    This is a large, thorny issue,
+    closely related to the question of how to translate discrete-population models
+    to continuous geography.
+    We've added a bit more \revref,
+    as well as some more discussion of this in the section on poplar data.
+    \llname{ll:discretization}
 }
@@
 \reply{
-    No. The proposal distribution includes taking the absolute value of parameters.
+    No. The proposal distribution includes taking the absolute value of parameters
+    (a standard method in MCMC to preserve reversibility).
+    We've added this note to the MCMC methods section. \revref
 }
@@
     into insights more meaningful to biologists.  The authors do a good job of
-    this in discussing results for asymmetric gene flow at lines 296-299.
+    this in discussing results for asymmetric gene flow at lines 296-299. % \llname{ll:dense3}
 \end{quote}
+
+Thanks for the encouragement.
+We've tried to improve accessibility in a number of places,
+in particular by adding a ``flag'' at the top of the methods {\llname{ll:flagging}},
+and improving the two bits of text referred to,
+at {\llname{ll:dense1}} and {\llname{ll:dense2}}.
 
@@
 
-\begin{point}{86:}
+\begin{point}{\revref}
       if the genome is selectively neutral ...
@@
 \reply{
+    Reworded.
 }
@@
 \reply{
-    added parenthetical clarification
+    Added parenthetical clarification.
 }
@@
 \reply{
+    Added explanatory note aobut how to estimate noise.
 }
@@
 \reply{
+    The job of enumerating \emph{all} assumptions involved in modeling a real population
+    with a system of discrete demes is daunting.
+    We've put in more discussion of these pitfalls (as described above),
+    but a comprehensive review is beyond the scope of this paper.
 }
@@
 \begin{point}{}
-    Eq 1:  what is $R_{ij}$ and how is $q$ calculated?
-\end{point}
-
-\reply{
+    Eq \eqref{eq:commute_approx}: what is $R_{ij}$ and how is $q$ calculated?
+\end{point}
+
+\reply{
+    We've added a clarifying sentence. \revref
 }
@@
 \reply{
+    A precise definition (in words and equations) is given where it is first introduced, at \revref.
+    We've rewritten it to be hopefully more clear.
 }
@@
 \reply{
-    Yes, this is one of the reasons that continuous space was used 
-    in the forwards time simulations. 
-    As long as the landscape behaves as if it is fairly continuous
-    and changing the boundary does not move a large number of individuals 
-    from one side to the other, 
-    the exact boundary in the discretization should not have a large effect.
-    Putting two isolated subpopulations into a single group would also be problematic.
+    This is an important point that applies not just to our method
+    but to most methods of population genetics inference,
+    and we view this disconnect as a major unsolved problem in the field.
+    Our simulations of continuous space are explicitly designed
+    to look at this problem.
+    We've added substantially more discussion of this point
+    in the interpretation of poplar results \llname{ll:continuous}.
 }
@@
             genes or SNPs.  To some extent one can trade off sampling
-            individuals vs loci, but there are limits to how effective that is
+            individuals vs loci, but there are limits to how effective that is.
         \item       The relative magnitudes of the three sources of error are
             not constant and can vary dramatically depending on parameter
-            values.  Process error is inversely related to Ne but otherwise
+            values.  Process error is inversely related to $N_e$ but otherwise
             fixed in magnitude; it does not change when sampling errors change.
@@
 \reply{
+    We certainly had omitted that source of noise,
+    but have added it to the list now \revref.
+    We agree it can be substantial, but aren't focusing on it here,
+    as the contribution of sampling noise is much less appreciated,
+    particularly in the context of continuous space (see our additional note on this).
+    We're not sure what you mean about SLiM reducing process noise:
+    on the contrary, since we use SLiM to simulate in continuous space,
+    we expect process noise to be larger than in a randomly mating population.
 }
@@
 \reply{
+    Good point, but to keep the paper relatively focused we're not going into this in detail
+    (but perhaps the pointer at {\revref} suffices?).
 }
@@
     require development of new inference methods that explicitly model
-    continuous geography. `` This seems like a fairly crucial paragraph.  But
+    continuous geography.''  This seems like a fairly crucial paragraph.  But
     what exactly is the take-home message for biologists interested in
@@
 \reply{
+    Good question!
+    As written, the take-home was ``wait until we make better methods'',
+    but we've added a better one. \revref
 }
@@
     rate is too low.  It is obvious that the coalescent rate is inversely
-    related to Ne, but in practical terms, what does it mean when the
+    related to $N_e$, but in practical terms, what does it mean when the
     coalescent rate is lowered from 10 to 0.1?
@@
 \reply{
-    Assuming that the movement parameters are of order one, 
-    a coalescence rate of 10 means that 
-    lineages would be more likely to coalesce the first time they are in the same location
-    and a coalescence rate of 0.1 means that lineages would likely meet many times
-    before finally coalescing. 
-    For example, if the gene flow rates out of a deme for both lineages sum to 8
-    (rate of 1 for each direction for each lineage), 
-    the chance of the lineages coalescing before one of them leaves 
-    would be 10/18 (~56\%) with a coalescence rate of 10 
-    but only 0.1/8.1 (~1.2\%) with a coalescence rate of 0.1.
+    We've added more context to this section. \revref
+    % Assuming that the movement parameters are of order one, 
+    % a coalescence rate of 10 means that 
+    % lineages would be more likely to coalesce the first time they are in the same location
+    % and a coalescence rate of 0.1 means that lineages would likely meet many times
+    % before finally coalescing. 
+    % For example, if the gene flow rates out of a deme for both lineages sum to 8
+    % (rate of 1 for each direction for each lineage), 
+    % the chance of the lineages coalescing before one of them leaves 
+    % would be 10/18 (~56\%) with a coalescence rate of 10 
+    % but only 0.1/8.1 (~1.2\%) with a coalescence rate of 0.1.
 }
@@
     ``we varied the standard deviation of the noise from 1/1000th to 1/50th of
-    the mean value of C for that grid.''  This sounds like a CV of 0.001 to
+    the mean value of $C$ for that grid.''  This sounds like a CV of 0.001 to
     0.02 for noise, both of which would be exceedingly small.  If that is not
@@
 \reply{
+    That is what we mean, and although small, we don't think that's unreasonable for genome-scale data
+    under a discrete population model (but see our later discussion of process noise).
+    We realize that many people are working with much smaller number of (ascertained) loci,
+    but there's enough bigger issues here related to continuous geography
+    that we've not explored in that direction.
 }
@@
 \reply{
+    We mean among individuals (and have added this).
 }
@@
     ``It is not required in this approach to have samples from every spatial
-    location ...''  Required for what?  Are they saying that unsampled but real
+    location \ldots''  Required for what?  Are they saying that unsampled but real
     (ghost) populations have no effect on precision or bias?
@@
 \reply{
-    In practice, the more locations with samples you have, the better the result will be.
-    We would not typically recommend doing inference that includes unsampled demes
-    if it could be avoided.
-    However, in the theoretical case of a population fitting the model perfectly
-    and knowing the other genetic distance values perfectly, 
-    it is possible to solve exactly for paramters relating to the unsampled populations
-    as long as there are as many unique equations as unknowns.
-    In a real situation, if data is very good for the sampled locations, 
-    inference should be reasonable for the unsampled ones.
+    We're simply saying that it is not a requirement to \emph{run} the method
+    to have samples in each location --
+    it is easy, in principle, to include ``ghost'' populations.
+    Of course, unsampled populations will affect accuracy of the result,
+    and we've added a note to clarify. \revref
+    % In practice, the more locations with samples you have, the better the result will be.
+    % We would not typically recommend doing inference that includes unsampled demes
+    % if it could be avoided.
+    % However, in the theoretical case of a population fitting the model perfectly
+    % and knowing the other genetic distance values perfectly, 
+    % it is possible to solve exactly for paramters relating to the unsampled populations
+    % as long as there are as many unique equations as unknowns.
+    % In a real situation, if data is very good for the sampled locations, 
+    % inference should be reasonable for the unsampled ones.
 }
@@
 \reply{
+    We sympathize, and have tried to improve this.
+    However, some of this we think is unavoidable --
+    the x-axis labels of figures like Figure~\ref{fig:5x3b_post_coalvcom}
+    cannot be made larger without overlapping.
+    Another approach would be to simply plot true versus observed values,
+    which is equivalent to the current plot without the x-axis labels.
+    We prefer to leave these labels, even if these are only really legible
+    in the digital version (after zooming in).
 }
@@
 \reply{
+    Good point, but we'd like to keep this bit more concise.
 }
@@
 \reply{
+    The truth. We've changed the phrasing. \revref
 }
@@
 \reply{
+    A lengthy name (that we've added). \revref
 }
@@
 \reply{
+    Clarified. \revref
 }
@@
 
-\begin{point}{184:}
+\begin{point}{\revref}
     why quotes for ``from the model''?
@@
 \reply{
+    Removed. \revref
 }
@@
     ``suppose that the spatial arrangement of populations is given by a graph
-    ...''  ``Graph'' is a very general term and not clear exactly what is meant
+    \ldots''  ``Graph'' is a very general term and not clear exactly what is meant
     here.  If the word has a more specific meaning in this context, please
@@
 \reply{
-}
-
-\begin{point}{Fig 6:}
+    Done. \revref
+}
+
+\begin{point}{Fig 6 caption:}
     the barrier is ``sufficiently thick'' to prohibit migration.  ``Thick''
@@
 \reply{
+    Yes, but this is the way we implemented it in the simulation.
 }
@@
 
+We've added more discussion along these lines
+particularly in {\llname{ll:history}} 
+and in relation to analysis of the poplar data.
+We've got quite a few figures already (and have added one for the poplars),
+so haven't tried to add a flow chart.
+
 \begin{point}{Line 7 of abstract and generally:}
@@
     this, or? It needs to be made clear how different study objectives will
-    benefit in landscape genetics and under what circumstances a 'more
+    benefit in landscape genetics and under what circumstances a `more
     accurate' predictor of genetic distance can be gained. It seems appropriate
@@
 \reply{
-}
-
-\begin{point}{Line 26-38:}
+    We're not sure which possible study aims are being referred to,
+    but one possible aim that is not possible currently with coalescence methods
+    is to produce high-resolution maps (as we discuss, \revref).
+    We made the statement that
+    genetic data are more accurately modeled by coalescence time than resistance distance
+    because while there is an extensive literature going back fourty years
+    describing the broad conditions under which coalescence is a good model 
+    for the process that produces genomic data,
+    we know of no such justification for resistance distance.
+    Perhaps the key disconnect here is that we're thinking about generative models?
+    We agree that since coalescence theory does not always apply,
+    resistance distance on some network could in principle be closer to the truth
+    than any coalescence times.
+    However, we're unaware of this claim having been made.
+    Furthermore, as we understand it,
+    the point of landscape genetics methods using resistance distance
+    is not merely to recapitulate observed genetic distances,
+    but to provide an accurate depiction of past or present gene flow,
+    i.e., movements of individuals across the landscape,
+    In our paper we've tried to describe how inaccuracies of the generative model
+    that underlies resistance distance
+    can lead to misleading results.
+    As discussed at \llname{ll:approx},
+    resistance distance is used only because it is a more computationally tractable
+    approxmation to coalescence time --
+    any method that uses resistance distances could in principle be 
+    done with coalescence times instead (e.g., to look at the effect of landscape features).
+    This is discussed at \llname{ll:circuitscape}.
+}
+
+\begin{point}{\revref}
     Extremely long sentence, please divide.
@@
 \reply{
-    Not sure what they mean.
+    Done.
 }
@@
 \reply{
+    This is an interesting idea,
+    and suggests looking at a statistic of genetic differentiation other than the mean,
+    perhaps IBD segment lengths (as done by \citet{alasadi2018estimating}).
+    We don't know enough about this to speculate, however.
 }
@@
 \reply{
+    Done, at \llname{ll:threepop}.
 }

### 2019-04-18 isolation_by_coalescence.tex
more responses

@@
 \newcommand{\responsefile}{review-responses.tex}  % name of the review reponses file
+
+\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
+  \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
+  \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
+  \renewenvironment{#1}%
+     {\linenomath\csname old#1\endcsname}%
+     {\csname oldend#1\endcsname\endlinenomath}}% 
+\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
+  \patchAmsMathEnvironmentForLineno{#1}%
+  \patchAmsMathEnvironmentForLineno{#1*}}%
+\AtBeginDocument{%
+\patchBothAmsMathEnvironmentsForLineno{equation}%
+\patchBothAmsMathEnvironmentsForLineno{align}%
+\patchBothAmsMathEnvironmentsForLineno{flalign}%
+\patchBothAmsMathEnvironmentsForLineno{alignat}%
+\patchBothAmsMathEnvironmentsForLineno{gather}%
+\patchBothAmsMathEnvironmentsForLineno{multline}%
+}
 
@@
 This measure therefore averages over all possible paths through the network between the two.
-For this reason, 
-if assigning local conductances based on the values of some landscape variable
-results in a significant positive correlation between 
+Suppose we
+assign local conductances based on the values of some landscape variable, 
+and compute the correlation of
 genetic distances between samples from different parts of the landscape
-and the effective resistances between those locations,
+with the effective resistances between the samples' locations.
+\revpoint{4}{2}
+If this results in a significant positive correlation,
 then this is taken as good evidence that the landscape variable is a good indicator of
@@
 although the two methods are conceptually similar,
-resistance-based inference may get the wrong answer if data are derived from a coalescent process,
-especially in the presence of gene flow asymmetry. \revpoint{3}{19}
+resistance-based inference may infer movement rates uncorrelated with the truth \revpoint{3}{19}
+if data are derived from a coalescent process,
+especially in the presence of gene flow asymmetry. 
 We also explore the question of identifiability.
@@
 which allows very fine geographic resolution \citep{shaffer2017desert}.
-Other methods such as EEMS \citep{petkova2016visualizing} infer conductances without such prior information; \revpoint{3}{20}
+Other methods such as EEMS \citep[``Estimating Effective Migration Surfaces'',][]{petkova2016visualizing} infer conductances without such prior information; \revpoint{3}{20}
 however, the geographical resolution is much coarser.
@@
 is that EEMS uses more computationally intensive Bayesian methods.
-However, we show that this is a more general problem, \revpoint{3}{21}
+However, we show that this coarse resolution reflects a more general problem, \revpoint{3}{21}
 both because of unmodeled ``process noise'' involved in discretization of geographical populations,
@@
 (which are reverse-time migration probabilities)
-from observations of coalescence time. \revpoint{1}{9}
+from observations of coalescence time, \revpoint{1}{9}
+and apply this to a real dataset.
+Although we have tried to highlight the intuitive aspects of this initial technical discussion,
+readers who are interested in practical differences
+more than the mathematical arguments should be able to skip to ``Test cases and simulations''
+without loss of continuity. \llabel{ll:flagging}
 
@@
 thus discretizing geographic space.
-If the genome is selectively neutral,
-then the motion of these lineages back through time
+It is common to assume that
+the motion of these lineages back through time
 forms a Markov process, the \emph{structured coalescent} \citep{wakeley2005coalescent}.
+(This assumption requires the effects of natural selection on each segment of the genome 
+to be generally weak.) \revpoint{3}{1}
 In this framework, each lineage performs a random walk across the populations,
@@
 \end{align}
+Concretely, this means finding 
+movement rates of a Markov chain whose matrix of commute times ($R$)
+yields a good approximation to the matrix of genetic distances,
+after adding local diversity parameters ($q$, which are also free parameters). \revpoint{3}{7}
 The factor of four is necessary to make commute time work on the same time scale as coalescence time:
@@
 see Figure \ref{fig:concept_coalcom}.
+
+A simple situation when the two clearly differ \llabel{ll:threepop}
+is in the presence of biased gene flow,
+as for instance between a source and a sink population:
+coalescence only requires that lineages move in \emph{one} direction
+between populations,
+while a commute requires movement in both.
 
@@
 The Markov chain is defined by its \emph{generator matrix}, denoted $G$,
-for which $G_{xy}$ gives the jump rate from $x$ to $y$.
-This means that if the Markov chain is at location $X_t$ at time $t$,
-then for $x \neq y$, the probability the chain jumps from $x$ to $y$ in the next unit of time
-is proportional to $G_{xy}$,
-i.e., $\PP{X_{t+\epsilon} = y \given X_t = y} = \epsilon G_{xy} + O(\epsilon^2)$,
-and $G_{xx} = - \sum_{y \neq x} G_{xy}$.
+for which $G_{xy}$ gives the ``jump rate'' from $x$ to $y$.
+The jump rate $G_{xy}$ is \revpoint{3}{8}
+the probability the chain jumps to position $y$ when at position $x$ per unit of time.
+(More precisely, if the Markov chain is at location $X_t$ at time $t$, 
+then for $x \neq y$, $\PP{X_{t+\epsilon} = y \given X_t = x} = \epsilon G_{xy} + O(\epsilon^2)$.)
+For mathematical convenience, the diagonal of this matrix is completed so that rows sum to zero:
+$G_{xx} = - \sum_{y \neq x} G_{xy}$.
 We will need to find the \emph{hitting times} of the chain --
@@
 equation \eqref{eqn:hitting_sum} can be rewritten in matrix form as
-\begin{align}
+\begin{align} \label{eqn:gh}
     G H = \diag(1/\pi) - \bone \bone^T ,
@@
 
-Although commute times are uniquely determined by movement rates, the reverse is not true.
-Commute times only depend on the symmetric part of $H$ --
+Although commute times are uniquely determined by movement rates, the reverse is not true --
+we can use equations \eqref{eqn:gh} and \eqref{eqn:R_from_H} to see how to modify movement rates
+without changing commute times.  \llabel{ll:dense1}
+A small concrete example of this is given in Appendix \ref{ss:simple_example}.
+More generally, commute times only depend on the symmetric part of $H$ --
 given a skew-symmetric matrix $Z$ (so that $Z + Z^T = 0$),
@@
 The resulting hitting times may not be valid --
-if we let $G_\epsilon$ be the matrix constructed by applying equation \eqref{eqn:G_from_H} to $H_\epsilon$,
+if $G_\epsilon$ is the matrix constructed by applying equation \eqref{eqn:G_from_H} to $H_\epsilon$,
 % (assuming this is invertible),
@@
 % (It is always true that rows of $G_\epsilon$ sum to zero.)
-We provide a small concrete example in Appendix \ref{ss:simple_example}.
 However, since $G$ and $H$ are continuous functions of each other,
@@
 This suggests that nonuniqueness of solutions may not be a problem in practice,
-as long as geography is discretized into sufficiently many locations
+as long as geography is discretized into sufficiently many regions
 and direct, long-distance migration is disallowed \revpoint{3}{2}
-(i.e., an individual is unlikely to live many demes away from their parents).
-However, another problem appears at finer geographic resolutions:
-although solutions may be in principle unique,
-the problems become \emph{ill-conditioned} 
-in the sense that arbitrarily small variations in the input times
+(i.e., parents and offspring live in the same or neighboring regions).
+However, finer geographic resolutions may present problems of nonidentifiability, \llabel{ll:dense2}
+similar to the problem of multiple linear regression with many collinear variables.
+The problem of inferring gene flow rates on fine geographic maps may become \emph{ill-conditioned},
+arbitrarily small variations in the data
 (even numerical instability)
-may produce large differences in the inferred rates.
-Roughly speaking, this occurs because as the spatial resolution increases,
+may produce large differences in the inferred ``best'' rates.
+Information about the landscape contained in higher-order modes
+(e.g., finer resolution changes in hitting times)
+is obscured by noise.
+(For the interested technical reader, 
+this could occur as the spatial resolution increases because,
 the matrix $G$ converges to a second-order elliptic differential operator \citep{stroock1997multidimensional};
 such operators are deformations of the Laplacian \citep{feller4},
-which is well-known to have eigenvalues that decay rapidly 
-\citep{hpmckean1967curvature,kuttler1984eigenvalues}.
-Information about the landscape contained in higher-order modes
-(e.g., finer resolution changes in hitting times)
-is then almost obliterated by the small eigenvalues of $G$,
-making the inverse problem possible in theory but not in practice.
+which has rapidly decaying eigenvalues \citep{hpmckean1967curvature,kuttler1984eigenvalues}.)
+% Information about the landscape contained in higher-order modes
+% (e.g., finer resolution changes in hitting times)
+% is then almost obliterated by the small eigenvalues of $G$,
+% making the inverse problem possible in theory but not in practice.
 This implies a fundamental limit to the geographical resolution of inferred maps.
@@
 except that $\epsilon_{ij} = \epsilon_{ji}$.
-(We assume that $C_{ij}$ is large compared to $\sigma_{\epsilon}$, \revpoint{3}{3}
-so that resulting values are still positive.)
+(We assume that coalescence times are measured to good accuracy 
+-- i.e., $C_{ij}$ is large compared to $\sigma_{\epsilon}$ -- \revpoint{3}{3}
+so that this model does not predict negative times.)
 Given $G$ and $\gamma$ we can solve equation \eqref{eqn:C_matrix} 
@@
 Concretely, suppose that the spatial arrangement of populations is given by a graph 
-(i.e., a collection of ``nodes'' connected by ``edges''),
-and only movement rates corresponding to edges in the graph are allowed to be nonzero.
+(i.e., a discrete collection of locations represented by ``nodes'',
+between which movement is possible only between those connected by ``edges''). \revpoint{3}{24}
+Only movement rates in $G$ that correspond to edges in the graph are allowed to be nonzero.
 Suppose there are $m_G$ edges in the graph,
@@
 
-It is not required in this approach to have samples from every spatial location, \revpoint{3}{16}
+It is not required in this approach to have samples from every spatial location,
 allowing inference on incompletely sampled spatial discretizations.
+(However, more complete sampling will produce more accurate results.) \revpoint{3}{16}
 To carry out inference in this situation,
@@
 we used R \citep{Rmanual} 
-to implement a standard Metropolis-Hastings algorithm with Gaussian proposals \citep{brooks2011handbook}.
+to implement a standard Metropolis-Hastings algorithm with Gaussian proposals 
+reflected into the positive quadrant \citep{brooks2011handbook}. \revpoint{2}{5}
 Starting locations were chosen by sampling from the prior distribution.
@@
 We compared inference under coalescence time and resistance distance models
-using both data ``from the model'' and from more realistic simulations.
+using both data from the model and from more realistic simulations. \revpoint{3}{23}
 For the first category, we designed a number of graphs to test particular aspects of inference,
@@
 so used the true standard deviation of the noise for $\sigma_\epsilon$ in the likelihood function.
+In practice, $\sigma_epsilon$ can be estimated by the standard deviation
+of genetic distances between different pairs of samples in the same region.
 To assess the impact on inference of the number of parameters,
@@
 \citep{wilkins2004separationoftimescales}.
+For instance, if between-population gene flow and coalescence rates are similar,
+then the lineages of two samples in the same population 
+are likely to coalesce before either leaves,
+leading to a strong signal of isolation by distance. \revpoint{3}{13}
+On the other hand, if coalescence rates are much smaller than gene flow,
+nearby samples are unlikely to share more recent common ancestors than distant ones.
 Indeed, coalescence rates had a strong effect on feasibility of the inference problem
@@
 increased mean absolute error of coalescence time inference from around 0.05 to 0.5.
+(Gene flow rates are of order 1, so this range of coalescence rates
+spans strong to weak isolation by distance.)
 Again, allowing multiple coalescence rates
@@
 Since lineages quickly move to the center and coalesce regardless of starting position, 
+\llabel{ll:dense3}
 coalescence time (and genetic diversity) will be relatively low between all individuals. \revpoint{4}{4}
@@
 each observed, empirical mean estimated using a sample of genotypes
-will deviate from this depending on both the individuals sampled (``sampling noise'')
+will deviate from this depending on the individuals sampled (``individual sampling noise'')
 \citep{ashander2018demographic},
+the genetic loci genotyped (``genome sampling noise'') \citep{waples2009modelling},
 and the stochasticity of the population history itself (``process noise'')
@@
 % This might make inference using a discretization of continuous space considerably more difficult.
-To quantify the relative contributions of the different sources of noise,
+Sequencing of many nonascertained loci genome-wide should minimize the second source of noise
+(although genome structure can affect this at large scales \citep{li2016local}). \revpoint{3}{11}
+Geographic models in continous space, 
+are expected to show much more process noise than discrete population models
+because of random fluctuations in local population density. \revpoint{3}{10}
+To quantify the relative contributions of the remaining sources of noise,
 we compared mean genetic distances, calculated in the same way,
 between two nonoverlapping sets of samples from each of three identical simulations.
-These simulations were done using SLiM as described in the Methods.
+These forwards-time, continuous-space simulations were done using SLiM as described in the Methods.
 We found that with these parameter values, process and sampling noise were of similar magnitude,
@@
 this step might not be straightforward. \revpoint{2}{4}
+To make parameters most easily interpretable,
+geography should probably be discretized along natural barriers
+and so that each region is of roughly the same geographic area.
 % For instance, we may wonder whether a river or other geographical feature
@@
 %%%%%%%%%%%%%
-\subsection*{Application to \textit{Populus trichocarpa}}
+\subsection*{Application to \textit{Populus}}
 
 Our discretization of the sampling area for the two poplar species
-is shown in Figure~\ref{fig:populus_results},
-along with
+is shown in Figure~\ref{fig:poplar_results},
+along with arrows depicting posterior mean estimates of gene flow.
+Posterior distributions of gene flow rates and coalescence rates
+are shown in Supplemental Figures~\ref{sfig:poplar_g} and \ref{sfig:poplar_gam}.
+The main features of the resulting model are:
+(a) low but nonzero gene flow between species,
+with the highest rate between regions 3 and 4;
+(b) strong bias in gene flow among (inland) \textit{balsamifera} regions,
+from southeast to northwest ($3 \to 5 \to 6 \to 1$); and 
+(c) strong bias in gene flow into regions 2 and 8.
+These results align with what is known about \textit{Populus} history and ecology in the region
+(personal communication, A.~Moreno Geraldes and Q.~Cronk, and \citet{geraldes2014landscape}).
+There is ongoing hybridization around region 4,
+and interspecific gene flow from 9 to 3 may be due to downstream transport along the Columbia river
+(A larger gene flow rate from 9 to 3 indicates that individuals in region 9
+are more likely to have recent ancestry in region 3 than vice-versa.)
+The glacial refugia for the two species are thought to be in western Alaska (\textit{balsamifera})
+and coastal British Columbia (\textit{trichocarpa}).
+Inferred gene flow leading back towards these regions (depicted by larger arrows on the map)
+are consistent with this --
+for instance, if region 5 was colonized from region 6 sufficiently recently,
+then individuals in region 5 should trace their ancestry to region 6 much more frequently
+than vice-versa, as indicated by the large $5 \to 6$ arrow 
+and small $6 \to 5$ arrow in Figure~\ref{fig:poplar_results}.
+
+Interpreting these results requires some nuance beyond what was necessary in our simulation results.
+Sampling locations are sparse, especially in the inland species,
+so how exactly should we think of the geographic extent of ``region 6'',
+which is represented by only two, distant samples?
+Is this region's low inferred coalescence rate
+a result of simply a large population (due to a large geographic region)
+or barriers to movement within the region?
+Although it is not entirely clear how to translate discrete-population models
+to continuous space, \llabel{ll:continuous}
+a ``gene flow rate'' from region $i$ to region $j$ as shown in Figure~\ref{fig:poplar_results}
+is best thought of as an estimate of the proportion of the individuals in region $i$
+that have a parent in region $j$, averaged over time and scaled by some unknown factor.
+Equivalently,
+this is the probability per unit time
+that a lineage traced back from an individual in region $i$ follows a line of descent to region $j$,
+again scaled by some factor.
+We therefore expect gene flow to be biased from an area of 
+lower population density (or, net fecundity)
+into a neighboring region of higher population density.
+We also expect gene flow to be biased 
+from a smaller region to a larger, neighboring one than vice-versa
+if a larger proprotion of the smaller region lies close to the boundary between the two.
+This may explain the net bias from many coastal populations to inland ones.
+We grouped samples so that each group represented roughly comparable geographic regions,
+to ameliorate these confounding factors. \llabel{ll:discretization}
+However, these difficulties in interpretation are common in any situation
+where discrete populations are used to model continuous geography.
+
+\begin{figure}
+\centering
+    \includegraphics[width=0.9\textwidth]{poplars/grid_populus_map}
+    \caption{
+        Model results for \textit{Populus} data,
+        superimposed on a map of northwestern North America.
+        Sampling locations of \textit{Populus} genomes are small circles,
+        with outer color indicating species 
+        (red-outlined, coastal samples are \textit{trichocarpa}, 
+        blue-outlined, inland samples are \textit{balsamifera});
+        circle fill color indicates how samples are grouped into discrete ``populations''.
+        Large circles are placed at group centroids, and given group labels.
+        Arrow width is proportional to the posterior mean gene flow rate inferred under
+        the coalescence time model (also given by the labels):
+        for instance, the inferred gene flow rate from 4 to 2 is 3.0,
+        while that from 2 to 4 is 0.7,
+        suggesting that poplars in region 4
+        have many more recent ancestors in region 2
+        than poplars in region 2 do in region 4.
+        Northwestern bias in gene flow between \textit{balsamifera} regions
+        may reflect postglacial expansion to the south and east out of a refugium in Alaska.
+        \label{fig:poplar_results}
+    }
+\end{figure}
 
@@
 describing Circuitscape \citep{mcrae2006isolation} and EEMS \citep{petkova2016visualizing}
-present the resistance approximation as a necessary step for computational feasibility.
+present the resistance approximation as a necessary step for computational feasibility. \llabel{ll:approx}
 Results of resistance-based methods should therefore be interpreted with caution:
@@
 that explicitly model continuous geography.
+In the meantime, comparison of results utilizing different discretizations of geography
+can help identify problems.  \revpoint{3}{12}
 
@@
 Circuitscape \citep{mcrae2008using} produces maps of much higher resolution 
-than is presently possible with coalescent times.
+than is presently possible with coalescent times. \revpoint{4}{1}
 \citet{hanks2013circuit} connects the problem to intrinsic conditional autoregressive models, 
@@
 
-\paragraph{Circuitscape}
+\paragraph{Circuitscape} \llabel{ll:circuitscape}
 Could we use coalescence time in place of commute time
@@
 
-\paragraph{The effect of history}
+\paragraph{The effect of history} \llabel{ll:history}
 The models we discuss here assume that population sizes and migration rates
@@
 but may not be appropriate.
-Even though the population dynamics in forwards time are Markov, 
+Even if the population dynamics in forwards time are Markov, 
 the dynamics of lineages traced back in time (the coalescent process) may not be. 
-For example, if fecundity is variable and population density is low,  \revpoint{3}{15}
+For example, if individual fecundity is variable and population density is low,  \revpoint{3}{15}
 two lineages near each other are more likely to share a recent common origin, 
@@
 and general input,
-and to Quentin Cronk and Armando Geraldes for consultations about the \emph{Populus} results.
+and to Quentin Cronk and Armando Moreno Geraldes for consultations about the \emph{Populus} results.
 We would also like to gratefully acknowledge the immense positive impact that Brad McRae \citep[1966--2017;][]{lawler2018tribute}
@@
 
+\begin{figure}
+\centering
+    \includegraphics[width=0.7\textwidth]{poplars/posterior_dists_g_populus}
+    \includegraphics[width=0.7\textwidth]{poplars/grid_populus}
+    \caption{
+        Posterior distributions of parameters for \textit{Populus} data,
+        using coalescence time inference with the model shown in Figure~\ref{fig:poplar_results}.
+        \textbf{(top)} Posterior medians (dark line),
+        interquartile range (box), and middle 95\% (whiskers),
+        for gene flow rates.
+        \textbf{(bottom)} Posterior medians, along with edge labels that match the upper figure.
+        \label{sfig:poplar_g}
+    }
+\end{figure}
+
+\begin{figure}
+\centering
+    \includegraphics[width=\textwidth]{poplars/posterior_dists_gam_populus}
+    \caption{
+        Posterior distributions of parameters for \textit{Populus} data,
+        using coalescence time inference with the model shown in Figure~\ref{fig:poplar_results}.
+        Plot shows posterior medians (dark line),
+        interquartile range (box), and middle 95\% (whiskers),
+        for coalescence rates, labeled as in Figure~\ref{fig:poplar_results}.
+        For instance, the coalescence rates for regions 4 and 8 (labeled ``gam4'' and ``gam8'')
+        are inferred to be much smaller than most other regions, 
+        perhaps indicating larger or older populations in southwestern British Columbia.
+        \label{sfig:poplar_gam}
+    }
+\end{figure}
+
 \ifsubmission

### 2019-04-19 review-responses.tex
done going through reviews

@@
 \reply{
+    Whoops, thanks! \revref
 }
@@
     \begin{align*}
-        H_xz &= dt + \sum_y Pr(X_{dt} = y | X_0 = x) H_{yz} \\
+        H_{xz} &= dt + \sum_y Pr(X_{dt} = y | X_0 = x) H_{yz} \\
                &= dt + \sum_{y != x} Q_{xy} dt H_{yz} + (Q_{xx} dt + 1) H_{xz}
@@
 \reply{
+    We wondered whether this would overload readers,
+    but have now included an explanation along these lines.
+    \revref
 }
@@
     \begin{align*}
-        C_xz &= dt + {} \\
+        C_{xz} &= dt + {} \\
                 & \qquad
@@
                 & \qquad
-             1/2 \sum_{y != x} Q_{xy} dt C_{yz} + (Q_{xx} dt + 1) C_{xz} + \\
+             1/2 \sum_{y \neq x} Q_{xy} dt C_{yz} + (Q_{xx} dt + 1) C_{xz} + \\
                 & \qquad
-             1/2 \sum_{y != x} Q_{zy} dt C_{xy} + (Q_{xx} dt + 1) C_{xz},
+             1/2 \sum_{y \neq x} Q_{zy} dt C_{xy} + (Q_{xx} dt + 1) C_{xz},
     \end{align*}
@@
 \reply{
+    Besides substituting $G$ for $Q$,
+    our equations only differ by the factor of 1/2.
+    The model we work under is that each lineage moves as an independent copy
+    of the Markov chain $X$, so that the total jump rate
+    when the two chains are at $x$ and $z$ is $-G_{xx}-G_{zz}$.
+    We think the factors of 1/2 are in error --
+    a more clear way to see this would be to say that $X$ and $Z$ are the locations of the two lineages,
+    that move independently until they coalesce,
+    and so
+    \begin{align*}
+        C_{xz} &= dt + {} \\
+                & \qquad
+             \sum_u \P(X_{dt} = u | X_0 = x) C_{uz} + {} \\
+                & \qquad
+             \sum_v \P(Z_{dt} = v | Z_0 = z) C_{xv}, \\
+           &= dt + {} \\
+                & \qquad
+             \sum_{u \neq x} G_{xu} dt C_{uz} + (G_{xx} dt + 1) C_{xz} + \\
+                & \qquad
+             \sum_{v \neq x} G_{zv} dt C_{xv} + (G_{xx} dt + 1) C_{xz} .
+    \end{align*}
 }
@@
     the data seems completely hidden. It is in fact conveyed by the coalescent
-    times $C_{ij}(G,\gamma)$ but a more detailed description of the likelihood
+    times $\mathcal{C}_{ij}(G,\gamma)$ but a more detailed description of the likelihood
     function where both genetic and spatial information appear explicitly would
@@
 \end{point}
+
+\reply{
+    We've added a clarifying sentence. \revref
+}
 
@@
 
+\reply{
+    In our model, spatial locations are (discrete and) assumed to be known without error,
+    and we do not want to model the sampling process,
+    so they do not enter into the likelihood.
+    Also recall that both $D$ and $C$ are \emph{complete} matrices of genetic distances 
+    and pairwise coalescence times -- i.e.,
+    between all pairs of populations --
+    so the locations do not enter at this stage (although they are certainly used in estimating $D$).
+    We have added a note reminding the reader of this. \revref
+}
+
 \begin{point}{}
@@
 \reply{
+    This is true; we've added a sentence emphasizing this. \revref
 }
@@
 \begin{point}{}
-    The sentence "with noise standard deviation equal to 1/1000th of the mean
-    coalescence time (left) and 1/100th of the mean coalescence time (right)"
-    should read "with noise standard deviation equal to 1/1000th of the mean
-    coalescence time (top) and 1/100th of the mean coalescence time (bottom)"
-\end{point}
-
-\reply{
+    The sentence ``with noise standard deviation equal to 1/1000th of the mean
+    coalescence time (left) and 1/100th of the mean coalescence time (right)''
+    should read ``with noise standard deviation equal to 1/1000th of the mean
+    coalescence time (top) and 1/100th of the mean coalescence time (bottom)''
+\end{point}
+
+\reply{
+    Whoops! Fixed. \revref
 }
@@
 \reply{
+    Certainly! We agree that the SLFV model is a promising way forward
+    to be able to use coalescent simulations in continuous space.
+    There's a lot of work left to explore how and when it's a good approximation
+    to reality also,
+    but we've added a bit about it to the ``Other methods'' section. \revref
 }
@@
 \reply{
-}
+    We agree, and very much hope to move in that direction.
+    We've added more discussion of this. \revref
+}

### 2019-04-19 isolation_by_coalescence.tex
done going through reviews

@@
 the probability the chain jumps to position $y$ when at position $x$ per unit of time.
-(More precisely, if the Markov chain is at location $X_t$ at time $t$, 
+(More precisely, if the Markov chain is at location $X_t$ at time $t$,  \revpoint{5}{1}
 then for $x \neq y$, $\PP{X_{t+\epsilon} = y \given X_t = x} = \epsilon G_{xy} + O(\epsilon^2)$.)
@@
 which implies the chain is connected and irreducible.
-By conditioning on $X_\epsilon$,
-one can use the Markov property to show that
+The Markov property lets us write a system of linear equations that can be solved for $H$, as follows
+(see, e.g., \citet{kemeny1983finite}):
+Suppose that the chain begins at $x \neq z$,
+and is at location $y \neq z$ after $dt$ units of time, without having visited $z$.
+Then the total time to hit $z$ is equal to the elapsed time $dt$
+plus the remaining time to hit $z$ from $y$.
+Formally, this says that \revpoint{5}{2}
+$H_{xz} \approx dt + \sum_y \P\{X_{dt} = y \given X_0 = x\} H_{yz}$.
+Replacing $\P\{X_{dt} = y \given X_0 = x\}$ with $dt G_{xy}$ and rearranging
+gives the system of equations:
 \begin{align} \label{eqn:hitting_sum}
-    \sum_y G_{xy} H_{yz} = -1 \qquad \text{for} \; x \neq z,
+    \sum_y G_{xy} H_{yz} = -1 \qquad \text{for} \; x \neq z.
 \end{align}
-i.e., if $G_{-z}$ is the matrix with the $z^\text{th}$ row and column removed,
+In other words, if $G_{-z}$ is the matrix with the $z^\text{th}$ row and column removed,
 and $\bone$ is the vector composed of all ones,
@@
 $D_{ij} = C_{ij} + \epsilon_{ij}$,
-where $\epsilon_{ij}$ are independent and Normally distributed
+where $\epsilon_{ij}$ are independent and Gaussian distributed
 with mean $\mathbf{0}$ and variance $\sigma_\epsilon^2$,
@@
 with mean $\mathcal{C}(G, \gamma)$.
+(This Gaussian stands in for the sampling noise of coalescence times about their mean.) \revpoint{5}{6}
 The maximum likelihood estimate of $G$ could be found directly;
@@
 respectively.
+Recall that both $D$ and $\mathcal{C}(G,\gamma)$ are full matrices of pairwise distances between discrete locations (in the same order), \revpoint{5}{5}
+the term $(\mathcal{C}_{ij}(G,\gamma) - D_{ij})^2$
+measures the deviation of the observed genetic distances
+from those predicted under the coalescent model on the graph. \revpoint{5}{4}
 
@@
 on this graph, with noise standard deviation equal to
-1/1000${}^\text{th}$ of the mean coalescence time (left)
-and 1/100${}^\text{th}$ of the mean coalescence time (right).
+1/1000${}^\text{th}$ of the mean coalescence time (top)
+and 1/100${}^\text{th}$ of the mean coalescence time (bottom). \revpoint{5}{7}
 Although increased noise in the data increased uncertainty in the estimated values,
@@
 
-Most promisingly, the recent method, MAPS \citep{alasadi2018estimating}
+Most promisingly, the recent method, MAPS \citep[``Migration and Population Surface estimation'',][]{alasadi2018estimating}
 is based on EEMS, but instead of resistance distances,
@@
 However, this is not motivated by a generative model for genetic data.
+
+All of the methods presented above work with a discretized model of space,
+for mathematical and computational reasons. \revpoint{5}{8}
+Development of new methods for continuous geography
+and assessment of how current methods behave when faced with continuous geography
+is an important challenge facing the field.
+This has been made substantially easier with the introduction of continuous space 
+into the simulator, SLiM \citep{haller2018forward},
+that we used to test the behavior of our method with more realistic data.
+Another promising approach centers around the Spatial Lambda-Fleming-Viot model
+\citep{barton2010continuum}, \revpoint{5}{9}
+which provides a mechanistic model in which coalescent simulations are possible.
+This has been used for inference \citep{guindon2016demographic},
+but more work remains to understand how it behaves as an approximation to real populations.
 

### 2019-04-19 poplars/populus.infer.R
updated poplar plots

@@ -189,14 +189,20 @@ cairo_pdf(filename=paste0(outpath,"/grid_",fname,".pdf"),width=14,height=14)
        edge.label=paste0("g",1:ng,"=",round(g_med*1000)/1000),main=paste0("Graph Structure"))
 dev.off()
 
-cairo_pdf(filename=paste0(outpath,"/posterior_dists_g_",fname,".pdf"),width=8,height=7)
-  boxplot(g[100*(1:(30e6/100)),order(g_med)],outline=FALSE,main=paste0("Posterior Distributions (g): ",name),
-        names=paste0("g",order(g_med)),xlab="Parameter Index",ylab="Parameter Value (rate)",las=2)
+#with helper function
+cairo_pdf(filename=paste0(outpath,"/grid_",fname,"2.pdf"),width=14,height=14)
+  plot.grid(g_med=g_med,gam_med=gam_med,name="Populus Habitat Graph Structure",
+            rect_grid = FALSE,use_centroids = TRUE,centroids=as.matrix(centroids[,1:2]),G_adj=G_adj)
 dev.off()
 
-cairo_pdf(filename=paste0(outpath,"/posterior_dists_gam_",fname,".pdf"),width=8,height=7)
+cairo_pdf(filename=paste0(outpath,"/posterior_dists_g_",fname,"2.pdf"),width=6,height=5)
+  boxplot(g[100*(1:(30e6/100)),order(g_med)],outline=FALSE,main=paste0("Posterior Distributions (g): ",name),
+        names=paste0("g",order(g_med)),xlab="Gene Flow Rate Parameter Label",ylab="Gene Flow Rate",las=2)
+dev.off()
+
+cairo_pdf(filename=paste0(outpath,"/posterior_dists_gam_",fname,"2.pdf"),width=6,height=5)
   boxplot(gam[100*(1:(30e6/100)),],outline=FALSE,main=paste0("Posterior Distributions (gamma): ",name),
-        names=paste0("gam",1:n),xlab="Parameter Index",ylab="Parameter Value (rate)",las=2)
+        names=paste0("gam",1:n),xlab="Coalescence Rate Parameter Label",ylab="Coalescence Rate",las=2)
 dev.off()
 
 

### 2019-04-20 review-responses.tex
final read-through

@@
 Comparing isolation by resistance to a new coalescent-based method.''
-(formerly titled ``Are populations like a circuit? The relationship between isolation by distance and isolation by resistance'').
-
-...
+(formerly titled 
+``Are populations like a circuit? 
+The relationship between isolation by distance and isolation by resistance'').
+
+Thanks to the five mostly anonymous reviewers for their many useful comments --
+we've addressed all of them, which we think has made the paper more generally understandable
+and suitable for this audience.
+In particular, we have included analysis of an empirical dataset 
+(which was fun, and produced good results),
+revised the figures for increased legibility,
+and added a substantial amount of additional explanation of the technical points.
+We hope that discussion of the empirical results should help make interpretation
+of the coalescent-based model more broadly understandable.
+And, at the suggestion of reviewer 1,
+we have changed the title.
+We think this has all improved the paper substantially,
+and hope you agree.
+
+We have included detailed responses to all the reviewer's comments,
+with page and line numbers that refer to the revised version of the manuscript.
+There is also a version of the manuscript showing differences to the previously submitted document.
 
@@
 
+Please find our responses below.
+We hope you agree we've addressed all four of these points.
+
 %%%%%%%%%%%%%%
@@
     stone models. They extensively compare their computational method to resistance
-    based methods using an own implementation of the latter, both "under the model"
+    based methods using an own implementation of the latter, both ``under the model''
     as well as using various quite realistic forward in time simulations.
@@
     Overall, I think this article makes a very valid contribution for Molecular
-    Ecology Ressources. The authors put their work into the right context and
+    Ecology Resources. The authors put their work into the right context and
     show good knowledge of the relevant literature. The writing is both clear
@@
     but in our testing, using sparse matrix methods didn't help at this scale.
-    We've added a point about this last thing to the discussion \revref
+    We've added a point about this last thing to the discussion \revref{}
     but haven't put in more detail (e.g., plots of scaling with $N$)
@@
 \begin{point}{\revref}
-    $F_ST \neq$ Absolute divergence!!!  $F_ST$ is more a
+    $F_{ST} \neq$ Absolute divergence!!!  $F_{ST}$ is more a
     co-variance measure, and for instance heavily influence by sampling a batch
@@
     We've added a note about this, although not a lot of detail,
-    because it seems outside the scope of this paper. \revref
+    because it seems outside the scope of this paper \revref{}.
     And, we thought we should mention $F_{ST}$ because it is commonly used
@@
 \begin{point}{}
-    Dimensionality arguments (Line 129 onward) can be a bit misleading when you
+    Dimensionality arguments (\revref{} and onward) can be a bit misleading when you
     already know the equations have a non-trivial kernel. Setting all
@@
 \reply{
-    We agree -- there's a lot of directions we could move in to improve our package,
-    and hope to work more on this.
+    True! We've tweaked the way the arrows are displayed,
+    and think they look better now.
 }
@@
     variation across a landscape. Also, it is inherently symmetric in $i$ and $j$, and
-    cannot accomodate asymmetric migration between demes. A more biologically
+    cannot accommodate asymmetric migration between demes. A more biologically
     realistic model is the structured coalescent. Hence, if data are generated under
@@
     number of simulated data sets. It seems that in general, this method and the
-    resistance-based approaches (as exemplified by EEMS) produce somewhat com-
-    parable results so long as migration rates are symmetric. The largest differences
-    emerge when gene flow is asymmetric, because this fundamentally violates the
-    resistance distance model used by EEMS. Here, the authors show convincingly
-    that inferring migration rates using the resistance distance approximation can
-    be wildly inaccurate compared to the same estimates obtained using coalescent
-    methods. To the extent that gene flow is more likely to be asymmetric in many
-    natural populations, this seems like an important finding.
+    resistance-based approaches (as exemplified by EEMS) produce somewhat
+    comparable results so long as migration rates are symmetric. The largest
+    differences emerge when gene flow is asymmetric, because this fundamentally
+    violates the resistance distance model used by EEMS. Here, the authors show
+    convincingly that inferring migration rates using the resistance distance
+    approximation can be wildly inaccurate compared to the same estimates
+    obtained using coalescent methods. To the extent that gene flow is more
+    likely to be asymmetric in many natural populations, this seems like an
+    important finding.
 
@@
 ``integrates over all possible paths that the organisms might take''.
-We think it is clear that \citet{mcrae2006isolation} saw resistance distance as an approximation
+We think \citet{mcrae2006isolation} saw resistance distance as an approximation
 to the more-accurate coalescence models; we're looking at the effects of this approximation.
@@
 This has also given us a chance to spend more time talking through how to interpret the results.
-(We didn't get to say this in the paper,
-but initially were worried by the southward bias in gene flow for the inland poplar species,
-since we would expect the opposite if poplars had expanded to the north following the glaciers;
-but then we learned that this species likely expanded from Alaksa, rather than from the south.)
+% (We didn't get to say this in the paper,
+% but initially were worried by the southward bias in gene flow for the inland poplar species,
+% since we would expect the opposite if poplars had expanded to the north following the glaciers;
+% but then we learned that this species likely expanded from Alaska, rather than from the south.)
 
@@
     due to the large number of parameters.
-    % Devoloping a way to reduce the number of parameters in a way similar to EEMS
+    Something along these lines might be useful for finer geographical resolutions, though.
+    % Developing a way to reduce the number of parameters in a way similar to EEMS
     % while still allowing asymmetric migration is a good direction for future implementations.
@@
 \reply{
-    We think this is correct (and reads reasonably well?) as it stands.
+    We've reworded this. \revref
 }
@@
 \reply{
-    Reworded.
+    Reworded. \revref
 }
@@
 \reply{
-    Added parenthetical clarification.
+    Added parenthetical clarification. \revref
 }
@@
 \reply{
-    Added explanatory note aobut how to estimate noise.
+    Added explanatory note about how to estimate noise. \revref
 }
@@
     differ?  If so it is just pairwise theta as used by Wakeley and others.  
-    Eq 1:  what is $R_{ij}$ and how is $q$ calculated?
-\end{point}
-
-\reply{
-    That's correct; we've reworded this to be more clear.
+\end{point}
+
+\reply{
+    That's correct; we've reworded this to be more clear. \revref
 }
@@
     0.02 for noise, both of which would be exceedingly small.  If that is not
-    what is meant, please clarify.
+    what is meant, please clarify. \revref
 \end{point}
@@
 \reply{
-    We mean among individuals (and have added this).
+    We mean among individuals (and have added this). \revref
 }
@@
 \reply{
-    We're simply saying that it is not a requirement to \emph{run} the method
+    We're saying that it is possible to \emph{run} the method
     to have samples in each location --
-    it is easy, in principle, to include ``ghost'' populations.
+    so it is easy, in principle, to include ``ghost'' populations.
     Of course, unsampled populations will affect accuracy of the result,
@@
 \reply{
-    We sympathize, and have tried to improve this.
-    However, some of this we think is unavoidable --
-    the x-axis labels of figures like Figure~\ref{fig:5x3b_post_coalvcom}
-    cannot be made larger without overlapping.
+    We sympathize, and have increased the sizes of most of the labels.
+    However, the x-axis labels of figures like Figure~\ref{fig:5x3b_post_coalvcom}
+    cannot be made larger without overlapping, and we'd like to keep them.
     Another approach would be to simply plot true versus observed values,
@@
 \reply{
-    A lengthy name (that we've added). \revref
+    A lengthy acronym (that we've added, as well as MAPS, later). \revref
 }
@@
 \reply{
-    Yes, but this is the way we implemented it in the simulation.
+    It could, but this is the way we implemented it in the simulation.
 }
@@
     genetic data are more accurately modeled by coalescence time than resistance distance
-    because while there is an extensive literature going back fourty years
+    because while there is an extensive literature % going back fourty years
     describing the broad conditions under which coalescence is a good model 
@@
     can lead to misleading results.
-    As discussed at \llname{ll:approx},
-    resistance distance is used only because it is a more computationally tractable
-    approxmation to coalescence time --
-    any method that uses resistance distances could in principle be 
+    The only downside to coalescence time inference is computational tractability
+    (discussed at \llname{ll:approx}):
+    otherwise, one can be swapped out for the other.
+    In fact, \emph{any} method that uses resistance distances could in principle be 
     done with coalescence times instead (e.g., to look at the effect of landscape features).
@@
 \reply{
-    Done.
+    Done. \revref
 }
@@
     coalescent-based technique, which applies to the analysis of pairs of
-    sampled individuals, is compared to an approach based on "resistance
-    distances" which aim at approximating coalescent distances. This work is
+    sampled individuals, is compared to an approach based on ``resistance
+    distances'' which aim at approximating coalescent distances. This work is
     interesting because resistance distances are at the core of several popular
@@
     a more clear way to see this would be to say that $X$ and $Z$ are the locations of the two lineages,
-    that move independently until they coalesce,
+    that move independently according to $G$ until they coalesce,
     and so

### 2019-04-20 isolation_by_coalescence.tex
updated figs
final read-through

@@
 Why should these be related to genetic distance? 
-First, we explain this connection,
-and then discuss their computation,
+We first explain this connection and discuss computation of both quantities,
 and finally develop a method that infers movement rates of the Markov chain
-(which are reverse-time migration probabilities)
-from observations of coalescence time, \revpoint{1}{9}
-and apply this to a real dataset.
-Although we have tried to highlight the intuitive aspects of this initial technical discussion,
+(which are rates of gene flow)
+from observations of coalescence time. \revpoint{1}{9}
+We apply the method to both realistic simulations in continuous space
+and to an empirical dataset.
+Although we have tried to highlight the intuitive aspects of the initial technical discussion,
 readers who are interested in practical differences
@@
 so that generally $D_{ii} > 0$: the diagonal elements measure local diversity.
-As discussed below, we will find commute times so that the mean commute time between
+As discussed below, 
+``resistance distance'' methods find commute times so that the mean commute time between
 $i$ and $j$ matches $D_{ij}$ -- but the commute time of a location to itself is zero,
-so we must account for local diversity.
+so a term is added to account for local diversity.
 We do this in the same way as \citet{petkova2016visualizing},
@@
 plus the remaining time to hit $z$ from $y$.
-Formally, this says that \revpoint{5}{2}
+Taking the average over possible paths, this says that \revpoint{5}{2}
 $H_{xz} \approx dt + \sum_y \P\{X_{dt} = y \given X_0 = x\} H_{yz}$.
@@
 (We prove that $H$ is invertible in Appendix~\ref{apx::hitting_calcs}.) \revpoint{1}{12}
-Therefore, $H$ uniquely determines $G$.
+Therefore, the pairwise mean hitting times, $H$, uniquely determine the jump rates, $G$.
 
 % \paragraph{Commute times}
-The \emph{commute times} are a symmetrized version of the hitting times
-(which need not be symmetric),
-defining $R = H + H^T$, i.e.:
+The \emph{commute time} between $x$ and $y$ is the mean time a chain started at $x$ takes 
+to first hit $y$ and then to return to $x$.
+We will write this as
 \begin{align} \label{eqn:R_from_H}
 \text{\bf (commute time)} \qquad
-    R_{xy} = H_{xy} + H_{yx}.
+    R_{xy} = H_{xy} + H_{yx},
 \end{align}
+or in matrix notation, $R = H + H^T$.
+Note that the commute times are symmetric,
+even though the hitting times may not be.
 
@@
 given a skew-symmetric matrix $Z$ (so that $Z + Z^T = 0$),
-any Markov chain that has hitting times given by $H_\epsilon = H + \epsilon Z$ for some $\epsilon$
+any Markov chain that has hitting times given by $H_\epsilon = H + \epsilon Z$ for some number $\epsilon$
 will have the same commute times.
@@
 that coalesce at rate $\gamma_x$ when $X$ and $Y$ are both at $x$.
-Then we define $\tau$ to be this coalescence time,
+Define $\tau$ to be this coalescence time,
 so that 
@@
 $\P\{\tau \le t + \epsilon \given \tau > t, \; X_t \neq Y_t\} = O(\epsilon^2)$.
-Then, the (mean) coalescence time is $C_{xy} = \E[\tau \given X_0 = x, \, Y_0 = y]$.
-A similar argument using the Markov property
-shows that $C$ satisfies the following equation:
+Then, the (mean) coalescence time is defined to be
+$C_{xy} = \E[\tau \given X_0 = x, \, Y_0 = y]$.
+An argument similar to the one behind equation \eqref{eqn:hitting_sum}
+shows that the coalescence times satisfy the following system of equations:
 \begin{align}
@@
 as the original chain.
-For $G + Z$ to be a valid generator, the rows of $Z$ must sum to zero,
+Furthermore, for $G + Z$ to be a valid generator of a Markov chain,
+the rows of $Z$ must sum to zero,
 and all offdiagonal entries of $G + Z$ must be nonnegative. \revpoint{1}{11}
@@
 The problem of inferring gene flow rates on fine geographic maps may become \emph{ill-conditioned},
-arbitrarily small variations in the data
+i.e., arbitrarily small variations in the data
 (even numerical instability)
@@
 (For the interested technical reader, 
-this could occur as the spatial resolution increases because,
+this could occur as the spatial resolution increases because
 the matrix $G$ converges to a second-order elliptic differential operator \citep{stroock1997multidimensional};
@@
 Such constraints amount to a form of regularization,
-since they look for solutions to the original problem with certain desireable properties
+since they look for solutions to the original problem with certain desirable properties
 (e.g., most of the values are the same). \revpoint{1}{4}
@@
 Recall that both $D$ and $\mathcal{C}(G,\gamma)$ are full matrices of pairwise distances between discrete locations (in the same order), \revpoint{5}{5}
-the term $(\mathcal{C}_{ij}(G,\gamma) - D_{ij})^2$
+and so the term $(\mathcal{C}_{ij}(G,\gamma) - D_{ij})^2$
 measures the deviation of the observed genetic distances
@@
 We also applied our method to genotyping data from 423 samples of \textit{Populus trichocarpa}
-\citet{geraldes2014landscape}
 and 8 samples of the closely related \textit{Populus balsamifera} (with which it hybridizes),
@@
 so used the true standard deviation of the noise for $\sigma_\epsilon$ in the likelihood function.
-In practice, $\sigma_epsilon$ can be estimated by the standard deviation
+In practice, $\sigma_\epsilon$ can be estimated by the standard deviation
 of genetic distances between different pairs of samples in the same region.
@@
 \centering
-     \includegraphics[scale=1]{figs/5x3b_post_coalvcom2}
+     \includegraphics[scale=1]{figs/5x3b_post_coalvcom}
     \caption{
@@
 and extreme in the biased asymmetric case.
-This difference suggests that resistance distance inference may be strongly misled \revpoint{2}{7}
+This difference suggests that inferences made with resistance distance 
+may be strongly misleading \revpoint{2}{7}
 in the asymmetric and biased situations,
@@
 \centering
-     \includegraphics[scale=1]{figs/4x4boxplots_paper2}
+     \includegraphics[scale=1]{figs/4x4coalvcom_paper}
     \caption{
@@
 
-The data we have used thus far are produced ``under the model'',
+The data we have used thus far are produced under the model used for inference,
 and should provide an accurate depiction of our method applied to discrete, randomly mating populations
@@
 (although genome structure can affect this at large scales \citep{li2016local}). \revpoint{3}{11}
-Geographic models in continous space, 
+Geographic models in continuous space, 
 are expected to show much more process noise than discrete population models
@@
 and interspecific gene flow from 9 to 3 may be due to downstream transport along the Columbia river
-(A larger gene flow rate from 9 to 3 indicates that individuals in region 9
+(a larger gene flow rate from 9 to 3 indicates that individuals in region 9
 are more likely to have recent ancestry in region 3 than vice-versa.)
@@
 from a smaller region to a larger, neighboring one than vice-versa
-if a larger proprotion of the smaller region lies close to the boundary between the two.
+if a larger proportion of the smaller region lies close to the boundary between the two.
 This may explain the net bias from many coastal populations to inland ones.
@@
         Large circles are placed at group centroids, and given group labels.
-        Arrow width is proportional to the posterior mean gene flow rate inferred under
-        the coalescence time model (also given by the labels):
-        for instance, the inferred gene flow rate from 4 to 2 is 3.0,
+        Arrows are labeled by the posterior mean gene flow rate inferred under
+        the coalescence time model (which also determines arrow width).
+        For instance, the inferred gene flow rate from 4 to 2 is 3.0,
         while that from 2 to 4 is 0.7,
@@
 providing a Bayesian model of an underlying Gaussian Markov random field.
-EEMS \citep{petkova2014visualizing} integrates over possible tessellations of the region
+EEMS \citep{petkova2016visualizing} integrates over possible tessellations of the region
 as a way of regularizing the inferred landscape,
@@
 and to Quentin Cronk and Armando Moreno Geraldes for consultations about the \emph{Populus} results.
+Thanks also go to five anonymous reviewers for numerous useful suggestions.
 We would also like to gratefully acknowledge the immense positive impact that Brad McRae \citep[1966--2017;][]{lawler2018tribute}

### 2019-04-20 poplars/populus.infer.R

@@ -171,8 +171,12 @@ g_med <- matrixStats::colMedians(g)
 gam_med <- matrixStats::colMedians(gam)
 
 #save medians
-save(file="g_med.txt",g_med)
-save(file="gam_med.txt",gam_med)
+write.table(g_med, file="g_med.tsv", col.names=FALSE, row.names=FALSE)
+write.table(gam_med, file="gam_med.tsv", col.names=FALSE, row.names=FALSE)
+
+# reload them
+if (!exists("g_med")) { g_med <- scan("g_med.tsv") }
+if (!exists("gam_med")) { gam_med <- scan("gam_med.tsv") }
 
 # for plotting
 library(sp)
@@ -224,6 +228,7 @@ cairo_pdf(filename=paste0(outpath,"/grid_",fname,"_map.pdf"),width=14,height=14)
        main=paste0("Graph Structure"), 
        vertex.size=200, vertex.label.cex=2,
        edge.arrow.size=2, edge.label.cex=1.5,
+       edge.color=rgb(colorRamp(c("#DD8000","gray","#9999FF"))(tanh(g_med/2)),max=255),
+       edge.label.color="#000000",
        add=TRUE, rescale=FALSE)
 dev.off()
-

### 2019-04-20 isolation_by_coalescence.tex
updated figure scaling
erik comments

@@
 so that $G + \epsilon Z$ remains the generator of a valid Markov chain.
+The coalescence rates, $\gamma$, provide additional degrees of freedom
+which may render solutions nonunique in a broader range of situations.
 
@@
 We simulated two other situations on a continuous landscape that produced an bias in effective migration.
-First, we simulated a \textbf{valley} in a landscape with a $5 \times 3$ aspect ratio as above, 
+First, we simulated a \textbf{valley} in a square landscape,
 in which fecundity of individuals in the middle 1/4 of the landscape was reduced by 50\%.
@@
 \centering
-     \includegraphics[scale=0.8]{figs/5x3boxplots}
+     \includegraphics[scale=1]{figs/5x3boxplots}
     \caption{Posterior distributions for values of $g$ 
@@
 \centering
-     \includegraphics[scale=.7]{figs/4x4_grids}
+     \includegraphics[scale=1]{figs/4x4_grids}
     \caption{
@@
 \centering
-    \includegraphics[width=0.45\textwidth]{figs/bias_coal_graph}
-    \includegraphics[width=0.45\textwidth]{figs/bias_commute_graph}
-    \includegraphics[width=0.45\textwidth]{figs/bias_coal_dists}
-    \includegraphics[width=0.45\textwidth]{figs/bias_commute_dists}
+    \includegraphics[width=0.49\textwidth]{figs/bias_coal_graph}
+    \includegraphics[width=0.49\textwidth]{figs/bias_commute_graph}
+    \includegraphics[width=0.49\textwidth]{figs/bias_coal_dists}
+    \includegraphics[width=0.49\textwidth]{figs/bias_commute_dists}
 \caption{
@@
 \centering
-    \includegraphics[width=0.45\textwidth]{figs/barrier_locations}
-    \includegraphics[width=0.45\textwidth]{figs/barrier_eems}
-    \includegraphics[width=0.45\textwidth]{figs/barrier_coal_graph}
-    \includegraphics[width=0.45\textwidth]{figs/barrier_commute_graph}
-    \includegraphics[width=0.45\textwidth]{figs/barrier_coal_dists}
-    \includegraphics[width=0.45\textwidth]{figs/barrier_commute_dists}
+    \includegraphics[width=0.4\textwidth]{figs/barrier_locations}
+    \includegraphics[width=0.4\textwidth]{figs/barrier_eems}
+    \includegraphics[width=0.4\textwidth]{figs/barrier_coal_graph}
+    \includegraphics[width=0.4\textwidth]{figs/barrier_commute_graph}
+    \includegraphics[width=0.4\textwidth]{figs/barrier_coal_dists}
+    \includegraphics[width=0.4\textwidth]{figs/barrier_commute_dists}
     \caption{
@@
 \centering
-    \includegraphics[width=0.45\textwidth]{figs/valley_locations}
-    \includegraphics[width=0.45\textwidth]{figs/valley_eems}
-    \includegraphics[width=0.45\textwidth]{figs/valley_coal_graph}
-    \includegraphics[width=0.45\textwidth]{figs/valley_commute_graph}
-    \includegraphics[width=0.45\textwidth]{figs/valley_coal_dists}
-    \includegraphics[width=0.45\textwidth]{figs/valley_commute_dists}
+    \includegraphics[width=0.4\textwidth]{figs/valley_locations}
+    \includegraphics[width=0.4\textwidth]{figs/valley_eems}
+    \includegraphics[width=0.4\textwidth]{figs/valley_coal_graph}
+    \includegraphics[width=0.4\textwidth]{figs/valley_commute_graph}
+    \includegraphics[width=0.4\textwidth]{figs/valley_coal_dists}
+    \includegraphics[width=0.4\textwidth]{figs/valley_commute_dists}
     \caption{
@@
 \centering
-    \includegraphics[width=0.45\textwidth]{figs/expansion_locations}
-    \includegraphics[width=0.45\textwidth]{figs/expansion_eems}
-    \includegraphics[width=0.45\textwidth]{figs/expansion_coal_graph}
-    \includegraphics[width=0.45\textwidth]{figs/expansion_commute_graph}
-    \includegraphics[width=0.45\textwidth]{figs/expansion_coal_dists}
-    \includegraphics[width=0.45\textwidth]{figs/expansion_commute_dists}
+    \includegraphics[width=0.4\textwidth]{figs/expansion_locations}
+    \includegraphics[width=0.4\textwidth]{figs/expansion_eems}
+    \includegraphics[width=0.4\textwidth]{figs/expansion_coal_graph}
+    \includegraphics[width=0.4\textwidth]{figs/expansion_commute_graph}
+    \includegraphics[width=0.4\textwidth]{figs/expansion_coal_dists}
+    \includegraphics[width=0.4\textwidth]{figs/expansion_commute_dists}
     \caption{
@@
 \centering
-    \includegraphics[width=0.7\textwidth]{poplars/posterior_dists_g_populus}
-    \includegraphics[width=0.7\textwidth]{poplars/grid_populus}
+    \includegraphics[width=0.65\textwidth]{poplars/posterior_dists_g_populus}
+    \includegraphics[width=0.65\textwidth]{poplars/grid_populus}
     \caption{

### 2019-04-21 isolation_by_coalescence.tex
igraph tweaks!

@@
     up to the nearest tenth.
+    Thickness and color of arrows reflect movement rates:
+    darker red arrows have smaller rates and darker blue arrows have larger rates.
     } \label{fig:5x3b_grid}
@@
         Arrows are labeled by the posterior mean gene flow rate inferred under
-        the coalescence time model (which also determines arrow width).
+        the coalescence time model (which also determines arrow width and color).
         For instance, the inferred gene flow rate from 4 to 2 is 3.0,
@@
         \textbf{(top)} Posterior medians (dark line),
-        interquartile range (box), and middle 95\% (whiskers),
+        interquartile range (box), and min--max (whiskers),
         for gene flow rates.
-        \textbf{(bottom)} Posterior medians, along with edge labels that match the upper figure.
+        \textbf{(bottom)} Posterior medians gene flow rates, 
+        both as a numerical label and color
+        (as in other figures, red is lower and blue is larger),
+        along with edge labels that match the upper figure.
+        Nodes are also colored according to posterior median coalescence rate:
+        grey is a high rate (small population)
+        and orange is a low rate (large population).
         \label{sfig:poplar_g}
@@
         Plot shows posterior medians (dark line),
-        interquartile range (box), and middle 95\% (whiskers),
+        interquartile range (box), and min--max (whiskers),
         for coalescence rates, labeled as in Figure~\ref{fig:poplar_results}.

### 2019-04-21 poplars/igraph_plot.R

@@ -1,0 +1,342 @@
+# copied version of igraph::plot.igraph
+#   that we can get the coordinates out of
+my_plot_igraph <- function (x, axes = FALSE, add = FALSE, xlim = c(-1, 1), ylim = c(-1, 
+    1), mark.groups = list(), mark.shape = 1/2, mark.col = rainbow(length(mark.groups), 
+    alpha = 0.3), mark.border = rainbow(length(mark.groups), 
+    alpha = 1), mark.expand = 15, ...) 
+{
+    graph <- x
+    if (!is_igraph(graph)) {
+        stop("Not a graph object")
+    }
+    params <- i.parse.plot.params(graph, list(...))
+    vertex.size <- 1/200 * params("vertex", "size")
+    label.family <- params("vertex", "label.family")
+    label.font <- params("vertex", "label.font")
+    label.cex <- params("vertex", "label.cex")
+    label.degree <- params("vertex", "label.degree")
+    label.color <- params("vertex", "label.color")
+    label.dist <- params("vertex", "label.dist")
+    labels <- params("vertex", "label")
+    shape <- igraph.check.shapes(params("vertex", "shape"))
+    edge.color <- params("edge", "color")
+    edge.width <- params("edge", "width")
+    edge.lty <- params("edge", "lty")
+    arrow.mode <- params("edge", "arrow.mode")
+    edge.labels <- params("edge", "label")
+    loop.angle <- params("edge", "loop.angle")
+    edge.label.font <- params("edge", "label.font")
+    edge.label.family <- params("edge", "label.family")
+    edge.label.cex <- params("edge", "label.cex")
+    edge.label.color <- params("edge", "label.color")
+    elab.x <- params("edge", "label.x")
+    elab.y <- params("edge", "label.y")
+    arrow.size <- params("edge", "arrow.size")[1]
+    arrow.width <- params("edge", "arrow.width")[1]
+    curved <- params("edge", "curved")
+    if (is.function(curved)) {
+        curved <- curved(graph)
+    }
+    layout <- params("plot", "layout")
+    margin <- params("plot", "margin")
+    margin <- rep(margin, length = 4)
+    rescale <- params("plot", "rescale")
+    asp <- params("plot", "asp")
+    frame <- params("plot", "frame")
+    main <- params("plot", "main")
+    sub <- params("plot", "sub")
[…]

### 2019-04-21 poplars/populus.infer.R

@@ -209,26 +209,58 @@ cairo_pdf(filename=paste0(outpath,"/posterior_dists_gam_",fname,"2.pdf"),width=6
         names=paste0("gam",1:n),xlab="Coalescence Rate Parameter Label",ylab="Coalescence Rate",las=2)
 dev.off()
 
+####
+# nice plot on a map
 
+gr <- igraph::graph_from_adjacency_matrix(G_adj,mode="directed",weighted=TRUE)
+igraph::E(gr)$curved <- TRUE
+igraph::V(gr)$color <- cols <- adjustcolor(RColorBrewer::brewer.pal(9, "Set3"), 0.95)
+edge_labels <- sprintf("%0.2f", g_med)
+
+# FAKE plot, really plots AFTER this
+# the point here it to get the edge label coords
+source("
 cairo_pdf(filename=paste0(outpath,"/grid_",fname,"_map.pdf"),width=14,height=14)
-  gr <- igraph::graph_from_adjacency_matrix(G_adj,mode="directed",weighted=TRUE)
-  igraph::E(gr)$curved <- TRUE
-  igraph::V(gr)$color <- cols <- adjustcolor(RColorBrewer::brewer.pal(9, "Set3"), 0.95)
   # set up map
-    
     plot(pop_coords, pch=21, cex=2, 
          col=adjustcolor(c("blue", "red")[as.numeric(info$Species)], 0.75),
          bg=cols[as.numeric(info$groups)], lwd=3)
     # points(centroids[,1:2], pch=21, col=adjustcolor(cols[1:nlevels(info$groups)], 0.25), cex=5)
-    maps::map(add=TRUE, col=grey(0.8))
-  # edge_labels <- paste0("g",1:ng,"=",round(g_med*1000)/1000)
-  edge_labels <- sprintf("%0.2f", g_med)
-  plot(gr,layout=as.matrix(centroids[,1:2]),edge.width=(3*g_med+.2),
+    maps::map(add=TRUE, col=grey(0.6))
+  ip <- my_plot_igraph(gr,layout=as.matrix(centroids[,1:2]),
+       edge.width=pmin(15, 3 * g_med + .1),
        edge.label=edge_labels,
        main=paste0("Graph Structure"), 
        vertex.size=200, vertex.label.cex=2,
[…]

### 2019-04-21 isolation_by_coalescence.tex
resubmission version

@@
         \textbf{(top)} Posterior medians (dark line),
-        interquartile range (box), and min--max (whiskers),
+        approximate middle 50\% (box), and 90\% (whiskers),
         for gene flow rates.
@@
         Plot shows posterior medians (dark line),
-        interquartile range (box), and min--max (whiskers),
+        approximate middle 50\% (box), and 90\% (whiskers),
         for coalescence rates, labeled as in Figure~\ref{fig:poplar_results}.

### 2019-04-22 review-responses.tex
actual submission

@@
 with page and line numbers that refer to the revised version of the manuscript.
-There is also a version of the manuscript showing differences to the previously submitted document.
+We are also including a file with changes made since the first submission highlighted
+(named \texttt{ibc-diff.pdf}).
+Finally, we are including an additional pdf (named \texttt{isolation\_by\_coalescence.pdf})
+that has the paper, the supplementary material, and the responses to reviewers all in one place
+in which the page/line numbers in the reviews are hyperlinks 
+to the correct location in the main document.
 

### 2019-04-22 isolation_by_coalescence.tex
actual submission

@@
 described in \citet{geraldes2014landscape}.
-These data \citep{geraldes2014data},
+These data 
 include geographical coordinates and genotypes from a genotyping chip targeting 34K SNPs.
@@
 %%%%%%%%%%%%%%%%
-\subsection*{Code availability}
+\subsection*{Code and data availability}
 
@@
 and \url{https://github.com/petrelharp/isolation_by_coalescence} (including the SLiM simulations).
+Sequencing data for \textit{Populus} is available from Data Dryad \citep{geraldes2014data}.
 

### 2019-05-03 isolation_by_coalescence.tex
final submission

@@
 \newif\ifreviewresponses
-\reviewresponsestrue  % include them
-% \reviewresponsesfalse  % don't include them
+% \reviewresponsestrue  % include them
+\reviewresponsesfalse  % don't include them
 \input{review-response-commands}
@@
 
-\linenumbers
+\nolinenumbers
 
@@
 without changing commute times.  \llabel{ll:dense1}
-A small concrete example of this is given in Appendix \ref{ss:simple_example}.
+A small concrete example of this is given in Appendix \ref{ss:simple_example},
+and depicted in Supplemental figure \ref{fig:valid_range}.
 More generally, commute times only depend on the symmetric part of $H$ --
@@
 Similar recursions are common in the literature,
-going back at least to \citet{hill1972effective} (see also \citet{whitlock1997effective}).
+going back at least to \citet{hill1972effective} 
+(see also \citet{tufto1998analysis} and \citet{whitlock1997effective}).
 In matrix notation, this is
@@
 To produce realistic data, we implemented forwards-time simulations using SLiM v3.1 
-\citep{haller2018forward,haller2018treesequence},
+\citep{haller2018forward,haller2019treesequence},
 with individuals living across continuous, two-dimensional geography (sometimes with barriers),
@@
 the mutation rate was set to 0.0 during the forwards simulation,
-while tree sequences were recorded \citep{haller2018treesequence},
+while tree sequences were recorded \citep{haller2019treesequence},
 and mutations were added after the fact with msprime,
@@
 Thanks to David Levin for useful suggestions regarding hitting time calculations,
-to Paul Marjoram for useful comments,
+to Paul Marjoram and Nick Barton for useful comments,
 to Brad Shaffer and Evan McCartney-Melstad for discussions about landscape modeling,
